### 2023-12-04 R/Paper_markdown.Rmd
To_Be_Cleaned

---
title: "MASCARA: coexpression analysis in data from designed experiments"
author: "Fred T.G. White, Anna Heintz-Buschart, Lemeng Dong, Harro J. Bouwmeester, Johan A. Westerhuis, Age K. Smilde"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  # powerpoint_presentation
  # ## keep_tex: true
  # bookdown::pdf_document2:
    bookdown::word_document2:
      number_sections: false
      toc: false
      classoption:
        twocolumn
 # # html_document:
#  #   toc: true
#  #   theme: united
#  #   toc_depth: 3
#  #   number_sections: true
# bibliography: MASCARA_refs.json
bibliography: "`r rbbt::bbt_write_bib('MASCARA_refs.json', overwrite = TRUE)`"
# csl: oxford-scimed.csl
#https://raw.githubusercontent.com/citation-style-language/styles/master/apa-6th-edition.csl
# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-6th-edition.csl
csl: apa-6th-edition.csl
keep_md: true
link-citations: yes
header-includes:
  - \usepackage{multicol}
  - \usepackage{amsmath}
  - \usepackage{float}
  - \usepackage{bm}
  - \newcommand{\hideFromPandoc}[1]{#1}
  -  \hideFromPandoc{
       \let\Begin\begin
          \let\End\end
          }

---

```{r setup, include=FALSE}
library(knitr)
# library(here)
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F, fig.pos = "H") #, root.dir=here()
#pdf_document
#powerpoint_presentation
```

```{r , echo=FALSE, include=FALSE}
#SOURCE FUNCTIONS FROM EXTERNAL FILE
source("DATA_SIM_FUNCS.R")
```

```{r, echo = FALSE, include=FALSE}
library(devtools)
library(ggplot2)
library(MetStaT)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)
library(MUVR)
library(pheatmap)
library(caret)
library(mixOmics)
library(cowplot)
library(reticulate)
library(biomformat)
library(MetStaT)
library(DiscriMiner)
library(viridis)
library(directlabels)
library(ggpmisc)
library(DescTools)
library(data.table)
library(plotly)
library(ggbreak)

```

# Abstract

<!-- Need to change this as it is repeated in introduction -->

Experiments in plant transcriptomics are usually designed to induce variation in a pathway of interest (POI). Harsh experimental conditions can cause widespread transcriptional changes between groups. Discovering coexpression within a POI (here strigolactones) in this context is hampered by the dominant variance induced by the design. Minor changes in experimental conditions not controlled for may affect the plants, leading to small coordinated differences in genes within POIs and related pathways (RPs) between replicate plants in the same controlled experimental condition. These systematic differences between replicates within the same experimental conditions can be used to detect genes from the same POI or RPs. We introduce a novel framework “MASCARA” which combines ANOVA simultaneous component analysis and partial least squares to remove the experimentally induced variance and investigate multivariate relationships in the non-designed variance. MASCARA is tested against a selection of competitors on simulated data, created to mimic a designed transcriptome study. In a coexpression analysis of a real dataset MASCARA detects several uncharacterised but relevant transcripts. Our results indicate that there is sufficient structure left in a typical dataset after correcting for experimental variance and that this residual information is useful to investigate coexpression.

<!-- \twocolumn -->

\Begin{multicols}{2}

# Introduction
<!-- DE or CoE to understand system + feature selection -->
Plant transcriptome studies typically involve the use of designed experiments which aim to induce variation in pathways of interest (POIs) by controlling one or more experimental factors. This can be done through, for example, varying the level of certain essential nutrients, or knocking out/down a gene within a POI. To explore which pathways are affected by the intervention, gene expression can be measured through RNA sequencing (RNAseq). Analysis of RNAseq data generally adheres to one of two approaches: differential expression analysis (DE; @andersDifferentialExpressionAnalysis2010; @robinsonEdgeRBioconductorPackage2010) or coexpression analysis (CoE; @langfelderWGCNAPackageWeighted2008; @tzfadiaCoExpNetVizComparativeCoExpression2016). DE is used to determine which genes are expressed differently between two or more experimental conditions, highlighting known or novel pathways affected by the experimental conditions. CoE, on the other hand, aims at discovering genes that are part of a partially characterized pathway. Here, the expression profiles of known genes of the POI (termed "baits") are used to detect novel pathway members with similar expression patterns (here termed "spikes").

```{r nettype, fig.cap= c("Pathway types exemplified, POI; pathway of interest, RP; related pathway, URP; unrelated pathway. A represents hypothetical biological pathways, B highlights the problem with detecting these pathways when dealing with data from designed experiments.")}
#load network types image

a <- knitr::include_graphics("networktypes.png")
a

```

Figure \@ref(fig:nettype) outlines pathway nomenclature that will be used throughout this text. *A* represents a true hypothetical pathway including a pathway of interest (POI), related pathway (RP) and an unrelated pathway (URP). All three types of pathway are assumed to be affected to varying degrees by the experimental design. When dealing with experimentally designed data, we have to consider two key sources of variance: between-group variance and within-group variance. Between-group variance pertains to the differences between experimental groups (or conditions). This type of variance is of interest when conducting DE studies. Within-group variance, on the other hand, is the variance within experimental groups, this is not interesting for DE studies but may contain information useful to CoE. It is the excessive between-group variance that leads to the densely connected nature of *B* from \@ref(fig:nettype). Figure \@ref(fig:withinbetween) exemplifies between-group and within-group driven correlations. The plots show the relationship between two genes of the POI (*A* and *C*) and the relationship between two genes that are not part of the same pathway (*B* and *D*).  The total correlations of both pairs of transcripts are high but the second pair’s correlation is driven by the between-group variation and not by correlations within experimental groups. The within-group correlation is lower than the total in both cases, but there is a clear trend in one condition (facet *C*: P- condition in red) for the first gene pair. Because of this, the within group correlation (r) is higher for the first transcript pair (*C*) than the second (*D*), which makes it possible to discern between genes within a POI or RP and genes that both respond similarly to a treatment (URP). Hence, we need to distinguish *total CoE* which is calculated on the total variance and *within-group CoE* where the focus is on the shared variance within experimental groups.

<!-- From a plant biology point of view, the between-group variance and the within-group variance can be quite different and related to different parts of the biological system. It is, in fact, expected that there are mechanistic changes that arise from the very different conditions that could alter the coexpression of genes.  -->

<!-- , examples include: @miettinenSecoiridoidPathwayCatharanthus2014;.  -->

From a plant biology point of view, the between-group and the within-group variance can be quite different. However, both may contain valuable biological information. The between-group variance is in principle larger because groups are the result of treatments that are usually selected because they affect the genes and/or pathways under investigation. Within-group variance (variance across replicates), if too large, can be an issue in DE analysis, we argue that in co-expression analysis the information present in this variance may be valuable. Variance between replicates is caused by non-controlled variations in plant development and environmental conditions. One replicate within a group may be slightly bigger than the others and therefore respond slightly different to a treatment, for example phosphate deficiency. A larger plant may start to suffer from the deficiency earlier than a smaller plant. This does not imply that the larger plant shows a different biological response to P deficiency. We can assume that the response is similar, but just occurs a bit faster. This in turn implies that the expression of the bait genes are induced a little bit faster than in the other replicates, but this will also be true for the genes that we want to identify with a co-expression analysis. There may be many other factors that are inducing this within-group variance, but it is quite likely that if a factor induces a change in the treatment response in one replicate, that this response difference occurs throughout the entire process or pathway in that one replicate. This means that the within-group variance likely contains valuable information and by averaging out this variance, as is done in differential expression analysis, we are losing information.

<!-- This phenomenon is well documented in @caldanaHighdensityKineticAnalysis2011 where an experiment was performed investigating the effects of temperature and light on arabidopsis plants over time. In this study the authors investigated variable relationships between metabolites and showed these were strongly dependent on the specific levels of the experimental factors, e.g., shikimic acid and phenylalanine were positively correlated at 21 degrees in high light condition, but negatively correlated at 32 degrees in the dark.  -->

<!-- Between experimental groups different pathways are active, whereas within groups more general (undocumented/uncontrolled) ambient factors such as temperature, light intensity and soil water content can affect the plant and its transcriptome in a concerted fashion. In experiments where the intervention causes large between-group differences to many genes, not making the distinction between types of variance can lead to results that contain total CoE associations driven by similar differences between treatment groups.  -->

Considering this within-group variance increases the strength of coexpression analysis, for example, in experiments where the intervention causes large between-group differences in many genes. Not making the distinction between types of variance results in *total CoE* associations driven by the differences between treatment groups as opposed to within group correlations. For example in @haiderTranscriptomeAnalysisPhosphate2023 a phosphate (P) starvation was applied to rice plants to induce strigolactone (SL) biosynthesis, to study the change in gene expression in comparison to control plants with normal supplies of P. The SL pathway is known to be triggered through P deficiency. However, as P is one of three key nutrients for plant growth, deficiency of this nutrient causes widespread differential expression throughout the transcriptome (@wangRoleStrigolactonesDeficiency2021). @haiderTranscriptomeAnalysisPhosphate2023 used ranked correlations, similar to MutRank (@poretskyMutRankShinyWebapplication2020), to determine (total) coexpression candidates. Due to such a strong effect of the experiment, many genes were affected on top of the desired SL genes. This masked the within-group correlation between the baits and the spikes (remaining unknown SL genes). This problematic issue was also highlighted recently by @saccentiWhatCanGo2023 where group sampling assumptions are investigated. Hence, both from a data analysis as well as from a biological perspective it is relevant to make a distinction between the different types of CoE. The focus of this paper will be on developing methods to investigate within-group CoE.

```{r withinbetween, fig.cap= c("Correlation examples, Pearson’s correlation coefficient (r) indicated in facet titles. X axes, expression levels of CCD8 gene in rice, Y axes expression levels of example coexpression candidates. Top row A-B; *total CoE*, bottom row C-D *within-group CoE*. Gene pairs in both A and B have strong total correlations. Correlation in A is also exhibited within a group, B has strong total correlations mainly due to the differences between groups. These examples are indistinguishable from eachother without variance partitioning which has been performed on the same genes in C-D.")}

a <- knitr::include_graphics("within_between_example_2.png")
a

```

For a successful coexpression analysis it is necessary to have a predefined set of baits that are already known to be involved in the mechanism under study. Methods like ranked correlations and partial least squares regression (PLS; @geladiPartialLeastsquaresRegression1986; @woldCollinearityProblemLinear1984) are able to investigate relationships between a set of baits and the other measured genes. In ranked correlations the relationships are quantified at a bivariate level i.e. bait 1 - candidate 1, bait 1 - candidate 2. In PLS regression the relationships can be investigated in a multivariate way, in its simplest form we can see which genes are most predictive of the expression of one bait. This can be extended to a PLS2 model where relationships are investigated between a set of multiple baits and all other genes. These methods do not by default account for the different variance types discussed above. Hence, they are not directly suitable for within-group CoE.

The most widely known and applied method that investigates variance sources is the analysis of variance (ANOVA; @searleLinearModels1971). ANOVA simultaneous component analysis (ASCA; @smildeANOVAsimultaneousComponentAnalysis2005) and its extension to imbalanced data via the generalised linear model (ASCA+; @thielASCAAPCAExtensions2017) have been developed to combine principal component analysis (PCA) with ANOVA and can therefore account for experimental design in dimensionality reduction. ASCA is of interest due to the variance partitioning, this is typically used to estimate the effect matrices that are calculated from the design and particularly the within-group variance structure which is maintained in the residuals of the model. Both PLS and ASCA have been used for DE (@augerGeneCoexpressionNetwork2022, @renRNABindingProteins2023, @nuedaDiscoveringGeneExpression2007, @jarmundALASCAPackageLongitudinal2022) and CoE (@guanSRGSSparsePartial2022, @nuedaFunctionalAssessmentTime2009). ASCA and PLS have been combined before in @thissenImprovingAnalysisDesigned2009 creating ANOVA-PLS, however the goal of their study was not coexpression analysis. Neither of these methods are optimal by themselves for CoE in the context of large between group variance with known baits.

The baits are assumed to be part of the same POI and the remaining uncharacterised spikes (POI or RP genes) are assumed to be numerous. Therefore, both the baits and the remainder of the data could be reduced to linear combinations in order to investigate relationships between a set of baits and putative spikes. We introduce a new data analysis method called multivariate ASCA residual analysis (MASCARA). MASCARA is motivated by four key factors; (1) the presence of dominant between-group variance caused by the setup of the experiment, as well as, (2) ambient variance (structured effects caused by undocumented environmental factors i.e. temperature, light intensity or soil water content) within the replicates of each experimental group, (3) the need to investigate multivariate relationships and (4) the low number of samples. The method capitalises on ASCA to account for an experimental design and the multivariate analysis capability of PLS2 to find uncharacterised POI genes using a set of baits. Instead of focusing on the variance induced by the experimental design, as is a typical use of ASCA, this variance is removed, and the remaining variance is further analysed to investigate the multivariate relationships between baits and the putative spikes in terms of within-group CoE.

To explore the utility of MASCARA, the method will be compared to ASCA, PLS2 and correlation analysis illustrated using an example RNAseq dataset (from @haiderTranscriptomeAnalysisPhosphate2023). Furthermore, this dataset is used as a guide to several simulation studies. For these simulations we emulate both different levels of between- and within-group variance. We test the effects of size of between group differences, number of URP differential genes, measurement noise (random unstructured technical variance), structured ambient variance and number of replicates. The following sections seek to investigate which statistical methods are best suited for biosynthetic POI elucidation when a set of baits is already known and illustrate the use benefit of MASCARA, particularly in data coming from experiments designed for differential expression analysis.

```{r, echo = FALSE, include=FALSE}
baits <- paste0("X_",c(1989:2000))   #501:510

meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
colnames(meta) <- c("growth_condition","time","interaction")
```
# Materials and Methods

<!-- ## Dataset: Experimental Design -->
```{r, echo = FALSE, include=FALSE}
META <- as.data.frame(meta[,1:2])
META$growth_condition[which(META$growth_condition == "1")] <- "P+"
META$growth_condition[which(META$growth_condition == "-1")] <- "P-"

```

```{r, echo = F}
META <- META[!duplicated(paste0(META$growth_condition, META$time)),]
rownames(META) <- NULL
# kable(META, format = "simple") #replace with CANDTAB like

fill <- rep(c("grey95", "grey90"), nrow(META))

theme1 <- gridExtra::ttheme_default(core = list(
  fg_params = list(fontface=c(rep("plain", nrow(META)))),
  bg_params = list(fill = fill)),
  base_size = 10, padding = unit(c(3, 3), "mm"))

META <- gridExtra::tableGrob(META, rows = NULL, theme = theme1)
title <- textGrob("Design",gp=gpar(fontsize=24))
META <- gtable::gtable_add_grob(META, title,1,1,1,ncol(META))

rdo <- readRDS("Data_Haider/RICE_DATA_OVERVIEW.RDS")
levels(rdo[[2]]$data$time) <- c("1","3","7","8")

# rdo <- list(META,rdo[[1]],rdo[[2]],rdo[[3]])
# plot_grid(META)
# # grid.newpage()
# # grid.draw(META)
```

```{r, echo = F}
# h <- readRDS("heat_comp.RDS")
# levels(h[[2]]$time) <- c("1","3","7","8")
# 
# ann_colours <- list(g_c = c("P-" = "#F8766D","P+" = "lightblue"),
#                     time = c("1" = "#fde725","3" = "#35b779","7" = "#31688e","8" = "#440154"))
# 
# heat <- pheatmap::pheatmap(t(scale(t(h[[1]]))),  #
#                              cluster_rows = T,
#                              cluster_cols = F, 
#                              show_rownames = T, 
#                              show_colnames = F,
#                              annotation_col = h[[2]],
#                            annotation_colors = ann_colours,
#                            treeheight_row = 10,
#                            fontsize = 8,
#                              # annotation_row = anno,
#                              main = "Strigolactone Gene Expression",silent = T)
#   
#   # heat
#   library(ggplotify)
#   heat2 <- as.ggplot(heat) + theme(plot.title = element_text(face = "plain"))

# heat2 <- readRDS("heatmap_MASCARA_Fig_2.RDS")

```

## RNAseq - Rice strigolactone dataset @haiderTranscriptomeAnalysisPhosphate2023

In order to guide the simulations we illustrate the characteristics of the simulated data with a real dataset. The real data are a subset of samples from a study in which the root transcriptome of rice plants is measured under two  growth condition (P+/P-) and measured at four time points (1, 3, 7 and 8  days post treatment). At each combination of growth condition and time point, the root transcriptome of three new plants was harvested for RNA sequencing. In this dataset, the curated list of SL pathway genes show a specific profile: low variance baseline expression across all time points in the P+ condition but activation and increasing expression over time with P starvation.

Figure \@ref(fig:Dataset1-Overview) shows the gene expression levels of 9 strigolactone pathway genes along with 9 highly differential genes that are not part of the SL pathway. While the expression is low in P+ condition a clear increasing time profile can be observed in the P- condition, indicating an interaction between the growth condition and the time factors.

What is notable here is the replicate specific patterns that are apparent amongst the SL genes and largely absent between the SLs (purple) and the non-SL differential genes (green). For example, at time point 3,  most SL genes are expressed lowest in replicate 1 and highest in replicate 3, while at time point 7 replicate 2 has a higher SL gene expression. These within-group patterns are either not shared at all between the SL genes and the other DE genes or to a much lesser extent. This is due to the other genes not directly being part of the same pathway. There are of course other pathways and thus there are still some subtle replicate patterns in the non-SL genes but this effect is much clearer within our pathway of interest. Figure \@ref(fig:ASCA-1) shows the output of the ASCA decomposition in which the PCA scores and loadings of the combined effect of the  Growth condition factor and the interaction of Growth condition and Time is explored. The score plot shows that the replicates of each condition are clustered while the difference between the conditions is much larger. The loading plot shows that although the 9 SL genes are mostly on the outside of the plot, many other genes are as well. Thus many differentially expressed genes are found important that are not part of the SL pathway, in other words ASCA does not detect spike SL genes from others that are similarly or more differentially expressed but not part of the SL pathway.

This marks the goal of this paper, which is to detect within-group coexpressed genes with a set of baits preferentially to other differentially expressed genes that are not part of the same pathway (with no within-group structure).

<!-- \onecolumn ### -->

\End{multicols}

```{r Dataset1-Overview, echo = F, fig.cap=c("Real data overview. RNAseq of rice root. Heatmap of SL pathway and example highly differential genes.")}

# (rdo[[1]] + heat2)/(rdo[[2]] + rdo[[3]])
# layout <- rbind(c(1,1,2,2,2,2),
#                 c(1,1,2,2,2,2),
#                 c(1,1,2,2,2,2),
#                 c(1,1,2,2,2,2),
#                 c(NA,NA,NA,NA,NA,NA),
#                 c(3,3,3,4,4,4),
#                 c(3,3,3,4,4,4),
#                 c(3,3,3,4,4,4),
#                 c(3,3,3,4,4,4))
# grid.arrange(rdo[[1]], heat2, rdo[[2]],rdo[[3]], layout_matrix = layout)
# rdo[[3]]$layers[[1]]$aes_params$colour <- "green"
# rdo[[3]]$layers[[2]]$aes_params$colour <- "purple"

##Error in discrete_range() : could not find function "discrete_range"
# rdo[[3]] <- rdo[[3]] + scale_color_manual(values = c("green","purple"), breaks = c("Other","Pathway"), labels = c("Other","SL")) + theme(legend.title=element_blank())

# withr::with_options(
#   list(ggplot2.discrete.fill = c("green","purple")),
#   print(rdo[[3]])
# )

# s <- readRDS("rice_ASCA_scores.RDS")
# l <- readRDS("rice_ASCA_loadings.RDS")

heat2 <- knitr::include_graphics("MASCARA_heatmap.png")
heat2
#/(rdo[[2]] + rdo[[3]])

# pdf("Rice_ASCA.pdf")
# heat2/(rdo[[2]] + rdo[[3]])
# dev.off()

```

```{r ASCA-1, echo = F, out.width = '100%', fig.cap=c("Scores and loadings of the ASCA model of the combined effect (condition + condition:time interaction).")}
knitr::include_graphics("rice_s_l.png")

```

\twocolumn

<!-- \Begin{multicols}{2} -->

## Simulations

The problem in the real dataset is that the nutrient stress applied causes a major effect on many other genes besides our baits and spikes. This widespread transcriptional regulation can mask the coexpression patterns of interest due to the large number of genes that are upregulated under the stress condition, making it more difficult to detect genes that are responding to a lesser extent, for example the SL pathway.

The simulations are built upon the results of the ASCA analysis of the combined effect of growth condition and the interaction of growth condition and time. The simulated data is created by simulating score and loading values (as in Figure \@ref(fig:simOverview)) for the combined effect ($\alpha + \alpha\beta$) ) explaining the difference in time profiles of the P+ and P- condition and the main time effect $\beta$, as well as, structured ambient variance ($S$). This is to mimic variance due to gene expression as a function of growth condition and time as well as variance due to coexpression of genes in the same  pathway on top of this experimentally induced DE variance.

## Data simulation 1: Combined effect size and number of differential genes

The first simulation aims to determine how well the methods perform to find coexpressed genes in the presence of a dominant between-group variance structure and increasing numbers of non-pathway differential genes (genes with high between-group correlation but no within-group correlation). The expression levels of differential transcripts is modeled with the scores and loadings of the growth condition and its interaction with time ( $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{P}_{\alpha + \alpha\beta}^T$ ). The disturbing time effect is modeled with scores and loadings ( $\mathbf{T}_{\beta}$  and $\mathbf{P}_{\beta}^T$ ). Also included is some random measurement noise, this is modeled as a matrix $\mathbf{E}$ with normally distributed random values $e_{ij} = \mathcal{N}(0,\sigma^2)$.

On top of the between-group effect there is an additional a correlation structure applied to the baits and spikes to create a model of coexpression that is independent to the variance induced by the experimental factors ($\mathbf{X}_S = \mathbf{T}_S\mathbf{P}_S^T$). In this simulation we have a set of 4 baits that have positive loadings for the combined effect along with a variable number of other genes - "Combined Effect Responders". For the structured ambient variance; 2 random sample offset vectors in \(\bm{T}_S\) (each column representing an effect caused by ambient conditions) which our baits and spikes contribute to (i.e. green and black bars in Figure \@ref(fig:simOverview) are non zero in \(\bm{P}_S\)). Similarly to the real data, the gene expression values for each experimental condition are simulated with three independent replicates per time-condition-combination. Each simulated data $\mathbf{Y}$ set thus consists of 24 samples and 2000 genes.

```{r, echo = FALSE}
X_funced_test2 <- Create_Core(nreps = 3, meta, irr_spikes = TRUE, struc_resid = FALSE,
                        a_sigma = c(1,1), b_sigma = c(1.5,1.5), e_sigma = c(1,0.3),
                        noise_sd = 0, EffectSize = c(X_a_ab = 1, time = 0, E = 0, mu = 1),
                        plot = TRUE)
```

The scores matrices $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{T}_{\beta}$ each consist of two orthogonal score vectors and their relative levels are indicated by the colour. The different colours in Figure \@ref(fig:simOverview) represent that the scores are orthogonal to each other. Each small block consists of the same score value for each of the replicates within that condition. Loadings $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$  are simulated to create the baits and spikes. As displayed in Figure \@ref(fig:simOverview), the green bar represents 4 baits (or known pathway genes), the black bar represents the 12 spikes (“unknown” genes of interest), yellow represents differentially expressed genes not in SL pathway, and purple; a set of differential genes not in SL pathway with an alternative treatment response.

In $\mathbf{P}_{\beta1}^T$ the baits, spikes and other DE genes along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model. The remaining elements of $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$ are populated with zeroes. The $\mathbf{T}$ matrices represent the simulated scores for their respective factors. For blues, reds, greys and golds we have 4 different shades representing 4 unique values per factor level (total 8 unique values where reds and golds are sign flipped blues and greys respectively). Similarly oranges and pinks represent a growth condition independent time profile with four unique values. $\mathbf{E}$ represents random unstructured technical noise and the additional $\bm{1\mu}^T$ term in \@ref(eq:Sim2-bp) represents a mean offset for each gene.

```{r simOverview, echo=FALSE, fig.cap="Coexpression simulation, simplified overview. Colours correspond to the same genes in the different loading vectors (*P*). Black; baits (known SL genes), green; spikes (unknown SL genes), yellow; other non-pathway DE genes, purple non-pathway DE genes with a different profile. In the score vectors (T); Blues, reds, greys and golds represent experimental conditions, oranges and pinks; time. Each of the textures in the ambient/residual part represent a distinct normally distributed set of random values.", out.width = '100%'}
knitr::include_graphics("Sim_SL_diagram_simplified.png")
```

\begin{equation}
\begin{aligned}
\bm{Y} &= \bm{1\mu}^T + \bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\
\text{where for each element ik or jk:}\\
&\bm{T}_{S_{ik}} \sim \mathcal{N}(0,1), \quad \bm{P}_{S_{jk}}^{\text{baits,spikes}} \sim \mathcal{N}(0.5,0.2) \\
\text{and:} &\\
\bm{P}_{S}^{\text{rest}} &= 0
\end{aligned}(\#eq:Sim2-bp)
\end{equation}

<!-- ### Combined effect size and number of differential genes -->

<!-- \begin{equation} -->
<!-- \begin{aligned} -->
<!-- \bm{Y} &= \bm{1\mu}^T + d(\bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T) + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\ -->
<!-- \text{where:} &\\ -->
<!-- \bm{P}_{\alpha + \alpha\beta}^T = \begin{bmatrix} -->
<!--        \bm{p}_{11} & \bm{p}_{12} \quad \bm{p}_{13} \\ -->
<!--        \bm{p}_{21} & \bm{p}_{22} \quad \bm{p}_{23} -->
<!--      \end{bmatrix} -->
<!-- \end{aligned}(\#eq:nCER) -->
<!-- \end{equation} -->

The issue to address is the presence of many non-pathway differential genes; those that have strong between group correlations but no within group correlation. Here the differential genes relate  to the yellow bar in \@ref(fig:simOverview). We incrementally increase the number of non-zero values and decrease the number of 0 values, which in figure \@ref(fig:simOverview) relates to increasing the size of the yellow bar, as well as varying the F-norm of the combined effect matrix through the scalar parameter *d* below;

\begin{equation}
\begin{aligned}
\bm{Y} &= \bm{1\mu}^T + d(\bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T) + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\
\end{aligned}(\#eq:nCER)
\end{equation}
<!-- equation \@ref(eq:nCER).  -->

<!-- \(\bm{p}_{2}\) remains the same with \(\bm{p}_{21}\) and \(\bm{p}_{23}\) being 0 and \(\bm{p}_{22}\) (purple bar in \@ref(fig:simOverview)) positive. -->

## Data simulation 2: Replication

In order to evaluate the generalisation of methods and indicate the levels of variance required for optimal performance we conduct a larger scale assay. Here we fix the number of non-pathway DE genes to 30 which corresponds (to the yellow bar size) and the size of the within correlation effect matrix (\(\|\bm{T}_S\bm{P}_S^T\|_F = 0.025\)). We investigate a range of values for sizes of between variance (*d*) and random noise (*l*) across random structures (which correspond to repeated simulations of randomised values for $\mathbf{T}$ matrices). Parameters were adjusted according to \@ref(eq:Test-2) on 20 randomised $\mathbf{T}$ matrices.

\begin{equation}
\begin{aligned}
\bm{Y} &= \bm{1\mu}^T + d(\bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T) + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + l\bm{E} \\
\end{aligned}(\#eq:Test-2)
\end{equation}

On these 20 datasets, per condition, the replicate number $n_r$ was also varied: the number of replicates $n_r \in 3:15$ per experimental condition were sampled from a maximum $n_r$ of 15.

```{r, echo = FALSE}
# X_funced_classic <- Create_Core(nreps = 3, meta = meta, irr_spikes = TRUE,
#                         struc_resid = FALSE, noise_sd = 0.75, plot = TRUE)   ## sum(X_funced[[1]] != X_core) = 0 with these parameters

X_funced_classic <- Create_Core(nreps = 3, meta, irr_spikes = FALSE, struc_resid = FALSE,
                        a_sigma = c(1,1), b_sigma = c(1.5,1.5), e_sigma = c(1,0.3),
                        noise_sd = 0, EffectSize = c(X_a_ab = 1, time = 0, E = 0, mu = 1),
                        plot = TRUE)

# alphaN <- 2
# betaN <- 4
# nreps <- 5
# 
# meta2 <- cbind.data.frame(rep(c(1,-1), each = betaN*nreps), rep(c(1:betaN), each = nreps), rep(c(-1:-betaN,1:betaN), each = nreps))
# 
# colnames(meta2) <- c("growth_condition","time","interaction")

X_funced_test <- Create_Core(nreps = 3, meta, irr_spikes = FALSE, struc_resid = FALSE,
                        a_sigma = c(1,1), b_sigma = c(1.5,1.5), e_sigma = c(1,0.3),
                        noise_sd = 0.75, EffectSize = c(X_a_ab = 1, time = 0, E = 1, mu = 1),
                        plot = TRUE)

X <- X_funced_classic[[1]]
P <- X_funced_classic[[2]]
Pb <- X_funced_classic[[3]]
```

```{r, echo = F}
#create "ref" for feature spike information

ref <- cbind.data.frame(Feature = colnames(X), P, Pb)

cols <- colnames(ref[,-1])
ref$Effect <- do.call(paste, c(ref[cols], sep = "_"))

ref$Effect[which(ref$Effect == "0.5_0_0_0.1")] <- "Low PC1"
ref$Effect[which(ref$Effect == "1_0_0_0.1")] <- "Medium PC1"
ref$Effect[which(ref$Effect == "1.5_0_0_0.1")] <- "High PC1"
ref$Effect[which(ref$Effect == "0_0.5_0.1_0")] <- "PC2"

ref$Effect[grep("^0", ref$Effect)] <- "None"

ref$Effect <- factor(ref$Effect, levels = c("Low PC1", "Medium PC1", "High PC1", "PC2", "None"))

```

## Analysis and feature selection

### Ranked correlations

Pearson's correlation coefficient is calculated between all columns of \(\bm{Y}\). 
For example the correlation between a SL gene $j_{sl}$ and a non-SL gene $j_q$ is calculated as:

\begin{equation}
\begin{aligned}
  r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}}
\end{aligned}(\#eq:Pearson)
\end{equation}

For ranked correlations *r* is calculated between each bait and all other genes, with the lowest rank for the highest correlation. Then for each bait the other genes are ranked based on *r*. These ranks are averaged across all baits, similarly to @poretskyMutRankShinyWebapplication2020.

### ASCA+

ANOVA simultaneous component analysis (@smildeANOVAsimultaneousComponentAnalysis2005) involves splitting data into effect matrices which are defined by the experimental design structure, after which a PCA is applied to the effect matrices or combination thereof. In ASCA+ this variance partitioning is achieved through the generalised linear model as described in @thielASCAAPCAExtensions2017. ASCA+ addresses the issue of biased effects matrices in unbalanced designs. The following is a summary of the algorithm description provided in @thielASCAAPCAExtensions2017.

The multivariate generalised linear model (GLM) can be expressed as:

\begin{equation}
\mathbf{Y} = \mathbf{X} \mathbf{\Theta} + \mathbf{E}
\end{equation}

where \(\mathbf{Y}\) is the response matrix of dimension \(I \times J\), \(\mathbf{X}\) is the model matrix of dimension \(I \times p\), containing the coded indicator levels for each of the levels for each factor and their interactions,
 \(\mathbf{\Theta}\) is the parameter matrix of dimension \(p \times J\) and
 \(\mathbf{E}\) is the error matrix of dimension \(I \times J\).

Through Ordinary Least Squares (OLS) we can obtain the (unbiased) estimators of the model parameters:

\begin{equation}
\hat{\mathbf{\Theta}} = (\mathbf{X}^\intercal \mathbf{X})^{-1} \mathbf{X}^\intercal \mathbf{Y}
\end{equation}

The effect matrices for different terms of the model can be obtained as:

\begin{equation}
\mathbf{M}_f = \mathbf{X}_f^* \hat{\mathbf{\Theta}}
\end{equation}

where: \(\mathbf{M}_f\) is the effect matrix corresponding to effect \(f\), \(\hat{\mathbf{\Theta}}\) is the matrix of estimated parameters obtained from OLS and \(\mathbf{X}_f^*\) is a new model matrix obtained by keeping in \(\mathbf{X}\) only the block \(\mathbf{X}_f\) and replacing all other columns with zeros as per:

\(\mathbf{M}_f=(0|\mathbf{X}_f |0)\mathbf{\hat{\Theta}} = \mathbf{X}_f \mathbf{\hat{\Theta}}_f\)

The above is achieved through the sum coded indicator matrix as illustrated further in @thielASCAAPCAExtensions2017.

Furthermore, the matrix \(\mathbf{E}\) is estimated as the residual matrix of the model:

\begin{equation}
\begin{aligned}
\hat{\mathbf{E}} = \mathbf{Y} - \mathbf{X} \hat{\mathbf{\Theta}}
\end{aligned}
\end{equation}

<!-- This approach provides a general solution to build the ASCA effect matrices for any fixed effects ANOVA model. -->

### PLS regression - canonical mode

Partial Least Squares canonical mode (PLS2) is a multivariate regression technique used for modeling relationships between two sets of variables, \( \mathbf{Y} \) and \( \mathbf{B} \).

After splitting the data into the response matrix (containing only baits) \( \mathbf{B} \) (the green bar in the loadings) and the predictor matrix (containing the rest of the data) \( \mathbf{Y} \), resulting in the model of the data:

\begin{equation}
\begin{aligned}
\mathbf{Y} = \mathbf{T}\mathbf{P}^T + \mathbf{E}_Y \\
\text{and} \\
\mathbf{B} = \mathbf{T}\mathbf{Q}^T + \mathbf{E}_B \\
\end{aligned}
\end{equation}

This model is estimated in our work as follows:

<!-- 2. Center and scale both \( \mathbf{X} \) and \( \mathbf{Y} \). -->
1. Initialize \( \mathbf{Y}_0 \) to \( \mathbf{Y} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \).
2. Iteratively perform the following steps for \( k = 1, 2, \ldots \):
   a. Compute the singular value decomposition (SVD) of \( \mathbf{Y}_{k}^{T} \mathbf{B}_{k} \) to obtain the left and right singular vectors \( \mathbf{u}_k \) and \( \mathbf{v}_k \) respectively.
   b. Project \( \mathbf{Y}_{k} \) and \( \mathbf{B}_{k} \) onto the first left and right singular vectors (\( \mathbf{u}_k \), \( \mathbf{v}_k \)) to obtain the scores \( \bm{t}_k = \mathbf{Y}_{k} \mathbf{u}_k \) and \( \bm{w}_k = \mathbf{B}_{k} \mathbf{v}_k \).
   c. Perform linear regression of \( \mathbf{Y}_k \) and \( \mathbf{B}_k \) onto their corresponding scores \( \bm{t}_k \) and \( \bm{w}_k \) to obtain the respective loadings \( \bm{p}_k \), and \( \bm{q}_k \).
   d. Deflate \( \mathbf{Y}_k \) and \( \mathbf{B}_k \) by removing the variance explained by the $k^{th}$ pair of scores (\( \bm{t}_k \), \( \bm{w}_k \)), and use the deflated matrices as inputs for the next iteration.

PLS2 aims to maximize the covariance between the scores \( \bm{t}_k \) and \( \bm{w}_k \) at each iteration. In the following simulations and application the number of components calculated by PLS2 is set to 2.

<!-- although there are methods to optimally select this number, such as; (REFS). -->

### Variable importance in projection (VIP)

In this work genes are selected in both ASCA+ and PLS through VIP which is a way to measure the contributions of each variable to the underlying models. According to:

$\text{VIP}_{j} = \sqrt{\frac{ \sum_{k=1}^{K} \mathbf{h}^2_{jk} \cdot \text{VE}_{k}}{\text{VE}_{total} \cdot K}}$

Where k=1...K represents the number of components in the respective models, j indexes a gene,  \(\bm{h}\) is substituted for the loadings or weights (for ASCA+ and PLS respectively) and VE is the variance explained, where $VE_{k}$ is the amount of variance explained by component k and $VE_{total}$ is the variance explained by the whole model. In ASCA+ we calculate the VIP scores in the latent space of \(\bm{Y}_{\alpha + \alpha\beta}\) as described in @thielASCAAPCAExtensions2017. In PLS we use the VE from the latent space of \(\bm{B}\) using the weights ($\mathbf{U}$) from the latent space of \(\bm{Y}\).

### Multivariate ASCA residual analysis (MASCARA)

MASCARA combines the effect estimation with GLM (from ASCA+) with PLS2. The method entails estimating effect matrices with the GLM step of ASCA+ as described above to obtain the estimated residual matrix \(\hat{\mathbf{E}}\). The variance explained by all effect matrices of the experimental factors and interactions is removed. What remains in this \(\hat{\mathbf{E}}\) matrix is individual variance of each plant that is assumed to contain not just random technical variance but some form of structured variance induced by ambient (uncontrolled) effects. The matrix \(\hat{\mathbf{E}}\) is used as input to PLS2  after being split into baits and remainder.

The genes are ranked based on target projection (@kvalheimInterpretationPartialLeast2010) of the loadings $\mathbf{P}$ onto the mean vector of the bait loadings $\mathbf{Q}$ in 2 components: $\mathbf{\mu}_{baits} = \begin{bmatrix} \mathbf{\bar{q}}_1 \\ \mathbf{\bar{q}}_2 \end{bmatrix}$ where $\mathbf{\bar{q}}_k$ is the mean of the $k^{th}$ component of the PLS2 model. The target projected loadings $\mathbf{P}_{TP}$ are calculated such that $\mathbf{P}_{TP} = \left( \frac{\mathbf{P} \cdot \mathbf{\mu}_{baits}}{\|\mathbf{\mu}_{baits}\|^2} \right)$. Higher values in the target projection equate to stronger positive associations with the center of the baits. Large negative values can also be interesting candidates for strong negative associations, although non positive associations are not considered here. This approach allows a general direction in the PLS space to be defined by multiple baits and assumes that these baits are highly correlated with one another. The MASCARA pipeline in exemplified in Figure \@ref(fig:MASCARAOVERVIEW).

```{r MASCARAOVERVIEW, echo = F, fig.cap="MASCARA pipeline overview"}
a <- knitr::include_graphics("MASCA_Pipeline_overview.png")
a

```

## Performance Evaluation

To evaluate performance we use log2 of the geometric mean rank ($\bar{\gamma}$), defined as: 
$\bar{\gamma} = log2((\prod_{i=1}^{n}\gamma_{i})^{1/n})$ 
for each spike *i* for n spikes.

For the real data, since the ground truth is not known we visualise relationships between baits and genes detected by the different methods rather than using a metric.

# Results

```{r, echo = FALSE}
#### VST inverse #######
#mean/sd relationship plots
##Power transformation of counts/concentration matrix
m_sd_n <- M_SD(X, meta)

plot_normal <- m_sd_n[[3]]
```

```{r, echo=FALSE, include=FALSE}
db <- cbind.data.frame("Sample" = c(1:24),meta[,1:2],X)
db[,1:3] <- apply(db[,1:3],2,as.character)
mc <- reshape2::melt(db)
colnames(mc)[4:5] <- c("Feature","Expression")
```

```{r, echo = FALSE}
dist_norm <- ggplot(mc, aes(x = Expression)) + geom_density(alpha = 0.4)+
  scale_y_continuous(labels = percent_format()) +
  theme_bw()+
  ggtitle("Core Distribution")
```

```{r, echo=FALSE}
dist_nb <- ggplot(mc, aes(x = Expression)) + geom_density(alpha = 0.4)+
  scale_y_continuous(labels = percent_format()) +
  scale_x_continuous(guide = guide_axis(check.overlap = TRUE)) +
  theme_bw()+
  ggtitle("NB transformed")

# (dist_norm + plot_normal)  # no noise data so cannot plot mean sd relationships here
```

```{r, echo = FALSE, include = F}
m <- meta
m[,1:2] <- lapply(meta[,1:2],factor)

res_ASCAplus_nb <- ASCA_decompose(d = m[,1:2], x = X_funced_test2[[1]],
                               f = "growth_condition + time + growth_condition:time")
```

```{r, echo = FALSE}
# comb_terms <- ASCA_combine_terms(res_ASCAplus_nb, c("growth_condition","growth_condition:time"))
# svd_test <- ASCA_svd(comb_terms$decomposition$`growth_condition+growth_condition:time`)
#

 minT <-  res_ASCAplus_nb$decomposition$growth_condition + res_ASCAplus_nb$decomposition$`growth_condition:time`
  PCD <- prcomp(minT)

  newdata <- res_ASCAplus_nb$decomposition$growth_condition + res_ASCAplus_nb$decomposition$`growth_condition:time` + res_ASCAplus_nb$residuals

  #this predict method does matrix multiplication of the input (new data) and the rotation (loading) matrix
  #projects the data into the PCA space
  PCDE <- predict(PCD, newdata)
  # PCDE <- cbind(m[,1:2], PCDE)

```

```{r, echo = F}
a_c.PCD <- PCD

######

SL_plot <- function(PCD,meta){

  a_c.PCD <- PCD
  meta <- as.data.frame(meta)
  a_c.PCD$x <- cbind.data.frame(time = meta$time, growth_condition = meta$growth_condition, a_c.PCD$x)

  a_c.PCD$rotation <- cbind.data.frame(paste0("X_",c(1:dim(X)[2])), a_c.PCD$rotation)
  colnames(a_c.PCD$rotation)[1] <- "Feature"
  a_c.PCD$rotation$Feature <- factor(a_c.PCD$rotation$Feature, levels = a_c.PCD$rotation$Feature)

  ## *Recreated* simulated data, scores and loadings in PCs 1 and 2

  t1_rec <- ggplot(a_c.PCD$x, aes(x = time, y = PC1, group=growth_condition, colour=growth_condition)) +
    geom_line() +
    theme_classic() +
    theme(legend.position = "none")

  t2_rec <- ggplot(a_c.PCD$x, aes(x = time, y = PC2, group=growth_condition, colour=growth_condition)) +
    geom_line() +
    theme_classic() +
    theme(legend.position = "none")

  ###

  p1_rec <- ggplot(a_c.PCD$rotation[which(abs(a_c.PCD$rotation$PC1) > 0.05),], aes(x = Feature, y = PC1)) +     #CHANGE to include candidate selection here
    geom_bar(stat="identity") +
    theme_classic() +
    theme(axis.text.x = element_blank())

  p2_rec <- ggplot(a_c.PCD$rotation[which(abs(a_c.PCD$rotation$PC2) > 0.05),], aes(x = Feature, y = PC2)) +
    geom_bar(stat="identity") +
    theme_classic() +
    theme(axis.text.x = element_blank())

  plot <- (t1_rec | t2_rec) / (p1_rec | p2_rec)
  return(plot)

}

######

SL <- SL_plot(a_c.PCD, meta)
```

```{r, echo = F}

ve1 <- paste0(round(summary(a_c.PCD)$importance[2,1:2] * 100, digits = 2), "%")
l <- loadingplot(a_c.PCD$rotation, meta, baits, ve1, ref, aes(x = PC1,y = PC2, colour = Effect))
s <- scoreplot(PCDE, meta, ve1, aes(x= PC1, y = PC2, colour = growth_condition, shape = time))

#+ guide_area()# + plot_layout(design = design)
qq <- qqplot(newdata, meta, aes(x=X_2000, y= X_1999, colour = growth_condition, shape = time))

ASCA_cands <- get_ASCA_cands(a_c.PCD,  distance_calc = T, baits = baits[9:12], ret_candN = 10)
# colnames(ASCA_cands) <- c("VIP", "PC1","PC2")
#kable(head(ASCA_cands,10))

# CAND_TAB <- round(ASCA_cands[1:10,], 4)
CAND_TAB <- cbind.data.frame(rownames(ASCA_cands), ASCA_cands)
CAND_TAB <- CAND_TAB[1:10,]

colnames(CAND_TAB)[1] <- "Feature"

fill <- rep(c("grey95", "grey90"), nrow(CAND_TAB))
fill[which(CAND_TAB$Feature %in% baits)] <- "yellow"

theme1 <- ttheme_default(core = list(
  fg_params = list(fontface=c(rep("plain", nrow(CAND_TAB)))),
  bg_params = list(fill = fill)),
  base_size = 5, padding = unit(c(2, 2), "mm"))

candtab <- gridExtra::tableGrob(CAND_TAB, rows = NULL, theme = theme1)

f <- F1_plot(ASCA_cands, baits, 500)[[2]]
```

```{r,echo = F}
# N_T <- readRDS("Noise_Sim_220909.RDS")

Noise_Tests <- readRDS("Noise_Sim_Coexp_23_07_26_11.RDS")  #Noise_Sim_Coexp_23_06_12.RDS -> updated Beta and VIP for PLS   ##Noise_Sim_Coexp_23_06_14_2.RDS

names(Noise_Tests) <- c("ASCA", "PLS", "Correlations", "MASCARA")

Noise_Tests <- Noise_Tests[c("ASCA", "PLS", "Correlations","MASCARA")]
# names(Noise_Tests) <- ms$Method
Noise_Tests <- lapply(Noise_Tests, narm, keep_col1 = T)

# y <- rep((c(0:20)/33.3333333333333333333333333333333))
# x <-  c(0:10) * 10
# 
# 
# x <- rep(x,each = length(y))
# y <- rep(y, length(unique(x)))
# 
# 
# z <- x*y
#

# y <- rep((c(0:20)/33.3333333333333333333333333333333),40)
# x <-  c(1:10) * 10
# x <- rep(x,each = 84)
# 
# z <- x*y

# y <- rep((c(0:20)/10))
# x <-  c(0:10) * 10 
# 
# 
# x <- rep(x,each = length(y))
# y <- rep(y, length(unique(x)))
# 
# 
# z <- x*y

# y <- rep((c(0:10)/33.3333333333333333333333333333333)*2)
# x <-  c(0:10) * 5
# 
# 
# x <- rep(x,each = length(y))
# y <- rep(y, length(unique(x)))
# 
# 
# z <- x*y

y <- rep((c(0:10)/50))
x <-  c(0:10) * 5  #5

x <- rep(x,each = length(y))
y <- rep(y, length(unique(x)))

z <- x*y

main_lev  <-  z

# E_lev <- c(0:10)/20
ERs <- x

# X_funced <- list()
# names <- c()
# i <- NULL
# l <- 1
# for(i in 1:length(ERs)){
# 
# 
#     
#     names[l] <- paste(ERs[i],main_lev[i], sep = "_")
#     l <- l + 1
#     
#     
# 
#   
#   
# }
```

```{r}
names <- paste(ERs,main_lev, sep = "_")

# Noise_Tests <- lapply(Noise_Tests, function(x) names(x) = names)

i <- NULL
for(i in 1:length(Noise_Tests)){

  colnames(Noise_Tests[[i]]) <- names

}
#########

i <- NULL
RES <- data.frame(matrix())
for(i in 1:1){
  res <- Noise_Tests[[i]][16,]
  RES <- cbind.data.frame(RES,t(res))

}

for(i in 2:4){
  res <- Noise_Tests[[i]][12,]
  RES <- cbind.data.frame(RES,t(res))

}

RES <- RES[,-1]
colnames(RES) <- names(Noise_Tests)
```

```{r, echo = F}
RES <- cbind.data.frame(rownames(RES), RES)
colnames(RES)[1] <- "Cond"

# RES$Noise <- ERs
# RES$Main <- main_lev

#####
# 
# RES$Noise <- gsub("_.*", "",RES$Cond)
# RES$Noise <- gsub("X","",RES$Noise)
# # RES$Resid <- gsub("E_","",RES$Resid)
# 
# # RES$Feature <- gsub(".*_","",RES$Cond)
# 
# RES$Main <- gsub(".*_","", RES$Cond)
# # RES$Main <- gsub("X","", RES$Main)
# 
mRES <- melt(RES)

colnames(mRES)[c(2,3)] <- c("Method","F1")

mRES$Noise <- rep(ERs, 4)
mRES$Main <- rep(main_lev, 4)
```

```{r, echo=FALSE}
# mRES$Noise  <- as.numeric(mRES$Noise)
# mRES$Main  <- as.numeric(mRES$Main)
mRES$F1 <- as.numeric(mRES$F1)

colnames(mRES)[3] <- "Rank_Product"
mRES$Rank_Product <- log2(mRES$Rank_Product)
# colnames(mRES)[5] <- "Rank_Product"

mRES$Size_num <- rep(y, 4)
```

```{r}

mRES <- mRES[-which(mRES$Main == 0),]
mRES <- mRES[-which(mRES$Size_num >= 0.15),]
mRES <- mRES[-which(mRES$Noise >=35),]

max <- max(mRES$Rank_Product)
min <- min(mRES$Rank_Product)

```

```{r, echo=FALSE}

i <- NULL
methods <- unique(mRES$Method)
plots <- list()
for(i in 1:length(methods))
  plots[[i]] <- ggplot(mRES[which(mRES$Method == methods[i]),], aes(x = Noise, y = Size_num))+ #, colour = Rank_Product
  # geom_point( alpha = 1, size = 10, shape = 15) + #position = "jitter",
    geom_tile( aes(fill = Rank_Product)) + #position = "jitter",

  # scale_x_discrete(guide = guide_axis(angle = 90, check.overlap = T))+
  # scale_color_continuous(type = "viridis", limits = c(min,max))+
    scale_fill_continuous(type = "viridis", limits = c(min,max))+

  ggtitle(methods[i]) +
  ylab("Effect Size") + 
  xlab("Num. DEs") +
  theme_bw() +
  theme(plot.title = element_text(size=8))

```

## Simulation 1: Combined effect size and number of differential genes

For this simulation we have some experimental variance for the combined effect and time ($\bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T$ and $\bm{T}_{\beta}\bm{P}_{\beta}^T$ respectively) a fixed amount of random noise \(\bm{E}\) and a fixed amount of structured residual variance $\bm{T}_S\bm{P}_{S}^T$, within which our 4 baits and 12 spikes share a correlation structure. This reflects the situation expected from the real data example - drastic effect caused by nutrient deficiency, which also activates the SL pathway. The pathway genes are expected to share some variance that is independent to the experiment - the 4 baits and 12 spikes load on to the structured part of the noise $\bm{T}_S\bm{P}_{S}^T$.

```{r tc, echo = F, fig.cap= "Results of simulation 1 for all 4 methods. The geometric mean rank as a function of the number of differentially expressed genes (x-axis) and combined effect size (y axis)."}
wrap_plots(plots) + plot_layout(guides = "collect")

# pdf("coexp.pdf")
# wrap_plots(plots) + plot_layout(guides = "collect")
# dev.off()

```

<!-- The results of simulation 1 were expected to show no effect for MASCARA,  due to the capturing and removal of the dominant experimental variance, but a decrease in performance for the other methods with increasing numbers of non-pathway DE genes and increasing combined effect size it becomes more difficult to detect the true coexpression pattern, as this is masked. -->

We can see in Figure \@ref(fig:tc) that MASCARA was neither affected by the presence of an increasing number of non-pathway DE genes nor the size of this effect. Correlations, ASCA and PLS, however, all showed the expected breakdown in performance due to the injection of more non-pathway DE genes. This was the expected result as the between group variance is removed with MASCARA.

## Simulation 2: Replication

The results from above were obtained from datasets with varying noise, with core aspects such as the underlying structure, the number of replicates, the strength of correlation between variables, and the size of the noise remained fixed. With the following, we test the feasibility range of MASCARA with respect to the ratio of 3 key variance parameters namely:

1. Within group correlation structure
2. Between group correlation structure
3. Random noise

Here we set a fixed size of within group correlation structure (\(\|\bm{T}_S\bm{P}_{S}^T\|_F = 0.025\)) as well as number of non-pathway DE genes to 30 (yellow bar size \@ref(fig:simOverview)) across all iterations of this test. We vary the sizes of the other two variance parameters (*d* and *l*). Furthermore we also show the effect of replicate number and demonstrate that MASCARA (along with the other methods) can be applied to data with generic profiles, not limited to the specific structures that were used to demonstrate the differences in coexpression analysis above. Here we create 20 datasets with random combined effect and within group correlation structures.

```{r, echo = F}
#30 random initial scores, similar noise 
#parameters, 3-30 replicates per condition. 
Rep_Tests <- readRDS("Noise_Sim_23_07_29_Replicate_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS
names(Rep_Tests) <- c("ASCA", "PLS","Correlations","MASCARA")

# names(Noise_Tests) <- ms$Method
Rep_Tests <- lapply(Rep_Tests, narm, keep_col1 = T)

names <- c()
l <- 1

# ERs <- c(5,25,50)

y <- rep(c(0.0004,0.002, 0.01),3)

x <- rep(25,9)

nse <- rep(c(0.1,0.2,0.3), each = 3)

DS <- c(1:9)

main_lev <- x*y

ERs <- DS

a <- NULL
for(a in 1: length(ERs)){
  i <- NULL
for(i in 1:20){
  j <- NULL
  for(j in 3:15){
    name <- paste0("Dataset_",i,"_Rep_",j,"_nCERS_",ERs[a])
        name <- paste0("Dataset_",i,"_Rep_",j,"_nCERS_",ERs[a])

    names <- c(names,name)
  }
}

}
```

```{r}

i <- NULL
for(i in 1:length(Rep_Tests)){
  colnames(Rep_Tests[[i]]) <- names
}

#

# Dm <- mean(data.matrix(RF_RES[12,]))
# Em <- mean(data.matrix(COEXP_HIGH_RES[8,]))

# get F1 scores for 12 and 8 respectively for linegraph with error

i <- NULL
RES <- data.frame(matrix())
for(i in 1){
  res <- Rep_Tests[[i]][16,]
  RES <- cbind.data.frame(RES,t(res))

}
for(i in 2:4){
  res <- Rep_Tests[[i]][12,]
  RES <- cbind.data.frame(RES,t(res))

}

RES <- RES[,-1]
colnames(RES) <- names(Rep_Tests)

```

```{r, echo = F}
RES <- cbind.data.frame("Replicate_Number" = gsub(".*Rep_","",rownames(RES)), RES)
RES <- cbind.data.frame("nCERs" = gsub(".*ERS_","",RES$Replicate_Number), RES)
RES$Replicate_Number <- gsub("_nCE.*","",RES$Replicate_Number)

mRES <- melt(RES)
colnames(mRES)[3:4] <- c("Method","F1_Score")
```

```{r, echo = F}

# mRES <- melt(new_res)

# colnames(mRES)[2:3] <- c("Method","F1_Score")
# mRES$F1_Score[which(mRES$F1_Score == 0)] <- 0.001

mRES$Replicate_Number <- as.numeric(mRES$Replicate_Number)#, c(3:30))
```

```{r}

# MR <- mRES
# 
# mRES <- mRES[-which(mRES$Method == "ASCA"),]
orders <- c(7,8,9,4,5,6,1,2,3)

max <- max(log2(mRES$F1_Score) + 1)
i <- NULL
l <- 1
One_Balance <- list()
for(i in orders){

  One_Balance[[i]] <- ggplot(mRES[which(mRES$nCERs == orders[i]),], aes(x = Replicate_Number, y = log2(F1_Score), colour = Method)) +
  # geom_point()+
  scale_color_viridis(discrete = T)+
  # geom_line() +
    # geom_point() +

    stat_summary() +

    # geom_errorbar(mappings = aes(x = Replicate_Number, y = log2(F1_Score)),stat = "summary",
    #               fun.min = function(z) {quantile(z,0.25)},
    #               fun.max = function(z) {quantile(z,0.75)},
    #               fun = median) +
  # geom_smooth( se = F)+    # aes(outfit=fit<<-..y..)
        ylim(0,max ) +
    # ggtitle("") +
  labs(subtitle = paste0("CE = ", main_lev[l], " E = ", nse[l]))+
  # ylab("Rank Product") + 
  theme_bw() +
      theme(axis.title = element_blank())

  l <- l + 1

}
```

```{r}

 Two_Balance <- ggplot(mRES, aes(x = Replicate_Number, y = log2(F1_Score), colour = Method)) +
  # geom_point()+
  scale_color_viridis(discrete = T)+
  # geom_line() +
    # geom_point() +

    stat_summary() +

    # geom_errorbar(mappings = aes(x = Replicate_Number, y = log2(F1_Score)),stat = "summary",
    #               fun.min = function(z) {quantile(z,0.25)},
    #               fun.max = function(z) {quantile(z,0.75)},
    #               fun = median) +
  # geom_smooth( se = F)+    # aes(outfit=fit<<-..y..)
        ylim(0,max ) +
    # ggtitle("") +
  labs(subtitle = paste0("CE = ", main_lev[l], " E = ", nse[l]))+
  # ylab("Rank Product") + 
  theme_bw() +
  facet_wrap(~ nCERs)
      # theme(axis.title = element_blank())

```

```{r repres, echo = F, fig.cap= "Simulation 3 results. Replications and variance type feasibility ranges. X axes; replicate number, y axes log2 transformed rank product. Each facet is created with set parameters for combined effect size ($CE = \\|\\mathbf{T}_{\\alpha + \\alpha\\beta}\\mathbf{P}_{\\alpha + \\alpha\\beta}^T\\|_F$) and random gaussian noise size ($E = \\|\\mathbf{E}\\|_F$). Dots are median performance across 20 random structures."}

# One_Balance# + AUCs #+ plot_layout(guides = "collect")

ps <- patchwork::patchworkGrob(wrap_plots(One_Balance) + plot_layout(guides = "collect"))  #, heights = unit(c(5), c("cm")), widths = c(5))

test_patch <- (One_Balance[[1]] + One_Balance[[2]] + One_Balance[[3]]  + ylab(NULL))
              # (One_Balance[[4]] + One_Balance[[5]] + One_Balance[[6]]  + ylab(NULL))/
              # (One_Balance[[7]] + One_Balance[[8]] + One_Balance[[9]]  + ylab(NULL)) # + plot_layout(guides = "collect")

# One_Balance[[1]] + One_Balance[[2]]/
#   One_Balance[[3]] + plot_layout(guides = "collect")

# ps1 <- grid.arrange(ps, left = "Rank Product", bottom = "Replicate Number                     ")

a <- knitr::include_graphics("C:/Users/fwhite/Documents/replication_res.png")
a

```

<!-- In Figure \@ref(fig:repres)  -->

It was expected that all geometric mean ranks converge towards a minimum with increasing numbers of replicates, see \@ref(fig:repres). Perfect detection of 12 spikes in top 12 candidates is achieved with a log2 geometric mean rank of 2.402955, MASCARA this was able to achieve this however the other methods were perturbed by the presence of the 30 non spike differential genes.With more replicates MASCARA always outperforms competitors. With higher combined effect size (CE) MASCARA outperforms the other methods but its performance is affected by random noise (E) size.

In simulation 2 we fixed the number of differential genes to 30, however this is less than 2% of the total number of genes in each dataset. In the real data example more than 10% of of the total number of genes are differentially expressed. As shown in simulation 1 (Figure \@ref(fig:tc)) higher numbers of DE genes results in higher geometric mean rank of the spikes for correlations, ASCA and PLS.  MASCARA is designed specifically to mitigate the presence of large numbers of differential genes. Here we show that this is achieved.

# Real Data Applications

It is apparent that MASCARA has a particular use case for coexpression analysis in data where there is a lot of experimentally induced between-group variance across many genes as well as some within-group correlation structure shared between pathway genes that is caused by independent non-controlled ambient factors.

Here we illustrate an application of MASCARA to the dataset from @haiderTranscriptomeAnalysisPhosphate2023. For this we take 4 core strigolactone genes (Os11t0587000, Os04t0550600, Os01t0746400 and Os01t0700900 of the 9 used for illustrations in figures \@ref(fig:Dataset1-Overview) and \@ref(fig:ASCA-1)) as baits and investigate their relationship with top candidate genes.

<!-- As well as, detection of 5 other genes known to be involved in the pathway (expression profiles shown in heatmap of Figure \@ref(fig:Dataset1-Overview)). -->

```{r real data import}

X <- readRDS("Data_Haider/Rice_Counts4ASCA.RDS")
meta <- readRDS("Data_Haider/Rice_Meta4ASCA.RDS")
colnames(meta)[c(6,7,11)] <- c("gc","time","growth_condition")

baits <- paste0(c("Os11t0587000","Os04t0550600","Os01t0746400","Os01t0700900"),"-01") 
#D27 CCD7 CCD8 CYP711A2
# baits2 <- c("Os01t0700300", "Os01t0701700")  
baits2 <- c("Os01t0700300-00", "Os02t0817900-01")  # identified methyltransferases

#remove RP samples
X <- X[-which(meta$growth_condition == "PR"),]
meta <- meta[-which(meta$growth_condition == "PR"),]

meta$time <- as.character(meta$time)

meta$time[which(meta$time == "8")] <- "1"
meta$time[which(meta$time == "10")] <- "2"
meta$time[which(meta$time == "14")] <- "3"
meta$time[which(meta$time == "15")] <- "4"

# REF <- readRDS("ref4ASCA_rice.RDS")
ref <- cbind.data.frame(Feature = colnames(X), feature = colnames(X), Baits = colnames(X))

# ref$Baits[-which(ref$Baits %in% c(baits,baits2))] <- "non-bait"
# ref$Baits[which(ref$Baits %in% baits)] <- "bait"
# ref$Baits[which(ref$Baits %in% baits2)] <- "test_bait"

REF <- fread("Rice_Annotation/IRGSP-1.0_representative_annotation_2022-09-01.tsv",sep = "\t", header = T)

ref <- ref[which(ref$Feature %in% REF$Transcript_ID),]
ref <- merge(ref,REF[,c("Transcript_ID","Description")], by.x = "Feature", by.y = "Transcript_ID")

X <- X[,which(colnames(X) %in% ref$Feature)]
ref <- ref[match(colnames(X),ref$Feature),]

SL_ref <- ref[grep("trigolact",ref$Description),]

SL_ref <- ref[grep("trigolact",ref$Description),]
SL_ref <- SL_ref[-which(SL_ref$Feature %in% c("Os01t0701400-01","Os03t0203200-01","Os01t0763200-01","Os04t0668900-01","Os06t0154200-01","Os11t0104300-01","Os08t0250900-01")),]

PSI_ref <-read.table("Rice_Annotation/PSI_genes.txt", sep = "\t", header = T)

# baits <- SL_ref$feature

ref$Baits[-which(ref$Baits %in% c(baits,baits2,SL_ref$Baits))] <- "non_pathway"
ref$Baits[which(ref$Baits %in% baits)] <- "bait"

ref$Baits[which(ref$Baits %in% c(SL_ref$Baits[-which(SL_ref$Baits %in% baits)],baits2))] <- "bait"

ref$Baits[which(ref$Baits != "bait")] <- "Other"
ref$Baits[which(ref$Baits == "bait")] <- "Pathway"

```

```{r}
meta <- meta[order(meta$time),]
meta <- meta[order(meta$gc),]
```

```{r}
X <- X[match(meta$ID, rownames(X)),]

```

```{r, echo = FALSE, include = F}
m <- meta
m[,c("growth_condition","time")] <- lapply(m[,c("growth_condition","time")],factor)

res_ASCAplus_nb <- ASCA_decompose(d = m[,c("growth_condition","time")], x = X,
                               f = "growth_condition + time + growth_condition:time")
```

```{r, echo = FALSE}

 minT <-  res_ASCAplus_nb$decomposition$growth_condition + res_ASCAplus_nb$decomposition$`growth_condition:time`
  PCD <- prcomp(minT)

  newdata <- res_ASCAplus_nb$decomposition$growth_condition + res_ASCAplus_nb$decomposition$`growth_condition:time` + res_ASCAplus_nb$residuals

  PCDE <- predict(PCD, newdata)

      F_norms <- c(do.call(c,lapply(res_ASCAplus_nb$decomposition,norm, type = "F")), 
                 residual = norm(res_ASCAplus_nb$residuals, type = "F"))

```

```{r, echo = F, include = FALSE}
a_c.PCD <- PCD
SL <- SL_plot(a_c.PCD, meta)
```

```{r, echo = F, inlcude = F}

#, fig.cap="Top: distribution and mean sd relationship. Bottom: scores and loadings of the combined effect; growth condition + interaction"

ve1 <- paste0(round(summary(a_c.PCD)$importance[2,1:2] * 100, digits = 2), "%")

loadingplot <- function(df, meta, baits, ve, ref, ...){
  #df is PCD$rotation

  #rematch REF, ref, df
  df <- as.data.frame(df)
  df <- cbind.data.frame(ref,df)

  ggplot(df, ...) +

    xlab(paste0("PC1 ",ve[1])) +
    ylab(paste0("PC2 ",ve[2])) +

    geom_point(data = df[which(df$Baits == "Other"),], alpha = 0.1) +
    geom_point(data = df[which(df$Baits == "Pathway"),], alpha = 1) +
    scale_x_continuous(guide = guide_axis(check.overlap = TRUE)) +
    guides(shape = guide_legend(override.aes = list(size = 0.5))) +
    ggtitle("Loadings") +
    theme_bw()+
        scale_colour_discrete(name=NULL)

}

l <- loadingplot(a_c.PCD$rotation, meta, c(baits,baits2), ve = ve1, ref = ref, aes(x = PC1,y = PC2, colour = Baits, label = Description))

meta2 <- meta
colnames(meta2)[11] <- "g_c"

scoreplot <- function(df, meta, ve, ...){

  #df <- as.data.frame(df)
  df <- cbind.data.frame(meta,df)
  df$time <- factor(df$time, levels = c(1:4))
  df$g_c <- factor(df$g_c)

  ggplot(df, ...)+
    geom_point() +
    xlab(paste0("PC1 ",ve[1])) +
    ylab(paste0("PC2 ",ve[2])) + 
    theme_bw()

}
s <- scoreplot(PCDE, meta2, ve1, aes(x= PC1, y = PC2, colour = g_c, shape = time)) + ggtitle("Scores")

#+ guide_area()# + plot_layout(design = design)
qq <- qqplot(newdata, meta, aes(x=X_2000, y= X_1999, colour = growth_condition, shape = time))

ASCA_cands <- get_ASCA_cands(a_c.PCD)
colnames(ASCA_cands) <- c("VIP", "PC1","PC2")
#kable(head(ASCA_cands,10))

CAND_TAB <- round(ASCA_cands[1:10,], 4)
CAND_TAB <- cbind.data.frame(rownames(CAND_TAB), CAND_TAB)
colnames(CAND_TAB)[1] <- "Feature"

fill <- rep(c("grey95", "grey90"), nrow(CAND_TAB))
fill[which(CAND_TAB$Feature %in% baits)] <- "yellow"

theme1 <- ttheme_default(core = list(
  fg_params = list(fontface=c(rep("plain", nrow(CAND_TAB)))),
  bg_params = list(fill = fill)),
  base_size = 5, padding = unit(c(2, 2), "mm"))

candtab <- gridExtra::tableGrob(CAND_TAB, rows = NULL, theme = theme1)

# (dist_norm + plot_normal)/(s + l)

```

```{r, include = F}
source("DATA_SIM_FUNCS.R")

pcd <- prcomp(res_ASCAplus_nb$residuals)
residual_cands <- get_ASCA_cands(pcd)

#HERE
```

```{r}
# image(cor(res_ASCAplus_nb$residuals[,c(baits,baits2)]))
```

```{r, inlcude = FALSE}

minT <-  res_ASCAplus_nb$decomposition$time
tcd <- prcomp(minT)

newdata <- res_ASCAplus_nb$decomposition$time + res_ASCAplus_nb$residuals
PCDE <- predict(PCD, newdata)

tcd <- prcomp(res_ASCAplus_nb$decomposition$time + res_ASCAplus_nb$residuals)

```

```{r, include = F}
#do permutation tests to see if the correlations in the "non experimentally interesting" data are significant between the strigolactone genes compared to random sets.

pathway <- ref[which(ref$Baits == "Pathway"),1]

p_n_b <- pathway[-which(pathway %in% baits)]

psi_ref <- PSI_ref[which(PSI_ref[,1] %in% colnames(res_ASCAplus_nb$residuals)),]

image(cor(res_ASCAplus_nb$residuals[,c(pathway,psi_ref[,1])]))
```

```{r}
get_ASCA_cands2 <- function(PCD, meta, distance_calc= FALSE, baits = NULL, spikes = NULL, ret_candN = nrow(PCD$rotation)){
  #############

  if(distance_calc== TRUE){

    cands <- ranked_dist(baits,PCD)

  }else{
    absload <- abs(data.matrix(PCD$rotation[,1:2])) %*% diag(summary(PCD)$importance[2,1:2])
    combscore <- rowSums(absload[,1:2])

    orderedload <- cbind(combscore, PCD$rotation[,1:2])
    cands <- as.data.frame(orderedload[order(orderedload[,1], decreasing = T),])
    colnames(cands) <- c("VIP", "PC1","PC2")

    cands <- round(cands[1:ret_candN,], 4)

  }

  return(cands)

}

ASCA_cands2 <- get_ASCA_cands2(a_c.PCD)
ASCA_cands2 <- ASCA_cands2[which(ASCA_cands2[,2] < 0),]
ASCA_cands2 <- ASCA_cands2[order(ASCA_cands2[,2]),]
```

In the real data from @haiderTranscriptomeAnalysisPhosphate2023 some plants were subjected to massive stress with the phosphate limitation. This is therefore reflected by large differences in the transcriptome compared to those with a normal supply of P. This limitation affected many more genes than those directly involved in SL biosynthesis. Furthermore, there appears to be a stronger experiment independent relationship amongst the SL genes compared to their correlations with other DE genes, as displayed in Figure \@ref(fig:residualdists) where the blue distribution indicates the 9 transcripts from figures \@ref(fig:Dataset1-Overview) and \@ref(fig:ASCA-1) and the red is the relationship between these 9 and other top differential genes.

```{r residualdists, fig.cap=c("Distributions of Fisher transformed Pearson correlations in residuals from the ASCA model. Blue indicates correlations between SL pathway genes, red indicates correlations between SL pathway genes and top 10% (non-SL) DE genes; upregulated in P-")}

cor_val <- function(x){

  # y <- y[which(y %in% colnames(x))]

  cor_res <- cor(x[])

  cor_res <- 0.5*log((1+cor_res)/(1-cor_res))

  # res <- (sum(cor_res))   /length(y)^2

  # res <- mean(cor_res[upper.tri(cor_res)])

  return(cor_res)

}

# plot(ASCA_cands2[1:1253,2], ASCA_cands2[1:1253,3])

test <- cor_val(res_ASCAplus_nb$residuals[,pathway])  #res_ASCAplus_nb$residuals
# hist(test[upper.tri(test)])

#define top DE ID vector, with no SL genes
ASCA_cands2 <- ASCA_cands2[-which(rownames(ASCA_cands2) %in% pathway),]
non_pathway <- rownames(ASCA_cands2)[1:1253]

test2 <- cor_val(res_ASCAplus_nb$residuals[,c(pathway,non_pathway)])
test2[upper.tri(test2, diag = T)] <- NA

mt <- melt(test2, na.rm = T)
mt$group <- as.character(mt$Var1)

mt$group[which(mt$Var1 %in% pathway & mt$Var2 %in% pathway)] <- "SL-SL"

mt$group[which(mt$Var1 %in% non_pathway & mt$Var2 %in% pathway)] <- "SL-Other"

mt <- mt[which(mt$group %in% c("SL-SL","SL-Other")),]
colnames(mt)[3] <- "Correlation"
dens <- ggplot(mt, aes(Correlation, colour = group)) + geom_density() + theme_bw()
dens
```

```{r extra plot making}
# ASCA_cands3 <- ASCA_cands2[-which(rownames(ASCA_cands2) %in% c("Os06t0570600-00","Os06t0570566-00", "Os09t0321200-00","Os06t0651900-01", "Os10t0444566-00")),]
# h <- readRDS("heat_comp.RDS")
# levels(h[[2]]$time) <- c("1","3","7","8")
# 
# # make row annotation
# 
# row_ann <- data.frame("Feature" = rep(c("SL","Other DE"),each = 9))
# rownames(row_ann) <- c(pathway,rownames(ASCA_cands3)[1:9])
# 
# ann_colours <- list(g_c = c("P-" = "#F8766D","P+" = "lightblue"),
#                     time = c("1" = "#fde725","3" = "#35b779","7" = "#31688e","8" = "#440154"),
#                     Feature = c("SL" = "purple", "Other DE" = "lightgreen"))
# 
# h[[1]] <- t(X[,c(pathway,rownames(ASCA_cands3)[1:9])])  #psi_ref[,1]
# 
# heat <- pheatmap::pheatmap(t(scale(t(h[[1]]))),  #
#                              cluster_rows = F,
#                              cluster_cols = F,
#                              show_rownames = T,
#                              show_colnames = F,
#                              annotation_col = h[[2]],
#                            annotation_row = row_ann,
#                            annotation_colors = ann_colours,
#                            treeheight_row = 10,
#                            fontsize = 8,
#                              # annotation_row = anno,
#                              main = "Scaled Gene Expression",silent = T)
# 
#   # heat
#   library(ggplotify)
#   heat2 <- as.ggplot(heat) + theme(plot.title = element_text(face = "plain"))
#
#   saveRDS(heat2, "heatmap_MASCARA_Fig_2.RDS")

# 
# qqplot <- function(data, meta, ...){
# 
#   df <- cbind.data.frame(meta,data)
# 
#   df$time <- factor(df$time, levels = unique(meta$time))
#   df$growth_condition <- factor(df$growth_condition)
# 
# 
#   ggplot(df, ...) +
#     geom_point() +
#     theme_bw()
# 
# 
# }
# 
# meta$time <- rep(c(1,3,7,8), each = 3, 2)
# 
# cand_b <- PSI_cands[10]
# cand_w <- baits[3]
# 
# cor_tw <- round(cor(X[,which(colnames(X) %in% cand_w)], X[,which(colnames(X) %in% baits[4])]), digits = 2)
# 
# cor_tb <- round(cor(X[,which(colnames(X) %in% cand_b)], X[,which(colnames(X) %in% baits[4])]), digits = 2)
# 
# cor_rw <- round(cor(res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% cand_w)], res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% baits[4])]), digits = 2)
# 
# cor_rb <- round(cor(res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% cand_b)], res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% baits[4])]), digits = 2)
# 
# 
# within <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% cand_w)],
#                         "Bait" = X[,which(colnames(X) %in% baits[4])]),meta,
#        aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + xlab(baits[4]) + ylab(cand_w) + labs(title = paste0("r = ",cor_tw))
# 
# PSI_cands <- rownames(ASCA_cands[which(rownames(ASCA_cands) %in% psi_ref[,1]),])
# PSI_cands <- head(rownames(ASCA_cands[-which(rownames(ASCA_cands) %in% pathway),]),20)
# 
# 
# between <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% cand_b)],
#                         "Bait" = X[,which(colnames(X) %in% baits[4])]),meta,
#        aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + xlab(baits[4]) + ylab(cand_b) + labs(title = paste0("r = ",cor_tb))
# 
# # within + between + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A')) + plot_layout(heights = unit(c(5), c("cm")), widths = c(5))
# 
# 
# 
# 
# within_r <- qqplot(cbind.data.frame("Candidate" = res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% cand_w)],
#                         "Bait" = res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% baits[4])]),meta,
#        aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + xlab(baits[4]) + ylab(cand_w)   + labs(title = paste0("r = ",cor_rw))#, alpha = (16- as.numeric(growth_condition))
# 
# 
# between_r <- qqplot(cbind.data.frame("Candidate" = res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% cand_b)],
#                         "Bait" = res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% baits[4])]),meta,
#        aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + xlab(baits[4]) + ylab(cand_b)  + labs(title = paste0("r = ",cor_rb))
# 
# (within + between)/(within_r + between_r) + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A','1')) #+ plot_layout(heights = unit(c(5), c("cm")), widths = c(5))

#^^ C:/Users/fwhite/Documents/within_between_example_2.png
```

```{r}
# ggp_rand_cor <- as.data.frame(rand_cor)
# colnames(ggp_rand_cor)[1] <- "Cor"
# dist <- ggplot(ggp_rand_cor) + geom_density(aes(x = Cor))
#

```

```{r}

all <- cbind.data.frame(meta[,c("growth_condition","time")], X)
residuals <- cbind.data.frame(meta[,c("growth_condition","time")], res_ASCAplus_nb$residuals)
```

```{r}
#get cands from residuals

rPCD <- prcomp(data.matrix(residuals[,-c(1:2)]))

# residual_cands_all <- ranked_dist(pathway, rPCD)
# residual_cands_phos <- ranked_dist(PSI_ref[,1], rPCD)

```

```{r}
#filter residuals for only n genes that show treatment effect i.e. set VIP cutoff value
#how to do this with density?...

#first quick and dirty attempt with top n candidates  -  then add density

cutoff <- round(quantile(c(1:nrow(ASCA_cands)),0.1))
res_filt <- residuals[,c(1:2,which(colnames(residuals) %in% rownames(ASCA_cands)[1:cutoff]))]

rpcd <- prcomp(data.matrix(res_filt[,-c(1:2)]))

ref$Baits <- as.character(ref$Baits)
ref$Baits[which(ref$Feature %in% PSI_ref[,1])] <- "PSI"
ref$Baits <- factor(ref$Baits, levels = c("Other","Pathway","PSI"))

Fref <- ref[match(rownames(rpcd$rotation),ref$Feature),]

```

```{r, include = F}
importance <- paste0(round(summary(rpcd)$importance[2,] * 100, digits = 2), "%")
```

```{r}
# resids <- residuals[,-c(1:2)]
resids <- res_filt[,-c(1:2)]
spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], resids[,which(colnames(resids) %in% baits)], all.outputs = T)

sPLS_cands <- cbind.data.frame(abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X, spls_res$loadings$X * spls_res$prop_expl_var$X)
sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = T),]

importance <- paste0(round(spls_res$prop_expl_var$X * 100, digits = 2), "%")

Fref2 <- Fref[-which(Fref$Feature %in% baits),]
Fref2 <- Fref2[match(rownames(sPLS_cands),Fref2$Feature),]

# sPLS_cands$Feature <- rownames(sPLS_cands)
# sPLS_cands <- merge(sPLS_cands, ref, by = "Feature")

# ggplotly(loadingplot(sPLS_cands, meta, baits = Fref2$Feature[which(Fref2$Baits %in% c("Pathway","PSI"))], 
            # importance, PCs = c(1,2), ref = Fref2, aes(x = comp1, y = comp2, colour = Baits, label = Description)))

```

```{r}
#calculating direction of baits to determine direction of candidate selection spls loading space

# loadingplot(sPLS_cands, meta, baits = Fref2$Feature[which(Fref2$Baits %in% c("Pathway","PSI"))], 
#             importance, PCs = c(1,2), ref = Fref2, aes(x = comp1, y = comp2, colour = Baits, label = Description))

angles <- rbind.data.frame(spls_res$loadings$Y, colMeans(spls_res$loadings$Y))

x <- angles[4,]
y <- angles[5,]

#adapted from https://stackoverflow.com/questions/1897704/angle-between-two-vectors-in-r
angle <- function(x,y){

  x <- c(as.matrix(x))
  y <- c(as.matrix(y))

  dot.prod <- x%*%y 
  norm.x <- norm(x,type="2")
  norm.y <- norm(y,type="2")
  theta <- acos(dot.prod / (norm.x * norm.y))
  return(as.numeric(theta))
}

angles_loadings <- apply(sPLS_cands[,-1],1,angle, y = y)

#define this better
sPLS_cands <- cbind.data.frame(sPLS_cands, "angle" = angles_loadings, "score" = 3 * sPLS_cands$`abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X` * (1-angles_loadings))

sPLS_cands <- sPLS_cands[order(sPLS_cands$score, decreasing = T),]
```

```{r}
avg_point <- angles[5,]/5

rot.mat <- matrix(c(0,-1,1,0),nrow = 2, byrow = T)  #90 degree clockwise rotation

rotated_point <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat)

ang <- angle(avg_point,c(1,0))

rot.mat2 <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation
rotated_point2 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat2)

# rot.mat3 <- matrix(c(cos(ang),sin(ang),-sin(ang),cos(ang)),nrow = 2, byrow = T) #anticlockwise rotation
# rotated_point3 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat3)

# ggplot() + 
#   geom_point(data = sPLS_cands, aes(x = comp1, y = comp2), alpha = 0.1) +
#   geom_point(data = sPLS_cands[1:100,], aes(x = comp1, y = comp2)) +
#   geom_segment(data = avg_point, aes(x = 0, y = 0, xend = comp1/10, yend = comp2/10)) +
#   # geom_segment(data = rotated_point, aes(x = 0, y = 0, xend = V1, yend = V2)) +
#   geom_segment(data = rotated_point2, aes(x = 0, y = 0, xend = V1/10, yend = V2/10)) +
#   # geom_segment(data = rotated_point3, aes(x = 0, y = 0, xend = V1, yend = V2)) +
# 
#   # xlim(c(-0.1,0.1))+
#   # ylim(c(-0.1,0.1)) +
# 
#   theme_bw()
#   # geom_smooth(method = "lm")

```

```{r}
#rotated loadings
ang <- angle(avg_point,c(1,0))

rot.mat2 <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation

rotated_point2 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat2)

rotate <- function(x,y){
  x <- c(as.matrix(x))
  x %*% y
}

loadings_r_baits <- t(apply(sPLS_cands[,c("comp1","comp2")],1,rotate, y = rot.mat2))
colnames(loadings_r_baits) <- c("comp1","comp2")
```

```{r}
sPLS_bait_cands <- loadings_r_baits[order(loadings_r_baits[,1], decreasing = T),]

```

```{r}
#U+03C0
#"\u03c0 / 4" #pi over 4

# rotate <- function(x,y){
#   x <- c(as.matrix(x))
#   x %*% y
# }

#loop over rotation matrices from -pi/4 -> pi/4
#resolution of 100 (?)

range.ang <- seq(from = -pi/6, to = pi/6, length.out = 60)

###

i <- NULL
L_R <- data.frame()
LOADINGS_R <- list()

for(i in 1:length(range.ang)){

ang <- range.ang[i]

rot.mat <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation
loadings_r <- t(apply(loadings_r_baits[,c("comp1","comp2")],1,rotate, y = rot.mat))
colnames(loadings_r) <- c("comp1","comp2")
L_R <- rbind(L_R,loadings_r)
  LOADINGS_R[[i]] <- loadings_r

}

# colnames(LOADINGS_R) <- c("comp1","comp2")
L_R$rad <- rep(range.ang, each = nrow(sPLS_cands))
L_R$frame <- rep(c(1:length(range.ang)), each = nrow(sPLS_cands))

```

```{r}
#get sd of original comp1 distribution
sigma <- sd(sPLS_cands$comp1)

big.T.sig <- function(x,y){
  big <- length(which(x > 2.5*y))
  return(big)
}

i <- NULL
res <- c()
for(i in 1:length(LOADINGS_R)){

  r <- big.T.sig(LOADINGS_R[[i]][,1],sigma)
  res <- c(res,r)
}

```

```{r}
RES <- cbind.data.frame("Rad" = range.ang, "Density" = res)

density <- ggplot(RES, aes(x = Rad, y = Density))+
  geom_smooth() + theme_bw() + ggtitle("Number of points above 2*sigma of original rotation")

# density

```

```{r}

#taken from https://stackoverflow.com/questions/6836409/finding-local-maxima-and-minima
inflect <- function(x, threshold = 1){
  up   <- sapply(1:threshold, function(n) c(x[-(seq(n))], rep(NA, n)))
  down <-  sapply(-1:-threshold, function(n) c(rep(NA,abs(n)), x[-seq(length(x), length(x) - abs(n) + 1)]))
  a    <- cbind(x,up,down)
  list(minima = which(apply(a, 1, min) == a[,1]), maxima = which(apply(a, 1, max) == a[,1]))
}

ggres <- ggplot_build(density)$data[[1]]
infs <- inflect(ggres$y)
rads <- ggres$x[infs$maxima]

```

```{r}
maxima.ang <- rads

###

i <- NULL
# L_R <- data.frame()
sPLS_maxima_cands <- list()

for(i in 1:length(maxima.ang)){

ang <- maxima.ang[i]

rot.mat <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation
loadings_r <- as.data.frame(t(apply(loadings_r_baits[,c("comp1","comp2")],1,rotate, y = rot.mat)))
colnames(loadings_r) <- c("comp1","comp2")
# L_R <- rbind(L_R,loadings_r)
  sPLS_maxima_cands[[i]] <- loadings_r[order(loadings_r$comp1, decreasing = T),]

}
```

```{r}
Fref2 <- Fref[-which(Fref$Feature %in% baits),]
Fref2 <- Fref2[match(rownames(sPLS_maxima_cands[[1]]),Fref2$Feature),]
```

```{r}

Fref2 <- Fref[-which(Fref$Feature %in% baits),]
Fref2 <- Fref2[match(rownames(sPLS_maxima_cands[[2]]),Fref2$Feature),]

```

```{r}
coexp_cands <- ranked_coexp(baits,X[,which(colnames(X) %in% colnames(resids))])

residual_cands <- ranked_dist(baits[1:3], rpcd)

residual_coexp <- ranked_coexp(baits, resids)

```

```{r}

#replace this to  get cands from original bait axis plus optimised maxima
sPLS_cands <- sPLS_maxima_cands[[2]]

Methods <- c("sPLS_ASCA","PCA_DIST_ASCA","Correlations_ASCA", "Correlations")

getrank <- function(RES,geneset,NOINFO = 26402){
  # geneset <- 
  # NOINFO = length(geneset) - length(ranks)
  ranks <- which(rownames(RES) %in% geneset)
  if(length(ranks) < length(geneset)){
    ranks <- c(ranks,rep(nrow(RES) + 1, length(geneset) - length(ranks))) #NOINFO
  }

  Ranks <- cbind.data.frame("Geneset" = geneset, "Rank" = ranks)
  return(Ranks)

}

psis <- PSI_ref[which(PSI_ref[,1] %in% colnames(resids)),1]

```

```{r}
GA_ref <- read_xlsx("Rice_Annotation/GA pathway genes-IDs.xlsx")
GA_ref$ID <- gsub("g","t",GA_ref$ID)
GAs <- ref[which(gsub("-.*","",ref$Feature) %in% GA_ref$ID),1]

GAs <- GAs[which(GAs %in% colnames(resids))]
sPLS_ASCA <- rbind.data.frame(getrank(sPLS_cands,p_n_b),getrank(sPLS_cands,psis),getrank(sPLS_cands,GAs))
sPLS_ASCA <- cbind.data.frame(sPLS_ASCA, "Method" = rep("sPLS_ASCA", nrow(sPLS_ASCA)))

#####

PCA_DIST_ASCA <- rbind.data.frame(getrank(residual_cands,p_n_b),getrank(residual_cands,psis),getrank(residual_cands,GAs))
PCA_DIST_ASCA <- cbind.data.frame(PCA_DIST_ASCA, "Method" = rep("PCA_DIST_ASCA", nrow(PCA_DIST_ASCA)))

Correlations <- rbind.data.frame(getrank(coexp_cands,p_n_b),getrank(coexp_cands,psis),getrank(coexp_cands,GAs))
Correlations <- cbind.data.frame(Correlations, "Method" = rep("Correlations", nrow(Correlations)))

Correlations_ASCA <- rbind.data.frame(getrank(residual_coexp,p_n_b),getrank(residual_coexp,psis),getrank(residual_coexp,GAs))
Correlations_ASCA <- cbind.data.frame(Correlations_ASCA, "Method" = rep("Correlations_ASCA", nrow(Correlations_ASCA)))

RES <- rbind.data.frame(sPLS_ASCA, PCA_DIST_ASCA, Correlations_ASCA, Correlations)

RES$Geneset[which(RES$Geneset %in% p_n_b)] <- "SLs"
RES$Geneset[which(RES$Geneset %in% psis)] <- "PSIs"
RES$Geneset[which(RES$Geneset %in% GAs)] <- "GAs"

RES$Method[which(RES$Method == "sPLS_ASCA")] <- "MASCARA"

RES <- RES[which(RES$Method %in% c("Correlations","MASCARA")),]

q <- ggplot(RES[-which(RES$Geneset == "GAs"),], aes(x = Method, y = Rank, fill = Geneset)) + geom_boxplot() + theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust=1))
```

<!-- ## Methods compared -->

```{r, fig.cap="Method Ranks on Real Data"}

library(rhmmer)
REF2 <- as.data.frame(read_tblout("Rice_Annotation/HMM_ITAG_1.txt"))
# unique(REF2$query_name[which(REF2$domain_name %in% rownames(sPLS_cands)[1:9])])

# REF2[which(REF2$domain_name %in% rownames(sPLS_cands)[1:9]),c(1,3,4,19)]

f_p_m <- REF2[which(REF2$domain_name %in% rownames(sPLS_cands)[1:25]),]
f_p_m <- f_p_m[order(f_p_m$sequence_evalue),]
fpm <- f_p_m[!duplicated(f_p_m$domain_name),]
fpm <- fpm[,c(1,3,19)]
fpm[1,3] <- "Diterpenoid phytoalexin biosynthesis"
fpm[12,3] <- "Vesicle-Associated Membrane Protein-Associated Protein"
cands <- rownames(sPLS_cands)[1:25]
cands <- cands[which(cands %in% fpm$domain_name)]
rownames(fpm) <- fpm[,1]
fpm <- fpm[cands,]

```

```{r}
rownames(fpm) <- NULL
# kable(fpm)

```

<!-- Ranked correlations outperforms the other methods at detecting SL genes, however it also detects more general phosphate starvation induced genes (PSIs) as top candidates, due to the strong between correlations.  -->

Some example top candidate genes are shown in Figure \@ref(fig:realcandidates). The between versus within problem is again exemplified here. With several of the top candidates detected by correlations (ASCA and PLS, results not shown) showing no indication of within group correlations. Candidates selected with MASCARA, on the other hand, consistently show within group correlations with the baits. Furthermore, the strigolactone pathway, which is still under active investigation, has several uncharacterised biosynthetic steps. These are thought to be carried out by cytochrome p450s. The top candidates from MASCARA contain at least three (putative) cytochrome p450 genes that have not yet been biologically investigated in SL biosynthesis, as well as, a methyltransferase and several putative carotenoid associated genes. MASCARA also detected genes linked to Gibberellins (GA) preferentially to general phosphate starvation genes. The GA pathway has been linked to the strigolactones (@itoRegulationStrigolactoneBiosynthesis2017, @marzecStrigolactonesGibberellinsNew2017). The interaction between these two pathways has been characterised recently under nitrogen limitation by @sunStrigolactoneGibberellinSignaling2023.

```{r realcandidates, fig.cap="Selected candidates from A-C ranked correlations and D-F MASCARA. X axes; expression level of one of the bait genes (CCD8), y axes; expression levels of three example coexpression candidates for each method."}
# x2 <- X
# 
# X <- res_ASCAplus_nb$residuals

qqplots_spls <- list()
qqplots_coexp <- list()
qqplots_pca <- list()
qqplots_resid_cor <- list()

qqplots_spls_SR <- list()

i <- NULL
for(i in 1:9){
  qqplots_spls[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(sPLS_cands)[i])],
                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + ylab(rownames(sPLS_cands)[i])

  qqplots_coexp[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(coexp_cands)[i])],
                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + ylab(rownames(coexp_cands)[i])

  # qqplots_pca[[i]] <- qqplot(cbind.data.frame("Bait" = X[,which(colnames(X) %in% rownames(residual_cands)[i])],
  #                       "Candidate" = X[,which(colnames(X) %in% baits[1])]),meta, 
  #      aes(x = Bait, y = Candidate, colour = growth_condition, shape = time))
  # 
  qqplots_resid_cor[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(residual_coexp)[i])],
                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + ylab(rownames(residual_coexp)[i])

    # qqplots_spls_SR[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(SR)[i])],
    #                     "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
    #    aes(x = Bait, y = Candidate, colour = growth_condition, shape = time))
    #

}

plot1 <- wrap_plots(qqplots_coexp[c(5,7,8)]) + plot_annotation(title = "Correlations")
plot2 <- wrap_plots(qqplots_spls[c(1,5,8)]) + plot_annotation(title = "MASCARA")

# plot1/plot2 + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A','1'))

a <- knitr::include_graphics("within_between_example.png")
a

```

# Discussion and conclusions

In this article we introduced multivariate ASCA residual analysis (MASCARA) as a method for coexpression analysis in data containing a dominant variance structure from a known (experimental) source. MASCARA estimates effects through the GLM with a design matrix to remove the variance between conditions, then the residual matrix can be decomposed through PLS2, with feature ranking through target projections. This framework enables the investigation of multivariate relationships between a set of baits and the rest of the data. MASCARA thrives on the fact that within the typical experimental setup small systematic variance exists between replicates due to variations in ambient factors that affect plant growth, and therefore the variance structures in the resulting data.

Through simulation studies we showed the conditions under which MASCARA is suitable for coexpression analysis and compared it to a selection of other coexpression methods (ASCA+, PLS2 and correlations). These methods were applied to data from @haiderTranscriptomeAnalysisPhosphate2023 exploring the strigolactone pathway and simulated data based on the structure of the Haider data. In a coexpression context, within certain assumed parameters of variance structure (dominant between-group, existing within-group), MASCARA outperformed competitors. Ranked correlations, ASCA (@smildeANOVAsimultaneousComponentAnalysis2005, @thielASCAAPCAExtensions2017) and PLS (@woldCollinearityProblemLinear1984) are all affected by dominant between-group variance, which is a common occurrence in experimental data, exemplified here by the dataset from @haiderTranscriptomeAnalysisPhosphate2023 but discussed elsewhere (@farahbodUntanglingEffectsCellular2020, @zhaoWeightedGeneCoexpression2010, @chowdhuryDifferentialCoExpressionAnalysis2020).

The feature selection stage of MASCARA uses an average vector constructed from the loadings of the baits onto which to project the loadings of other genes, which is similar to target projection in @kvalheimInterpretationPartialLeast2010 although we construct the target vector from averaging the loadings in a PLS2 model as opposed to already having it defined with the univariate response in PLS1. There are many established methods for feature selection in PLS which include Variable Importance in Projection (VIP) and Selectivity Ratio (@rajalahtiBiomarkerDiscoveryMass2009), however these methods focus on the importance of variables in the construction of the model rather than a directional association with the response block. With the averaged target projection approach we can use the information in a set of baits to define a direction in the latent space which can then be used as an axis to determine the relatedness of other genes. This assumes moderate to high correlation in the residuals of the baits, which is a realistic assumption in coexpression analysis although should be verified in the selection of baits.

Our analysis framework does not consider different correlation structures within the residual variance, this could be an issue especially in the cases similar to @caldanaHighdensityKineticAnalysis2011 where different combinations of experimental factors show differential within-group correlation structure. With MASCARA these could be missed, ideally there would be enough samples to calculate robust correlations per group, to account for possible rewiring under different conditions, but in our case we focus on experiments where there are generally not enough replicates to do this. Experiments with higher replication number are always desirable from the point of view of data analysis. However, this is not always feasible due to time and space constraints. We tested the effects of replication number and found that MASCARA can outperform its competitors with lower replication, especially in data containing many differential genes between experimental conditions. This advantage holds as long as there is a realistic amount of structured variance caused by undocumented ambient factors.

In coexpression analysis, there is no need for class balance, it is in fact better to have more data points in the class where the pathway is induced (P- in the example dataset). When the mechanism of interest is only active in the case condition then the control samples are mostly redundant. With more samples in the condition of interest the reliability of the coexpression analysis will improve. The MASCARA approach indicates that the typical experimental design used for DE is not optimal for coexpression studies. The fact that the residuals need to be isolated to find within group variance confirms the notion that coexpression analysis warrants different experiments than those used for DE studies. If quantification of differential expression is also needed, for example to confirm the pathway under study is affected by the treatment, but the main focus of research is coexpression, then it is an option to have just a few samples in a control condition to confirm the experimental treatment has caused the desired effect. This was not tested explicitly here, however if pathway genes are not active in a given condition then this set of samples does not provide any useful coexpression information.

This work has outlined and tested a basic use of MASCARA where the relationship between baits and unknown pathway genes is predefined to be positive. The method is not limited to the calculation of positive associations, however performance in detecting negative correlations (indicative of inhibitory or negative feedback processes) or more complex relationships has not been tested within the current simulations. Our approach was based off the structure of the Haider RNAseq dataset, however the method is applicable across other omics types. This holds under the assumption that the set of baits have a multivariate relationship between themselves and multiple other (unknown) genes. These genes are also related to the baits, as in our partially characterised strigolactone pathway example. In untargeted metabolomics datasets these assumptions are also likely to be met.

Taken together our results indicate that the lack of ability to discern different types of variance in a dataset can compromise the detection of true coexpression patterns. MASCARA accounts for the different types of (experimental and ambient) variance and enables finer scale detection of within group variance structure.

<!-- Future research on this topic would involve further development of the data simulator.  -->

# Acknowledgements

FW would like to thank Frans van der Kloet, Fentaw Abegaz, Roel van der Ploeg for statistical discussions and Imran Haider for consultation on the strigolactone pathway. FW thanks the Data Science Centre of the University of Amsterdam for partial financial support.

# References

<div id="refs"></div>

# Supplementary

Target projection in PLS2 models has not been described in the literature, we therefore tested our approach against the most common methods for feature selection in PLS namely variable importance in projection (VIP) and selectivity ratio (SR). Here we have similar but reduced simulation set up to the replicate test section; 3 levels of unstructured noise; matrix F-norm sizes (as indicated in the panels). In each of the 3 tests the $\mathbf{T}_S$ is randomised 30 times resulting in different correlation structures between baits and spikes. Target projection using $\mathbf{\bar{q}}$ significantly outperforms the conventional approaches.

```{r}

#30 random initial scores, similar noise 
#parameters, 3-30 replicates per condition. 
Rep_Tests <- readRDS("Noise_Sim_23_07_30_FS_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS
names(Rep_Tests) <- c("TP","SR","VIP","SVD1")

RES <- cbind.data.frame(t(Rep_Tests[[1]][1,]), t(Rep_Tests[[2]][1,]), t(Rep_Tests[[3]][1,]), t(Rep_Tests[[4]][1,]))

RES <- cbind.data.frame("E" = c(rep(c(5,10,15), each = 30)), RES)
colnames(RES) <- c("E", names(Rep_Tests))

mRES <- pivot_longer(RES,c("TP","SR","VIP","SVD1"))
colnames(mRES) <- c("E","Method","RP")

mRES <- mRES[-which(mRES$Method == "SVD1"),]

ggplot(mRES, aes(x = Method, y = RP)) +
  geom_boxplot() +
  facet_grid( ~E)

```

<!-- MASCARA example run on $params: -->

```{r}

```

<!-- Data was simulated: -->

<!-- \begin{equation} -->
<!-- \begin{aligned} -->
<!-- \bm{Y} &= \bm{1\mu}^T + \bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T + \bm{Y}_{\beta} + \bm{E} \\ -->
<!-- \text{where:} &\\ -->
<!-- \bm{T}_{\alpha + \alpha\beta} &= \text{vec}(\bm{ab}^T) \quad \text{where:} \\ -->
<!-- \bm{c}_{i,1:4} &\sim \mathcal{N}(0,1), \quad \bm{a}_{1:r} = [1,\ldots,1], \quad \bm{b}_{1:8} = [\bm{c}_{i},-\bm{c}_{i}]^T \\ -->
<!-- \bm{T}_{\beta} &= \text{vec}(\bm{de}^T) \quad \text{where:} \\ -->
<!-- \bm{f}_{1:4} &\sim \mathcal{N}(0,1.5), \quad \bm{d}_{1:r} = [1,\ldots,1], \quad \bm{e}_{1:8} = [\bm{f},\bm{f}]^T \\ -->
<!-- \bm{P}_{\alpha + \alpha\beta}^{1:2} &= \begin{bmatrix}1.5_{1:4} & 1_{5:8} & 0.5_{9:12} & 0_{13:2000}\end{bmatrix} \\ -->
<!-- \bm{P}_{\beta}^{1:2} &= \begin{bmatrix}0.1_{1:1000} & 0_{1001:2000}\end{bmatrix} \\ -->
<!-- \bm{E} &\sim \mathcal{N}(0,0.75) -->
<!-- \end{aligned}(\#eq:suppl-Sim-bp) -->
<!-- \end{equation} -->

<!-- genes 1:12 have non-zero loadings on PC1 of $svd({\bm{Y}_{\alpha + \alpha\beta}})$, these are considered genes of interest. These 12 genes correspond to the baits and spikes (spikes represent "unknown" genes of interest, used for method evaluation). Which of these genes are baits and which are spikes depends on the methods being tested and will be discussed in the Analysis and Feature Selection section. genes 501:510 have a value of 0.5 on PC2 of $svd({\bm{Y}_{\alpha + \alpha\beta}})$ and are considered to display a nonsense effect. -->

<!-- genes 1:1000 have a loadings of 0.1 on PC1 of $svd({\bm{Y}_\beta})$ and genes 1001:2000 have loadings of 0.1 on PC2 of $svd({\bm{Y}_\beta})$. To create $\bm{Y}$ the dot product of the scores and transposed loadings is calculated per factor, and added to a general mean value for each feature ($\bm{1\mu}^T$) as well as some normally distributed noise \(\bm{E}\) to create differences between the replicates, see equation \@ref(eq:Sim-bp) for an overview. -->

<!-- <!-- \End{multicols} -->

<!-- sim2:  -->
<!-- For this simulation we create 2 components in the combined effect (${\bm{Y}_{\alpha + \alpha\beta}}$), 2 for time ($\bm{Y}_\beta$). For the time effect, we create this similarly to simulation 1. For the combined effect: 2 random orthogonal scores. The loadings for these components contain several sets of genes, some with positive and some with negative loadings onto the 4 components.  -->

<!-- Y = TP (ambient structure) This is regressed onto the concatenated scores of the combined effect and time to ensure orthogonality, creating relationships that are independent of the experimentally induced variance. -->

### 2023-12-04 R/run_first/1.pattern_examples.Rmd
To_Be_Cleaned

---
title: "MASCARA Playground"
author: "Fred White"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(devtools)
library(ggplot2)
library(MetStaT)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)    
library(MUVR) 
library(pheatmap)
library(caret)
library(data.table)
library(plotly)
library(mixOmics)
library(DiscriMiner)
library(pracma)
# library(NineteenEightyR)
library(matlib)
library(truncnorm)
# library(sf)

source("DATA_SIM_FUNCS.R")

```

```{r}

meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
colnames(meta) <- c("growth_condition","time","interaction")
meta <- as.data.frame(apply(meta,2,factor))  #9

Experiment_responders <- 30

set.seed(3)
rnd_seed <- abs(rnorm(3, mean = 0.7, sd = 0.5))
#c(1,0.8,0.3)
# Create_Core_DEV
sim_data <- Create_Core_DEV_2(nreps = 3, meta = meta, plot = TRUE, 
                        EffectSize = c(X_a_ab = 5, time = 1, E = 10, mu = 1, Struc_E = 1), 
                        struc_resid = TRUE, e_sigma = c(1.5,0.8,0.0), a_sigma = c(1,0.7,0.6,0.1),
                        b_sigma = c(1,0.9), irr_spikes = TRUE, SCORE_SEED = 12, Experiment_responders = Experiment_responders,
                        struc_seed = 127)  #12

#list(X,P,Pb,a_ab,PLOT, X_a_ab,time, Struc_E )

###plot error in sim_data[[5]]

```

```{r}

sim_data <- Create_Core_DEV_2(nreps = 3, meta = meta, 
                        EffectSize = c(X_a_ab = 5, time = 1, E = 1, mu = 1, Struc_E = 0.4),    # E = 10, Struc_E = 1
                        struc_resid = TRUE, e_sigma = c(1.5,0.8,0.0), a_sigma = c(1,0.7,0.6,0.1),
                        b_sigma = c(1,0.9), irr_spikes = TRUE, SCORE_SEED = 12, Experiment_responders = Experiment_responders,
                        struc_seed = 127, score_plot = FALSE, plot = FALSE)  #12

# sim_data[[1]] <- sim_data[[1]][,1850:2000]

# image(sim_data)
```

```{r}
baits <- paste0("X_",c(1985:1988))   #501:510
spikes <- paste0("X_",c(1989:2000))

ref <- cbind.data.frame(Feature = colnames(sim_data[[1]]), sim_data[[2]], sim_data[[3]])
cols <- colnames(ref[,-1])
ref$Effect <- do.call(paste, c(ref[cols], sep = "_"))
ref$Effect[which(ref$Effect == "0.5_0_0_0.1")] <- "Low PC1"
ref$Effect[which(ref$Effect == "1_0_0_0.1")] <- "Medium PC1"
ref$Effect[which(ref$Effect == "1.5_0_0_0.1")] <- "High PC1"
ref$Effect[which(ref$Effect == "0_0.5_0.1_0")] <- "PC2"
ref$Effect[grep("^0", ref$Effect)] <- "None"
ref$Effect <- factor(ref$Effect, levels = c("Low PC1", "Medium PC1", "High PC1", "PC2", "None"))
colnames(ref)[8] <- "Baits"

ref$Description <- ref$Baits

ref$Baits[1985:1988] <- "Baits"
ref$Baits[1989:2000] <- "Candidates"

ref$Baits[(2000 - (16 + Experiment_responders)):1984] <- "ab1+ve"
# ref$Baits[1969:1976] <- "ab1+ve"
ref$Baits[501:515] <- "ab2+ve"
ref$Baits[486:500] <- "ab2-ve"
# ref$Baits[1969:1976] <- "Expt_Independent"

ref$Baits[1:24] <- "ab1-ve"

ref$Description <- ref$Baits

ar <- ASCA(sim_data[[1]], m = meta, Baits = baits, spikes = spikes, ncands = ncands, distance_calc = FALSE, Return_Model = FALSE)

```

```{r}

gc <- scale(ar[[4]]$decomposition$growth_condition[,1985:1988])  #0.1*
time <- sim_data$Effects[[2]][,1985:1988] #15 *

int <- scale(sim_data$Effects[[1]][,1985:1988])  #0.08*

ambient <- sim_data$Effects[[4]][,1985:1988]  #1.2 * 
ambient[13:24,] <- 0

noise <- sim_data$Effects[[5]][,1985:1988]  #1.2 *

rownames(gc) <- paste0("X_",1:nrow(gc))
rownames(time) <- paste0("X_",1:nrow(time))
rownames(int) <- paste0("X_",1:nrow(int))
rownames(ambient) <- paste0("X_",1:nrow(ambient))
rownames(noise) <- paste0("X_",1:nrow(noise))

```

```{r}
max <- max(c(gc,time,int,ambient,noise))
min <- min(c(gc,time,int,ambient,noise))

mData <- reshape2::melt(gc)
colnames(mData)[1:2] <- c("Samples","Features")
# mData$Features <- fct_rev(mData$Features)

library(viridis)
library(hrbrthemes)
library(ggplot2)
p1 <- ggplot(mData, aes(Samples, Features, fill= value)) + 
  geom_tile() +
  ggtitle("Growth Condition") +
  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
  theme_bw() +
  theme(axis.text.x=element_blank(), #remove x axis labels
        axis.ticks.x=element_blank(), #remove x axis ticks
        axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

mData <- reshape2::melt(time)
colnames(mData)[1:2] <- c("Samples","Features")
# mData$Features <- fct_rev(mData$Features)

library(viridis)
library(hrbrthemes)
library(ggplot2)
p2 <- ggplot(mData, aes(Samples, Features, fill= value)) + 
  geom_tile() +
  ggtitle('Time') +
  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
  theme_bw() +
  theme(axis.text.x=element_blank(), #remove x axis labels
        axis.ticks.x=element_blank(), #remove x axis ticks
        axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

interaction <- int - time

mData <- reshape2::melt(interaction)
colnames(mData)[1:2] <- c("Samples","Features")
# mData$Features <- fct_rev(mData$Features)

library(viridis)
library(hrbrthemes)
library(ggplot2)
p3 <- ggplot(mData, aes(Samples, Features, fill= value)) + 
  geom_tile() +
  ggtitle('Interaction') +
  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
  theme_bw() +
  theme(axis.text.x=element_blank(), #remove x axis labels
        axis.ticks.x=element_blank(), #remove x axis ticks
        axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

mData <- reshape2::melt(ambient)
colnames(mData)[1:2] <- c("Samples","Features")
# mData$Features <- fct_rev(mData$Features)

library(viridis)
library(hrbrthemes)
library(ggplot2)
p4 <- ggplot(mData, aes(Samples, Features, fill= value)) + 
  geom_tile() +
  ggtitle('Ambient') +
  scale_fill_viridis(discrete=FALSE ) +   #, limits = c(min,max)
  theme_bw() +
  theme(axis.text.x=element_blank(), #remove x axis labels
        axis.ticks.x=element_blank(), #remove x axis ticks
        axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

mData <- reshape2::melt(noise)
colnames(mData)[1:2] <- c("Samples","Features")
# mData$Features <- fct_rev(mData$Features)

library(viridis)
library(hrbrthemes)
library(ggplot2)
p5 <- ggplot(mData, aes(Samples, Features, fill= value)) + 
  geom_tile() +
  ggtitle('Residuals') +
  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
  theme_bw() +
  theme(axis.text.x=element_blank(), #remove x axis labels
        axis.ticks.x=element_blank(), #remove x axis ticks
        axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

data <- (0.1 * interaction) + (1.5*time) + ambient + noise
mData <- reshape2::melt(data)
colnames(mData)[1:2] <- c("Samples","Genes")
# mData$Features <- fct_rev(mData$Features)

library(viridis)
library(hrbrthemes)
library(ggplot2)
p0 <- ggplot(mData, aes(Samples, Genes, fill= value)) + 
  geom_tile() +
  ggtitle('Data') +
  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
  theme_bw() +
  theme(axis.text.x=element_blank(), #remove x axis labels
        axis.ticks.x=element_blank(), #remove x axis ticks
        axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank())#,
        #axis.title.x = element_blank(),
        #axis.title.y = element_blank())

plots <- list(p1, p2, p3, p4, p5)

p5
```

```{r}
ps <- patchwork::patchworkGrob(wrap_plots(plots) + plot_layout(guides = "collect"))  #, heights = unit(c(5), c("cm")), 
ps1 <- grid.arrange(ps)#, left = "Features", bottom = "Replicate Number                     ")

# p5
```

```{r}

mData <- reshape2::melt(noise + ambient)
colnames(mData)[1:2] <- c("Samples","Features")
# mData$Features <- fct_rev(mData$Features)

library(viridis)
library(hrbrthemes)
library(ggplot2)
p5 <- ggplot(mData, aes(Samples, Features, fill= value)) + 
  geom_tile() +
  ggtitle('Residuals') +
  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
  theme_bw() +
  theme(axis.text.x=element_blank(), #remove x axis labels
        axis.ticks.x=element_blank(), #remove x axis ticks
        axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

p5
```

```{r}

# data <- (0.1 * interaction) + (1.5*time) + (0.5*ambient) + (0*noise)
# 
input <- data
# 
# input <- cbind.data.frame(input,input,input)

```

```{r}
# data <- (0.1 * interaction) + (1.5*time) + ambient + noise

# input <- input + rnorm(nrow(input) * ncol(input),mean = 0, sd = 1)
#replace with simulation input
# colnames(input) <- make.names(colnames(input))

spls_res <- simpls.fit(data.matrix(input[,1:2]), data.matrix(input[,-c(1:2)]), ncomp = 2)

scores <- spls_res$scores

X_P <- spls_res$loadings
Y_P <- spls_res$Yloadings

X_TP <- (spls_res$scores %*% t(spls_res$loadings))
Y_TP <- (spls_res$scores %*% t(spls_res$Yloadings))

X_E <- spls_res$residuals[,1,]
Y_E <- spls_res$residuals[,2,]

```

```{r}
####PLS pattern visualisation change this

data <- scores[]    # data.matrix(sim_data[[4]][,3:4])  #
mData <- reshape2::melt(data)
colnames(mData)[1:2] <- c("Samples","Features")
# mData$Features <- fct_rev(mData$Features)

library(viridis)
library(hrbrthemes)
library(ggplot2)
p0 <- ggplot(mData, aes(Features,Samples, fill= value)) + 
  geom_tile() +
  ggtitle('Patterns (T)') +
  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
  theme_bw() +
  theme(axis.text.x=element_blank(), #remove x axis labels
        axis.ticks.x=element_blank(), #remove x axis ticks
        axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none")

# p0

data <- input[,1:2]
mData <- reshape2::melt(data)
colnames(mData)[1:2] <- c("Samples","Features")
# mData$Features <- fct_rev(mData$Features)

library(viridis)
library(hrbrthemes)
library(ggplot2)
p1 <- ggplot(mData, aes(Samples,Features, fill= value)) + 
  geom_tile() +
  ggtitle('Data; X') +
  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
  theme_bw() +
  theme(axis.text.x=element_blank(), #remove x axis labels
        axis.ticks.x=element_blank(), #remove x axis ticks
        axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none")

data <- input[,3:4]
mData <- reshape2::melt(data)
colnames(mData)[1:2] <- c("Samples","Features")
# mData$Features <- fct_rev(mData$Features)

library(viridis)
library(hrbrthemes)
library(ggplot2)
p2 <- ggplot(mData, aes(Samples,Features, fill= value)) + 
  geom_tile() +
  ggtitle('Data; Y') +
  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
  theme_bw() +
  theme(axis.text.x=element_blank(), #remove x axis labels
        axis.ticks.x=element_blank(), #remove x axis ticks
        axis.text.y=element_blank(),  #remove y axis labels
        axis.ticks.y=element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none")

p0 + (p1 / p2) + theme(legend.position = "none")
```

```{r}
MASCARA4_test_1 <- function(resids,ref, baits, spikes, ncands){

  Fref <- ref

  spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], 
                   resids[,which(colnames(resids) %in% baits)], all.outputs = TRUE)

  # spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)], 
  #                  resids[,which(colnames(resids) %in% baits)], all.outputs = TRUE)

  sPLS_cands <- cbind.data.frame(abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X, spls_res$loadings$X) #  * spls_res$prop_expl_var$X
  sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = TRUE),]

  importance <- paste0(round(spls_res$prop_expl_var$X * 100, digits = 2), "%")

  Fref2 <- Fref[-which(Fref$Feature %in% baits),]
  Fref2 <- Fref2[match(rownames(sPLS_cands),Fref2$Feature),]

  #####

  y <- colMeans(spls_res$loadings$Y)

  U <- spls_res$loadings$X

  # UTP <- ((U %*% y)/norm(y, "2")) * y
  # UTP <- ((U %*% y)/dot(y,y)) * y   #_9

  # UTP <- (dot(t(U),y)/dot(y,y)) * y  #_10

  UTP <- (dot(t(U),y)/dot(y,y))

  sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])

  # sPLS_bait_cands <- as.data.frame(UTP[order(UTP[,1], decreasing = TRUE),])  #_9 and below

  sPLS_maxima_cands <- sPLS_bait_cands

  Fref2 <- Fref[-which(Fref$Feature %in% baits),]
  Fref2 <- Fref2[match(rownames(sPLS_maxima_cands),Fref2$Feature),]

  RP <- prod(which(rownames(sPLS_maxima_cands) %in% spikes))^(1/length(spikes))

  return(list(RP, sPLS_maxima_cands,UTP))
}

MASCARA4_test_2 <- function(resids,ref, baits, spikes, ncands){

  Fref <- ref

  spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], 
                   resids[,which(colnames(resids) %in% baits)], all.outputs = TRUE)

  # spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)], 
  #                  resids[,which(colnames(resids) %in% baits)], all.outputs = TRUE)

  sPLS_cands <- cbind.data.frame(abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$Y, spls_res$loadings$X) #  * spls_res$prop_expl_var$X
  sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = TRUE),]

  importance <- paste0(round(spls_res$prop_expl_var$X * 100, digits = 2), "%")

  Fref2 <- Fref[-which(Fref$Feature %in% baits),]
  Fref2 <- Fref2[match(rownames(sPLS_cands),Fref2$Feature),]

  #####

  y <- colMeans(spls_res$loadings$Y)

  U <- spls_res$loadings$X

  # UTP <- ((U %*% y)/norm(y, "2")) * y
  # UTP <- ((U %*% y)/dot(y,y)) * y   #_9

  # UTP <- (dot(t(U),y)/dot(y,y)) * y  #_10

  UTP <- (dot(t(U),y)/dot(y,y))

  sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])

  # sPLS_bait_cands <- as.data.frame(UTP[order(UTP[,1], decreasing = TRUE),])  #_9 and below

  sPLS_maxima_cands <- sPLS_bait_cands

  Fref2 <- Fref[-which(Fref$Feature %in% baits),]
  Fref2 <- Fref2[match(rownames(sPLS_maxima_cands),Fref2$Feature),]

  RP <- prod(which(rownames(sPLS_maxima_cands) %in% spikes))^(1/length(spikes))

  return(list(RP, sPLS_maxima_cands,UTP))
}

```

```{r}
##feature selection alternatives
source("MASCARA_Test_Funcs.R")
#default - target projection onto average Q in 2 dimensions
resids <- ar[[3]]

MASCARA_1 <- MASCARA4_test_1(ar[[3]], baits = baits, spikes = spikes)
which(rownames(MASCARA_1[[2]]) %in% spikes)

# # replacing mixomics with SIMPLS from pls package

#1 component, absolute projections
spls_res2.1 <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
                   resids[,which(colnames(resids) %in% baits)], ncomp = 1)

res_prj <- cbind.data.frame(rownames(spls_res2.1$projection),spls_res2.1$projection)
res_VIP <- res_prj[order(abs(res_prj[,2]), decreasing = TRUE),]
rownames(res_VIP) <- res_VIP[,1]

which(rownames(res_VIP) %in% spikes)

## 2 component target projection
MASCARA_2 <- MASCARA4_test_2(ar[[3]], baits, spikes = spikes)
which(rownames(MASCARA_2[[2]]) %in% spikes)

#PLS1 model with svd(baits)$T as response
MASCARA_3 <- MASCARA4_test_3(ar[[3]], baits, spikes = spikes)
which(rownames(MASCARA_3[[2]]) %in% spikes)

# 
# 
#

#PLS1 model with svd(baits)$T as response and target projected deflation

# MASCARA_4 <- MASCARA4_test_4(ar[[3]], baits, spikes = spikes)
# which(rownames(MASCARA_4[[2]]) %in% spikes)

# #PLS2 model with target projected deflation
# 
# 
#

```

```{r}
resids <- ar[[3]]

spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], 
                 resids[,which(colnames(resids) %in% baits)], all.outputs = TRUE)

spls_res2 <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
                resids[,which(colnames(resids) %in% baits)], ncomp = 2)

spls_res2.1 <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
                   resids[,which(colnames(resids) %in% baits)], ncomp = 1)

res_prj <- cbind.data.frame(rownames(spls_res2.1$projection),spls_res2.1$projection)
res_VIP <- res_prj[order(abs(res_prj[,2]), decreasing = TRUE),]
rownames(res_VIP) <- res_VIP[,1]

which(rownames(res_VIP) %in% spikes)

```

```{r}
source("MASCARA_Test_Funcs.R")
#classic
MASCARA_2 <- MASCARA4_test_2(ar[[3]], baits, spikes = spikes)
which(rownames(MASCARA_2[[2]]) %in% spikes)

#u1 of baits as pls1 response
MASCARA_3 <- MASCARA4_test_3(ar[[3]], baits, spikes = spikes)
which(rownames(MASCARA_3[[2]]) %in% spikes)

#sr
MASCARA_4 <- MASCARA4_test_7(ar[[3]], baits, spikes = spikes)
which(rownames(MASCARA_4[[2]]) %in% spikes)

#vip

MASCARA_5 <- MASCARA4_test_8(ar[[3]], baits, spikes = spikes)
which(rownames(MASCARA_5[[2]]) %in% spikes)

#tpsr

```

```{r}
#plotting qbar TP for demo

# str(MASCARA_2[[4]])

loadings <- MASCARA_2[[4]]$loadings[]
yloadings <- MASCARA_2[[4]]$Yloadings[]

# yloadings[3,2] <- -0.5*yloadings[3,2]   #pure data manipulation for demo purposes

scores <- MASCARA_2[[4]]$scores[]
colnames(scores) <- gsub(" ","_", colnames(scores))

colnames(loadings) <- gsub(" ","_", colnames(loadings))

line <- colMeans(yloadings)

  #adapted from https://stackoverflow.com/questions/1897704/angle-between-two-vectors-in-r
  angle <- function(x,y){

    x <- c(as.matrix(x))
    y <- c(as.matrix(y))

    dot.prod <- x%*%y 
    norm.x <- norm(x,type="2")
    norm.y <- norm(y,type="2")
    theta <- acos(dot.prod / (norm.x * norm.y))
    return(as.numeric(theta))
  }

  #rotated loadings
  ang <- angle(line,c(1,-0.8))

  rot.mat2 <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = TRUE) #clockwise rotation

  rotated_point2 <- as.data.frame(c(as.matrix(line)) %*% rot.mat2)

  rotate <- function(x,y){
    x <- c(as.matrix(x))
    x %*% y
  }

  loadings_r_baits <- t(apply(loadings,1,rotate, y = rot.mat2))

  scores_r_baits <- t(apply(scores,1,rotate, y = rot.mat2))

  colnames(loadings_r_baits) <- c("Comp_1","Comp_2")
  colnames(scores_r_baits) <- c("Comp_1","Comp_2")

scores_r_baits <- cbind.data.frame(scores_r_baits, meta)

yloadings <- t(apply(yloadings,1,rotate, y = rot.mat2))

# loadings_r_baits <- rbind(loadings_r_baits, yloadings)

```

```{r}
line <- colMeans(yloadings)

LINE <- rbind(line, c(0,0))
LINE <- t(LINE)
colnames(LINE) <- c("to","from")
# scores <- scores %*% 
scores <- cbind.data.frame(scores,meta[,1:2])

a <- ggplot(scores_r_baits, aes(x = Comp_1, y = Comp_2)) +
  geom_point(aes(colour = growth_condition, shape = time, size = 20)) +
  ggtitle("Scores") + 
  theme_bw()

colnames(yloadings) <- colnames(loadings_r_baits)

loadings_r_baits <- rbind(loadings_r_baits, yloadings)
loadings_r_baits <- cbind.data.frame("ID" = rownames(loadings_r_baits), loadings_r_baits)
```

```{r}
loadings_r_baits$ID[-which(loadings_r_baits$ID %in% c(baits,spikes))] <- "Other"

loadings_r_baits$ID[which(loadings_r_baits$ID %in% spikes)] <- "Pathway"
loadings_r_baits$ID[which(loadings_r_baits$ID %in% baits)] <- "Baits"
```

```{r}
b <- ggplot(loadings_r_baits, aes(x = Comp_1, y = Comp_2, colour = ID, shape = ID)) +
  geom_point() +
    ggtitle("Loadings") +

  geom_segment(x = 0, y = 0, xend = line[1], yend = line[2], lineend = "round", linewidth = 2, colour = "purple", arrow = arrow()) +    
  scale_shape_manual(values=c(3, 16, 17))+

    scale_color_manual(values=c('#E69F00',"black", '#56B4E9'))+
      geom_point(data = loadings_r_baits[which(loadings_r_baits$ID %in% c("Baits","Pathway")),], aes(size = 15)) +
  # geom_point(data = yloadings, shape = 4, aes(x = Comp_1, y = Comp_2, size = 3 )) + 
  theme_bw()

b
```

```{r}

#multiple PLS1 models with svd(baits) deflated

```

```{r}
#remake simpls to deflate with target projected loadings

#first recreate simpls such that the output is the same as simpls.fit()

#$projection -> T = XR

```

### 2023-12-04 R/run_first/2.plot_pathway_net_example.Rmd
To_Be_Cleaned

---
title: "network_plot_work"
author: "Fred White"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggrepel)
library(readxl)
library(extrafont)
library(cowplot)
library(graphlayouts)
library(gganimate)
library(animation)
```

```{r}
# font_import() # download once 
#windowsFonts() # to check which are loaded
```

```{r}
loadfonts(device = "win", quiet = TRUE)

###partial correlations

library(ppcor)
#library(matlib)
library(psych)

#
# plot residuals and full

baits <- c(baits,spikes[1:5])
spikes <- spikes[-c(1:5)]

residuals <- ar[[3]]
data <- sim_data[[1]]

cand_vals <- residuals[,1970:2000]    #rgs[which(rownames(rgs) %in% names(g_b)[1:50]),]

# cand_vals <- data[,1970:2000]

##CHANGE THIS 
#make p3 interconnections stronger, 
#make p3 connect weakly to p2 through one or two nodes, 
#sparsify p2 and p3 connections

adj <- cor(cand_vals)

rownames(adj) <- colnames(cand_vals)
colnames(adj) <- colnames(cand_vals)

```

```{r}
set.seed(12)
pways <- sum(length(baits),length(spikes))
W <- matrix(sample(c(0,0,0,0,1),pways*pways, replace = T), nrow = pways)

adj[which(rownames(adj) %in% c(baits,spikes)),which(colnames(adj) %in% c(baits,spikes))] <- adj[which(rownames(adj) %in% c(baits,spikes)),which(colnames(adj) %in% c(baits,spikes))] * W

#adj[which(rownames(adj) %in% baits),which(colnames(adj) %in% spikes)] <- adj[which(rownames(adj) %in% baits),which(colnames(adj) %in% c(spikes))] * W
```

```{r}
set.seed(345)
W <- matrix(sample(c(1,0,0,0,0),15*15, replace = T), nrow = 15)
adj[-which(rownames(adj) %in% c(baits,spikes)),-which(colnames(adj) %in% c(baits,spikes))]  <- 0.9

adj[-which(rownames(adj) %in% c(baits,spikes)),-which(colnames(adj) %in% c(baits,spikes))] <- adj[-which(rownames(adj) %in% c(baits,spikes)),-which(colnames(adj) %in% c(baits,spikes))] * W

```

```{r}
#sparsify p1 p2 connections

set.seed(22)

W <- matrix(sample(c(1,0,0,0,0),length(baits) * length(spikes), replace = T), nrow = length(spikes))

adj[which(rownames(adj) %in% spikes),which(colnames(adj) %in% baits)] <- adj[which(rownames(adj) %in% spikes),which(colnames(adj) %in% baits)] * W

```

```{r}

##need to know which nodes are in at least one of the conditions before here

library(igraph)

g <- graph.adjacency(
  as.matrix(as.dist(adj)),
  mode="undirected",
  weighted=TRUE,
  diag=FALSE
)

g <- simplify(g, remove.multiple=TRUE, remove.loops=FALSE)

# Colour negative correlation edges as blue
E(g)[which(E(g)$weight<0)]$association <- "-ve"
# Colour positive correlation edges as red
E(g)[which(E(g)$weight>0)]$association <- "+ve"

# Convert edge weights to absolute values
E(g)$weight <- abs(E(g)$weight)
```

```{r}
# Remove edges    make this general to both types of correlations

###
g <- delete_edges(g, E(g)[which(E(g)$weight<0.55)])

# Remove any vertices remaining that have no edges
g <- delete.vertices(g, degree(g)==0)

# Assign names to the graph vertices (optional)
V(g)$name <- V(g)$name

# Change shape of graph vertices
V(g)$shape <- "sphere"

# Change colour of graph vertices
V(g)$color <- "skyblue"

```

```{r}

coords <- layout_with_kk(g)  #to fix coordinates of nodes

```

```{r}

########
library(tidygraph)
library(ggraph)

#define correctly ordered type type
type <- V(g)$name

# ref$Feature <- gsub("_","",ref$Feature)

```

```{r}
id <- c(1:length(type))

#get metabolite and taxa info

###CHANGE THIS
#define correctly ordered VIP values  -- not actually VIP - replace with ???
# VIP <- g_b[type]

VIP <- rep(5,length(V(g)))
#VIP <-

type[which(type %in% baits)] <- "POI"
type[which(type %in% spikes)] <- "RP"

type[-which(type %in% c("POI","RP"))] <- "URP"

tidynet <- as_tbl_graph(g)

tidynet <- tidynet %>%
  activate(nodes) %>%
  mutate(VIP = VIP)

tidynet <- tidynet %>%
  activate(nodes) %>%
  mutate(type = type)

colours = c("POI" = "purple", "RP" = "orange", "URP" = "lightgreen")   ##adapt this for 3 different pathway types

edge_colours <- c("+ve" = "grey", "-ve" = "blue")

realnet <- tidynet

#edge_linetype <- round(E(g)$weight, digits = 1)

#cols_f <- colorRampPalette(RColorBrewer::brewer.pal(11, 'Spectral'))

# pdf(paste0(Sys.Date(),"_net_test.pdf"))
realplot <- ggraph(tidynet, layout = coords) + 
  geom_edge_link(aes(colour = association, alpha = weight), show.legend=FALSE) + #, linetype = association)) + 
  scale_edge_color_manual(values = edge_colours, guide = "none") +
  #scale_edge_linetype_continuous(values = edge_linetype) +
  geom_node_point(
    aes(size = VIP, color = type), show.legend=FALSE
    #size = 4
  ) +
  geom_node_point(
    aes(color = type)
    #size = 4
  ) +
  scale_color_manual(values = colours) + 
  # geom_node_text(label = id, alpha = 1, family = "mono", fontface = "bold", size = 5)+  #, fontface = "italic"

  theme_void() #+
 # theme(legend.position = "none")
# dev.off()

################################################################################

```

```{r}
#only take nodes and layout from above

#remake graph or replace weights and edges with between group variance added

#remake adj with only names(V(g)) and calculate weights on data not residuals

cand_vals <- data[,which(colnames(data) %in% names(V(g)))]

adj <- cor(cand_vals)

rownames(adj) <- colnames(cand_vals)
colnames(adj) <- colnames(cand_vals)

g <- graph.adjacency(
  as.matrix(as.dist(adj)),
  mode="undirected",
  weighted=TRUE,
  diag=FALSE
)

g <- simplify(g, remove.multiple=TRUE, remove.loops=FALSE)

# Colour negative correlation edges as blue
E(g)[which(E(g)$weight<0)]$association <- "-ve"
# Colour positive correlation edges as red
E(g)[which(E(g)$weight>0)]$association <- "+ve"

# Convert edge weights to absolute values
E(g)$weight <- abs(E(g)$weight)

###
g <- delete_edges(g, E(g)[which(E(g)$weight<0.4)])

# Remove any vertices remaining that have no edges
g <- delete.vertices(g, degree(g)==0)

# Assign names to the graph vertices (optional)
V(g)$name <- V(g)$name

# Change shape of graph vertices
V(g)$shape <- "sphere"

# Change colour of graph vertices
V(g)$color <- "skyblue"

#define correctly ordered type type
type <- V(g)$name

# ref$Feature <- gsub("_","",ref$Feature)

```

```{r}
id <- c(1:length(type))

#get metabolite and taxa info

###CHANGE THIS
#define correctly ordered VIP values  -- not actually VIP - replace with ???
# VIP <- g_b[type]

VIP <- rep(5,length(V(g)))
# #VIP <- 
# 
# 
type[which(type %in% baits)] <- "POI"
type[which(type %in% spikes)] <- "RP"

type[-which(type %in% c("POI","RP"))] <- "URP"

tidynet <- as_tbl_graph(g)

tidynet <- tidynet %>%
  activate(nodes) %>%
  mutate(VIP = VIP)

tidynet <- tidynet %>%
  activate(nodes) %>%
  mutate(type = type)

colours = c("POI" = "purple", "RP" = "orange", "URP" = "lightgreen")   ##adapt this for 3 different pathway types

edge_colours <- c("+ve" = "grey", "-ve" = "blue")

badnet <- tidynet

```

```{r}

badplot <- ggraph(tidynet, layout = coords) + 
  geom_edge_link(aes(colour = association, alpha = weight), show.legend=FALSE) + #, linetype = association)) + 
  scale_edge_color_manual(values = edge_colours, guide = "none") +
  #scale_edge_linetype_continuous(values = edge_linetype) +
  geom_node_point(
    aes(size = VIP, color = type), show.legend=FALSE
    #size = 4
  ) +
  geom_node_point(
    aes(color = type)
    #size = 4
  ) +
  scale_color_manual(values = colours) + 
  # geom_node_text(label = id, alpha = 1, family = "mono", fontface = "bold", size = 5)+  #, fontface = "italic"

  theme_void() #+
 # theme(legend.position = "none")
# dev.off()

  # 
  # 
  # ggraph(tidynet, layout = coords) + 
  # geom_edge_link(aes(colour = association, alpha = weight)) + #, linetype = association)) + 
  # scale_edge_color_manual(values = edge_colours) +
  # #scale_edge_linetype_continuous(values = edge_linetype) +
  # geom_node_point(
  #   aes(size = VIP, color = type),
  #   #size = 4
  # ) +
  # scale_color_manual(values = colours) + 
  # # geom_node_text(label = id, alpha = 1, family = "mono", fontface = "bold", size = 5)+  #, fontface = "italic"
  # 
  # theme_void()# +
  # # theme(legend.position = "none")

```

```{r}

png("networktypes.png", height = 360)
realplot + badplot + plot_annotation("Pathway Types", tag_levels = 'A') + plot_layout(guides = "collect")
dev.off()

```

<!-- Animations  -->

<!-- ```{r} -->
<!-- s50 <- list(badnet, realnet) -->

<!-- ##CHANGE THIS -->
<!-- #adapting from https://www.r-bloggers.com/2021/09/animating-network-evolutions-with-gganimate/ -->
<!-- #losing type type  -->
<!-- xy <- layout_as_dynamic(s50, alpha = 0.2) -->

<!-- nodes_lst <- lapply(1:length(s50), function(i) { -->
<!--   cbind(igraph::as_data_frame(s50[[i]], "vertices"), -->
<!--     x = xy[[i]][, 1], y = xy[[i]][, 2], frame = i -->
<!--   ) -->
<!-- }) -->

<!-- edges_lst <- lapply(1:length(s50), function(i) cbind(igraph::as_data_frame(s50[[i]], "edges"), frame = i)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- edges_lst <- lapply(1:length(s50), function(i) { -->
<!--   edges_lst[[i]]$x <- nodes_lst[[i]]$x[match(edges_lst[[i]]$from, nodes_lst[[i]]$name)] -->
<!--   edges_lst[[i]]$y <- nodes_lst[[i]]$y[match(edges_lst[[i]]$from, nodes_lst[[i]]$name)] -->
<!--   edges_lst[[i]]$xend <- nodes_lst[[i]]$x[match(edges_lst[[i]]$to, nodes_lst[[i]]$name)] -->
<!--   edges_lst[[i]]$yend <- nodes_lst[[i]]$y[match(edges_lst[[i]]$to, nodes_lst[[i]]$name)] -->
<!--   edges_lst[[i]]$id <- paste0(edges_lst[[i]]$from, "-", edges_lst[[i]]$to) -->
<!--   edges_lst[[i]]$status <- TRUE -->
<!--   # edges_lst[[i]]$association <-  -->
<!--   edges_lst[[i]] -->
<!-- }) -->

<!-- head(edges_lst[[1]]) -->
<!-- ``` -->

<!-- ```{r} -->

<!-- ##CHANGE THIS -->
<!-- #need to get weight and association into all_edges -->

<!-- get_attributes <- function(x){ -->
<!--  do.call(cbind,get.edge.attribute(x)) -->
<!-- } -->

<!-- edg_attr <- do.call(rbind,lapply(s50,get_attributes)) -->

<!-- # edg_attr <- edg_attr[!duplicated(edg_attr),] -->

<!-- all_edges <- do.call("rbind", lapply(s50, get.edgelist)) -->
<!-- all_edges <- cbind(all_edges,edg_attr) -->

<!-- all_edges <- all_edges[!duplicated(all_edges), ] -->
<!-- all_edges <- cbind.data.frame(all_edges, paste0(all_edges[, 1], "-", all_edges[, 2])) -->

<!-- colnames(all_edges)[c(1,2,5)] <- c("from","to","id") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ##CHANGE THIS -->
<!-- #weight and association missing from temp -->

<!-- edges_lst <- lapply(1:length(s50), function(i) { -->
<!--   idx <- which(!all_edges[, 5] %in% edges_lst[[i]]$id) -->
<!--   if (length(idx != 0)) { -->
<!--     tmp <- all_edges[idx,]#data.frame(from = all_edges[idx, 1], to = all_edges[idx, 2], id = all_edges[idx, 5]) -->
<!--     tmp$x <- nodes_lst[[i]]$x[match(tmp$from, nodes_lst[[i]]$name)] -->
<!--     tmp$y <- nodes_lst[[i]]$y[match(tmp$from, nodes_lst[[i]]$name)] -->
<!--     tmp$xend <- nodes_lst[[i]]$x[match(tmp$to, nodes_lst[[i]]$name)] -->
<!--     tmp$yend <- nodes_lst[[i]]$y[match(tmp$to, nodes_lst[[i]]$name)] -->
<!--     tmp$frame <- i -->
<!--     tmp$status <- FALSE -->
<!--     edges_lst[[i]] <- rbind(edges_lst[[i]], tmp) -->
<!--   } -->
<!--   edges_lst[[i]] -->
<!-- }) -->

<!-- edges_df <- do.call("rbind", edges_lst) -->
<!-- nodes_df <- do.call("rbind", nodes_lst) -->
<!-- ``` -->

<!-- ```{r} -->

<!-- ##need to apply coordinates form realnet to badnet -->
<!-- gif <- ggplot() + -->
<!--   geom_segment( -->
<!--     data = edges_df, -->
<!--     aes(x = x, xend = xend, y = y, yend = yend, group = id, alpha = status),  -->
<!--     show.legend = FALSE -->
<!--   ) + -->
<!--   geom_point( -->
<!--     data = nodes_df, aes(x, y, group = name, fill = type), -->
<!--     shape = 21, size = 4, show.legend = FALSE -->
<!--   ) + -->
<!--   scale_fill_manual(values = colours) + -->
<!--   scale_alpha_manual(values = c(0, 1)) + -->
<!--   ease_aes("linear") + -->
<!--   transition_states(frame, state_length = 0.5, wrap = FALSE) + -->
<!--   # labs(title = "Wave {closest_state}") + -->
<!--   theme_void() #%>% animate(end_pause = 0) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- anim_save("network_test.gif", gif, end_pause = 0) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ##CHANGE THIS -->
<!-- ##add these parameters into the code snippet above -->
<!-- ggraph(tidynet, layout = coords) +  -->
<!--   geom_edge_link(aes(colour = association, alpha = weight)) + #, linetype = association)) +  -->
<!--   scale_edge_color_manual(values = edge_colours) + -->
<!--   #scale_edge_linetype_continuous(values = edge_linetype) + -->
<!--   geom_node_point( -->
<!--     aes(size = VIP, color = type), -->
<!--     #size = 4 -->
<!--   ) + -->
<!--   scale_color_manual(values = colours) +  -->
<!--   geom_node_text(label = id, alpha = 1, family = "mono", fontface = "bold", size = 5)+  #, fontface = "italic" -->

<!--   theme_void() -->

<!-- ``` -->

### 2023-12-04 R/DATA_SIM_FUNCS.R

@@ -1,0 +1,2747 @@
+#######
+Create_Core <- function(nreps, meta, irr_spikes = TRUE, struc_resid = FALSE, 
+                        a_sigma = c(1,1), b_sigma = c(1.5,1.5), e_sigma = c(1,0.3),
+                        noise_sd = 0.75, EffectSize = c(X_a_ab = 1, time = 1, E = 1, mu = 1), 
+                        SCORE_SEED = 1000, plot = FALSE, score_plot = FALSE){
+  
+  ########### A_AB
+  ##CREATE SCORES FOR PC1 top half
+  set.seed(SCORE_SEED)
+  a_ab_PC1 <- rnorm(length(unique(meta$time)))
+  a_ab_PC1 <- abs(a_ab_PC1)
+  a_ab_PC1 <- a_ab_PC1[order(a_ab_PC1, decreasing = FALSE)]
+  a_ab_PC1_small <- a_ab_PC1
+  
+  ##CREATE SCORES FOR PC2 top half
+  set.seed(1)
+  a_ab_PC2 <- Create_Orthogonal(a_ab_PC1_small)
+  a_ab_PC1 %*% a_ab_PC2 == 0
+  a_ab_PC1 <- rep(a_ab_PC1, each = nreps)
+  a_ab_PC2 <- rep(a_ab_PC2, each = nreps)
+  
+  ##CREATE FULL SCORES
+  a_ab_PC1 <- matrix(c(a_ab_PC1, a_ab_PC1*(-1)))
+  a_ab_PC2 <- matrix(c(a_ab_PC2, a_ab_PC2*(-1)))
+  a_ab <- cbind.data.frame(meta[,1:2], a_ab_PC1, a_ab_PC2)
+  colnames(a_ab)[c(3,4)] <- c("PC1","PC2")
+  a_ab$PC1 %*% a_ab$PC2
+  
+  #CREATE LOADINGS 
+  # a_ab_p1 <- c(rep(0,1988), rep(0.5,4), rep(1,4), rep(1.5,4))
+  a_ab_p1 <- c(rep(0,1984), rep(0.5,6), rep(1,6), rep(1.5,4))
+  
+  
+  if(irr_spikes == FALSE){
+    a_ab_p2 <- c(rep(0,length(a_ab_p1)))
+  }
+  else{
+    a_ab_p2 <- c(rep(0,500), rep(0.5,10), rep(0,1490))
+  }
[…]

### 2023-12-04 R/run_first/3.sim1_CE_size.R

@@ -1,0 +1,363 @@
+library(devtools)
+library(ggplot2)
+library(MetStaT)
+library(gASCA)
+library(patchwork)
+library(readxl)
+library(ggfortify)
+library(ggrepel)
+library(grid)
+library(gridExtra)
+library(tidyr)
+library(tidyverse)
+library(reshape2)
+library(scales)
+library(matrixStats)
+library(DESeq2)
+library(MASS)
+library(glmnet)
+library(foreach)
+library(doParallel)    
+library(MUVR) 
+library(pheatmap)
+library(caret)
+library(mixOmics)
+library(pracma)
+library(pls)
+
+source("DATA_SIM_FUNCS_DEV.R")
+F1_plot <- F1_plot_no_plot
+nreps <- 3
+ncands <- 500
+
+Create_Core_DEV <- Create_Core_DEV_2
+
+sPLSr <- function(X, meta, baits, spikes, ncands, TITLE = NULL){
+  
+  
+  splsR <- spls(X[,-which(colnames(X) %in% baits)],X[,which(colnames(X) %in% baits)], ncomp = 2)
+  
+  # splsR <- simpls.fit(X[,-which(colnames(X) %in% baits)],X[,which(colnames(X) %in% baits)], ncomp = 2)
+  
+  sPLS_cands <- cbind.data.frame(abs(splsR$loadings$X) %*% splsR$prop_expl_var$X, splsR$loadings$X)
+  sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = T),]
+  
+  F1_scores <- F1_plot(sPLS_cands, spikes, ncands, TITLE = paste0(": ",TITLE))
+  
+  return(list(F1_scores,sPLS_cands))
+  
+}
+
+
+
+MASCARA4_test <- function(resids,ref, baits, spikes, ncands){
+  
+  Fref <- ref
+  
+  spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], 
+                   resids[,which(colnames(resids) %in% baits)], all.outputs = T)
+  
+  # spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)], 
+  #                  resids[,which(colnames(resids) %in% baits)], all.outputs = T)
+  
[…]

### 2023-12-04 R/run_first/4.sim2_rep.R

@@ -1,0 +1,355 @@
+library(devtools)
+library(ggplot2)
+library(MetStaT)
+library(gASCA)
+library(patchwork)
+library(readxl)
+library(ggfortify)
+library(ggrepel)
+library(grid)
+library(gridExtra)
+library(tidyr)
+library(tidyverse)
+library(reshape2)
+library(scales)
+library(matrixStats)
+library(DESeq2)
+library(MASS)
+library(glmnet)
+library(foreach)
+library(doParallel)    
+library(MUVR) 
+library(pheatmap)
+library(caret)
+library(mixOmics)
+library(pracma)
+
+source("DATA_SIM_FUNCS_DEV.R")
+
+Create_Core_DEV <- Create_Core_DEV_2
+
+F1_plot <- F1_plot_no_plot
+# nreps <- 3
+alphaN <- 2
+betaN <- 4
+ncands <- 500
+# baits <- paste0("X_",c(1989:1996))   #501:510,
+# baits <- paste0("X_",c(1985:1996))   #501:510
+
+
+
+MASCARA4_test <- function(resids,ref, baits, spikes, ncands){
+  
+  Fref <- ref
+  
+  spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], 
+                   resids[,which(colnames(resids) %in% baits)], all.outputs = T)
+  
+  sPLS_cands <- cbind.data.frame(abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X, spls_res$loadings$X) #  * spls_res$prop_expl_var$X
+  sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = T),]
+  
+  importance <- paste0(round(spls_res$prop_expl_var$X * 100, digits = 2), "%")
+  
+  Fref2 <- Fref[-which(Fref$Feature %in% baits),]
+  Fref2 <- Fref2[match(rownames(sPLS_cands),Fref2$Feature),]
+  
+  #####
+  
+  y <- colMeans(spls_res$loadings$Y)
+  
+  U <- spls_res$loadings$X
+  
+  # UTP <- ((U %*% y)/norm(y, "2")) * y
+  # UTP <- ((U %*% y)/dot(y,y)) * y   #_9
+  
+  # UTP <- (dot(t(U),y)/dot(y,y)) * y  #_10
+  
+  UTP <- (dot(t(U),y)/dot(y,y))
+  
[…]

### 2023-12-04 R/run_first/5.sim3_feature_select.R

@@ -1,0 +1,278 @@
+library(devtools)
+library(ggplot2)
+library(MetStaT)
+library(gASCA)
+library(patchwork)
+library(readxl)
+library(ggfortify)
+library(ggrepel)
+library(grid)
+library(gridExtra)
+library(tidyr)
+library(tidyverse)
+library(reshape2)
+library(scales)
+library(matrixStats)
+library(DESeq2)
+library(MASS)
+library(glmnet)
+library(foreach)
+library(doParallel)    
+library(MUVR) 
+library(pheatmap)
+library(caret)
+library(mixOmics)
+library(pracma)
+
+source("DATA_SIM_FUNCS_DEV.R")
+source("MASCARA_Test_Funcs.R")
+
+Create_Core_DEV <- Create_Core_DEV_2
+
+F1_plot <- F1_plot_no_plot
+# nreps <- 3
+alphaN <- 2
+betaN <- 4
+ncands <- 500
+# baits <- paste0("X_",c(1989:1996))   #501:510,
+# baits <- paste0("X_",c(1985:1996))   #501:510
+
+
+
+
+
+# baits <- paste0("X_",c(1997:2000))
+# spikes <- paste0("X_",c(1985:1996))
+
+baits <- paste0("X_",c(1985:1988))   #501:510
+spikes <- paste0("X_",c(1989:2000))
+
+
+Baits <- c(baits,spikes)
+
+Experiment_responders <- 12
+
+nreps <- 3
+meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = nreps))
+colnames(meta) <- c("growth_condition","time")
+meta$ID <- paste(meta$growth_condition, meta$time, sep = "_")
+
+# sim_data <- Create_Core(nreps = 3, meta, irr_spikes = TRUE, struc_resid = FALSE, 
+#                         a_sigma = c(1,1), b_sigma = c(1.5,1.5), e_sigma = c(1,0.3),
+#                         noise_sd = 0.75, EffectSize = c(X_a_ab = 1, time = 1, E = 1, mu = 1), 
+#                         plot = FALSE)
+
+sim_data <- Create_Core_DEV(nreps = 3, meta = meta, plot = F, 
[…]

### 2023-12-04 R/run_first/2.plot_pathway_net_example.Rmd
2BCleaned

@@
 ```
-
-<!-- Animations  -->
-
-<!-- ```{r} -->
-<!-- s50 <- list(badnet, realnet) -->
-
-<!-- ##CHANGE THIS -->
-<!-- #adapting from https://www.r-bloggers.com/2021/09/animating-network-evolutions-with-gganimate/ -->
-<!-- #losing type type  -->
-<!-- xy <- layout_as_dynamic(s50, alpha = 0.2) -->
-
-<!-- nodes_lst <- lapply(1:length(s50), function(i) { -->
-<!--   cbind(igraph::as_data_frame(s50[[i]], "vertices"), -->
-<!--     x = xy[[i]][, 1], y = xy[[i]][, 2], frame = i -->
-<!--   ) -->
-<!-- }) -->
-
-<!-- edges_lst <- lapply(1:length(s50), function(i) cbind(igraph::as_data_frame(s50[[i]], "edges"), frame = i)) -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- edges_lst <- lapply(1:length(s50), function(i) { -->
-<!--   edges_lst[[i]]$x <- nodes_lst[[i]]$x[match(edges_lst[[i]]$from, nodes_lst[[i]]$name)] -->
-<!--   edges_lst[[i]]$y <- nodes_lst[[i]]$y[match(edges_lst[[i]]$from, nodes_lst[[i]]$name)] -->
-<!--   edges_lst[[i]]$xend <- nodes_lst[[i]]$x[match(edges_lst[[i]]$to, nodes_lst[[i]]$name)] -->
-<!--   edges_lst[[i]]$yend <- nodes_lst[[i]]$y[match(edges_lst[[i]]$to, nodes_lst[[i]]$name)] -->
-<!--   edges_lst[[i]]$id <- paste0(edges_lst[[i]]$from, "-", edges_lst[[i]]$to) -->
-<!--   edges_lst[[i]]$status <- TRUE -->
-<!--   # edges_lst[[i]]$association <-  -->
-<!--   edges_lst[[i]] -->
-<!-- }) -->
-
-<!-- head(edges_lst[[1]]) -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-
-<!-- ##CHANGE THIS -->
-<!-- #need to get weight and association into all_edges -->
-
-<!-- get_attributes <- function(x){ -->
-<!--  do.call(cbind,get.edge.attribute(x)) -->
-<!-- } -->
-
-<!-- edg_attr <- do.call(rbind,lapply(s50,get_attributes)) -->
-
-<!-- # edg_attr <- edg_attr[!duplicated(edg_attr),] -->
-
-<!-- all_edges <- do.call("rbind", lapply(s50, get.edgelist)) -->
-<!-- all_edges <- cbind(all_edges,edg_attr) -->
-
-<!-- all_edges <- all_edges[!duplicated(all_edges), ] -->
-<!-- all_edges <- cbind.data.frame(all_edges, paste0(all_edges[, 1], "-", all_edges[, 2])) -->
-
-<!-- colnames(all_edges)[c(1,2,5)] <- c("from","to","id") -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- ##CHANGE THIS -->
-<!-- #weight and association missing from temp -->
-
-<!-- edges_lst <- lapply(1:length(s50), function(i) { -->
-<!--   idx <- which(!all_edges[, 5] %in% edges_lst[[i]]$id) -->
-<!--   if (length(idx != 0)) { -->
-<!--     tmp <- all_edges[idx,]#data.frame(from = all_edges[idx, 1], to = all_edges[idx, 2], id = all_edges[idx, 5]) -->
-<!--     tmp$x <- nodes_lst[[i]]$x[match(tmp$from, nodes_lst[[i]]$name)] -->
-<!--     tmp$y <- nodes_lst[[i]]$y[match(tmp$from, nodes_lst[[i]]$name)] -->
-<!--     tmp$xend <- nodes_lst[[i]]$x[match(tmp$to, nodes_lst[[i]]$name)] -->
-<!--     tmp$yend <- nodes_lst[[i]]$y[match(tmp$to, nodes_lst[[i]]$name)] -->
-<!--     tmp$frame <- i -->
-<!--     tmp$status <- FALSE -->
-<!--     edges_lst[[i]] <- rbind(edges_lst[[i]], tmp) -->
-<!--   } -->
-<!--   edges_lst[[i]] -->
-<!-- }) -->
-
-<!-- edges_df <- do.call("rbind", edges_lst) -->
-<!-- nodes_df <- do.call("rbind", nodes_lst) -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-
-<!-- ##need to apply coordinates form realnet to badnet -->
-<!-- gif <- ggplot() + -->
-<!--   geom_segment( -->
-<!--     data = edges_df, -->
-<!--     aes(x = x, xend = xend, y = y, yend = yend, group = id, alpha = status),  -->
-<!--     show.legend = FALSE -->
-<!--   ) + -->
-<!--   geom_point( -->
-<!--     data = nodes_df, aes(x, y, group = name, fill = type), -->
-<!--     shape = 21, size = 4, show.legend = FALSE -->
-<!--   ) + -->
-<!--   scale_fill_manual(values = colours) + -->
-<!--   scale_alpha_manual(values = c(0, 1)) + -->
-<!--   ease_aes("linear") + -->
-<!--   transition_states(frame, state_length = 0.5, wrap = FALSE) + -->
-<!--   # labs(title = "Wave {closest_state}") + -->
-<!--   theme_void() #%>% animate(end_pause = 0) -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- anim_save("network_test.gif", gif, end_pause = 0) -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- ##CHANGE THIS -->
-<!-- ##add these parameters into the code snippet above -->
-<!-- ggraph(tidynet, layout = coords) +  -->
-<!--   geom_edge_link(aes(colour = association, alpha = weight)) + #, linetype = association)) +  -->
-<!--   scale_edge_color_manual(values = edge_colours) + -->
-<!--   #scale_edge_linetype_continuous(values = edge_linetype) + -->
-<!--   geom_node_point( -->
-<!--     aes(size = VIP, color = type), -->
-<!--     #size = 4 -->
-<!--   ) + -->
-<!--   scale_color_manual(values = colours) +  -->
-<!--   geom_node_text(label = id, alpha = 1, family = "mono", fontface = "bold", size = 5)+  #, fontface = "italic" -->
-
-<!--   theme_void() -->
-
-<!-- ``` -->

### 2023-12-11 R/Paper_markdown.Rmd
231211

@@
 
-<!-- Need to change this as it is repeated in introduction -->
-
 Experiments in plant transcriptomics are usually designed to induce variation in a pathway of interest (POI).
@@
 # Introduction
-<!-- DE or CoE to understand system + feature selection -->
 Plant transcriptome studies typically involve the use of designed experiments which aim to induce variation in pathways of interest (POIs) by controlling one or more experimental factors.
@@
 DE is used to determine which genes are expressed differently between two or more experimental conditions, highlighting known or novel pathways affected by the experimental conditions.
-CoE, on the other hand, aims at discovering genes that are part of a partially characterized pathway.
+CoE, on the other hand, aims at discovering genes that are part of a partially characterized POI.
 Here, the expression profiles of known genes of the POI (termed "baits") are used to detect novel pathway members with similar expression patterns (here termed "spikes").
@@
 A represents hypothetical biological pathways, B highlights the problem with detecting these pathways when dealing with data from designed experiments.")}
-#load network types image
 
@@
 a
-
 ```
@@
 *A* represents a true hypothetical pathway including a pathway of interest (POI), related pathway (RP) and an unrelated pathway (URP).
-All three types of pathway are assumed to be affected to varying degrees by the experimental design.
+All three types of pathway are assumed to be affected (dis-)proportionally by the experimental design.
+Note; URPs are not necessarily completely unlinked to POIs or RPs, in the biological sense, but are assumed here to be several degrees away.
+Far enough that the effect can be approximated orthogonally to the coexpression, when modelling.
 When dealing with experimentally designed data, we have to consider two key sources of variance: between-group variance and within-group variance.
@@
 It is the excessive between-group variance that leads to the densely connected nature of *B* from \@ref(fig:nettype).
-Figure \@ref(fig:withinbetween) exemplifies between-group and within-group driven correlations.
+Figure \@ref(fig:withinbetween) exemplifies between-group and within-group driven correlations from a real dataset (@haiderTranscriptomeAnalysisPhosphate2023).
 The plots show the relationship between two genes of the POI (*A* and *C*) and the relationship between two genes that are not part of the same pathway (*B* and *D*).
-The total correlations of both pairs of transcripts are high but the second pair’s correlation is driven by the between-group variation and not by correlations within experimental groups.
-The within-group correlation is lower than the total in both cases, but there is a clear trend in one condition (facet *C*: P- condition in red) for the first gene pair.
-Because of this, the within group correlation (r) is higher for the first transcript pair (*C*) than the second (*D*), which makes it possible to discern between genes within a POI or RP and genes that both respond similarly to a treatment (URP).
+The total correlations in both *A* and *B* are high but correlation in *B* is driven by the between-group variation and not by correlations within experimental groups.
+The within-group correlation is lower than the total in both cases, but there is a clear trend in *C*: P- condition in red.
+The within group correlation (r) is higher for the first transcript pair (*C*) than the second (*D*), which makes it possible to discern between genes within a POI or RP and genes that both respond similarly to a treatment (URP).
 Hence, we need to distinguish *total CoE* which is calculated on the total variance and *within-group CoE* where the focus is on the shared variance within experimental groups.
-
-<!-- From a plant biology point of view, the between-group variance and the within-group variance can be quite different and related to different parts of the biological system.
-It is, in fact, expected that there are mechanistic changes that arise from the very different conditions that could alter the coexpression of genes.
--->
-
-<!-- , examples include: @miettinenSecoiridoidPathwayCatharanthus2014;.  -->
 
@@
 
-<!-- This phenomenon is well documented in @caldanaHighdensityKineticAnalysis2011 where an experiment was performed investigating the effects of temperature and light on arabidopsis plants over time.
-In this study the authors investigated variable relationships between metabolites and showed these were strongly dependent on the specific levels of the experimental factors, e.g., shikimic acid and phenylalanine were positively correlated at 21 degrees in high light condition, but negatively correlated at 32 degrees in the dark.
--->
-
-<!-- Between experimental groups different pathways are active, whereas within groups more general (undocumented/uncontrolled) ambient factors such as temperature, light intensity and soil water content can affect the plant and its transcriptome in a concerted fashion.
-In experiments where the intervention causes large between-group differences to many genes, not making the distinction between types of variance can lead to results that contain total CoE associations driven by similar differences between treatment groups.
--->
-
 Considering this within-group variance increases the strength of coexpression analysis, for example, in experiments where the intervention causes large between-group differences in many genes.
@@
 
-<!-- ## Dataset: Experimental Design -->
 ```{r, echo = FALSE, include=FALSE}
@@
 The loading plot shows that although the 9 SL genes are mostly on the outside of the plot, many other genes are as well.
-Thus many differentially expressed genes are found important that are not part of the SL pathway, in other words ASCA does not detect spike SL genes from others that are similarly or more differentially expressed but not part of the SL pathway.
-
-This marks the goal of this paper, which is to detect within-group coexpressed genes with a set of baits preferentially to other differentially expressed genes that are not part of the same pathway (with no within-group structure).
+Thus many differentially expressed genes are found important that are not part of the SL pathway, in other words ASCA does not detect SL (POI) genes from others that are similarly or more differentially expressed but not part of the SL pathway (URPs).
+
+This marks the goal of this paper, which is to detect within-group coexpressed genes (POI/RP) with a set of baits preferentially to other differentially expressed genes that are not part of the same pathway (URPs, with no shared within-group structure).
 
@@
 
-```{r Dataset1-Overview, echo = F, fig.cap=c("Real data overview. RNAseq of rice root. Heatmap of SL pathway and example highly differential genes.")}
+```{r Dataset1-Overview, echo = F, fig.cap=c("Real data overview.
+RNAseq of rice root.
+Heatmap of SL pathway and example highly differential genes.
+Data and preprocessing from @haiderTranscriptomeAnalysisPhosphate2023, genes autoscaled i.e.
+blue indicates no/low expression and red indicates higher expression.")}
 
@@
 
-```{r ASCA-1, echo = F, out.width = '100%', fig.cap=c("Scores and loadings of the ASCA model of the combined effect (condition + condition:time interaction).")}
+```{r ASCA-1, echo = F, out.width = '100%', fig.cap=c("Scores and loadings of the ASCA model of the combined effect (condition + condition:time interaction), model indicates samples and genes relative positions in the designed variance.")}
 knitr::include_graphics("rice_s_l.png")
@@
 
-The simulations are built upon the results of the ASCA analysis of the combined effect of growth condition and the interaction of growth condition and time.
+The simulations are built upon the results of the ASCA analysis of the combined effect of growth condition and the interaction of growth condition and time (see Figure \@ref(fig:ASCA-1) for overview).
 The simulated data is created by simulating score and loading values (as in Figure \@ref(fig:simOverview)) for the combined effect ($\alpha + \alpha\beta$) ) explaining the difference in time profiles of the P+ and P- condition and the main time effect $\beta$, as well as, structured ambient variance ($S$).
@@
 
-The first simulation aims to determine how well the methods perform to find coexpressed genes in the presence of a dominant between-group variance structure and increasing numbers of non-pathway differential genes (genes with high between-group correlation but no within-group correlation).
-The expression levels of differential transcripts is modeled with the scores and loadings of the growth condition and its interaction with time ( $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{P}_{\alpha + \alpha\beta}^T$ ).
-The disturbing time effect is modeled with scores and loadings ( $\mathbf{T}_{\beta}$  and $\mathbf{P}_{\beta}^T$ ).
+The first simulation aims to determine how well the methods perform to find coexpressed genes in the presence of a dominant between-group variance structure and increasing numbers of differential URP genes (genes with high between-group correlation but no within-group correlation).
+The expression levels of differential transcripts is modeled with the scores and loadings of the growth condition and its interaction with time; termed here the combined effect ( $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{P}_{\alpha + \alpha\beta}^T$ ).
+The time effect is modeled with scores and loadings ( $\mathbf{T}_{\beta}$  and $\mathbf{P}_{\beta}^T$ ).
 Also included is some random measurement noise, this is modeled as a matrix $\mathbf{E}$ with normally distributed random values $e_{ij} = \mathcal{N}(0,\sigma^2)$.
 
-On top of the between-group effect there is an additional a correlation structure applied to the baits and spikes to create a model of coexpression that is independent to the variance induced by the experimental factors ($\mathbf{X}_S = \mathbf{T}_S\mathbf{P}_S^T$).
-In this simulation we have a set of 4 baits that have positive loadings for the combined effect along with a variable number of other genes - "Combined Effect Responders".
-For the structured ambient variance; 2 random sample offset vectors in \(\bm{T}_S\) (each column representing an effect caused by ambient conditions) which our baits and spikes contribute to (i.e.
+On top of the between-group effect there is an additional correlation structure applied to the baits and spikes to create a model of coexpression that is independent to the variance induced by the experimental factors ($\mathbf{X}_S = \mathbf{T}_S\mathbf{P}_S^T$).
+This $\mathbf{X}_S$ matrix contains two component sources of variation which links the baits and spikes.
+In this simulation we have a set of 4 baits that have positive loadings for the combined effect along with a variable number of other genes - "Combined Effect Responders", which are the genes from a simulated URP (indicated by the yellow bar in Figure \@ref(fig:simOverview)).
+For the structured ambient variance; 2 random vectors in \(\bm{T}_S\) (each column representing an effect caused by ambient conditions) which our baits and spikes contribute to (i.e.
 green and black bars in Figure \@ref(fig:simOverview) are non zero in \(\bm{P}_S\)).
@@
 Loadings $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$  are simulated to create the baits and spikes.
-As displayed in Figure \@ref(fig:simOverview), the green bar represents 4 baits (or known pathway genes), the black bar represents the 12 spikes (“unknown” genes of interest), yellow represents differentially expressed genes not in SL pathway, and purple; a set of differential genes not in SL pathway with an alternative treatment response.
-
-In $\mathbf{P}_{\beta1}^T$ the baits, spikes and other DE genes along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model.
+As displayed in Figure \@ref(fig:simOverview), the green bar represents 4 baits (or known pathway genes), the black bar represents the 12 spikes (“unknown” genes from POI or RP), yellow represents differential URP genes, and purple; a set of differential URPs with an alternative treatment response.
+
+In $\mathbf{P}_{\beta1}^T$ the baits, spikes and other DE genes (all genes from POI, RP and URP) along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model.
 The remaining elements of $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$ are populated with zeroes.
@@
 Similarly oranges and pinks represent a growth condition independent time profile with four unique values.
-$\mathbf{E}$ represents random unstructured technical noise and the additional $\bm{1\mu}^T$ term in \@ref(eq:Sim2-bp) represents a mean offset for each gene.
+$\mathbf{E}$ represents random unstructured technical noise and the additional $\bm{1\mu}^T$ term in \@ref(eq:Sim2-bp) represents a mean value for each gene.
 
@@
 Colours correspond to the same genes in the different loading vectors (*P*).
-Black; baits (known SL genes), green; spikes (unknown SL genes), yellow; other non-pathway DE genes, purple non-pathway DE genes with a different profile.
+Black; baits (known POI genes), green; spikes (unknown POI genes), yellow; DE URP genes, purple DE URP genes with a different profile.
 In the score vectors (T); Blues, reds, greys and golds represent experimental conditions, oranges and pinks; time.
@@
 \bm{Y} &= \bm{1\mu}^T + \bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\
-\text{where for each element ik or jk:}\\
+\text{where for each element}\\
 &\bm{T}_{S_{ik}} \sim \mathcal{N}(0,1), \quad \bm{P}_{S_{jk}}^{\text{baits,spikes}} \sim \mathcal{N}(0.5,0.2) \\
@@
 
-<!-- ### Combined effect size and number of differential genes -->
-
-<!-- \begin{equation} -->
-<!-- \begin{aligned} -->
-<!-- \bm{Y} &= \bm{1\mu}^T + d(\bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T) + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\ -->
-<!-- \text{where:} &\\ -->
-<!-- \bm{P}_{\alpha + \alpha\beta}^T = \begin{bmatrix} -->
-<!--        \bm{p}_{11} & \bm{p}_{12} \quad \bm{p}_{13} \\ -->
-<!--        \bm{p}_{21} & \bm{p}_{22} \quad \bm{p}_{23} -->
-<!--      \end{bmatrix} -->
-<!-- \end{aligned}(\#eq:nCER) -->
-<!-- \end{equation} -->
-
-The issue to address is the presence of many non-pathway differential genes; those that have strong between group correlations but no within group correlation.
-Here the differential genes relate  to the yellow bar in \@ref(fig:simOverview).
-We incrementally increase the number of non-zero values and decrease the number of 0 values, which in figure \@ref(fig:simOverview) relates to increasing the size of the yellow bar, as well as varying the F-norm of the combined effect matrix through the scalar parameter *d* below;
+The issue to address is the presence of many differential URPs; those that have strong between group correlations but no within group correlation.
+Here these differential genes relate  to the yellow bar in Figure \@ref(fig:simOverview).
+We incrementally increase the number of non-zero values and decrease the number of 0 values, which in figure \@ref(fig:simOverview) relates to increasing the size of the yellow bar, as well as varying the size (as quantified by the Frobenius norm) of the combined effect matrix through the scalar parameter *d* below;
 
@@
 
-<!-- \(\bm{p}_{2}\) remains the same with \(\bm{p}_{21}\) and \(\bm{p}_{23}\) being 0 and \(\bm{p}_{22}\) (purple bar in \@ref(fig:simOverview)) positive. -->
-
 ## Data simulation 2: Replication
 
-In order to evaluate the generalisation of methods and indicate the levels of variance required for optimal performance we conduct a larger scale assay.
-Here we fix the number of non-pathway DE genes to 30 which corresponds (to the yellow bar size) and the size of the within correlation effect matrix (\(\|\bm{T}_S\bm{P}_S^T\|_F = 0.025\)).
-We investigate a range of values for sizes of between variance (*d*) and random noise (*l*) across random structures (which correspond to repeated simulations of randomised values for $\mathbf{T}$ matrices).
+In order to evaluate the methods and indicate the levels of variance required for optimal performance we conduct a larger scale assay.
+Here we fix the number of DE URP genes to 30 which corresponds to the yellow bar size and the size of the within correlation effect matrix (\(\|\bm{T}_S\bm{P}_S^T\|_F = 0.025\)).
+We investigate a range of values for sizes of between variance (*d*) and random noise (*l*) across random structures (which correspond to repeated simulations of randomised values for all $\mathbf{T}$ matrices).
 Parameters were adjusted according to \@ref(eq:Test-2) on 20 randomised $\mathbf{T}$ matrices.
@@
 ```{r, echo = FALSE}
-# X_funced_classic <- Create_Core(nreps = 3, meta = meta, irr_spikes = TRUE,
-#                         struc_resid = FALSE, noise_sd = 0.75, plot = TRUE)   ## sum(X_funced[[1]] != X_core) = 0 with these parameters
 
@@
 
-# alphaN <- 2
-# betaN <- 4
-# nreps <- 5
-# 
-# meta2 <- cbind.data.frame(rep(c(1,-1), each = betaN*nreps), rep(c(1:betaN), each = nreps), rep(c(-1:-betaN,1:betaN), each = nreps))
-# 
-# colnames(meta2) <- c("growth_condition","time","interaction")
-
 X_funced_test <- Create_Core(nreps = 3, meta, irr_spikes = FALSE, struc_resid = FALSE,
@@
 Pearson's correlation coefficient is calculated between all columns of \(\bm{Y}\).
-For example the correlation between a SL gene $j_{sl}$ and a non-SL gene $j_q$ is calculated as:
-
-\begin{equation}
-\begin{aligned}
-  r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}}
-\end{aligned}(\#eq:Pearson)
-\end{equation}
-
-For ranked correlations *r* is calculated between each bait and all other genes, with the lowest rank for the highest correlation.
-Then for each bait the other genes are ranked based on *r*.
+For ranked correlations Pearson's *r* is calculated between each bait and all other genes.
+Ranks are assigned to all non-baits transcripts based on *r* with the lowest rank for the highest correlation.
 These ranks are averaged across all baits, similarly to @poretskyMutRankShinyWebapplication2020.
+
+<!-- For example the correlation between a SL gene $j_{sl}$ and a non-SL gene $j_q$ is calculated as: -->
+
+<!-- \begin{equation} -->
+<!-- \begin{aligned} -->
+<!--   r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}} -->
+<!-- \end{aligned}(\#eq:Pearson) -->
+<!-- \end{equation} -->
 
@@
 
-<!-- This approach provides a general solution to build the ASCA effect matrices for any fixed effects ANOVA model. -->
-
-### PLS regression - canonical mode
-
-Partial Least Squares canonical mode (PLS2) is a multivariate regression technique used for modeling relationships between two sets of variables, \( \mathbf{Y} \) and \( \mathbf{B} \).
-
-After splitting the data into the response matrix (containing only baits) \( \mathbf{B} \) (the green bar in the loadings) and the predictor matrix (containing the rest of the data) \( \mathbf{Y} \), resulting in the model of the data:
+<!-- In this work we calculate the model: -->
+
+### PLS regression
+
+Partial Least Squares (PLS2) is a multivariate regression technique used for modeling relationships between two sets of variables.
+Here we split our data matrix $\mathbf{Y}$ into sub-matrices; $\mathbf{Y} = \left[\mathbf{\tilde{Y}} : \mathbf{B}\right]$
+
+The response matrix \( \mathbf{B} \) contains only the baits (relating to the green bar in the loadings of Figure \@ref(fig:simOverview)) and the predictor matrix \( \mathbf{\tilde{Y}} \) contains the rest of the data, this can then be expressed as:
 
@@
 \begin{aligned}
-\mathbf{Y} = \mathbf{T}\mathbf{P}^T + \mathbf{E}_Y \\
+\mathbf{\tilde{Y}} = \mathbf{T}\mathbf{P}^T + \mathbf{E}_\tilde{Y} \\
 \text{and} \\
@@
 
-This model is estimated in our work as follows:
-
-<!-- 2. Center and scale both \( \mathbf{X} \) and \( \mathbf{Y} \). -->
-1. Initialize \( \mathbf{Y}_0 \) to \( \mathbf{Y} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \).
+This model is estimated in our work using the SIMPLS algorithm (@dejongSIMPLSAlternativeApproach1993) as follows:
+
+1. Initialize \( \mathbf{\tilde{Y}}_0 \) to \( \mathbf{\tilde{Y}} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \).
 2. Iteratively perform the following steps for \( k = 1, 2, \ldots \):
-   a. Compute the singular value decomposition (SVD) of \( \mathbf{Y}_{k}^{T} \mathbf{B}_{k} \) to obtain the left and right singular vectors \( \mathbf{u}_k \) and \( \mathbf{v}_k \) respectively.
+   a.
+Compute the singular value decomposition (SVD) of \( \mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k} \) to obtain the left and right singular vectors \( \mathbf{u}_k \) and \( \mathbf{v}_k \) respectively.
    b.
-Project \( \mathbf{Y}_{k} \) and \( \mathbf{B}_{k} \) onto the first left and right singular vectors (\( \mathbf{u}_k \), \( \mathbf{v}_k \)) to obtain the scores \( \bm{t}_k = \mathbf{Y}_{k} \mathbf{u}_k \) and \( \bm{w}_k = \mathbf{B}_{k} \mathbf{v}_k \).
+Project \( \mathbf{\tilde{Y}}_{k} \) and \( \mathbf{B}_{k} \) onto the first left and right singular vectors (\( \mathbf{u}_k \), \( \mathbf{v}_k \)) to obtain the scores \( \bm{t}_k = \mathbf{\tilde{Y}}_{k} \mathbf{u}_k \) and \( \bm{w}_k = \mathbf{B}_{k} \mathbf{v}_k \).
    c.
-Perform linear regression of \( \mathbf{Y}_k \) and \( \mathbf{B}_k \) onto their corresponding scores \( \bm{t}_k \) and \( \bm{w}_k \) to obtain the respective loadings \( \bm{p}_k \), and \( \bm{q}_k \).
+Perform linear regression of \( \mathbf{\tilde{Y}}_k \) and \( \mathbf{B}_k \) onto their corresponding scores \( \bm{t}_k \) and \( \bm{w}_k \) to obtain the respective loadings \( \bm{p}_k \), and \( \bm{q}_k \).
    d.
-Deflate \( \mathbf{Y}_k \) and \( \mathbf{B}_k \) by removing the variance explained by the $k^{th}$ pair of scores (\( \bm{t}_k \), \( \bm{w}_k \)), and use the deflated matrices as inputs for the next iteration.
+Deflate \( \mathbf{\tilde{Y}}_k \) and \( \mathbf{B}_k \) by removing the variance explained by the $k^{th}$ set of scores in the space of $\mathbf{\tilde{Y}}$\( \bm{t}_k \) and use the deflated matrices as inputs for the next iteration.
+
+\begin{equation}
+\begin{aligned}
+\mathbf{\tilde{Y}_{k}} = \mathbf{\tilde{Y}}_{k-1} - \mathbf{t}_{k}\mathbf{p}^T_{k} \\
+\text{and} \\
+\mathbf{B_{k}} = \mathbf{B}_{k-1} - \mathbf{t}_{k}\mathbf{q}^T_{k} \\
+\end{aligned}
+\end{equation}
 
 PLS2 aims to maximize the covariance between the scores \( \bm{t}_k \) and \( \bm{w}_k \) at each iteration.
-In the following simulations and application the number of components calculated by PLS2 is set to 2.
-
-<!-- although there are methods to optimally select this number, such as; (REFS). -->
+In the following simulations and application the number of components calculated by PLS2 is set to 2 to create an interpretable model.
 
@@
 
-$\text{VIP}_{j} = \sqrt{\frac{ \sum_{k=1}^{K} \mathbf{h}^2_{jk} \cdot \text{VE}_{k}}{\text{VE}_{total} \cdot K}}$
-
-Where k=1...K represents the number of components in the respective models, j indexes a gene,  \(\bm{h}\) is substituted for the loadings or weights (for ASCA+ and PLS respectively) and VE is the variance explained, where $VE_{k}$ is the amount of variance explained by component k and $VE_{total}$ is the variance explained by the whole model.
+$\text{VIP}_{j} = \sqrt{\frac{ \sum_{k=1}^{K} \mathbf{H}^2_{jk} \cdot \text{VE}_{k}}{\text{VE}_{total} \cdot K}}$
+
+Where k=1...K represents the number of components in the respective models, j indexes a gene,  \(\bm{H}\) is substituted for the loadings or weights (for ASCA+ and PLS respectively) and VE is the variance explained, where $VE_{k}$ is the amount of variance explained by component k and $VE_{total}$ is the variance explained by the whole model.
 In ASCA+ we calculate the VIP scores in the latent space of \(\bm{Y}_{\alpha + \alpha\beta}\) as described in @thielASCAAPCAExtensions2017.
@@
 
-The genes are ranked based on target projection (@kvalheimInterpretationPartialLeast2010) of the loadings $\mathbf{P}$ onto the mean vector of the bait loadings $\mathbf{Q}$ in 2 components: $\mathbf{\mu}_{baits} = \begin{bmatrix} \mathbf{\bar{q}}_1 \\ \mathbf{\bar{q}}_2 \end{bmatrix}$ where $\mathbf{\bar{q}}_k$ is the mean of the $k^{th}$ component of the PLS2 model.
-The target projected loadings $\mathbf{P}_{TP}$ are calculated such that $\mathbf{P}_{TP} = \left( \frac{\mathbf{P} \cdot \mathbf{\mu}_{baits}}{\|\mathbf{\mu}_{baits}\|^2} \right)$.
+The genes are ranked based on target projection (@kvalheimInterpretationPartialLeast2010) of the loadings $\mathbf{P}$ onto the mean vector of the bait loadings $\mathbf{Q}$ in 2 components: $\mathbf{\bar{q}} = \begin{bmatrix} \bar{q}_1 \\ \bar{q}_2 \end{bmatrix}$ where $\bar{q}_k$ is the mean of the $k^{th}$ component of the PLS2 model.
+The target projected loadings $\mathbf{P}_{TP}$ are calculated such that $\mathbf{P}_{TP} = \left( \frac{\mathbf{P} \cdot \mathbf{\bar{q}}}{\|\mathbf{\bar{q}}\|^2} \right)$.
 Higher values in the target projection equate to stronger positive associations with the center of the baits.
@@
 ```{r MASCARAOVERVIEW, echo = F, fig.cap="MASCARA pipeline overview"}
-a <- knitr::include_graphics("MASCA_Pipeline_overview.png")
+a <- knitr::include_graphics("MASCARA_Pipeline_overview.png")
 a
-
-```
-
-## Performance Evaluation
-
-To evaluate performance we use log2 of the geometric mean rank ($\bar{\gamma}$), defined as: 
-$\bar{\gamma} = log2((\prod_{i=1}^{n}\gamma_{i})^{1/n})$ 
-for each spike *i* for n spikes.
-
-For the real data, since the ground truth is not known we visualise relationships between baits and genes detected by the different methods rather than using a metric.
+```
+
+## Performance evaluation; Log2 geometric mean rank
+
+To evaluate performance we use log2 geometric mean rank (GMR) ($\bar{\gamma}$), defined as: 
+$\bar{\gamma} = log2((\prod_{j_{spike~}=1}^{J_{spike}}\gamma_{j_{spike}})^{1/J_{spike}})$ 
+for each spike $j_{spike}$ for $J_{spike}$ spikes. A lower $\gamma$ indicates better performance.
 
 # Results
-
-```{r, echo = FALSE}
-#### VST inverse #######
-#mean/sd relationship plots
-##Power transformation of counts/concentration matrix
-m_sd_n <- M_SD(X, meta)
-
-plot_normal <- m_sd_n[[3]]
-```
-
-```{r, echo=FALSE, include=FALSE}
-db <- cbind.data.frame("Sample" = c(1:24),meta[,1:2],X)
-db[,1:3] <- apply(db[,1:3],2,as.character)
-mc <- reshape2::melt(db)
-colnames(mc)[4:5] <- c("Feature","Expression")
-```
-
-```{r, echo = FALSE}
-dist_norm <- ggplot(mc, aes(x = Expression)) + geom_density(alpha = 0.4)+
-  scale_y_continuous(labels = percent_format()) +
-  theme_bw()+
-  ggtitle("Core Distribution")
-```
-
-```{r, echo=FALSE}
-dist_nb <- ggplot(mc, aes(x = Expression)) + geom_density(alpha = 0.4)+
-  scale_y_continuous(labels = percent_format()) +
-  scale_x_continuous(guide = guide_axis(check.overlap = TRUE)) +
-  theme_bw()+
-  ggtitle("NB transformed")
-
-# (dist_norm + plot_normal)  # no noise data so cannot plot mean sd relationships here
-```
 
@@
 ```{r, echo = FALSE}
-# comb_terms <- ASCA_combine_terms(res_ASCAplus_nb, c("growth_condition","growth_condition:time"))
-# svd_test <- ASCA_svd(comb_terms$decomposition$`growth_condition+growth_condition:time`)
-#
 
@@
 
-######
-
-SL_plot <- function(PCD,meta){
-
-  a_c.PCD <- PCD
-  meta <- as.data.frame(meta)
-  a_c.PCD$x <- cbind.data.frame(time = meta$time, growth_condition = meta$growth_condition, a_c.PCD$x)
-
-  a_c.PCD$rotation <- cbind.data.frame(paste0("X_",c(1:dim(X)[2])), a_c.PCD$rotation)
-  colnames(a_c.PCD$rotation)[1] <- "Feature"
-  a_c.PCD$rotation$Feature <- factor(a_c.PCD$rotation$Feature, levels = a_c.PCD$rotation$Feature)
-
-  ## *Recreated* simulated data, scores and loadings in PCs 1 and 2
-
-  t1_rec <- ggplot(a_c.PCD$x, aes(x = time, y = PC1, group=growth_condition, colour=growth_condition)) +
-    geom_line() +
-    theme_classic() +
-    theme(legend.position = "none")
-
-  t2_rec <- ggplot(a_c.PCD$x, aes(x = time, y = PC2, group=growth_condition, colour=growth_condition)) +
-    geom_line() +
-    theme_classic() +
-    theme(legend.position = "none")
-
-  ###
-
-  p1_rec <- ggplot(a_c.PCD$rotation[which(abs(a_c.PCD$rotation$PC1) > 0.05),], aes(x = Feature, y = PC1)) +     #CHANGE to include candidate selection here
-    geom_bar(stat="identity") +
-    theme_classic() +
-    theme(axis.text.x = element_blank())
-
-  p2_rec <- ggplot(a_c.PCD$rotation[which(abs(a_c.PCD$rotation$PC2) > 0.05),], aes(x = Feature, y = PC2)) +
-    geom_bar(stat="identity") +
-    theme_classic() +
-    theme(axis.text.x = element_blank())
-
-  plot <- (t1_rec | t2_rec) / (p1_rec | p2_rec)
-  return(plot)
-
-}
-
-######
-
-SL <- SL_plot(a_c.PCD, meta)
 ```
@@
 
-#+ guide_area()# + plot_layout(design = design)
 qq <- qqplot(newdata, meta, aes(x=X_2000, y= X_1999, colour = growth_condition, shape = time))
@@
 ASCA_cands <- get_ASCA_cands(a_c.PCD,  distance_calc = T, baits = baits[9:12], ret_candN = 10)
-# colnames(ASCA_cands) <- c("VIP", "PC1","PC2")
-#kable(head(ASCA_cands,10))
-
-# CAND_TAB <- round(ASCA_cands[1:10,], 4)
+
 CAND_TAB <- cbind.data.frame(rownames(ASCA_cands), ASCA_cands)
@@
 
-f <- F1_plot(ASCA_cands, baits, 500)[[2]]
+# f <- F1_plot(ASCA_cands, baits, 500)[[2]]
 ```
@@
 ```{r,echo = F}
-# N_T <- readRDS("Noise_Sim_220909.RDS")
-
-Noise_Tests <- readRDS("Noise_Sim_Coexp_23_07_26_11.RDS")  #Noise_Sim_Coexp_23_06_12.RDS -> updated Beta and VIP for PLS   ##Noise_Sim_Coexp_23_06_14_2.RDS
+
+Noise_Tests <- readRDS("Noise_Sim_Coexp_23_07_26_11.RDS")
 
@@
 Noise_Tests <- Noise_Tests[c("ASCA", "PLS", "Correlations","MASCARA")]
-# names(Noise_Tests) <- ms$Method
 Noise_Tests <- lapply(Noise_Tests, narm, keep_col1 = T)
 
-# y <- rep((c(0:20)/33.3333333333333333333333333333333))
-# x <-  c(0:10) * 10
-# 
-# 
-# x <- rep(x,each = length(y))
-# y <- rep(y, length(unique(x)))
-# 
-# 
-# z <- x*y
-#
-
-# y <- rep((c(0:20)/33.3333333333333333333333333333333),40)
-# x <-  c(1:10) * 10
-# x <- rep(x,each = 84)
-# 
-# z <- x*y
-
-# y <- rep((c(0:20)/10))
-# x <-  c(0:10) * 10 
-# 
-# 
-# x <- rep(x,each = length(y))
-# y <- rep(y, length(unique(x)))
-# 
-# 
-# z <- x*y
-
-# y <- rep((c(0:10)/33.3333333333333333333333333333333)*2)
-# x <-  c(0:10) * 5
-# 
-# 
-# x <- rep(x,each = length(y))
-# y <- rep(y, length(unique(x)))
-# 
-# 
-# z <- x*y
-
 y <- rep((c(0:10)/50))
-x <-  c(0:10) * 5  #5
+x <-  c(0:10) * 5
 
@@
 y <- rep(y, length(unique(x)))
-
 z <- x*y
@@
 
-# E_lev <- c(0:10)/20
 ERs <- x
-
-# X_funced <- list()
-# names <- c()
-# i <- NULL
-# l <- 1
-# for(i in 1:length(ERs)){
-# 
-# 
-#     
-#     names[l] <- paste(ERs[i],main_lev[i], sep = "_")
-#     l <- l + 1
-#     
-#     
-# 
-#   
-#   
-# }
 ```
@@
 names <- paste(ERs,main_lev, sep = "_")
-
-# Noise_Tests <- lapply(Noise_Tests, function(x) names(x) = names)
 
@@
 for(i in 1:length(Noise_Tests)){
-
   colnames(Noise_Tests[[i]]) <- names
-
-}
-#########
+}
 
@@
   RES <- cbind.data.frame(RES,t(res))
-
 }
@@
 
-# RES$Noise <- ERs
-# RES$Main <- main_lev
-
-#####
-# 
-# RES$Noise <- gsub("_.*", "",RES$Cond)
-# RES$Noise <- gsub("X","",RES$Noise)
-# # RES$Resid <- gsub("E_","",RES$Resid)
-# 
-# # RES$Feature <- gsub(".*_","",RES$Cond)
-# 
-# RES$Main <- gsub(".*_","", RES$Cond)
-# # RES$Main <- gsub("X","", RES$Main)
-# 
 mRES <- melt(RES)
@@
 ```{r, echo=FALSE}
-# mRES$Noise  <- as.numeric(mRES$Noise)
-# mRES$Main  <- as.numeric(mRES$Main)
+
 mRES$F1 <- as.numeric(mRES$F1)
 
-colnames(mRES)[3] <- "Rank_Product"
-mRES$Rank_Product <- log2(mRES$Rank_Product)
-# colnames(mRES)[5] <- "Rank_Product"
+colnames(mRES)[3] <- "GMR"
+mRES$GMR <- log2(mRES$GMR)
+# colnames(mRES)[5] <- "GMR"
 
@@
 ```{r}
-
 mRES <- mRES[-which(mRES$Main == 0),]
@@
 
-max <- max(mRES$Rank_Product)
-min <- min(mRES$Rank_Product)
+max <- max(mRES$GMR)
+min <- min(mRES$GMR)
 
@@
 ```{r, echo=FALSE}
-
 i <- NULL
@@
 for(i in 1:length(methods))
-  plots[[i]] <- ggplot(mRES[which(mRES$Method == methods[i]),], aes(x = Noise, y = Size_num))+ #, colour = Rank_Product
-  # geom_point( alpha = 1, size = 10, shape = 15) + #position = "jitter",
-    geom_tile( aes(fill = Rank_Product)) + #position = "jitter",
-
-  # scale_x_discrete(guide = guide_axis(angle = 90, check.overlap = T))+
-  # scale_color_continuous(type = "viridis", limits = c(min,max))+
+  plots[[i]] <- ggplot(mRES[which(mRES$Method == methods[i]),], aes(x = Noise, y = Size_num))+ 
+    geom_tile( aes(fill = GMR)) + #position = "jitter",
+
     scale_fill_continuous(type = "viridis", limits = c(min,max))+
@@
   theme(plot.title = element_text(size=8)) 
-
 ```
@@
 This reflects the situation expected from the real data example - drastic effect caused by nutrient deficiency, which also activates the SL pathway.
-The pathway genes are expected to share some variance that is independent to the experiment - the 4 baits and 12 spikes load on to the structured part of the noise $\bm{T}_S\bm{P}_{S}^T$.
+The POI/RP genes are expected to share some variance that is independent to the experiment - the 4 baits and 12 spikes load on to the structured part of the noise $\bm{T}_S\bm{P}_{S}^T$.
 
 ```{r tc, echo = F, fig.cap= "Results of simulation 1 for all 4 methods.
-The geometric mean rank as a function of the number of differentially expressed genes (x-axis) and combined effect size (y axis)."}
+Log2 geometric mean rank (GMR) as a function of the number of differentially expressed genes (x-axis) and combined effect size (y axis)."}
 wrap_plots(plots) + plot_layout(guides = "collect")
@@
 
-<!-- The results of simulation 1 were expected to show no effect for MASCARA,  due to the capturing and removal of the dominant experimental variance, but a decrease in performance for the other methods with increasing numbers of non-pathway DE genes and increasing combined effect size it becomes more difficult to detect the true coexpression pattern, as this is masked.
--->
-
 We can see in Figure \@ref(fig:tc) that MASCARA was neither affected by the presence of an increasing number of non-pathway DE genes nor the size of this effect.
-Correlations, ASCA and PLS, however, all showed the expected breakdown in performance due to the injection of more non-pathway DE genes.
+Correlations, ASCA and PLS, however, all showed the expected breakdown in performance due to the injection of more DE URP genes.
 This was the expected result as the between group variance is removed with MASCARA.
@@
 
-# ERs <- c(5,25,50)
-
 y <- rep(c(0.0004,0.002, 0.01),3)      
-
 x <- rep(25,9)
-
 nse <- rep(c(0.1,0.2,0.3), each = 3)
-
 DS <- c(1:9)
@@
 ```{r}
-
 i <- NULL
@@
 }
-
-#
-
-# Dm <- mean(data.matrix(RF_RES[12,]))
-# Em <- mean(data.matrix(COEXP_HIGH_RES[8,]))
-
-# get F1 scores for 12 and 8 respectively for linegraph with error
 
@@
   RES <- cbind.data.frame(RES,t(res))
-
-}
+}
+
 for(i in 2:4){
@@
 colnames(RES) <- names(Rep_Tests)
-
 ```
@@
 ```{r, echo = F}
-
-# mRES <- melt(new_res)
-
-# colnames(mRES)[2:3] <- c("Method","F1_Score")
-# mRES$F1_Score[which(mRES$F1_Score == 0)] <- 0.001
-
 mRES$Replicate_Number <- as.numeric(mRES$Replicate_Number)#, c(3:30))
@@
 
-# MR <- mRES
-# 
-# mRES <- mRES[-which(mRES$Method == "ASCA"),]
 orders <- c(7,8,9,4,5,6,1,2,3)
@@
   One_Balance[[i]] <- ggplot(mRES[which(mRES$nCERs == orders[i]),], aes(x = Replicate_Number, y = log2(F1_Score), colour = Method)) +
-  # geom_point()+
   scale_color_viridis(discrete = T)+
-  # geom_line() +
-    # geom_point() +
-
     stat_summary() +
-
-    # geom_errorbar(mappings = aes(x = Replicate_Number, y = log2(F1_Score)),stat = "summary",
-    #               fun.min = function(z) {quantile(z,0.25)},
-    #               fun.max = function(z) {quantile(z,0.75)},
-    #               fun = median) +
-  # geom_smooth( se = F)+    # aes(outfit=fit<<-..y..)
     ylim(0,max ) +
-    # ggtitle("") +
   labs(subtitle = paste0("CE = ", main_lev[l], " E = ", nse[l]))+
-  # ylab("Rank Product") + 
   theme_bw() +
@@
 
-```{r}
-
- Two_Balance <- ggplot(mRES, aes(x = Replicate_Number, y = log2(F1_Score), colour = Method)) +
-  # geom_point()+
-  scale_color_viridis(discrete = T)+
-  # geom_line() +
-    # geom_point() +
-
-    stat_summary() +
-
-    # geom_errorbar(mappings = aes(x = Replicate_Number, y = log2(F1_Score)),stat = "summary",
-    #               fun.min = function(z) {quantile(z,0.25)},
-    #               fun.max = function(z) {quantile(z,0.75)},
-    #               fun = median) +
-  # geom_smooth( se = F)+    # aes(outfit=fit<<-..y..)
-        ylim(0,max ) +
-    # ggtitle("") +
-  labs(subtitle = paste0("CE = ", main_lev[l], " E = ", nse[l]))+
-  # ylab("Rank Product") + 
-  theme_bw() +
-  facet_wrap(~ nCERs)
-      # theme(axis.title = element_blank())
-
-```
-
 ```{r repres, echo = F, fig.cap= "Simulation 3 results.
 Replications and variance type feasibility ranges.
-X axes; replicate number, y axes log2 transformed rank product.
+X axes; replicate number, y axes log2 transformed geometric mean rank.
 Each facet is created with set parameters for combined effect size ($CE = \\|\\mathbf{T}_{\\alpha + \\alpha\\beta}\\mathbf{P}_{\\alpha + \\alpha\\beta}^T\\|_F$) and random gaussian noise size ($E = \\|\\mathbf{E}\\|_F$).
 Dots are median performance across 20 random structures."}
-
-# One_Balance# + AUCs #+ plot_layout(guides = "collect")
 
@@
 
-# ps1 <- grid.arrange(ps, left = "Rank Product", bottom = "Replicate Number                     ")
+# ps1 <- grid.arrange(ps, left = "Log2(GMR)", bottom = "Replicate Number                     ")
 
@@
 
-<!-- In Figure \@ref(fig:repres)  -->
-
 It was expected that all geometric mean ranks converge towards a minimum with increasing numbers of replicates, see \@ref(fig:repres).
-Perfect detection of 12 spikes in top 12 candidates is achieved with a log2 geometric mean rank of 2.402955, MASCARA this was able to achieve this however the other methods were perturbed by the presence of the 30 non spike differential genes.With more replicates MASCARA always outperforms competitors.
+Perfect detection of 12 spikes in top 12 candidates is achieved with a log2 geometric mean rank of 2.4, MASCARA was able to achieve this however the other methods were perturbed by the presence of the 30 differential URP genes.
+With more replicates MASCARA always outperforms competitors.
 With higher combined effect size (CE) MASCARA outperforms the other methods but its performance is affected by random noise (E) size.
 
-In simulation 2 we fixed the number of differential genes to 30, however this is less than 2% of the total number of genes in each dataset.
+In simulation 2 we fixed the number of differential URP genes to 30, however this is less than 2% of the total number of genes in each dataset.
 In the real data example more than 10% of of the total number of genes are differentially expressed.
-As shown in simulation 1 (Figure \@ref(fig:tc)) higher numbers of DE genes results in higher geometric mean rank of the spikes for correlations, ASCA and PLS.
-MASCARA is designed specifically to mitigate the presence of large numbers of differential genes.
+As shown in simulation 1 (Figure \@ref(fig:tc)) higher numbers of DE URP genes results in higher geometric mean rank of the spikes for correlations, ASCA and PLS.
+MASCARA is designed specifically to mitigate the presence of large numbers of DE URP genes.
 Here we show that this is achieved.
@@
 For this we take 4 core strigolactone genes (Os11t0587000, Os04t0550600, Os01t0746400 and Os01t0700900 of the 9 used for illustrations in figures \@ref(fig:Dataset1-Overview) and \@ref(fig:ASCA-1)) as baits and investigate their relationship with top candidate genes.
-
-<!-- As well as, detection of 5 other genes known to be involved in the pathway (expression profiles shown in heatmap of Figure \@ref(fig:Dataset1-Overview)). -->
 
@@
 #D27 CCD7 CCD8 CYP711A2
-# baits2 <- c("Os01t0700300", "Os01t0701700")  
 baits2 <- c("Os01t0700300-00", "Os02t0817900-01")  # identified methyltransferases
@@
 
-# ref$Baits[-which(ref$Baits %in% c(baits,baits2))] <- "non-bait"
-# ref$Baits[which(ref$Baits %in% baits)] <- "bait"
-# ref$Baits[which(ref$Baits %in% baits2)] <- "test_bait"
-
 REF <- fread("Rice_Annotation/IRGSP-1.0_representative_annotation_2022-09-01.tsv",sep = "\t", header = T)
@@
 X <- X[match(meta$ID, rownames(X)),]
-
 ```
@@
 a_c.PCD <- PCD
-SL <- SL_plot(a_c.PCD, meta)
 ```
@@
 ```{r, echo = F, inlcude = F}
-
-#, fig.cap="Top: distribution and mean sd relationship. Bottom: scores and loadings of the combined effect; growth condition + interaction"
-
 ve1 <- paste0(round(summary(a_c.PCD)$importance[2,1:2] * 100, digits = 2), "%")
@@
 
-# (dist_norm + plot_normal)/(s + l)
-
 ```
@@
 
-#HERE
-```
-
-```{r}
-# image(cor(res_ASCAplus_nb$residuals[,c(baits,baits2)]))
 ```
@@
 tcd <- prcomp(res_ASCAplus_nb$decomposition$time + res_ASCAplus_nb$residuals)
-
 ```
@@
 ```{r, include = F}
-#do permutation tests to see if the correlations in the "non experimentally interesting" data are significant between the strigolactone genes compared to random sets.
 
@@
 This limitation affected many more genes than those directly involved in SL biosynthesis.
-Furthermore, there appears to be a stronger experiment independent relationship amongst the SL genes compared to their correlations with other DE genes, as displayed in Figure \@ref(fig:residualdists) where the blue distribution indicates the 9 transcripts from figures \@ref(fig:Dataset1-Overview) and \@ref(fig:ASCA-1) and the red is the relationship between these 9 and other top differential genes.
+Furthermore, there appears to be a stronger experiment-independent relationship amongst the SL genes compared to their correlations with other DE genes, as displayed in Figure \@ref(fig:residualdists) where the blue distribution indicates the 9 transcripts from figures \@ref(fig:Dataset1-Overview) and \@ref(fig:ASCA-1) and the red is the relationship between these 9 and other top differential genes.
 
@@
 cor_val <- function(x){
-
-  # y <- y[which(y %in% colnames(x))]
-
   cor_res <- cor(x[])
-
   cor_res <- 0.5*log((1+cor_res)/(1-cor_res))
-
-  # res <- (sum(cor_res))   /length(y)^2
-
-  # res <- mean(cor_res[upper.tri(cor_res)])
-
   return(cor_res)
-
-}
-
-# plot(ASCA_cands2[1:1253,2], ASCA_cands2[1:1253,3])
+}
 
 test <- cor_val(res_ASCAplus_nb$residuals[,pathway])  #res_ASCAplus_nb$residuals
-# hist(test[upper.tri(test)])
 
@@
 ```{r}
-
 all <- cbind.data.frame(meta[,c("growth_condition","time")], X)
@@
 #get cands from residuals
-
 rPCD <- prcomp(data.matrix(residuals[,-c(1:2)]))
-
-# residual_cands_all <- ranked_dist(pathway, rPCD)
-# residual_cands_phos <- ranked_dist(PSI_ref[,1], rPCD)
-
 ```
@@
 
-# sPLS_cands$Feature <- rownames(sPLS_cands)
-# sPLS_cands <- merge(sPLS_cands, ref, by = "Feature")
-
-# ggplotly(loadingplot(sPLS_cands, meta, baits = Fref2$Feature[which(Fref2$Baits %in% c("Pathway","PSI"))], 
-            # importance, PCs = c(1,2), ref = Fref2, aes(x = comp1, y = comp2, colour = Baits, label = Description)))
-
 ```
@@
 #calculating direction of baits to determine direction of candidate selection spls loading space
-
-# loadingplot(sPLS_cands, meta, baits = Fref2$Feature[which(Fref2$Baits %in% c("Pathway","PSI"))], 
-#             importance, PCs = c(1,2), ref = Fref2, aes(x = comp1, y = comp2, colour = Baits, label = Description))
 
@@
 ```{r}
-#U+03C0
-#"\u03c0 / 4" #pi over 4
-
-# rotate <- function(x,y){
-#   x <- c(as.matrix(x))
-#   x %*% y
-# }
-
-#loop over rotation matrices from -pi/4 -> pi/4
-#resolution of 100 (?)
-
 range.ang <- seq(from = -pi/6, to = pi/6, length.out = 60)
@@
   geom_smooth() + theme_bw() + ggtitle("Number of points above 2*sigma of original rotation")
-
-# density
 
@@
 i <- NULL
-# L_R <- data.frame()
 sPLS_maxima_cands <- list()
@@
 colnames(loadings_r) <- c("comp1","comp2")
-# L_R <- rbind(L_R,loadings_r)
 sPLS_maxima_cands[[i]] <- loadings_r[order(loadings_r$comp1, decreasing = T),]
@@
 ```{r}
-
 Fref2 <- Fref[-which(Fref$Feature %in% baits),]
@@
 
-#replace this to  get cands from original bait axis plus optimised maxima
 sPLS_cands <- sPLS_maxima_cands[[2]]
@@
 getrank <- function(RES,geneset,NOINFO = 26402){
-  # geneset <- 
-  # NOINFO = length(geneset) - length(ranks)
+
   ranks <- which(rownames(RES) %in% geneset)
@@
 
-<!-- ## Methods compared -->
-
 ```{r, fig.cap="Method Ranks on Real Data"}
@@
 REF2 <- as.data.frame(read_tblout("Rice_Annotation/HMM_ITAG_1.txt"))
-# unique(REF2$query_name[which(REF2$domain_name %in% rownames(sPLS_cands)[1:9])])
-
-# REF2[which(REF2$domain_name %in% rownames(sPLS_cands)[1:9]),c(1,3,4,19)]
 
@@
 rownames(fpm) <- NULL
-# kable(fpm)
-
-```
-
-<!-- Ranked correlations outperforms the other methods at detecting SL genes, however it also detects more general phosphate starvation induced genes (PSIs) as top candidates, due to the strong between correlations.
--->
+
+```
 
@@
 The between versus within problem is again exemplified here.
-With several of the top candidates detected by correlations (ASCA and PLS, results not shown) showing no indication of within group correlations.
+Several of the top candidates detected by correlations (ASCA and PLS, results not shown) show no indication of within group correlations.
 Candidates selected with MASCARA, on the other hand, consistently show within group correlations with the baits.
@@
 X axes; expression level of one of the bait genes (CCD8), y axes; expression levels of three example coexpression candidates for each method."}
-# x2 <- X
-# 
-# X <- res_ASCAplus_nb$residuals
 
@@
 
-  # qqplots_pca[[i]] <- qqplot(cbind.data.frame("Bait" = X[,which(colnames(X) %in% rownames(residual_cands)[i])],
-  #                       "Candidate" = X[,which(colnames(X) %in% baits[1])]),meta, 
-  #      aes(x = Bait, y = Candidate, colour = growth_condition, shape = time))
-  # 
   qqplots_resid_cor[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(residual_coexp)[i])],
@@
        aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + ylab(rownames(residual_coexp)[i])
-
-    # qqplots_spls_SR[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(SR)[i])],
-    #                     "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
-    #    aes(x = Bait, y = Candidate, colour = growth_condition, shape = time))
-    #
 
@@
 The feature selection stage of MASCARA uses an average vector constructed from the loadings of the baits onto which to project the loadings of other genes, which is similar to target projection in @kvalheimInterpretationPartialLeast2010 although we construct the target vector from averaging the loadings in a PLS2 model as opposed to already having it defined with the univariate response in PLS1.
-There are many established methods for feature selection in PLS which include Variable Importance in Projection (VIP) and Selectivity Ratio (@rajalahtiBiomarkerDiscoveryMass2009), however these methods focus on the importance of variables in the construction of the model rather than a directional association with the response block.
-With the averaged target projection approach we can use the information in a set of baits to define a direction in the latent space which can then be used as an axis to determine the relatedness of other genes.
+There are many established methods for feature selection in PLS which include Variable Importance in Projection (VIP) and Selectivity Ratio (@rajalahtiBiomarkerDiscoveryMass2009), however these methods focus on the importance of variables in the construction of the model rather than a directional association with the response block as with the $\mathbf{\bar{q}}$ target projection stage.
+With this averaged target projection approach we can use the information in a set of baits to define a direction in the latent space which can then be used as an axis to determine the relatedness of other genes.
 This assumes moderate to high correlation in the residuals of the baits, which is a realistic assumption in coexpression analysis although should be verified in the selection of baits.
@@
 Our analysis framework does not consider different correlation structures within the residual variance, this could be an issue especially in the cases similar to @caldanaHighdensityKineticAnalysis2011 where different combinations of experimental factors show differential within-group correlation structure.
-With MASCARA these could be missed, ideally there would be enough samples to calculate robust correlations per group, to account for possible rewiring under different conditions, but in our case we focus on experiments where there are generally not enough replicates to do this.
+With MASCARA these will not be preferentially detected, as the focus is on detecting postively associating features.
+Ideally there would be enough samples to calculate robust correlations per group, to account for possible rewiring under different conditions, but in our case we focus on experiments where there are generally not enough replicates to do this.
 Experiments with higher replication number are always desirable from the point of view of data analysis.
@@
 These genes are also related to the baits, as in our partially characterised strigolactone pathway example.
-In untargeted metabolomics datasets these assumptions are also likely to be met.
+In untargeted metabolomics datasets under similar experimental conditions these assumptions are also likely to be met.
 
@@
 MASCARA accounts for the different types of (experimental and ambient) variance and enables finer scale detection of within group variance structure.
-
-<!-- Future research on this topic would involve further development of the data simulator.  -->
 
@@
 ```{r}
-
-#30 random initial scores, similar noise 
-#parameters, 3-30 replicates per condition. 
 Rep_Tests <- readRDS("Noise_Sim_23_07_30_FS_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS

### 2023-12-11 R/run_first/1.pattern_examples.Rmd
231211

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = TRUE, root.dir = "../")
 ```
@@
 library(pracma)
-# library(NineteenEightyR)
 library(matlib)
 library(truncnorm)
-# library(sf)
 
-source("DATA_SIM_FUNCS.R")
-
+source("../DATA_SIM_FUNCS.R")
 ```
@@
 rnd_seed <- abs(rnorm(3, mean = 0.7, sd = 0.5))
-#c(1,0.8,0.3)
-# Create_Core_DEV
+
 sim_data <- Create_Core_DEV_2(nreps = 3, meta = meta, plot = TRUE, 
@@
                         b_sigma = c(1,0.9), irr_spikes = TRUE, SCORE_SEED = 12, Experiment_responders = Experiment_responders,
-                        struc_seed = 127)  #12
-
-#list(X,P,Pb,a_ab,PLOT, X_a_ab,time, Struc_E )
-
-###plot error in sim_data[[5]]
-
+                        struc_seed = 127)  
 ```
@@
 
-# sim_data[[1]] <- sim_data[[1]][,1850:2000]
-
-# image(sim_data)
 ```
@@
 
-# data <- (0.1 * interaction) + (1.5*time) + (0.5*ambient) + (0*noise)
-# 
 input <- data
-# 
-# input <- cbind.data.frame(input,input,input)
 
@@
 ```{r}
-# data <- (0.1 * interaction) + (1.5*time) + ambient + noise
-
-# input <- input + rnorm(nrow(input) * ncol(input),mean = 0, sd = 1)
-#replace with simulation input
-# colnames(input) <- make.names(colnames(input))
 
@@
 ```{r}
-MASCARA4_test_1 <- function(resids,ref, baits, spikes, ncands){
-
-  Fref <- ref
-
-  spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], 
-                   resids[,which(colnames(resids) %in% baits)], all.outputs = TRUE)
-
-  # spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)], 
-  #                  resids[,which(colnames(resids) %in% baits)], all.outputs = TRUE)
-
-  sPLS_cands <- cbind.data.frame(abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X, spls_res$loadings$X) #  * spls_res$prop_expl_var$X
-  sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = TRUE),]
-
-  importance <- paste0(round(spls_res$prop_expl_var$X * 100, digits = 2), "%")
-
-  Fref2 <- Fref[-which(Fref$Feature %in% baits),]
-  Fref2 <- Fref2[match(rownames(sPLS_cands),Fref2$Feature),]
-
-  #####
-
-  y <- colMeans(spls_res$loadings$Y)
-
-  U <- spls_res$loadings$X
-
-  # UTP <- ((U %*% y)/norm(y, "2")) * y
-  # UTP <- ((U %*% y)/dot(y,y)) * y   #_9
-
-  # UTP <- (dot(t(U),y)/dot(y,y)) * y  #_10
-
-  UTP <- (dot(t(U),y)/dot(y,y))
-
-  sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])
-
-  # sPLS_bait_cands <- as.data.frame(UTP[order(UTP[,1], decreasing = TRUE),])  #_9 and below
-
-  sPLS_maxima_cands <- sPLS_bait_cands
-
-  Fref2 <- Fref[-which(Fref$Feature %in% baits),]
-  Fref2 <- Fref2[match(rownames(sPLS_maxima_cands),Fref2$Feature),]
-
-  RP <- prod(which(rownames(sPLS_maxima_cands) %in% spikes))^(1/length(spikes))
-
-  return(list(RP, sPLS_maxima_cands,UTP))
-}
-
-MASCARA4_test_2 <- function(resids,ref, baits, spikes, ncands){
-
-  Fref <- ref
-
-  spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], 
-                   resids[,which(colnames(resids) %in% baits)], all.outputs = TRUE)
-
-  # spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)], 
-  #                  resids[,which(colnames(resids) %in% baits)], all.outputs = TRUE)
-
-  sPLS_cands <- cbind.data.frame(abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$Y, spls_res$loadings$X) #  * spls_res$prop_expl_var$X
-  sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = TRUE),]
-
-  importance <- paste0(round(spls_res$prop_expl_var$X * 100, digits = 2), "%")
-
-  Fref2 <- Fref[-which(Fref$Feature %in% baits),]
-  Fref2 <- Fref2[match(rownames(sPLS_cands),Fref2$Feature),]
-
-  #####
-
-  y <- colMeans(spls_res$loadings$Y)
-
-  U <- spls_res$loadings$X
-
-  # UTP <- ((U %*% y)/norm(y, "2")) * y
-  # UTP <- ((U %*% y)/dot(y,y)) * y   #_9
-
-  # UTP <- (dot(t(U),y)/dot(y,y)) * y  #_10
-
-  UTP <- (dot(t(U),y)/dot(y,y))
-
-  sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])
-
-  # sPLS_bait_cands <- as.data.frame(UTP[order(UTP[,1], decreasing = TRUE),])  #_9 and below
-
-  sPLS_maxima_cands <- sPLS_bait_cands
-
-  Fref2 <- Fref[-which(Fref$Feature %in% baits),]
-  Fref2 <- Fref2[match(rownames(sPLS_maxima_cands),Fref2$Feature),]
-
-  RP <- prod(which(rownames(sPLS_maxima_cands) %in% spikes))^(1/length(spikes))
-
-  return(list(RP, sPLS_maxima_cands,UTP))
-}
-
-```
-
-```{r}
 ##feature selection alternatives
-source("MASCARA_Test_Funcs.R")
+source("../MASCARA_FUNCS.R")
 #default - target projection onto average Q in 2 dimensions
@@
 which(rownames(MASCARA_2[[2]]) %in% spikes)
-
-#PLS1 model with svd(baits)$T as response
-MASCARA_3 <- MASCARA4_test_3(ar[[3]], baits, spikes = spikes)
-which(rownames(MASCARA_3[[2]]) %in% spikes)
-
-# 
-# 
-#
-
-#PLS1 model with svd(baits)$T as response and target projected deflation
-
-# MASCARA_4 <- MASCARA4_test_4(ar[[3]], baits, spikes = spikes)
-# which(rownames(MASCARA_4[[2]]) %in% spikes)
-
-# #PLS2 model with target projected deflation
-# 
-# 
-#
 
@@
 ```{r}
-source("MASCARA_Test_Funcs.R")
+source("../MASCARA_FUNCS.R")
 #classic
@@
 which(rownames(MASCARA_2[[2]]) %in% spikes)
-
-#u1 of baits as pls1 response
-MASCARA_3 <- MASCARA4_test_3(ar[[3]], baits, spikes = spikes)
-which(rownames(MASCARA_3[[2]]) %in% spikes)
 
@@
 #vip
-
 MASCARA_5 <- MASCARA4_test_8(ar[[3]], baits, spikes = spikes)
 which(rownames(MASCARA_5[[2]]) %in% spikes)
-
-#tpsr
 
@@
 yloadings <- MASCARA_2[[4]]$Yloadings[]
-
-# yloadings[3,2] <- -0.5*yloadings[3,2]   #pure data manipulation for demo purposes
 
@@
 
-# loadings_r_baits <- rbind(loadings_r_baits, yloadings)
-
 ```
@@
       geom_point(data = loadings_r_baits[which(loadings_r_baits$ID %in% c("Baits","Pathway")),], aes(size = 15)) +
-  # geom_point(data = yloadings, shape = 4, aes(x = Comp_1, y = Comp_2, size = 3 )) + 
   theme_bw()
@@
 ```
-
-```{r}
-
-#multiple PLS1 models with svd(baits) deflated
-
-```
-
-```{r}
-#remake simpls to deflate with target projected loadings
-
-#first recreate simpls such that the output is the same as simpls.fit()
-
-#$projection -> T = XR
-
-```

### 2023-12-11 R/run_first/1.pattern_examples_MASCARA_pipeline.Rmd
231211

--- 2023-12-11 R/run_first/1.pattern_examples.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, root.dir = "../")
+knitr::opts_chunk$set(echo = TRUE)
 ```
@@
 source("../DATA_SIM_FUNCS.R")
+
 ```
@@
 
-set.seed(3)
-rnd_seed <- abs(rnorm(3, mean = 0.7, sd = 0.5))
-
-sim_data <- Create_Core_DEV_2(nreps = 3, meta = meta, plot = TRUE, 
-                        EffectSize = c(X_a_ab = 5, time = 1, E = 10, mu = 1, Struc_E = 1), 
-                        struc_resid = TRUE, e_sigma = c(1.5,0.8,0.0), a_sigma = c(1,0.7,0.6,0.1),
-                        b_sigma = c(1,0.9), irr_spikes = TRUE, SCORE_SEED = 12, Experiment_responders = Experiment_responders,
-                        struc_seed = 127)  
 ```
@@
 sim_data <- Create_Core_DEV_2(nreps = 3, meta = meta, 
-                        EffectSize = c(X_a_ab = 5, time = 1, E = 1, mu = 1, Struc_E = 0.4),    # E = 10, Struc_E = 1
+                        EffectSize = c(X_a_ab = 5, time = 1, E = 10, mu = 1, Struc_E = 1),    # E = 1, Struc_E = 0.4
                         struc_resid = TRUE, e_sigma = c(1.5,0.8,0.0), a_sigma = c(1,0.7,0.6,0.1),
                         b_sigma = c(1,0.9), irr_spikes = TRUE, SCORE_SEED = 12, Experiment_responders = Experiment_responders,
-                        struc_seed = 127, score_plot = FALSE, plot = FALSE)  #12
+                        struc_seed = 127, score_plot = FALSE, plot = FALSE)
 
@@
 spikes <- paste0("X_",c(1989:2000))
-
-ref <- cbind.data.frame(Feature = colnames(sim_data[[1]]), sim_data[[2]], sim_data[[3]])
-cols <- colnames(ref[,-1])
-ref$Effect <- do.call(paste, c(ref[cols], sep = "_"))
-ref$Effect[which(ref$Effect == "0.5_0_0_0.1")] <- "Low PC1"
-ref$Effect[which(ref$Effect == "1_0_0_0.1")] <- "Medium PC1"
-ref$Effect[which(ref$Effect == "1.5_0_0_0.1")] <- "High PC1"
-ref$Effect[which(ref$Effect == "0_0.5_0.1_0")] <- "PC2"
-ref$Effect[grep("^0", ref$Effect)] <- "None"
-ref$Effect <- factor(ref$Effect, levels = c("Low PC1", "Medium PC1", "High PC1", "PC2", "None"))
-colnames(ref)[8] <- "Baits"
-
-ref$Description <- ref$Baits
-
-ref$Baits[1985:1988] <- "Baits"
-ref$Baits[1989:2000] <- "Candidates"
-
-ref$Baits[(2000 - (16 + Experiment_responders)):1984] <- "ab1+ve"
-# ref$Baits[1969:1976] <- "ab1+ve"
-ref$Baits[501:515] <- "ab2+ve"
-ref$Baits[486:500] <- "ab2-ve"
-# ref$Baits[1969:1976] <- "Expt_Independent"
-
-ref$Baits[1:24] <- "ab1-ve"
-
-ref$Description <- ref$Baits
 
@@
 library(ggplot2)
-p1 <- ggplot(mData, aes(Samples, Features, fill= value)) + 
+p1 <- ggplot(mData, aes(Features, Samples, fill= value), show.legend = FALSE) + 
   geom_tile() +
   ggtitle("Growth Condition") +
-  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
+  scale_fill_viridis(discrete=FALSE) +  
   theme_bw() +
@@
 colnames(mData)[1:2] <- c("Samples","Features")
-# mData$Features <- fct_rev(mData$Features)
 
@@
 library(ggplot2)
-p2 <- ggplot(mData, aes(Samples, Features, fill= value)) + 
+p2 <- ggplot(mData, aes(Features, Samples, fill= value)) + 
   geom_tile() +
@@
 library(ggplot2)
-p3 <- ggplot(mData, aes(Samples, Features, fill= value)) + 
+p3 <- ggplot(mData, aes(Features, Samples, fill= value)) + 
   geom_tile() +
@@
 library(ggplot2)
-p4 <- ggplot(mData, aes(Samples, Features, fill= value)) + 
+p4 <- ggplot(mData, aes(Features, Samples, fill= value)) + 
   geom_tile() +
@@
 library(ggplot2)
-p5 <- ggplot(mData, aes(Samples, Features, fill= value)) + 
+p5 <- ggplot(mData, aes(Features, Samples, fill= value)) + 
   geom_tile() +
@@
 colnames(mData)[1:2] <- c("Samples","Genes")
-# mData$Features <- fct_rev(mData$Features)
 
@@
 library(ggplot2)
-p0 <- ggplot(mData, aes(Samples, Genes, fill= value)) + 
+p0 <- ggplot(mData, aes(Genes, Samples, fill= value)) + 
   geom_tile() +
@@
         axis.text.y=element_blank(),  #remove y axis labels
-        axis.ticks.y=element_blank())#,
-        #axis.title.x = element_blank(),
-        #axis.title.y = element_blank())
+        axis.ticks.y=element_blank(),
+        axis.title.x = element_text(size = 16),
+        axis.title.y = element_text(size = 16))
 
@@
 p5
+
 ```
@@
 ```{r}
-ps <- patchwork::patchworkGrob(wrap_plots(plots) + plot_layout(guides = "collect"))  #, heights = unit(c(5), c("cm")), 
-ps1 <- grid.arrange(ps)#, left = "Features", bottom = "Replicate Number                     ")
+ps <- patchwork::patchworkGrob(wrap_plots(plots) )    #+ plot_layout(guides = "collect") -
 
-# p5
+ps1 <- grid.arrange(ps)
+
 ```
@@
 
+# 
 input <- data
+#
 
@@
 ```{r}
-####PLS pattern visualisation change this
+source("../MASCARA_FUNCS.R")
 
-data <- scores[]    # data.matrix(sim_data[[4]][,3:4])  #
-mData <- reshape2::melt(data)
-colnames(mData)[1:2] <- c("Samples","Features")
-# mData$Features <- fct_rev(mData$Features)
-
-library(viridis)
-library(hrbrthemes)
-library(ggplot2)
-p0 <- ggplot(mData, aes(Features,Samples, fill= value)) + 
-  geom_tile() +
-  ggtitle('Patterns (T)') +
-  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
-  theme_bw() +
-  theme(axis.text.x=element_blank(), #remove x axis labels
-        axis.ticks.x=element_blank(), #remove x axis ticks
-        axis.text.y=element_blank(),  #remove y axis labels
-        axis.ticks.y=element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.position = "none")
-
-# p0
-
-data <- input[,1:2]
-mData <- reshape2::melt(data)
-colnames(mData)[1:2] <- c("Samples","Features")
-# mData$Features <- fct_rev(mData$Features)
-
-library(viridis)
-library(hrbrthemes)
-library(ggplot2)
-p1 <- ggplot(mData, aes(Samples,Features, fill= value)) + 
-  geom_tile() +
-  ggtitle('Data; X') +
-  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
-  theme_bw() +
-  theme(axis.text.x=element_blank(), #remove x axis labels
-        axis.ticks.x=element_blank(), #remove x axis ticks
-        axis.text.y=element_blank(),  #remove y axis labels
-        axis.ticks.y=element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.position = "none")
-
-data <- input[,3:4]
-mData <- reshape2::melt(data)
-colnames(mData)[1:2] <- c("Samples","Features")
-# mData$Features <- fct_rev(mData$Features)
-
-library(viridis)
-library(hrbrthemes)
-library(ggplot2)
-p2 <- ggplot(mData, aes(Samples,Features, fill= value)) + 
-  geom_tile() +
-  ggtitle('Data; Y') +
-  scale_fill_viridis(discrete=FALSE) +  #, limits = c(min,max)
-  theme_bw() +
-  theme(axis.text.x=element_blank(), #remove x axis labels
-        axis.ticks.x=element_blank(), #remove x axis ticks
-        axis.text.y=element_blank(),  #remove y axis labels
-        axis.ticks.y=element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.position = "none")
-
-p0 + (p1 / p2) + theme(legend.position = "none")
-```
-
-```{r}
-##feature selection alternatives
-source("../MASCARA_FUNCS.R")
-#default - target projection onto average Q in 2 dimensions
-resids <- ar[[3]]
-
-MASCARA_1 <- MASCARA4_test_1(ar[[3]], baits = baits, spikes = spikes)
-which(rownames(MASCARA_1[[2]]) %in% spikes)
-
-# # replacing mixomics with SIMPLS from pls package
-
-#1 component, absolute projections
-spls_res2.1 <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
-                   resids[,which(colnames(resids) %in% baits)], ncomp = 1)
-
-res_prj <- cbind.data.frame(rownames(spls_res2.1$projection),spls_res2.1$projection)
-res_VIP <- res_prj[order(abs(res_prj[,2]), decreasing = TRUE),]
-rownames(res_VIP) <- res_VIP[,1]
-
-which(rownames(res_VIP) %in% spikes)
-
-## 2 component target projection
+#classic; q bar target projection
 MASCARA_2 <- MASCARA4_test_2(ar[[3]], baits, spikes = spikes)
@@
 
-```
-
-```{r}
-resids <- ar[[3]]
-
-spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], 
-                 resids[,which(colnames(resids) %in% baits)], all.outputs = TRUE)
-
-spls_res2 <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
-                resids[,which(colnames(resids) %in% baits)], ncomp = 2)
-
-spls_res2.1 <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
-                   resids[,which(colnames(resids) %in% baits)], ncomp = 1)
-
-res_prj <- cbind.data.frame(rownames(spls_res2.1$projection),spls_res2.1$projection)
-res_VIP <- res_prj[order(abs(res_prj[,2]), decreasing = TRUE),]
-rownames(res_VIP) <- res_VIP[,1]
-
-which(rownames(res_VIP) %in% spikes)
-
-```
-
-```{r}
-source("../MASCARA_FUNCS.R")
-#classic
-MASCARA_2 <- MASCARA4_test_2(ar[[3]], baits, spikes = spikes)
-which(rownames(MASCARA_2[[2]]) %in% spikes)
+#u1 of baits as pls1 response
+MASCARA_3 <- MASCARA4_test_3(ar[[3]], baits, spikes = spikes)
+which(rownames(MASCARA_3[[2]]) %in% spikes)
 
@@
 
-# str(MASCARA_2[[4]])
-
 loadings <- MASCARA_2[[4]]$loadings[]
@@
 line <- colMeans(yloadings)
+
+#rotation of scores and loadings for demo purposes only
 
@@
 colnames(LINE) <- c("to","from")
-# scores <- scores %*% 
+
 scores <- cbind.data.frame(scores,meta[,1:2])
@@
 
-loadings_r_baits$ID[which(loadings_r_baits$ID %in% spikes)] <- "Pathway"
+loadings_r_baits$ID[which(loadings_r_baits$ID %in% spikes)] <- "Spikes"
 loadings_r_baits$ID[which(loadings_r_baits$ID %in% baits)] <- "Baits"
@@
     scale_color_manual(values=c('#E69F00',"black", '#56B4E9'))+
-      geom_point(data = loadings_r_baits[which(loadings_r_baits$ID %in% c("Baits","Pathway")),], aes(size = 15)) +
+      geom_point(data = loadings_r_baits[which(loadings_r_baits$ID %in% c("Baits","Spikes")),], aes(size = 15)) +
   theme_bw()

### 2023-12-11 R/run_first/2.plot_pathway_net_example.Rmd
231211

@@
 library(ppcor)
-#library(matlib)
 library(psych)
@@
 #
-# plot residuals and full
+# plot residuals and full (with tweaks for problem demonstration)
 
@@
 cand_vals <- residuals[,1970:2000]    #rgs[which(rownames(rgs) %in% names(g_b)[1:50]),]
-
-# cand_vals <- data[,1970:2000]
-
-##CHANGE THIS 
-#make p3 interconnections stronger, 
-#make p3 connect weakly to p2 through one or two nodes, 
-#sparsify p2 and p3 connections
 
@@
 
-#adj[which(rownames(adj) %in% baits),which(colnames(adj) %in% spikes)] <- adj[which(rownames(adj) %in% baits),which(colnames(adj) %in% c(spikes))] * W
 ```
@@
 ```{r}
-#sparsify p1 p2 connections
+#randomly sparsify p1 p2 connections
 
@@
 ```{r}
-
-##need to know which nodes are in at least one of the conditions before here
-
 library(igraph)
@@
 
-# Colour negative correlation edges as blue
 E(g)[which(E(g)$weight<0)]$association <- "-ve"
-# Colour positive correlation edges as red
 E(g)[which(E(g)$weight>0)]$association <- "+ve"
@@
 ```{r}
-# Remove edges    make this general to both types of correlations
-
 ###
@@
 
-# Remove any vertices remaining that have no edges
+# Remove nodes with no edges
 g <- delete.vertices(g, degree(g)==0)
 
-# Assign names to the graph vertices (optional)
 V(g)$name <- V(g)$name
-
-# Change shape of graph vertices
-V(g)$shape <- "sphere"
-
-# Change colour of graph vertices
-V(g)$color <- "skyblue"
 
@@
 
-#define correctly ordered type type
 type <- V(g)$name
-
-# ref$Feature <- gsub("_","",ref$Feature)
-
 ```
@@
 
-#get metabolite and taxa info
-
-###CHANGE THIS
-#define correctly ordered VIP values  -- not actually VIP - replace with ???
-# VIP <- g_b[type]
-
 VIP <- rep(5,length(V(g)))
-#VIP <-
 
@@
 
-colours = c("POI" = "purple", "RP" = "orange", "URP" = "lightgreen")   ##adapt this for 3 different pathway types
+colours = c("POI" = "purple", "RP" = "orange", "URP" = "lightgreen")
 
@@
 
-#edge_linetype <- round(E(g)$weight, digits = 1)
-
-#cols_f <- colorRampPalette(RColorBrewer::brewer.pal(11, 'Spectral'))
-
-# pdf(paste0(Sys.Date(),"_net_test.pdf"))
 realplot <- ggraph(tidynet, layout = coords) + 
-  geom_edge_link(aes(colour = association, alpha = weight), show.legend=FALSE) + #, linetype = association)) + 
+  geom_edge_link(aes(colour = association, alpha = weight), show.legend=FALSE) +  
   scale_edge_color_manual(values = edge_colours, guide = "none") +
-  #scale_edge_linetype_continuous(values = edge_linetype) +
   geom_node_point(
     aes(size = VIP, color = type), show.legend=FALSE
-    #size = 4
   ) +
@@
     aes(color = type)
-    #size = 4
   ) +
   scale_color_manual(values = colours) +
-  # geom_node_text(label = id, alpha = 1, family = "mono", fontface = "bold", size = 5)+  #, fontface = "italic"
 
-  theme_void() #+
- # theme(legend.position = "none")
-# dev.off()
-
-################################################################################
+  theme_void()
 
@@
 
-# Remove any vertices remaining that have no edges
 g <- delete.vertices(g, degree(g)==0)
 
-# Assign names to the graph vertices (optional)
 V(g)$name <- V(g)$name
 
-# Change shape of graph vertices
 V(g)$shape <- "sphere"
-
-# Change colour of graph vertices
-V(g)$color <- "skyblue"
 
@@
 type <- V(g)$name
-
-# ref$Feature <- gsub("_","",ref$Feature)
 
@@
 
-#get metabolite and taxa info
+VIP <- rep(5,length(V(g)))
 
-###CHANGE THIS
-#define correctly ordered VIP values  -- not actually VIP - replace with ???
-# VIP <- g_b[type]
-
-VIP <- rep(5,length(V(g)))
-# #VIP <- 
-# 
-# 
 type[which(type %in% baits)] <- "POI"
@@
 badplot <- ggraph(tidynet, layout = coords) + 
-  geom_edge_link(aes(colour = association, alpha = weight), show.legend=FALSE) + #, linetype = association)) + 
+  geom_edge_link(aes(colour = association, alpha = weight), show.legend=FALSE) + 
   scale_edge_color_manual(values = edge_colours, guide = "none") +
-  #scale_edge_linetype_continuous(values = edge_linetype) +
   geom_node_point(
     aes(size = VIP, color = type), show.legend=FALSE
-    #size = 4
   ) +
@@
     aes(color = type)
-    #size = 4
   ) +
   scale_color_manual(values = colours) + 
-  # geom_node_text(label = id, alpha = 1, family = "mono", fontface = "bold", size = 5)+  #, fontface = "italic"
-
-  theme_void() #+
- # theme(legend.position = "none")
-# dev.off()
-
-  # 
-  # 
-  # ggraph(tidynet, layout = coords) + 
-  # geom_edge_link(aes(colour = association, alpha = weight)) + #, linetype = association)) + 
-  # scale_edge_color_manual(values = edge_colours) +
-  # #scale_edge_linetype_continuous(values = edge_linetype) +
-  # geom_node_point(
-  #   aes(size = VIP, color = type),
-  #   #size = 4
-  # ) +
-  # scale_color_manual(values = colours) + 
-  # # geom_node_text(label = id, alpha = 1, family = "mono", fontface = "bold", size = 5)+  #, fontface = "italic"
-  # 
-  # theme_void()# +
-  # # theme(legend.position = "none")
+  theme_void()
 

### 2023-12-11 R/MASCARA_FUNCS.R

@@ -1,0 +1,114 @@
+MASCARA4_test_1 <- function(resids, baits, spikes){
+  #MIXOMICS
+  spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], 
+                   resids[,which(colnames(resids) %in% baits)], all.outputs = TRUE)
+  
+  #####
+  y <- colMeans(spls_res$loadings$Y)
+  U <- spls_res$loadings$X
+  UTP <- (dot(t(U),y)/dot(y,y))
+  
+  sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])
+  RP <- prod(which(rownames(sPLS_bait_cands) %in% spikes))^(1/length(spikes))
+  
+  return(list(RP, sPLS_bait_cands,UTP, spls_res))
+}
+
+
+
+MASCARA4_test_2 <- function(resids, baits, spikes){
+  #target projection
+  spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
+                          resids[,which(colnames(resids) %in% baits)], ncomp = 2)
+  
+  #####
+  y <- colMeans(spls_res$Yloadings)
+  U <- spls_res$projection
+  UTP <- (dot(t(U),y)/dot(y,y))
+  
+  sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])
+  RP <- prod(which(rownames(sPLS_bait_cands) %in% spikes))^(1/length(spikes))
+  
+  return(list(RP, sPLS_bait_cands,UTP,spls_res,y))
+}
+
+
+
+
+MASCARA4_test_8 <- function(resids, baits, spikes){
+  #vip
+  spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
+                         resids[,which(colnames(resids) %in% baits)], ncomp = 2)
+  
+  
+  #####
+  y <- colMeans(spls_res$Yloadings)
+  U <- spls_res$projection
+  
+  #CHECK spls_res$residuals for variance explained 
+  
+  
+  x <- Yloadings(spls_res)
+  ve_y <- colSums(x^2)/nrow(x)
+  
+
+
+  sPLS_cands <- cbind.data.frame(abs(spls_res$projection) %*% ve_y, spls_res$projection)
[…]

### 2023-12-11 R/run_first/3.sim1_CE_size.R

@@ -25,7 +25,7 @@ library(mixOmics)
 library(pracma)
 library(pls)
 
-source("DATA_SIM_FUNCS_DEV.R")
+source("../DATA_SIM_FUNCS.R")
 F1_plot <- F1_plot_no_plot
 nreps <- 3
 ncands <- 500
@@ -73,17 +73,9 @@ MASCARA4_test <- function(resids,ref, baits, spikes, ncands){
   y <- colMeans(spls_res$loadings$Y)
 
   U <- spls_res$loadings$X
-  
-  # UTP <- ((U %*% y)/norm(y, "2")) * y
-  # UTP <- ((U %*% y)/dot(y,y)) * y   #_9
-  
-  # UTP <- (dot(t(U),y)/dot(y,y)) * y  #_10
-  
   UTP <- (dot(t(U),y)/dot(y,y))
   
   sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = T)])
-  
-  # sPLS_bait_cands <- as.data.frame(UTP[order(UTP[,1], decreasing = T),])  #_9 and below
   
   sPLS_maxima_cands <- sPLS_bait_cands
   
@@ -101,10 +93,7 @@ MASCARA4_test <- function(resids,ref, baits, spikes, ncands){
 
 
 
-# baits <- paste0("X_",c(1989:1996))   #501:510,
-# baits <- paste0("X_",c(1985:1996))   #501:510
 
-# baits <- paste0("X_",c(1989:1992))
 # spikes <- paste0("X_",c(1977:1988))
 baits <- paste0("X_",c(1985:1988))   #501:510
 spikes <- paste0("X_",c(1989:2000))
@@ -112,9 +101,6 @@ spikes <- paste0("X_",c(1989:2000))
 
 meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
 colnames(meta) <- c("growth_condition","time","interaction")
-# meta <- as.data.frame(apply(meta,2,factor))
-
-# meta <- as.data.frame(apply(meta,2,factor))
 
 
 sim_data <- Create_Core_DEV(nreps = 3, meta = meta, plot = F, 
@@ -153,34 +139,7 @@ ref$Baits[1:24] <- "ab1-ve"
 ref$Description <- ref$Baits
 
 
[…]

### 2023-12-11 R/run_first/4.sim2_rep.R

@@ -24,7 +24,7 @@ library(caret)
 library(mixOmics)
 library(pracma)
 
-source("DATA_SIM_FUNCS_DEV.R")
+source("../DATA_SIM_FUNCS.R")
 
 Create_Core_DEV <- Create_Core_DEV_2
 

### 2023-12-11 R/run_first/5.sim3_feature_select.R

@@ -24,8 +24,8 @@ library(caret)
 library(mixOmics)
 library(pracma)
 
-source("DATA_SIM_FUNCS_DEV.R")
-source("MASCARA_Test_Funcs.R")
+source("../DATA_SIM_FUNCS.R")
+source("../MASCARA_FUNCS.R")
 
 Create_Core_DEV <- Create_Core_DEV_2
 
@@ -34,15 +34,7 @@ F1_plot <- F1_plot_no_plot
 alphaN <- 2
 betaN <- 4
 ncands <- 500
-# baits <- paste0("X_",c(1989:1996))   #501:510,
-# baits <- paste0("X_",c(1985:1996))   #501:510
 
-
-
-
-
-# baits <- paste0("X_",c(1997:2000))
-# spikes <- paste0("X_",c(1985:1996))
 
 baits <- paste0("X_",c(1985:1988))   #501:510
 spikes <- paste0("X_",c(1989:2000))
@@ -235,13 +227,7 @@ while(i < length(X_funced) + 1){
     SVD1_RES[,i] <- MASCARAh[[1]]
     SVD1_CANDS[,i] <- rownames(MASCARAh[[2]])
     
-    # MASCARAh <- MASCARA4_test(ar[[3]], ref, baits, spikes = spikes, ncands)
-    # TPSR_RES[,i] <- MASCARAh[[1]]
-    # TPSR_CANDS[,i] <- rownames(MASCARAh[[2]])
-    # 
-    # 
-    
-    
+
     i <- i + 1
     
   },

### 2023-12-19 R/23_02_02_real_data_rice.Rmd
23_12_19

---
title: "Haider Data Demo"
author: "Fred White"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  bookdown::html_document2:
    toc: true
    theme: united
    toc_depth: 3
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, echo = F)
#pdf_document
#powerpoint_presentation
```

```{r , echo=FALSE, include=FALSE}
#SOURCE FUNCTIONS FROM EXTERNAL FILE
source("DATA_SIM_FUNCS.R")
```

```{r, echo = FALSE, include=FALSE}
library(devtools)
library(ggplot2)
library(MetStaT)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)    
library(MUVR) 
library(pheatmap)
library(caret)
library(data.table)
library(plotly)
library(mixOmics)
library(DiscriMiner)
```

```{r}
Date <- Sys.Date()

species <- "Oryza sativa"
# ncores = 16
# registerDoParallel(cores = ncores)

baits <- c("Os11g0587000","Os04g0550600","Os01g0746400","Os01g0700900","Os06g0565100","Os01g0700500","Os08g0547900")

sls <- c("Os01t0700300-00", "Os01t0700900-01", "Os01t0746400-01", "Os03t0263300-01", "Os03t0408600-01", "Os02t0817900-01", "Os04t0550600-01", "Os11t0587000-01", "Os12t0405200-01")
URPS <- c("Os08t0299400-01","Os03t0406100-01","Os10t0116800-01","Os11t0658900-01","Os02t0609000-01","Os11t0151700-01","Os08t0564000-01","Os08t0280100-01","Os05t0542150-01")

ref <- read.table("Rice_Annotation/Oryza_sativa.IRGSP-1.0.50.gtf", sep = "\t")      
raw_counts <- read.table("Data_Haider/raw_counts.tsv", sep = "\t", header = T)

meta <- readRDS("Data_Haider/RICE_META.rds")
```

```{r}

ref <- ref[which(ref$V3 == "transcript"),]
ref$ID <- gsub(".*gene_id ","",ref$V9)
ref$ID <- gsub(";.*","",ref$ID)

ref$transcript <- gsub(".*transcript_id ","",ref$V9)
ref$transcript <- gsub(";.*","",ref$transcript)

ref$biotype <- gsub(".*transcript_biotype ","",ref$V9)
ref$biotype <- gsub(";.*","",ref$biotype)

baits_transcript <- ref[which(ref$ID %in% baits),c(10:11)]
```

```{r}
meta <- meta[match(colnames(raw_counts[,-c(1:6)]),rownames(meta)),]

meta$growth_condition[which(meta$growth_condition == "-P")] <- "P-"
meta$growth_condition[which(meta$growth_condition == "+P")] <- "P+"

```

```{r}
#filtering low/no expression genes
counts <- raw_counts

colnames(counts)[-c(1:6)] <- meta$ID

conds <- unique(gsub("_1_.*","_",colnames(counts)[-c(1:6)]))

####filter per experimental condition on number of zeroes
i <- NULL
cond_zeroes <- data.frame(matrix())
for(i in 1:length(conds)){

  #print(grep(conds[i], colnames(samples)))

  zeroes <- rowSums(counts[,grep(conds[i], colnames(counts), fixed = T)] == 0)
  cond_zeroes <- cbind(cond_zeroes,zeroes)

}

cond_zeroes <- cond_zeroes[,-1]
colnames(cond_zeroes) <- conds

counts2 <- counts[rowSums(cond_zeroes==0) >= 3,]

cond_zeroes <- cond_zeroes[rowSums(cond_zeroes==0) >= 3,]

counts3 <- counts2[which(rowMax(data.matrix(counts2[,-c(1:6)])) < 5000),]

```

```{r}
rownames(meta) <- NULL

dds <- DESeqDataSetFromMatrix(countData = counts3[,-c(1:6)], 
                              colData = meta, 
                              design = ~ ID)

vsd <- varianceStabilizingTransformation(object = dds, 
                                         blind = TRUE,           # do not take the design formula into account. 
                                         # best practice for sample-level QC
                                         fitType = "parametric")

# extract the matrix of variance stabilised counts
variance_stabilised_counts <- assay(vsd)

saveRDS(vsd, "Data_Haider/Rice_Counts4ASCA.RDS")
saveRDS(meta, "Data_Haider/Rice_Meta4ASCA.RDS")
```

```{r real data import}

X <- readRDS("Data_Haider/Rice_Counts4ASCA.RDS")
meta <- readRDS("Data_Haider/Rice_Meta4ASCA.RDS")
colnames(meta)[3] <- c("time")

baits <- paste0(c("Os11t0587000","Os04t0550600","Os01t0746400","Os01t0700900"),"-01") 
#D27 CCD7 CCD8 CYP711A2
# baits2 <- c("Os01t0700300", "Os01t0701700")  
baits2 <- c("Os01t0700300-00", "Os02t0817900-01")  # identified methyltransferases
```

```{r}
meta$time <- as.character(meta$time)

meta$time[which(meta$time == "8")] <- "1"
meta$time[which(meta$time == "10")] <- "3"
meta$time[which(meta$time == "14")] <- "7"
meta$time[which(meta$time == "15")] <- "8"

X <- t(assay(X))
ref <- cbind.data.frame(Feature = colnames(X), feature = colnames(X), Baits = colnames(X))

REF <- fread("Rice_Annotation/IRGSP-1.0_representative_annotation_2022-09-01.tsv",sep = "\t", header = T)

ref <- ref[which(ref$Feature %in% REF$Transcript_ID),]
ref <- merge(ref,REF[,c("Transcript_ID","Description")], by.x = "Feature", by.y = "Transcript_ID")

```

```{r}
meta <- meta[order(meta$time),]
meta <- meta[order(meta$growth_condition),]
```

```{r}
X <- X[match(meta$ID, rownames(X)),]
```

```{r heatmap}
# h <- readRDS("heat_comp.RDS")
# levels(h[[2]]$time) <- c("1","3","7","8")

# make row annotation

row_ann <- data.frame("Gene" = rep(c("SL","URP"),each = 9))
rownames(row_ann) <-  c(sls,URPS)         #c(pathway,rownames(ASCA_cands3)[1:9])

ann_colours <- list(g_c = c("P-" = "#F8766D","P+" = "lightblue"),
                    time = c("1" = "#fde725","3" = "#35b779","7" = "#31688e","8" = "#440154"),
                    Gene = c("SL" = "purple", "URP" = "lightgreen"))

h <- t(X[,c(sls,URPS)])#t(X[,c(pathway,rownames(ASCA_cands3)[1:9])])  #psi_ref[,1]

col_ann <- meta
rownames(col_ann) <- col_ann$ID
col_ann <- col_ann[,-c(1)]
colnames(col_ann)[1] <- "g_c"

# png("MASCARA_heatmap.png")

heat <- pheatmap::pheatmap(t(scale(t(h))),  #
                             cluster_rows = F,
                             cluster_cols = F,
                             show_rownames = T,
                             show_colnames = F,
                             annotation_col = col_ann,
                           annotation_row = row_ann,
                           annotation_colors = ann_colours,
                           treeheight_row = 10,
                           fontsize = 8,
                             # annotation_row = anno,
                             main = "Scaled Gene Expression",silent = T)

  # # heat
  # library(ggplotify)
  # heat2 <- as.ggplot(heat) + theme(plot.title = element_text(face = "plain"))

  # saveRDS(heat2, "heatmap_MASCARA_Fig_2.RDS")

# png("MASCARA_heatmap.png")
# heat
# dev.off()

```

```{r supplementary heatmap id function table}
library(writexl)

supplementary <- REF[which(REF$Transcript_ID %in% c(sls,URPS)),]
supplementary <- as.data.frame(supplementary)
rownames(supplementary) <- supplementary$Transcript_ID
supplementary <- supplementary[c(sls,URPS),]
rownames(supplementary) <- NULL
write_xlsx(supplementary, "S_Table_1_Heatmap_ID_functions.xlsx")

```

```{r}
X <- X[,which(colnames(X) %in% ref$Feature)]
ref <- ref[match(colnames(X),ref$Feature),]

SL_ref <- ref[grep("trigolact",ref$Description),]

SL_ref <- ref[grep("trigolact",ref$Description),]
SL_ref <- SL_ref[-which(SL_ref$Feature %in% c("Os01t0701400-01","Os03t0203200-01","Os01t0763200-01","Os04t0668900-01","Os06t0154200-01","Os11t0104300-01","Os08t0250900-01", "Os03t0408600-01")),]

PSI_ref <-read.table("Rice_Annotation/PSI_genes.txt", sep = "\t", header = T)

GA_ref <- read_xlsx("Rice_Annotation/GA pathway genes-IDs.xlsx")
GA_ref$ID <- gsub("g","t",GA_ref$ID)
GAs <- ref[which(gsub("-.*","",ref$Feature) %in% GA_ref$ID),1]
```

```{r}

ref$Baits[-which(ref$Baits %in% c(sls,URPS))] <- "Other"
ref$Baits[which(ref$Baits %in% sls)] <- "POI"
ref$Baits[which(ref$Baits %in% URPS)] <- "URP"

ref$Baits <- factor(ref$Baits, levels = c("Other","POI","URP"))

```

```{r, echo = FALSE}
#### VST inverse #######
#mean/sd relationship plots
##Power transformation of counts/concentration matrix
m_sd_n <- M_SD(X, meta)

plot_normal <- m_sd_n[[3]]
```

```{r, echo=FALSE, include=FALSE}
# db <- cbind.data.frame(rownames(meta),meta[,c(7,11)],X)

db <- cbind.data.frame(meta,X)

# db[,1:3] <- apply(db[,1:3],2,as.character)
mc <- reshape2::melt(db)
colnames(mc)[4:5] <- c("Feature","Expression")
```

```{r, echo = FALSE}
dist_norm <- ggplot(mc, aes(x = Expression)) + geom_density(alpha = 0.4)+
  scale_y_continuous(labels = percent_format()) + 
  theme_bw()+
  ggtitle("Distribution")
```

```{r, echo=FALSE}
dist_nb <- ggplot(mc, aes(x = Expression)) + geom_density(alpha = 0.4)+
  scale_y_continuous(labels = percent_format()) + 
  scale_x_continuous(guide = guide_axis(check.overlap = TRUE)) +
  theme_bw()+
  ggtitle("NB transformed")

```

```{r, echo = FALSE, include = F}
m <- meta
m[,c("growth_condition","time")] <- lapply(m[,c("growth_condition","time")],factor)

res_ASCAplus_nb <- ASCA_decompose(d = m[,c("growth_condition","time")], x = X,
                               f = "growth_condition + time + growth_condition:time")
```

```{r, echo = FALSE}

 minT <-  res_ASCAplus_nb$decomposition$growth_condition + res_ASCAplus_nb$decomposition$`growth_condition:time`
  PCD <- prcomp(minT)

  newdata <- res_ASCAplus_nb$decomposition$growth_condition + res_ASCAplus_nb$decomposition$`growth_condition:time` + res_ASCAplus_nb$residuals

  PCDE <- predict(PCD, newdata)

```

```{r, echo = F, include = FALSE}
a_c.PCD <- PCD
SL <- SL_plot(a_c.PCD, meta)
```

# Background

The biological aim of the dataset is to find novel strigolactone (SL) genes. SLs are induced by phosphate starvation, but there are also a lot of more general phosphate deficiency related genes (PSIs) that are strongly affected by the experimental design.

In this case where SLs are the pathway of interest that can make it difficult to detect novel pathway genes, since the experimental design is affecting the pathway of interest along with many other genes.

Some SL genes are already characterised, we should use these to find novel pathway genes.

# Methodological aim

Find method that can preferentially detect members of a sub-network rather than more general experiment responders.

The genes should respond to the experiment but also show some experiment independent association.

# Data overview

```{r, echo = F, fig.cap="Top: distribution and mean sd relationship. Bottom: scores and loadings of the main effect; growth condition + interaction"}
ve1 <- paste0(round(summary(a_c.PCD)$importance[2,1:2] * 100, digits = 2), "%")

loadingplot <- function(df, meta, ve, ref, ...){
  #df is PCD$rotation

  #rematch REF, ref, df
  df <- as.data.frame(df)
  df <- cbind.data.frame(ref,df)

  ggplot(df, ...) +

    xlab(paste0("PC1 ",ve[1])) +
    ylab(paste0("PC2 ",ve[2])) +
    # geom_point(data = df[which(df$Baits == "Other"),], alpha = 0.5) +
    # geom_point(data = df[which(df$Baits == "POI"),], alpha = 1) +
    # geom_point(data = df[which(df$Baits == "URP"),], alpha = 1) +
geom_point(data = df) +
    scale_x_continuous(guide = guide_axis(check.overlap = TRUE)) +
    guides(shape = guide_legend(override.aes = list(size = 0.5))) +
    ggtitle("Loadings") +
    theme_bw()

}

l <- loadingplot(a_c.PCD$rotation, meta, ve = ve1, ref = ref, aes(x = PC1,y = PC2, colour = Baits, label = Description, alpha = Baits))  + theme(legend.title=element_blank())

l <- l + scale_color_manual(values = c("Other" = "black", "POI" = "purple","URP" = "green"))# + scale_alpha(c(0.1,1,1))

meta2 <- meta
colnames(meta2)[2] <- "g_c"

scoreplot <- function(df, meta, ve, ...){

  #df <- as.data.frame(df)
  df <- cbind.data.frame(meta,df)
  df$time <- factor(df$time) #, levels = c(1:4)
  df$g_c <- factor(df$g_c)

  ggplot(df, ...)+
    geom_point() +
    xlab(paste0("PC1 ",ve[1])) +
    ylab(paste0("PC2 ",ve[2])) + 
    theme_bw()

}
s <- scoreplot(PCDE, meta2, ve1, aes(x= PC1, y = PC2, colour = g_c, shape = time)) + ggtitle("Scores")

#+ guide_area()# + plot_layout(design = design)
qq <- qqplot(newdata, meta, aes(x=X_2000, y= X_1999, colour = growth_condition, shape = time))

ASCA_cands <- get_ASCA_cands(a_c.PCD)
colnames(ASCA_cands) <- c("VIP", "PC1","PC2")
#kable(head(ASCA_cands,10))

CAND_TAB <- round(ASCA_cands[1:10,], 4)
CAND_TAB <- cbind.data.frame(rownames(CAND_TAB), CAND_TAB)
colnames(CAND_TAB)[1] <- "Feature"

fill <- rep(c("grey95", "grey90"), nrow(CAND_TAB))
fill[which(CAND_TAB$Feature %in% baits)] <- "yellow"

theme1 <- ttheme_default(core = list(
  fg_params = list(fontface=c(rep("plain", nrow(CAND_TAB)))),
  bg_params = list(fill = fill)),
  base_size = 5, padding = unit(c(2, 2), "mm"))

candtab <- gridExtra::tableGrob(CAND_TAB, rows = NULL, theme = theme1)

(dist_norm + plot_normal)/(s + l)

```

<!-- ```{r} -->
<!-- png("rice_s_l.png") -->
<!-- (s + l) -->
<!-- dev.off() -->
<!-- ``` -->

```{r, include = F}
source("DATA_SIM_FUNCS.R")

pcd <- prcomp(res_ASCAplus_nb$residuals)
residual_cands <- get_ASCA_cands(pcd)

#HERE

image(cor(res_ASCAplus_nb$residuals[,c(baits,baits2)]))
```

```{r, inlcude = FALSE}

minT <-  res_ASCAplus_nb$decomposition$time
tcd <- prcomp(minT)

newdata <- res_ASCAplus_nb$decomposition$time + res_ASCAplus_nb$residuals
PCDE <- predict(PCD, newdata)

tcd <- prcomp(res_ASCAplus_nb$decomposition$time + res_ASCAplus_nb$residuals)

```

```{r}
#do permutation tests to see if the correlations in the "non experimentally interesting" data are significant between the strigolactone genes compared to random sets.

pathway <- ref[which(ref$Baits == "POI"),1]

p_n_b <- pathway[-which(pathway %in% baits)]

# sum(cor(res_ASCAplus_nb$residuals[,pathway]))
# 
# psi_ref <- PSI_ref[which(PSI_ref[,1] %in% colnames(res_ASCAplus_nb$residuals)),]

cor_val <- function(x,y){

  y <- y[which(y %in% colnames(x))]
  res <- (sum(cor(x[,y]))- length(y))/2   /length(y)
  return(res)

}

p_cor <- cor_val(res_ASCAplus_nb$residuals,sls)
phos_cor <- cor_val(res_ASCAplus_nb$residuals, URPS)

phos_p_cor <- cor_val(res_ASCAplus_nb$residuals, c(sls,URPS))

rand_cor <- c()
i <- NULL
for(i in 1:1000){

  rc <- cor_val(res_ASCAplus_nb$residuals, sample(colnames(res_ASCAplus_nb$residuals),13))
  rand_cor <- c(rand_cor, rc)

  }
```

The pathway genes are responding to the experiment, but they are not the only responders, nor are they the strongest responders. Therefore it is interesting to see if there is any significant correlation in the residuals of the ASCA model.

# Results

## Permutation test: average correlations in residuals

```{r, fig.cap="red: pathway genes, blue: General phosphate genes"}
hist(rand_cor, xlim = c(-0.5, 2), main = "Residual correlation permutation test", xlab = "Average correlation")
abline(v = c(p_cor,phos_cor), col = c("red","blue"))
# abline(v = phos_p_cor)

```

The SL genes (in red) have a much higher average correlation than a random set in the residuals, however this is also true for the PSIs (in blue).

```{r}

all <- cbind.data.frame(meta[,c("growth_condition","time")], X)
residuals <- cbind.data.frame(meta[,c("growth_condition","time")], res_ASCAplus_nb$residuals)
```

```{r}
#get cands from residuals

rPCD <- prcomp(data.matrix(residuals[,-c(1:2)]))

# residual_cands_all <- ranked_dist(pathway, rPCD)
# residual_cands_phos <- ranked_dist(PSI_ref[,1], rPCD)

```

```{r}
#filter residuals for only n genes that show treatment effect i.e. set VIP cutoff value
#how to do this with density?...

#first quick and dirty attempt with top n candidates  -  then add density

cutoff <- round(quantile(c(1:nrow(ASCA_cands)),0.1))
res_filt <- residuals[,c(1:2,which(colnames(residuals) %in% rownames(ASCA_cands)[1:cutoff]))]

rpcd <- prcomp(data.matrix(res_filt[,-c(1:2)]))

ref$Baits <- as.character(ref$Baits)
# ref$Baits[which(ref$Feature %in% PSI_ref[,1])] <- "PSI"
ref$Baits <- factor(ref$Baits, levels = c("Other","POI","URP"))

Fref <- ref[match(rownames(rpcd$rotation),ref$Feature),]

```

For the following analyses the data is prefiltered using the top 10% VIPs of the main effect model, since we expect the pathway to be induced by the experimental design.

## PCA on ASCA residuals

Since there is clearly an experiment independent relationship between the SL genes, it is interesting to see if there is any obvious structure in the residuals.
The structure relating to the experimental design, in the samples, is not interesting since we should have removed that in the main effect. 
Here we would like to see if the SL genes show any indication of clustering, and if that is in some way different to the PSIs.

```{r}

plots <- list()
plots_p <- list()
i <- NULL

k <- c(1:16)
l <- k[c(F,T)]
k <- k[c(T,F)]

importance <- paste0(round(summary(rpcd)$importance[2,] * 100, digits = 2), "%")

for(i in 1:8){

  plots[[i]] <- loadingplot(rpcd$rotation, meta, baits = Fref$Feature[which(Fref$Baits %in% c("POI","URP"))], c(importance[k[i]], importance[l[i]]), PCs =c(k[i],l[i]),  Fref, aes(x = !!sym(paste0("PC",k[i])),y = !!sym(paste0("PC",l[i])), colour = Baits, label = Description))

}
```

```{r}
wrap_plots(plots) + plot_layout(guides = "collect")
```

There isn't an obvious way to visually distinguish the two sets. But since the residual correlation analysis showed significant results we can investigate other methods on the residual matrix.

## sPLS on ASCA residuals

```{r}
# resids <- residuals[,-c(1:2)]
resids <- res_filt[,-c(1:2)]
spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], resids[,which(colnames(resids) %in% baits)], all.outputs = T)

sPLS_cands <- cbind.data.frame(abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X, spls_res$loadings$X)
sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = T),]

importance <- paste0(round(spls_res$prop_expl_var$X * 100, digits = 2), "%")

Fref2 <- Fref[-which(Fref$Feature %in% baits),]
Fref2 <- Fref2[match(rownames(sPLS_cands),Fref2$Feature),]

p <- loadingplot(sPLS_cands, meta, baits = Fref2$Feature[which(Fref2$Baits %in% c("POI","URP"))], 
            importance, PCs = c(1,2), ref = Fref2, aes(x = comp1, y = comp2, colour = Baits, label = Description))

# ggplotly(p)

```

<!-- ```{r} -->
<!-- #calculating direction of baits to determine direction of candidate selection spls loading space -->

<!-- # loadingplot(sPLS_cands, meta, baits = Fref2$Feature[which(Fref2$Baits %in% c("Pathway","PSI"))],  -->
<!-- #             importance, PCs = c(1,2), ref = Fref2, aes(x = comp1, y = comp2, colour = Baits, label = Description)) -->

<!-- angles <- rbind.data.frame(spls_res$loadings$Y, colMeans(spls_res$loadings$Y)) -->

<!-- x <- angles[4,] -->
<!-- y <- angles[5,] -->

<!-- #adapted from https://stackoverflow.com/questions/1897704/angle-between-two-vectors-in-r -->
<!-- angle <- function(x,y){ -->

<!--   x <- c(as.matrix(x)) -->
<!--   y <- c(as.matrix(y)) -->

<!--   dot.prod <- x%*%y  -->
<!--   norm.x <- norm(x,type="2") -->
<!--   norm.y <- norm(y,type="2") -->
<!--   theta <- acos(dot.prod / (norm.x * norm.y)) -->
<!--   return(as.numeric(theta)) -->
<!-- } -->

<!-- angles_loadings <- apply(sPLS_cands[,-1],1,angle, y = y) -->

<!-- #define this better -->
<!-- sPLS_cands <- cbind.data.frame(sPLS_cands, "angle" = angles_loadings, "score" = 3 * sPLS_cands$`abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X` * (1-angles_loadings)) -->

<!-- sPLS_cands <- sPLS_cands[order(sPLS_cands$score, decreasing = T),] -->
<!-- ``` -->

<!-- ```{r} -->
<!-- avg_point <- angles[5,]/5 -->

<!-- rot.mat <- matrix(c(0,-1,1,0),nrow = 2, byrow = T)  #90 degree clockwise rotation -->

<!-- rotated_point <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat) -->

<!-- ang <- angle(avg_point,c(1,0)) -->

<!-- rot.mat2 <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation -->
<!-- rotated_point2 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat2) -->

<!-- # rot.mat3 <- matrix(c(cos(ang),sin(ang),-sin(ang),cos(ang)),nrow = 2, byrow = T) #anticlockwise rotation -->
<!-- # rotated_point3 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat3) -->

<!-- ggplot() +  -->
<!--   geom_point(data = sPLS_cands, aes(x = comp1, y = comp2), alpha = 0.1) + -->
<!--   geom_point(data = sPLS_cands[1:100,], aes(x = comp1, y = comp2)) + -->
<!--   geom_segment(data = avg_point, aes(x = 0, y = 0, xend = comp1, yend = comp2)) + -->
<!--   # geom_segment(data = rotated_point, aes(x = 0, y = 0, xend = V1, yend = V2)) + -->
<!--   geom_segment(data = rotated_point2, aes(x = 0, y = 0, xend = V1, yend = V2)) + -->
<!--   # geom_segment(data = rotated_point3, aes(x = 0, y = 0, xend = V1, yend = V2)) + -->

<!--   # xlim(c(-0.1,0.1))+ -->
<!--   # ylim(c(-0.1,0.1)) + -->

<!--   theme_bw() -->
<!--   # geom_smooth(method = "lm") -->

<!-- ``` -->

<!-- ```{r} -->
<!-- #rotated loadings -->
<!-- ang <- angle(avg_point,c(1,0)) -->

<!-- rot.mat2 <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation -->

<!-- rotated_point2 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat2) -->

<!-- rotate <- function(x,y){ -->
<!--   x <- c(as.matrix(x)) -->
<!--   x %*% y -->
<!-- } -->

<!-- loadings_r_baits <- t(apply(sPLS_cands[,c("comp1","comp2")],1,rotate, y = rot.mat2)) -->
<!-- colnames(loadings_r_baits) <- c("comp1","comp2") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ggplot() +  -->
<!--   geom_point(data = loadings_r_baits, aes(x = comp1, y = comp2), alpha = 0.1) + -->
<!--   geom_point(data = loadings_r_baits[1:100,], aes(x = comp1, y = comp2)) + -->
<!--   theme_bw() -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #density  -->
<!-- density <- kde2d(loadings_r[,1], loadings_r[,2], n = 200) -->
<!-- image(as.matrix(density$z), col = hcl.colors(20)) -->

<!-- ``` -->

<!-- ```{r} -->
<!-- range.ang <- seq(from = -pi/6, to = pi/6, length.out = 60) -->

<!-- ### -->

<!-- i <- NULL -->
<!-- L_R <- data.frame() -->
<!-- LOADINGS_R <- list() -->

<!-- for(i in 1:length(range.ang)){ -->

<!-- ang <- range.ang[i] -->

<!-- rot.mat <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation -->
<!-- loadings_r <- t(apply(loadings_r_baits[,c("comp1","comp2")],1,rotate, y = rot.mat)) -->
<!-- colnames(loadings_r) <- c("comp1","comp2") -->
<!-- L_R <- rbind(L_R,loadings_r) -->
<!--   LOADINGS_R[[i]] <- loadings_r -->

<!-- } -->

<!-- # colnames(LOADINGS_R) <- c("comp1","comp2") -->
<!-- L_R$rad <- rep(range.ang, each = nrow(sPLS_cands)) -->
<!-- L_R$frame <- rep(c(1:length(range.ang)), each = nrow(sPLS_cands)) -->

<!-- ``` -->

<!-- ```{r} -->
<!-- #get sd of original comp1 distribution -->
<!-- sigma <- sd(sPLS_cands$comp1) -->

<!-- big.T.sig <- function(x,y){ -->
<!--   big <- length(which(x > 2*y)) -->
<!--   return(big) -->
<!-- } -->

<!-- i <- NULL -->
<!-- res <- c() -->
<!-- for(i in 1:length(LOADINGS_R)){ -->

<!--   r <- big.T.sig(LOADINGS_R[[i]][,1],sigma) -->
<!--   res <- c(res,r) -->
<!-- } -->

<!-- ``` -->

<!-- ```{r} -->
<!-- RES <- cbind.data.frame("Rad" = range.ang, "Density" = res) -->

<!-- density <- ggplot(RES, aes(x = Rad, y = Density))+ -->
<!--   geom_smooth() + theme_bw() + ggtitle("Number of points above 2*sigma of original rotation") -->

<!-- density -->

<!-- ``` -->

<!-- ```{r} -->

<!-- #taken from https://stackoverflow.com/questions/6836409/finding-local-maxima-and-minima -->
<!-- inflect <- function(x, threshold = 1){ -->
<!--   up   <- sapply(1:threshold, function(n) c(x[-(seq(n))], rep(NA, n))) -->
<!--   down <-  sapply(-1:-threshold, function(n) c(rep(NA,abs(n)), x[-seq(length(x), length(x) - abs(n) + 1)])) -->
<!--   a    <- cbind(x,up,down) -->
<!--   list(minima = which(apply(a, 1, min) == a[,1]), maxima = which(apply(a, 1, max) == a[,1])) -->
<!-- } -->

<!-- ggres <- ggplot_build(density)$data[[1]] -->
<!-- infs <- inflect(ggres$y) -->
<!-- rads <- ggres$x[infs$maxima] -->

<!-- ``` -->

<!-- ```{r} -->
<!-- maxima.ang <- rads -->

<!-- ### -->

<!-- i <- NULL -->
<!-- # L_R <- data.frame() -->
<!-- sPLS_maxima_cands <- list() -->

<!-- for(i in 1:length(maxima.ang)){ -->

<!-- ang <- maxima.ang[i] -->

<!-- rot.mat <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation -->
<!-- loadings_r <- as.data.frame(t(apply(sPLS_cands[,c("comp1","comp2")],1,rotate, y = rot.mat))) -->
<!-- colnames(loadings_r) <- c("comp1","comp2") -->
<!-- # L_R <- rbind(L_R,loadings_r) -->
<!--   sPLS_maxima_cands[[i]] <- loadings_r[order(loadings_r$comp1, decreasing = T),] -->

<!-- } -->
<!-- ``` -->

<!-- ```{r} -->

<!-- ggplotly( -->
<!--   ggplot() +  -->
<!--   geom_point(data = sPLS_maxima_cands[[1]], aes(x = comp1, y = comp2), alpha = 0.1) + -->
<!--   geom_point(data = sPLS_maxima_cands[[1]][1:20,], aes(x = comp1, y = comp2)) + -->
<!--   theme_bw() -->
<!-- ) -->

<!-- ``` -->

```{r}
coexp_cands <- ranked_coexp(baits,X[,which(colnames(X) %in% colnames(resids))])

residual_cands <- ranked_dist(baits[1:3], rpcd)

residual_coexp <- ranked_coexp(baits, resids)

```

```{r}
Methods <- c("sPLS_ASCA","PCA_DIST_ASCA","Correlations_ASCA", "Correlations")

getrank <- function(RES,geneset,NOINFO = 26402){
  # geneset <- 
  # NOINFO = length(geneset) - length(ranks)
  ranks <- which(rownames(RES) %in% geneset)
  if(length(ranks) < length(geneset)){
    ranks <- c(ranks,rep(nrow(RES) + 1, length(geneset) - length(ranks))) #NOINFO
  }

  Ranks <- cbind.data.frame("Geneset" = geneset, "Rank" = ranks)
  return(Ranks)

}

psis <- PSI_ref[which(PSI_ref[,1] %in% colnames(resids)),1]
GAs <- GAs[which(GAs %in% colnames(resids))]

sPLS_ASCA <- rbind.data.frame(getrank(sPLS_cands,p_n_b),getrank(sPLS_cands,psis),getrank(sPLS_cands,GAs))
sPLS_ASCA <- cbind.data.frame(sPLS_ASCA, "Method" = rep("sPLS_ASCA", nrow(sPLS_ASCA)))

#####

PCA_DIST_ASCA <- rbind.data.frame(getrank(residual_cands,p_n_b),getrank(residual_cands,psis),getrank(residual_cands,GAs))
PCA_DIST_ASCA <- cbind.data.frame(PCA_DIST_ASCA, "Method" = rep("PCA_DIST_ASCA", nrow(PCA_DIST_ASCA)))

Correlations <- rbind.data.frame(getrank(coexp_cands,p_n_b),getrank(coexp_cands,psis),getrank(coexp_cands,GAs))
Correlations <- cbind.data.frame(Correlations, "Method" = rep("Correlations", nrow(Correlations)))

Correlations_ASCA <- rbind.data.frame(getrank(residual_coexp,p_n_b),getrank(residual_coexp,psis),getrank(residual_coexp,GAs))
Correlations_ASCA <- cbind.data.frame(Correlations_ASCA, "Method" = rep("Correlations_ASCA", nrow(Correlations_ASCA)))

RES <- rbind.data.frame(sPLS_ASCA, PCA_DIST_ASCA, Correlations_ASCA, Correlations)

RES$Geneset[which(RES$Geneset %in% p_n_b)] <- "SLs"
RES$Geneset[which(RES$Geneset %in% psis)] <- "Phosphate"
RES$Geneset[which(RES$Geneset %in% GAs)] <- "GAs"

q <- ggplot(RES[-which(RES$Geneset == "GAs"),], aes(x = Method, y = Rank, fill = Geneset)) + geom_boxplot() + theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust=1))
```

## Methods compared

```{r, fig.cap="Method Ranks on Real Data"}
q

```

Ranked correlations outperforms the other methods at detecting SL genes, however it also detects PSIs as top candidates, due to the strong between correlations. Using the ASCA residuals to investigate the experiment independent associations helps to overcome this. Ranked correlations on ASCA residuals and sPLS on ASCA residuals both perform well at detecting the SLs and not detect the PSIs. The PSIs are ranked as least important with sPLS_ASCA, this would suggest the method is performing the required task best.

We can explain the larger variation, and slightly worse ranking of sPLS_ASCA compared to Correlations_ASCA since with sPLS we rank the features based on VIP scores and therefore do not take the direction of the relationship into account, where at least 2 of the features used to evaluate the methods were detected through directional ranked correlations.

Since looking at the distance between the SL genes in the PCA space did not yield the required results we drop this method at this point for simplicity.

# QQplots of some top candidates

Through assessing the relationship between one bait and the 9 top candidates of each method tested we highlight the problem with basic correlations, with 4 of the top 9 candidates showing little to no within group correlations. However it is difficult to determine which of the two residual based methods are performing best (within the top 20 candidates of the two there are 10 common features). Both are able to detect more complex relationships, after accounting for variance explained by the experimental design (plots below).

It makes more sense to use sPLS as this models the response in a multivariate way as opposed to averaged pairwise rankings in the correlation approach.

In this limited view we can see that sPLS seems to detect features that have a nonlinear relationship across the experiment, through linear relations in the ASCA residuals. However, sPLS as implemented through mixOmics shows an artefact (still uncharacterised) in previous simulations.

# Questions

- What metric to use in place of visualisation of pairwise relationships to measure method performance in this context?
- If the relationship between SLs in the PCA of the ASCA residuals is unclear, how to simulate their relationship?
- Could the filtering be automated or turned into an optimisation problem, instead of using arbitrary cutoff values?
- Would automated selection of an "axis of interest" in the main effect model improve candidate selection - i.e. features need to behave in a more strictly similar manner (positive or negative) to baits?
- Is it possible and would it make sense to project the baits into the latent space of the PLS model to define an axis/direction?

```{r}

qqplots_spls <- list()
qqplots_coexp <- list()
qqplots_pca <- list()
qqplots_resid_cor <- list()

qqplots_spls_SR <- list()

i <- NULL
for(i in 1:9){
  qqplots_spls[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(sPLS_cands)[i])],
                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time))

  qqplots_coexp[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(coexp_cands)[i])],
                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time))

  qqplots_resid_cor[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(residual_coexp)[i])],
                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time))

}

wrap_plots(qqplots_coexp) + plot_layout(guides = "collect") + plot_annotation(title = "Correlations")
wrap_plots(qqplots_spls) + plot_layout(guides = "collect") + plot_annotation(title = "sPLS_ASCA")
wrap_plots(qqplots_resid_cor) + plot_layout(guides = "collect") + plot_annotation(title = "Correlations_ASCA")

```

### 2023-12-19 R/Haider_preproc.Rmd
23_12_19

---
title: "Haider_preproc"
author: "Fred White"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(VennDiagram)
library(gplots)
library(ggplot2)
library(GGally)
library(knitr)
library(limma)
library(reshape2)
library(RColorBrewer)
library(WGCNA)
library(DESeq2)
library(edgeR)
library(igraph)
library(scales)
library(doParallel)
library(ggfortify)
library(biomaRt)
library(GenomicRanges)
library(pheatmap)
library(tidyverse)
library(ggforce)
library(grid)
library(devtools)
library(readxl)
```

```{r}
Date <- Sys.Date()

species <- "Oryza sativa"
# ncores = 16
# registerDoParallel(cores = ncores)

baits <- c("Os11g0587000","Os04g0550600","Os01g0746400","Os01g0700900","Os06g0565100","Os01g0700500","Os08g0547900")

ref <- read.table("Rice_Annotation/Oryza_sativa.IRGSP-1.0.50.gtf", sep = "\t")      
raw_counts <- read.table("Data_Haider/raw_counts.tsv", sep = "\t", header = T)

meta <- readRDS("Data_Haider/RICE_META.rds")
```

```{r}

ref <- ref[which(ref$V3 == "transcript"),]
ref$ID <- gsub(".*gene_id ","",ref$V9)
ref$ID <- gsub(";.*","",ref$ID)

ref$transcript <- gsub(".*transcript_id ","",ref$V9)
ref$transcript <- gsub(";.*","",ref$transcript)

ref$biotype <- gsub(".*transcript_biotype ","",ref$V9)
ref$biotype <- gsub(";.*","",ref$biotype)

baits_transcript <- ref[which(ref$ID %in% baits),c(10:11)]
```

```{r}
meta <- meta[match(colnames(raw_counts[,-c(1:6)]),rownames(meta)),]
```

```{r}
#filtering low/no expression genes
counts <- raw_counts

colnames(counts)[-c(1:6)] <- meta$ID

conds <- unique(gsub("_1_.*","_",colnames(counts)[-c(1:6)]))

####filter per experimental condition on number of zeroes
i <- NULL
cond_zeroes <- data.frame(matrix())
for(i in 1:length(conds)){

  #print(grep(conds[i], colnames(samples)))

  zeroes <- rowSums(counts[,grep(conds[i], colnames(counts), fixed = T)] == 0)
  cond_zeroes <- cbind(cond_zeroes,zeroes)

}

cond_zeroes <- cond_zeroes[,-1]
colnames(cond_zeroes) <- conds

counts2 <- counts[rowSums(cond_zeroes==0) >= 4,]

cond_zeroes <- cond_zeroes[rowSums(cond_zeroes==0) >= 4,]

counts3 <- counts2[which(rowMax(data.matrix(counts2[,-c(1:6)])) < 5000),]

```

```{r}
rownames(meta) <- NULL

dds <- DESeqDataSetFromMatrix(countData = counts3[,-c(1:6)], 
                              colData = meta, 
                              design = ~ ID)

vsd <- varianceStabilizingTransformation(object = dds, 
                                         blind = TRUE,           # do not take the design formula into account. 
                                         # best practice for sample-level QC
                                         fitType = "parametric")

# extract the matrix of variance stabilised counts
variance_stabilised_counts <- assay(vsd)

saveRDS(vsd, "Data_Haider/Rice_Counts4ASCA.RDS")
saveRDS(meta, "Data_Haider/Rice_Meta4ASCA.RDS")
```

```{r}
###DATA Preparation
#
#remove low variance and lowly expressed genes - this increases robustness of network
variance_stabilised_counts <- cbind(counts3[,1:6], variance_stabilised_counts)
counts_filt <- variance_stabilised_counts[apply(variance_stabilised_counts[,-c(1:6)], 1, var) > quantile(apply(variance_stabilised_counts[,-c(1:6)], 1, var),VARLIM),]

baits_transcript <- baits_transcript[which(baits_transcript$transcript %in% rownames(counts_filt)),]

meta_save <- meta

counts_filt <- counts_filt[,-c(1:6)]

#########
```

```{r}
##heatmap of strigolactone genes

```

```{r}

##Co-expression network construction from filtered variance stabilised counts

#colnames(counts_filt) <- meta$sample_identifier

adj_matrix <- cor(t(counts_filt))

corclean <- function(cor_mat){

  ####get rid of upper triangle of correlation matrix 
  #this line is problematic for downstream analysis
  #cor_mat[upper.tri(cor_mat)] <- 999

  melt_adj <- melt(cor_mat)

  #melt_adj <- melt_adj[-which(melt_adj$value == 999),]
  melt_adj <- melt_adj[-which(melt_adj$Var1 == melt_adj$Var2),]

  return(melt_adj)

}

melt_adj <- corclean(adj_matrix)

rm(adj_matrix)

#madj <- melt_adj

############
#melt_adj <- melt_adj[which(melt_adj$value > CC_Cutoff),]

#take baits and return top n (user defined) coexpression candidates
#make function that takes each bait, and it's top coexpression candidates
baiter <- function(x, melt_adj){

  if (x %in% gsub("\\..*","",melt_adj$Var1) == F){
    print(paste0(x," doesn't pass filters"))
    CoExpCands <- data.frame(matrix(c(x,"doesn't pass filters", 0), ncol = 3))
  } else{

    CoExpCands <- melt_adj[which(gsub("\\..*","",melt_adj$Var1) == x),]
    CoExpCands <- CoExpCands[order(CoExpCands$value, decreasing=T),]

    #write.table(CoExpCands, paste0(x,"_Top_CoExpCands.txt"), row.names = F, quote = F)

  }
  rownames(CoExpCands) <- NULL
  colnames(CoExpCands) <- c("Bait","Target","PCC")

  return(CoExpCands)

}

####
####changed to madj from melt_adj below to ignore CC_Cutoff
####

baits_transcript <- baits_transcript[c(2,3,4),]

#RES contains data frame of ordered coexpression candidates for each bait
#with no CC value cutoff this is very slow
i <- NULL
RES <- list()
for(i in 1:dim(baits_transcript)[1]){

  RES[[i]] <- baiter(baits_transcript[i,2], melt_adj)
  print(i)

}

rm(melt_adj)

names(RES) <- baits_transcript[,2]

##########
#####THROW OUT RANKS FOR THE TRANSCRIPTS THAT DONT PASS FILTERS
#change thsi to make it reusable !!!!!!!!!!
RES <- RES[c(which(do.call(rbind, lapply(RES,dim))[,1] > 1))]

##make list before and name it accordingly
l <- lapply(RES, "[[",2)
l <- lapply(l, as.character)

common <- Reduce(intersect, l)

############
#####heatmaps can still be plotted with lower CCs
############

#find top 200 coexpression candidate based on rank
#very big table so need to first optimise
#find out number of coexpression candidates for each bait

com_filt <- function(res,common){
  res <- res[which(res$Target %in% common),]
  res <- cbind(res, c(1:dim(res)[1]))
  res <- res[order(res$Target),]
  colnames(res)[4] <- "Rank"
  return(res)
}

rres <- lapply(RES, com_filt, common = common)

####cbind(?) ranks together 
i <- NULL
ranked <- data.frame()
for(i in 1:length(rres)){

  if(i == 1){

    ranked <- rres[[i]][,c(2,4)]

  }else{

    ranked <- cbind(ranked,rres[[i]][,4])

  }

}

#change this with RES
colnames(ranked)[-1] <- baits_transcript[c(which(do.call(rbind, lapply(RES,dim))[,1] > 1)),2]

ranked$comb_rank <- rowSums(ranked[,-1])/(dim(ranked)[2] - 1)
ranked <- ranked[order(ranked$comb_rank),]

##################

###need to cleanup colnames of counts_filt

colnames(counts_filt) <- meta$ID

#######

forhm <- counts_filt[which(rownames(counts_filt) %in% ranked[1:200,1]),]
baitsforhm <- counts_filt[which(gsub("\\..*","",rownames(counts_filt)) %in% baits_transcript[,2]),]

meta$day <- factor(as.numeric(as.character(meta$day)) - 7)

anno <- cbind(meta$growth_condition, as.character(meta$day), meta$genotype)

#anno <- gsub("WT_","",meta$gc)
anno <- as.data.frame(anno)
rownames(anno) <- meta$ID

colnames(anno) <- c("growth_condition","day","genotype")
anno$day <- factor(anno$day, level = c("1","3","7","8"))

#colnames(anno) <- "Condition"

#translate labels from IDs to gene names

fornet <- rbind(baitsforhm,forhm)
labels <- rownames(fornet)

SL_genes <- read.table("~/SL_genes.txt", sep = "\t",header =T)
PSI_genes <- read.table("~/PSI_genes.txt", sep = "\t",header =T)

btg <- baits_transcript
rownames(btg) <- btg$transcript

btg$Gene <- c("MAX1-Os900","OsCCD8","OsCCD7")

rownames(btg) <- btg$transcript

labels[-which(gsub("\\..*","",labels) %in% baits_transcript[,2])] <- ""

#labels[grep(paste(labels,collapse = "|"), btg$transcript)] == btg$transcript
labels[1:3] <- btg$Gene

meta$finid <- paste0(meta$genotype, "_",meta$growth_condition, "_Day",meta$day,"_R")

meta$order <- c(1:nrow(meta))
meta <- meta[order(meta$finid),]
meta$finid <- paste0(meta$finid, c(rep(1:3,7)))

meta <- meta[order(meta$order),]

colnames(fornet) <- meta$finid

fornet <- cbind(rownames(fornet), rownames(fornet), fornet)
colnames(fornet)[1:2] <- c("transcript_id","gene_id")
fornet$gene_id <- gsub("-.*","",fornet$gene_id)
fornet$gene_id <- gsub("t","g",fornet$gene_id)

rownames(anno) <- colnames(fornet[,-c(1:2)])

heat <- pheatmap::pheatmap(data.matrix(fornet[,-c(1:2)]), 
                           #                           color = brewer.pal(name = "RdYlBu"),
                           cluster_rows = TRUE,
                           cluster_cols = TRUE, 
                           show_rownames = TRUE, 
                           show_colnames = FALSE,
                           labels_row = labels,
                           annotation_col = anno,
                           main = "Top 200 Ranked Targets + Baits",silent = T)
#pdf("2021-05-17Coexpression/hm6_3.pdf")
pdf(paste0(Date,"Coexpression_Combined/BaitsCandidates_hm.pdf"))
add.flag(heat,
         kept.labels = labels,
         repel.degree = 0.5)
#a$grobs[[6]]$gp$fontsize <- 8
#plot(a)

dev.off()

#saveRDS(fornet, "~/spectral_decomposition/RICE_CANDS.rds")
#saveRDS(meta, "~/spectral_decomposition/RICE_META.rds")

##change granscripg to transcript id 
fornet$gene_id[which(fornet$gene_id == "granscripg")] <- fornet$transcript_id[which(fornet$gene_id == "granscripg")]

#merge with ref$biotype
bionet <- merge(ref[,c(11,12)], fornet, by.x = "transcript", by.y = "transcript_id")

library(ape)

locusref <- read.gff("/zfs/omics/personal/fwhite/REFS/IRGSP-1.0_representative/locus.gff")
locusref$note <- gsub(".*Note=","",locusref$attributes)
locusref$note <- gsub("\\..*","",locusref$note)
locusref$note <- gsub(" \\(Os.*","",locusref$note)

locusref$note <- gsub("%2C",",",locusref$note)
locusref$note <- gsub("%3B",";",locusref$note)

locusref$ID <- gsub("ID=","",locusref$attributes)
locusref$ID <- gsub(";.*","",locusref$ID)

t <- locusref[which(locusref$ID %in% fornet$gene_id),]
u <- t[,c(10,11)]

v <- fornet$gene_id[-which(fornet$gene_id %in% t$ID)]
v <- cbind.data.frame(c(rep("NA",length(v))),v)

colnames(v) <- colnames(u)

w <- rbind(u,v)

bionet <- merge(w,fornet, by.x = "ID", by.y = "gene_id")

write.table(fornet, paste0(Date,"Coexpression_Combined/Baits_Candidates_Normalised_Expression.txt"), quote = F, row.names = F, sep = "\t")

write.table(fornet[heat$tree_row$order,], paste0(Date,"Coexpression_Combined/Baits_Candidates_Normalised_Expression_ORDERED.txt"), quote = F, row.names = F, sep = "\t")

write.table(bionet[heat$tree_row$order,], paste0(Date,"Coexpression_Combined/Baits_Candidates_Normalised_Expression_ORDERED_biotype_task1.txt"), quote = F, row.names = F, sep = "\t")

PCD <- prcomp(t(counts_filt), scale. = F)

#force uncertainty ellipses 
pdf(paste0(Date,"Coexpression_Combined/pcaTOMreduced_prescaled3.pdf"))
autoplot(PCD, data = meta, colour = "growth_condition", shape = "genotype", size = "day") + #, frame = T) + #, frame.type = 'norm') + too few points to calculate ellipse
  #scale_shape_manual(values=1:nlevels(as.factor(meta$compartment))) +
  #ggforce::geom_mark_ellipse(aes(color = compartment)) + #fill
  theme_bw()
dev.off()

###PCAfor just task 1 samples######################################################

#get rid of the task2 and 3 samples from counts_filt and then meta

to <- unique(c(grep("GR24", meta$growth_condition),grep("d10",meta$genotype)))

counts_filt <- counts_filt[,-to]
meta <- meta[-to,]

PCD <- prcomp(t(counts_filt), scale. = F)

#force uncertainty ellipses 
pdf(paste0(Date,"Coexpression_Combined/pcaTOMreduced_prescaled1.pdf"))
autoplot(PCD, data = meta, colour = "growth_condition", shape = "day") + #, frame = T) + #, frame.type = 'norm') + too few points to calculate ellipse
  #scale_shape_manual(values=1:nlevels(as.factor(meta$compartment))) +
  #ggforce::geom_mark_ellipse(aes(color = compartment)) + #fill
  theme_bw()
dev.off()

#########

#plot density on normal log2 data and vsd data
densityplot <- function(counts){

  mc <- melt(counts)
  colnames(mc) <- c("Gene_ID","Sample_ID","Counts")

  plot <- ggplot(mc, aes(x = Counts)) + geom_density(alpha = 0.5)+
    scale_y_continuous(labels = percent_format()) + theme_bw()+
    ggtitle("Count Distribution")
  ggsave(paste0(Date,"Coexpression/Variance_stabilised_Distribution.pdf"), plot = plot)

}

densityplot(counts_filt)

df <- PCD$rotation#[c(1,147,294,442,589,736),]
#df <- melt(df)
#colnames(df) <- c("feature","component","loading")
#df <- df[which(df$component %in% c("PC1","PC2","PC3","PC4","PC5","PC6")),]

#df$feature <- as.character(df$feature)

df <- cbind.data.frame(rownames(df), df)
colnames(df)[1] <- "metabolite"

#remove 1900 and 5100 from baits
#replot the above with the rest of the baits labelled 
baits_transcript$Gene <- c("MAX1-Os900","OsCCD8","OsCCD7")

df$colour <- df$metabolite
df$colour[-which(df$colour %in% baits_transcript$transcript)] <- "non-baits"
df$colour[which(df$colour %in% baits_transcript$transcript)] <- "baits"

pdf(paste0(Date,"Coexpression_Combined/RICE_feature_loadings_normal_PCA.pdf"))
ggplot(df, aes(x = PC1,y = PC2, colour = colour))+#, colour = metabolite)) + 
  geom_point() +
  geom_label(data = df[which(df$metabolite %in% baits_transcript$transcript),], aes(label = metabolite), nudge_y = 0.003, nudge_x = 0.001) +
  geom_point(data = df[which(df$metabolite %in% baits_transcript$transcript),]) +
  theme_bw()
dev.off()

```

### 2023-12-19 R/Paper_markdown.Rmd
23_12_19

@@
 # csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-6th-edition.csl
-csl: apa-6th-edition.csl
+csl: vancouver.csl
 keep_md: true
@@
 #powerpoint_presentation
+
+# author: Fred T.G. White^[1,2], Anna Heintz-Buschart^[1], Lemeng Dong^[2], Harro J. Bouwmeester^[2], Johan A. Westerhuis^[1], Age K. Smilde^[1]
+
+# author: "Fred T.G. White, Anna Heintz-Buschart, Lemeng Dong, Harro J. Bouwmeester, Johan A. Westerhuis, Age K. Smilde"
+
+# Fred T.G.
+White^[Biosystems Data Analysis Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands, Plant Hormone Biology Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands *], Anna Heintz-Buschart^[Biosystems Data Analysis Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands], Lemeng Dong^[Plant Hormone Biology Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands], Harro J.
+Bouwmeester^[Plant Hormone Biology Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands], Johan A.
+Westerhuis^[Biosystems Data Analysis Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands], Age K.
+Smilde^[Biosystems Data Analysis Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands]
+
+# - name: Fred T.G. White
+#     email: f.t.g.white@uva.nl
+#     affiliation: 
+#       - University of Amsterdam
+#       - Another example
+#     corresponding: alice@example.com
+#   - name: Bob Security
+#     email: bob@example.com
+#     affiliation: 
+#       - Another University
+#       - Some Institute of Technology
+# address:
+#   - code: Univeristy of Amsterdam
+#     address: Department 1, Street, City, State, Zip
+#   - code: Another example
+#     address: Department 2, Street, City, State, Zip
 ```
@@
 Plant transcriptome studies typically involve the use of designed experiments which aim to induce variation in pathways of interest (POIs) by controlling one or more experimental factors.
-This can be done through, for example, varying the level of certain essential nutrients, or knocking out/down a gene within a POI.
+This can be done through, for example, varying the level of certain essential nutrients, knocking out/down a gene within a POI.
 To explore which pathways are affected by the intervention, gene expression can be measured through RNA sequencing (RNAseq).
@@
 
+<!-- For the rest of the manuscript we take the strigolactones as our example biological pathway. -->
+
 ```{r nettype, fig.cap= c("Pathway types exemplified, POI; pathway of interest, RP; related pathway, URP; unrelated pathway.
@@
 Figure \@ref(fig:nettype) outlines pathway nomenclature that will be used throughout this text.
-*A* represents a true hypothetical pathway including a pathway of interest (POI), related pathway (RP) and an unrelated pathway (URP).
+Panel *A* represents a true hypothetical pathway including a pathway of interest (POI), related pathway (RP) and an unrelated pathway (URP).
 All three types of pathway are assumed to be affected (dis-)proportionally by the experimental design.
@@
 It is the excessive between-group variance that leads to the densely connected nature of *B* from \@ref(fig:nettype).
+Concretising this with a biological example; in @haiderTranscriptomeAnalysisPhosphate2023 a phosphate (P) starvation was applied to rice plants to induce strigolactone (SL) biosynthesis, to study the change in gene expression in comparison to control plants with normal supplies of P.
+The SL pathway is known to be triggered through P deficiency.
+However, as P is one of three key nutrients for plant growth, deficiency of this nutrient causes widespread differential expression throughout the transcriptome (@wangRoleStrigolactonesDeficiency2021).
+
 Figure \@ref(fig:withinbetween) exemplifies between-group and within-group driven correlations from a real dataset (@haiderTranscriptomeAnalysisPhosphate2023).
@@
 However, both may contain valuable biological information.
-The between-group variance is in principle larger because groups are the result of treatments that are usually selected because they affect the genes and/or pathways under investigation.
-Within-group variance (variance across replicates), if too large, can be an issue in DE analysis, we argue that in co-expression analysis the information present in this variance may be valuable.
+The between-group variance is in principle larger because groups are the result of treatments that are usually selected because they affect the POI.
+Within-group variance (variance across replicates), if too large, can be an issue in DE analysis, we argue that in co-expression analysis the information present in this variance can be valuable.
 Variance between replicates is caused by non-controlled variations in plant development and environmental conditions.
-One replicate within a group may be slightly bigger than the others and therefore respond slightly different to a treatment, for example phosphate deficiency.
-A larger plant may start to suffer from the deficiency earlier than a smaller plant.
-This does not imply that the larger plant shows a different biological response to P deficiency.
-We can assume that the response is similar, but just occurs a bit faster.
-This in turn implies that the expression of the bait genes are induced a little bit faster than in the other replicates, but this will also be true for the genes that we want to identify with a co-expression analysis.
+One replicate within a group may be slightly bigger than the others and therefore respond slightly different to a treatment.
+This could results in slight differences in bait genes between the replicates, but also the genes that we want to identify with a co-expression analysis.
 There may be many other factors that are inducing this within-group variance, but it is quite likely that if a factor induces a change in the treatment response in one replicate, that this response difference occurs throughout the entire process or pathway in that one replicate.
@@
 Not making the distinction between types of variance results in *total CoE* associations driven by the differences between treatment groups as opposed to within group correlations.
-For example in @haiderTranscriptomeAnalysisPhosphate2023 a phosphate (P) starvation was applied to rice plants to induce strigolactone (SL) biosynthesis, to study the change in gene expression in comparison to control plants with normal supplies of P.
-The SL pathway is known to be triggered through P deficiency.
-However, as P is one of three key nutrients for plant growth, deficiency of this nutrient causes widespread differential expression throughout the transcriptome (@wangRoleStrigolactonesDeficiency2021).
-@haiderTranscriptomeAnalysisPhosphate2023 used ranked correlations, similar to MutRank (@poretskyMutRankShinyWebapplication2020), to determine (total) coexpression candidates.
+@haiderTranscriptomeAnalysisPhosphate2023 used ranked Pearson correlations to determine (total) coexpression candidates.
 Due to such a strong effect of the experiment, many genes were affected on top of the desired SL genes.
@@
 
+<!-- , similar to MutRank (@poretskyMutRankShinyWebapplication2020), -->
+
 ```{r withinbetween, fig.cap= c("Correlation examples, Pearson’s correlation coefficient (r) indicated in facet titles.
-X axes, expression levels of CCD8 gene in rice, Y axes expression levels of example coexpression candidates.
-Top row A-B; *total CoE*, bottom row C-D *within-group CoE*.
+Data from @haiderTranscriptomeAnalysisPhosphate2023.
+X axes, expression levels of MAX1 gene (Os01t0700900–01) in rice, Y axes expression levels of example coexpression candidates, aribitrary units.
+Top row A-B; *total CoE*, bottom row C-D *within-group CoE* from variance partitioning the data in A-B.
 Gene pairs in both A and B have strong total correlations.
 Correlation in A is also exhibited within a group, B has strong total correlations mainly due to the differences between groups.
-These examples are indistinguishable from eachother without variance partitioning which has been performed on the same genes in C-D.")}
+These examples are indistinguishable from eachother without variance partitioning.")}
 
@@
 Methods like ranked correlations and partial least squares regression (PLS; @geladiPartialLeastsquaresRegression1986; @woldCollinearityProblemLinear1984) are able to investigate relationships between a set of baits and the other measured genes.
-In ranked correlations the relationships are quantified at a bivariate level i.e.
-bait 1 - candidate 1, bait 1 - candidate 2.
+In ranked correlations the relationships are quantified at a bivariate level.
 In PLS regression the relationships can be investigated in a multivariate way, in its simplest form we can see which genes are most predictive of the expression of one bait.
@@
 To explore the utility of MASCARA, the method will be compared to ASCA, PLS2 and correlation analysis illustrated using an example RNAseq dataset (from @haiderTranscriptomeAnalysisPhosphate2023).
-Furthermore, this dataset is used as a guide to several simulation studies.
+Furthermore, this dataset is used as a guide to several simulation studies to validate the development of a novel approach to coexpression analysis.
 For these simulations we emulate both different levels of between- and within-group variance.
 We test the effects of size of between group differences, number of URP differential genes, measurement noise (random unstructured technical variance), structured ambient variance and number of replicates.
-The following sections seek to investigate which statistical methods are best suited for biosynthetic POI elucidation when a set of baits is already known and illustrate the use benefit of MASCARA, particularly in data coming from experiments designed for differential expression analysis.
+The following sections seek to investigate which statistical methods are best suited for POI elucidation when a set of baits is already known and illustrate the benefit of MASCARA.
+Especially in data coming from experiments designed for differential expression analysis.
 
@@
 
-In order to guide the simulations we illustrate the characteristics of the simulated data with a real dataset.
+In order to guide the simulations we illustrate the desired characteristics with a real dataset.
 The real data are a subset of samples from a study in which the root transcriptome of rice plants is measured under two  growth condition (P+/P-) and measured at four time points (1, 3, 7 and 8  days post treatment).
 At each combination of growth condition and time point, the root transcriptome of three new plants was harvested for RNA sequencing.
-In this dataset, the curated list of SL pathway genes show a specific profile: low variance baseline expression across all time points in the P+ condition but activation and increasing expression over time with P starvation.
-
-Figure \@ref(fig:Dataset1-Overview) shows the gene expression levels of 9 strigolactone pathway genes along with 9 highly differential genes that are not part of the SL pathway.
+In this dataset, the curated list of SL pathway genes (contained in Supplementary Table 1)  show a specific profile: low variance baseline expression across all time points in the P+ condition but activation and increasing expression over time with P starvation, data were generated and preprocessed to variance stabilised counts as per @haiderTranscriptomeAnalysisPhosphate2023.
+
+Figure \@ref(fig:Dataset1-Overview) shows the gene expression levels of 9 strigolactone pathway genes along with 9 highly differential genes that are induced by P deficiency, but are not part of the SL pathway.
 While the expression is low in P+ condition a clear increasing time profile can be observed in the P- condition, indicating an interaction between the growth condition and the time factors.
@@
 Figure \@ref(fig:ASCA-1) shows the output of the ASCA decomposition in which the PCA scores and loadings of the combined effect of the  Growth condition factor and the interaction of Growth condition and Time is explored.
-The score plot shows that the replicates of each condition are clustered while the difference between the conditions is much larger.
-The loading plot shows that although the 9 SL genes are mostly on the outside of the plot, many other genes are as well.
+The score plot shows that the replicates of each condition are clustered while the difference between the conditions is much larger, as expected.
+While the 9 SL transcripts have higher loadings on the 1st PC, there are many non-SL transcripts with similarly (or more) extreme loadings.
 Thus many differentially expressed genes are found important that are not part of the SL pathway, in other words ASCA does not detect SL (POI) genes from others that are similarly or more differentially expressed but not part of the SL pathway (URPs).
 
-This marks the goal of this paper, which is to detect within-group coexpressed genes (POI/RP) with a set of baits preferentially to other differentially expressed genes that are not part of the same pathway (URPs, with no shared within-group structure).
+This marks the goal of this paper, which is to detect within-group coexpressed genes (POI/RP) with a set of baits preferentially to other differentially expressed genes that are not part of the same pathway (URPs, with a different within-group structure).
 
@@
 Data and preprocessing from @haiderTranscriptomeAnalysisPhosphate2023, genes autoscaled i.e.
-blue indicates no/low expression and red indicates higher expression.")}
-
-# (rdo[[1]] + heat2)/(rdo[[2]] + rdo[[3]])
-# layout <- rbind(c(1,1,2,2,2,2),
-#                 c(1,1,2,2,2,2),
-#                 c(1,1,2,2,2,2),
-#                 c(1,1,2,2,2,2),
-#                 c(NA,NA,NA,NA,NA,NA),
-#                 c(3,3,3,4,4,4),
-#                 c(3,3,3,4,4,4),
-#                 c(3,3,3,4,4,4),
-#                 c(3,3,3,4,4,4))
-# grid.arrange(rdo[[1]], heat2, rdo[[2]],rdo[[3]], layout_matrix = layout)
-# rdo[[3]]$layers[[1]]$aes_params$colour <- "green"
-# rdo[[3]]$layers[[2]]$aes_params$colour <- "purple"
-
-##Error in discrete_range() : could not find function "discrete_range"
-# rdo[[3]] <- rdo[[3]] + scale_color_manual(values = c("green","purple"), breaks = c("Other","Pathway"), labels = c("Other","SL")) + theme(legend.title=element_blank())
-
-# withr::with_options(
-#   list(ggplot2.discrete.fill = c("green","purple")),
-#   print(rdo[[3]])
-# )
-
-# s <- readRDS("rice_ASCA_scores.RDS")
-# l <- readRDS("rice_ASCA_loadings.RDS")
+blue indicates no/low expression and red indicates higher expression.
+IDs with functional annotations in supplementary table 1.")}
 
@@
 heat2
-#/(rdo[[2]] + rdo[[3]])
-
-# pdf("Rice_ASCA.pdf")
-# heat2/(rdo[[2]] + rdo[[3]])
-# dev.off()
-
-```
-
-```{r ASCA-1, echo = F, out.width = '100%', fig.cap=c("Scores and loadings of the ASCA model of the combined effect (condition + condition:time interaction), model indicates samples and genes relative positions in the designed variance.")}
+```
+
+```{r ASCA-1, echo = F, out.width = '100%', fig.cap=c("Scores and loadings of the ASCA model of the combined effect (condition + condition:time interaction), model indicates samples and genes relative positions in the designed variance.
+*g_c* indicates growth condition, *time* indicates days after starvation induction.")}
 knitr::include_graphics("rice_s_l.png")
@@
 \end{equation}
-<!-- equation \@ref(eq:nCER).  -->
 
@@
 
-## Analysis and feature selection
+## Selection of genes of interest
 
@@
 
-Pearson's correlation coefficient is calculated between all columns of \(\bm{Y}\).
-For ranked correlations Pearson's *r* is calculated between each bait and all other genes.
-Ranks are assigned to all non-baits transcripts based on *r* with the lowest rank for the highest correlation.
-These ranks are averaged across all baits, similarly to @poretskyMutRankShinyWebapplication2020.
+Pearson's correlation coefficient *r* is calculated between all columns of \(\bm{Y}\).
+Ranks are assigned to all non-bait transcripts for each bait based on *r*, with the lowest rank for the highest correlation.
+These ranks are averaged across all baits.
+
+<!-- (see supplementary equation \@ref(eq:rankcor)). -->
 
@@
 \begin{equation}
+\begin{aligned}
 \mathbf{Y} = \mathbf{X} \mathbf{\Theta} + \mathbf{E}
+\end{aligned}(\#eq:glm)
 \end{equation}
@@
 \begin{equation}
+\begin{aligned}
 \hat{\mathbf{\Theta}} = (\mathbf{X}^\intercal \mathbf{X})^{-1} \mathbf{X}^\intercal \mathbf{Y}
+\end{aligned}(\#eq:ols)
 \end{equation}
@@
 \begin{equation}
+\begin{aligned}
 \mathbf{M}_f = \mathbf{X}_f^* \hat{\mathbf{\Theta}}
+\end{aligned}(\#eq:effects)
 \end{equation}
 
-where: \(\mathbf{M}_f\) is the effect matrix corresponding to effect \(f\), \(\hat{\mathbf{\Theta}}\) is the matrix of estimated parameters obtained from OLS and \(\mathbf{X}_f^*\) is a new model matrix obtained by keeping in \(\mathbf{X}\) only the block \(\mathbf{X}_f\) and replacing all other columns with zeros as per:
-
-\(\mathbf{M}_f=(0|\mathbf{X}_f |0)\mathbf{\hat{\Theta}} = \mathbf{X}_f \mathbf{\hat{\Theta}}_f\)
+where: $\mathbf{M}_f$ is the effect matrix corresponding to effect $f$, $\hat{\mathbf{\Theta}}$ is the matrix of estimated parameters obtained from OLS and $\mathbf{X}_f^*$ is a new model matrix obtained by keeping in $\mathbf{X}$ only the block $\mathbf{X}_f$ and replacing all other columns with zeros as per:
+
+\begin{equation}
+\begin{aligned}
+\mathbf{M}_f=\left[0|\mathbf{X}_f |0\right]\mathbf{\hat{\Theta}} = \mathbf{X}_f \mathbf{\hat{\Theta}}_f
+\end{aligned}(\#eq:scim)
+\end{equation}
 
@@
 
-Furthermore, the matrix \(\mathbf{E}\) is estimated as the residual matrix of the model:
+Furthermore, the matrix $\mathbf{E}$ is estimated as the residual matrix of the model:
 
@@
 \hat{\mathbf{E}} = \mathbf{Y} - \mathbf{X} \hat{\mathbf{\Theta}}
-\end{aligned}
+\end{aligned}(\#eq:ascaresid)
 \end{equation}
 
-<!-- In this work we calculate the model: -->
+Throughout this text when referring to the calculation of ASCA/ASCA+ model on simulated or real data the following model is calculated:
+
+\begin{equation}
+\begin{aligned}
+\mathbf{Y} = \mathbf{M}_{\alpha} + \mathbf{M}_{\beta} + \mathbf{M}_{\alpha\beta} + \mathbf{E}
+\end{aligned}(\#eq:ascamodel)
+\end{equation}
+
+With genes selected from:
+
+\begin{equation}
+\begin{aligned}
+\mathbf{\hat{Y}} =& \mathbf{M}_{\alpha} + \mathbf{M}_{\alpha\beta}
+                =& \mathbf{T}_{\alpha}\mathbf{P}_{\alpha}^T + \mathbf{T}_{\alpha\beta}\mathbf{P}_{\alpha\beta}^T
+                =& \mathbf{T}_{\alpha + \alpha\beta}\mathbf{P}_{\alpha + \alpha\beta}^T
+\end{aligned}(\#eq:ascagenes)
+\end{equation}
 
@@
 
-Partial Least Squares (PLS2) is a multivariate regression technique used for modeling relationships between two sets of variables.
-Here we split our data matrix $\mathbf{Y}$ into sub-matrices; $\mathbf{Y} = \left[\mathbf{\tilde{Y}} : \mathbf{B}\right]$
-
-The response matrix \( \mathbf{B} \) contains only the baits (relating to the green bar in the loadings of Figure \@ref(fig:simOverview)) and the predictor matrix \( \mathbf{\tilde{Y}} \) contains the rest of the data, this can then be expressed as:
+Partial Least Squares (PLS2) is a multivariate regression technique used for modeling relationships between two sets of variables. Here we split our data matrix $\mathbf{Y}$ into sub-matrices;
+
+\begin{equation}
+\begin{aligned}
+\mathbf{Y} = \left[\mathbf{\tilde{Y}} | \mathbf{B}\right]
+\end{aligned}(\#eq:plspart)
+\end{equation}
+
+The response matrix $\mathbf{B}$ contains only the baits (relating to the green bar in the loadings of Figure \@ref(fig:simOverview)) and the predictor matrix $\mathbf{\tilde{Y}}$ contains the rest of the data, this can then be expressed as:
 
@@
 \mathbf{B} = \mathbf{T}\mathbf{Q}^T + \mathbf{E}_B \\
-\end{aligned}
+\end{aligned}(\#eq:plsmodel)
 \end{equation}
@@
    a.
-Compute the singular value decomposition (SVD) of \( \mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k} \) to obtain the left and right singular vectors \( \mathbf{u}_k \) and \( \mathbf{v}_k \) respectively.
+Compute the singular value decomposition (SVD) of \( \mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k} \) to obtain the left and right singular vectors; weights \( \mathbf{u}_k \) and \( \mathbf{v}_k \) respectively.
    b.
 Project \( \mathbf{\tilde{Y}}_{k} \) and \( \mathbf{B}_{k} \) onto the first left and right singular vectors (\( \mathbf{u}_k \), \( \mathbf{v}_k \)) to obtain the scores \( \bm{t}_k = \mathbf{\tilde{Y}}_{k} \mathbf{u}_k \) and \( \bm{w}_k = \mathbf{B}_{k} \mathbf{v}_k \).
-   c.
+
+Under the conditions of orthogonal scores; $\mathbf{t}_{k}^{T}\mathbf{t}_{k} = 0$ and normalised weights; $\mathbf{u}_{k}^T\mathbf{u}_{k{}} = 1$ and $\mathbf{v}_{k}^T\mathbf{v}_{k{}} = 1$.
+
+   <!-- c.
 Perform linear regression of \( \mathbf{\tilde{Y}}_k \) and \( \mathbf{B}_k \) onto their corresponding scores \( \bm{t}_k \) and \( \bm{w}_k \) to obtain the respective loadings \( \bm{p}_k \), and \( \bm{q}_k \).
-   d.
+-->
+   <!-- d.
 Deflate \( \mathbf{\tilde{Y}}_k \) and \( \mathbf{B}_k \) by removing the variance explained by the $k^{th}$ set of scores in the space of $\mathbf{\tilde{Y}}$\( \bm{t}_k \) and use the deflated matrices as inputs for the next iteration.
+-->
+
+<!-- \begin{equation} -->
+<!-- \begin{aligned} -->
+<!-- \mathbf{\tilde{Y}_{k}} = \mathbf{\tilde{Y}}_{k-1} - \mathbf{t}_{k}\mathbf{p}^T_{k} \\ -->
+<!-- \text{and} \\ -->
+<!-- \mathbf{B_{k}} = \mathbf{B}_{k-1} - \mathbf{t}_{k}\mathbf{q}^T_{k} \\ -->
+<!-- \end{aligned} -->
+<!-- \end{equation} -->
+
+PLS2 aims to maximize the covariance between the scores \( \bm{t}_k \) and \( \bm{w}_k \) at each iteration.
+In the following simulations and application the number of components calculated by PLS2 is set to 2 to create an interpretable model.
+
+### Variable importance in projection (VIP)
+
+In this work genes are selected in both ASCA+ and PLS through VIP which is a way to measure the contributions of each variable to the underlying models. According to:
 
@@
 \begin{aligned}
-\mathbf{\tilde{Y}_{k}} = \mathbf{\tilde{Y}}_{k-1} - \mathbf{t}_{k}\mathbf{p}^T_{k} \\
-\text{and} \\
-\mathbf{B_{k}} = \mathbf{B}_{k-1} - \mathbf{t}_{k}\mathbf{q}^T_{k} \\
-\end{aligned}
+\text{VIP}_{j} = \sqrt{\frac{ \sum_{k=1}^{K} H^2_{jk} \cdot \text{VE}_{k}}{\text{VE}_{total} \cdot K}}
+\end{aligned}(\#eq:vip)
 \end{equation}
 
-PLS2 aims to maximize the covariance between the scores \( \bm{t}_k \) and \( \bm{w}_k \) at each iteration.
-In the following simulations and application the number of components calculated by PLS2 is set to 2 to create an interpretable model.
-
-### Variable importance in projection (VIP)
-
-In this work genes are selected in both ASCA+ and PLS through VIP which is a way to measure the contributions of each variable to the underlying models. According to:
-
-$\text{VIP}_{j} = \sqrt{\frac{ \sum_{k=1}^{K} \mathbf{H}^2_{jk} \cdot \text{VE}_{k}}{\text{VE}_{total} \cdot K}}$
-
-Where k=1...K represents the number of components in the respective models, j indexes a gene,  \(\bm{H}\) is substituted for the loadings or weights (for ASCA+ and PLS respectively) and VE is the variance explained, where $VE_{k}$ is the amount of variance explained by component k and $VE_{total}$ is the variance explained by the whole model.
+Where k=1...K represents the number of components in the respective models, j indexes a gene,  $H$ is substituted for the loadings $\mathbf{P}_{\alpha + \alpha\beta}^T$ or weights $\mathbf{U}$ (for ASCA+ and PLS respectively) and VE is the variance explained, where $VE_{k}$ is the amount of variance explained by component k and $VE_{total}$ is the variance explained by the whole model.
 In ASCA+ we calculate the VIP scores in the latent space of \(\bm{Y}_{\alpha + \alpha\beta}\) as described in @thielASCAAPCAExtensions2017.
@@
 What remains in this \(\hat{\mathbf{E}}\) matrix is individual variance of each plant that is assumed to contain not just random technical variance but some form of structured variance induced by ambient (uncontrolled) effects.
-The matrix \(\hat{\mathbf{E}}\) is used as input to PLS2  after being split into baits and remainder.
+The matrix \(\hat{\mathbf{E}}\) from equation \@ref(eq:ascaresid) is substituted as $\mathbf{Y}$ into PLS2  after being split into baits and remainder as per equation \@ref(eq:plspart).
 
@@
 This approach allows a general direction in the PLS space to be defined by multiple baits and assumes that these baits are highly correlated with one another.
-The MASCARA pipeline in exemplified in Figure \@ref(fig:MASCARAOVERVIEW).
-
-```{r MASCARAOVERVIEW, echo = F, fig.cap="MASCARA pipeline overview"}
-a <- knitr::include_graphics("MASCARA_Pipeline_overview.png")
-a
-```
+
+<!-- The MASCARA pipeline in exemplified in Figure \@ref(fig:MASCARAOVERVIEW). -->
+
+<!-- ```{r MASCARAOVERVIEW, echo = F, fig.cap="MASCARA pipeline overview"} -->
+<!-- a <- knitr::include_graphics("MASCARA_Pipeline_overview.png") -->
+<!-- a -->
+<!-- ``` -->
 
@@
 
-To evaluate performance we use log2 geometric mean rank (GMR) ($\bar{\gamma}$), defined as: 
-$\bar{\gamma} = log2((\prod_{j_{spike~}=1}^{J_{spike}}\gamma_{j_{spike}})^{1/J_{spike}})$ 
-for each spike $j_{spike}$ for $J_{spike}$ spikes. A lower $\gamma$ indicates better performance.
+To evaluate performance we use log2 geometric mean rank (GMR; $\bar{\gamma}$), defined as:
+
+\begin{equation}
+\begin{aligned}
+\bar{\gamma} = log2((\prod_{j_{spike}=1}^{J_{spike}}\gamma_{j_{spike}})^{1/J_{spike}})
+\end{aligned}(\#eq:GMR)
+\end{equation}
+
+for $J_{spike}$ spikes. A lower $\gamma$ indicates better performance.
 
@@
 ```{r tc, echo = F, fig.cap= "Results of simulation 1 for all 4 methods.
-Log2 geometric mean rank (GMR) as a function of the number of differentially expressed genes (x-axis) and combined effect size (y axis)."}
+Log2 geometric mean rank (GMR) as a function of the number of differentially expressed URP genes (x-axis) and combined effect size (y axis); ($CE = \\|\\mathbf{T}_{\\alpha + \\alpha\\beta}\\mathbf{P}_{\\alpha + \\alpha\\beta}^T\\|_F$), lower GMR; dark blue, indicates better performance.
+ASCA model calculated per equation \\@ref(eq:ascamodel), PLS model per equation \\@ref(eq:plsmodel)."}
 wrap_plots(plots) + plot_layout(guides = "collect")
-
-# pdf("coexp.pdf")
-# wrap_plots(plots) + plot_layout(guides = "collect")
-# dev.off()
 
@@
 
-```{r repres, echo = F, fig.cap= "Simulation 3 results.
+```{r repres, echo = F, fig.cap= "Simulation 2 results.
 Replications and variance type feasibility ranges.
@@
 Each facet is created with set parameters for combined effect size ($CE = \\|\\mathbf{T}_{\\alpha + \\alpha\\beta}\\mathbf{P}_{\\alpha + \\alpha\\beta}^T\\|_F$) and random gaussian noise size ($E = \\|\\mathbf{E}\\|_F$).
-Dots are median performance across 20 random structures."}
+Dots are median performance across 20 random structures.
+ASCA model calculated per equation \\@ref(eq:ascamodel), PLS model per equation \\@ref(eq:plsmodel)."}
 
@@
 As shown in simulation 1 (Figure \@ref(fig:tc)) higher numbers of DE URP genes results in higher geometric mean rank of the spikes for correlations, ASCA and PLS.
-MASCARA is designed specifically to mitigate the presence of large numbers of DE URP genes.
+MASCARA is designed specifically to mitigate the effect of large numbers of DE URP genes.
 Here we show that this is achieved.
@@
 meta <- readRDS("Data_Haider/Rice_Meta4ASCA.RDS")
-colnames(meta)[c(6,7,11)] <- c("gc","time","growth_condition")
+# colnames(meta)[c(6,7,11)] <- c("gc","time","growth_condition")
+
+X <- t(assay(X))
+
+colnames(meta)[3] <- "time"
 
@@
 
-#remove RP samples
-X <- X[-which(meta$growth_condition == "PR"),]
-meta <- meta[-which(meta$growth_condition == "PR"),]
-
+# #remove RP samples
+# X <- X[-which(meta$growth_condition == "PR"),]
+# meta <- meta[-which(meta$growth_condition == "PR"),]
+# 
 meta$time <- as.character(meta$time)
@@
 meta$time[which(meta$time == "8")] <- "1"
-meta$time[which(meta$time == "10")] <- "2"
-meta$time[which(meta$time == "14")] <- "3"
-meta$time[which(meta$time == "15")] <- "4"
+meta$time[which(meta$time == "10")] <- "3"
+meta$time[which(meta$time == "14")] <- "7"
+meta$time[which(meta$time == "15")] <- "8"
 
@@
 meta <- meta[order(meta$time),]
-meta <- meta[order(meta$gc),]
+meta <- meta[order(meta$growth_condition),]
 ```
@@
 meta2 <- meta
-colnames(meta2)[11] <- "g_c"
+colnames(meta2)[2] <- "g_c"
 
@@
 This limitation affected many more genes than those directly involved in SL biosynthesis.
-Furthermore, there appears to be a stronger experiment-independent relationship amongst the SL genes compared to their correlations with other DE genes, as displayed in Figure \@ref(fig:residualdists) where the blue distribution indicates the 9 transcripts from figures \@ref(fig:Dataset1-Overview) and \@ref(fig:ASCA-1) and the red is the relationship between these 9 and other top differential genes.
-
-```{r residualdists, fig.cap=c("Distributions of Fisher transformed Pearson correlations in residuals from the ASCA model.
-Blue indicates correlations between SL pathway genes, red indicates correlations between SL pathway genes and top 10% (non-SL) DE genes; upregulated in P-")}
+Furthermore, there appears to be a stronger experiment-independent relationship amongst the SL genes compared to their correlations with other DE genes, as displayed in Figure \@ref(fig:residualdists) where the blue distribution indicates residual correlations between the 9 POI transcripts from figures \@ref(fig:Dataset1-Overview) and \@ref(fig:ASCA-1) and the red indicates residual correlations between these 9 and top DE URP genes.
+
+```{r residualdists, fig.cap=c("Distributions of Fisher transformed Pearson correlations (supplementary equation \\@ref(eq:Fisher)) in residuals from the ASCA model on the data from @haiderTranscriptomeAnalysisPhosphate2023.
+Blue indicates correlations between SL pathway genes, red indicates correlations between SL pathway genes and top 10% DE URP genes (upregulated in P- over P+).")}
 
@@
 ```{r}
-# ggp_rand_cor <- as.data.frame(rand_cor)
-# colnames(ggp_rand_cor)[1] <- "Cor"
-# dist <- ggplot(ggp_rand_cor) + geom_density(aes(x = Cor))
-#
-
-```
-
-```{r}
 all <- cbind.data.frame(meta[,c("growth_condition","time")], X)
@@
 ```{r}
-#filter residuals for only n genes that show treatment effect i.e. set VIP cutoff value
-#how to do this with density?...
-
-#first quick and dirty attempt with top n candidates  -  then add density
-
 cutoff <- round(quantile(c(1:nrow(ASCA_cands)),0.1))
@@
 rotated_point2 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat2)
-
-# rot.mat3 <- matrix(c(cos(ang),sin(ang),-sin(ang),cos(ang)),nrow = 2, byrow = T) #anticlockwise rotation
-# rotated_point3 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat3)
-
-# ggplot() + 
-#   geom_point(data = sPLS_cands, aes(x = comp1, y = comp2), alpha = 0.1) +
-#   geom_point(data = sPLS_cands[1:100,], aes(x = comp1, y = comp2)) +
-#   geom_segment(data = avg_point, aes(x = 0, y = 0, xend = comp1/10, yend = comp2/10)) +
-#   # geom_segment(data = rotated_point, aes(x = 0, y = 0, xend = V1, yend = V2)) +
-#   geom_segment(data = rotated_point2, aes(x = 0, y = 0, xend = V1/10, yend = V2/10)) +
-#   # geom_segment(data = rotated_point3, aes(x = 0, y = 0, xend = V1, yend = V2)) +
-# 
-#   # xlim(c(-0.1,0.1))+
-#   # ylim(c(-0.1,0.1)) +
-# 
-#   theme_bw()
-#   # geom_smooth(method = "lm")
 
@@
 Some example top candidate genes are shown in Figure \@ref(fig:realcandidates).
-The between versus within problem is again exemplified here.
+The problem of dominant between group variation is again exemplified here.
 Several of the top candidates detected by correlations (ASCA and PLS, results not shown) show no indication of within group correlations.
@@
 ```{r realcandidates, fig.cap="Selected candidates from A-C ranked correlations and D-F MASCARA.
-X axes; expression level of one of the bait genes (CCD8), y axes; expression levels of three example coexpression candidates for each method."}
+X axes; transformed expression level of one of the bait genes (CCD8), y axes; expression levels of three example coexpression candidates for each method (see @haiderTranscriptomeAnalysisPhosphate2023 for preprocessing details)."}
 
@@
 This framework enables the investigation of multivariate relationships between a set of baits and the rest of the data.
-MASCARA thrives on the fact that within the typical experimental setup small systematic variance exists between replicates due to variations in ambient factors that affect plant growth, and therefore the variance structures in the resulting data.
+MASCARA thrives on the fact that within the typical experimental setup small systematic variance exists between replicates due to variations in ambient factors that affect the plant, and therefore the variance structures in the resulting data.
 
@@
 There are many established methods for feature selection in PLS which include Variable Importance in Projection (VIP) and Selectivity Ratio (@rajalahtiBiomarkerDiscoveryMass2009), however these methods focus on the importance of variables in the construction of the model rather than a directional association with the response block as with the $\mathbf{\bar{q}}$ target projection stage.
-With this averaged target projection approach we can use the information in a set of baits to define a direction in the latent space which can then be used as an axis to determine the relatedness of other genes.
+With this averaged target projection approach, as an extension of Kvalheim's target projection in PLS1 models, we can use the information in a set of baits to define a direction in the latent space which can then be used as an axis to determine the relatedness of other genes.
 This assumes moderate to high correlation in the residuals of the baits, which is a realistic assumption in coexpression analysis although should be verified in the selection of baits.
@@
 The method is not limited to the calculation of positive associations, however performance in detecting negative correlations (indicative of inhibitory or negative feedback processes) or more complex relationships has not been tested within the current simulations.
-Our approach was based off the structure of the Haider RNAseq dataset, however the method is applicable across other omics types.
+Our approach was based off the structure of the Haider RNAseq dataset, however the method is applicable across other organisms and omics types.
 This holds under the assumption that the set of baits have a multivariate relationship between themselves and multiple other (unknown) genes.
@@
 
+We acknowledge funding by the Dutch Research Council (NWO/OCW) for the MiCRop Consortium program, Harnessing the second genome of plants (Grant number 024.004.014; to HB, LD, AKS, JAW), the Dutch Research Council (NWO-TTW grant 16873 Holland Innovative Potato; to HB and LD) and the Data Science Centre of the University of Amsterdam (to FW).
 FW would like to thank Frans van der Kloet, Fentaw Abegaz, Roel van der Ploeg for statistical discussions and Imran Haider for consultation on the strigolactone pathway.
-FW thanks the Data Science Centre of the University of Amsterdam for partial financial support.
 
@@
 ```
+
+<!-- Ranked correlations are calculated from the correlation matrix $\mathbf{C} as: -->
+
+<!-- \begin{equation} -->
+<!-- \begin{aligned} -->
+<!-- RANKS  -->
+<!-- \end{aligned}(\#eq:rankcor) -->
+<!-- \end{equation} -->
+
+To calculate Fisher transformed Pearson correlations (z) with Pearson's r:
+
+\begin{equation}
+\begin{aligned}
+z = \frac{1}{2}ln(\frac{1 + r}{1-r})
+\end{aligned}(\#eq:Fisher)
+\end{equation}
+
+<!--   r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}} -->
 

### 2023-12-19 R/DATA_SIM_FUNCS.R

@@ -1140,7 +1140,7 @@ qqplot <- function(data, meta, ...){
   
   df <- cbind.data.frame(meta,data)
   
-  df$time <- factor(df$time, levels = c(1:4))
+  df$time <- factor(df$time)#, levels = factor(df$time))
   df$growth_condition <- factor(df$growth_condition)
   
   

### 2023-12-19 R/Paper_markdown.Rmd
23_11_12_2

@@
 All three types of pathway are assumed to be affected (dis-)proportionally by the experimental design.
-Note; URPs are not necessarily completely unlinked to POIs or RPs, in the biological sense, but are assumed here to be several degrees away.
-Far enough that the effect can be approximated orthogonally to the coexpression, when modelling.
+URPs defined to be pathways not directly interacting with the POI.
 When dealing with experimentally designed data, we have to consider two key sources of variance: between-group variance and within-group variance.
@@
 Within-group variance, on the other hand, is the variance within experimental groups, this is not interesting for DE studies but may contain information useful to CoE.
-It is the excessive between-group variance that leads to the densely connected nature of *B* from \@ref(fig:nettype).
+It is the excessive between-group variance that leads to the densely connected nature of panel *B* from \@ref(fig:nettype).
 Concretising this with a biological example; in @haiderTranscriptomeAnalysisPhosphate2023 a phosphate (P) starvation was applied to rice plants to induce strigolactone (SL) biosynthesis, to study the change in gene expression in comparison to control plants with normal supplies of P.
@@
 Ranks are assigned to all non-bait transcripts for each bait based on *r*, with the lowest rank for the highest correlation.
-These ranks are averaged across all baits.
-
-<!-- (see supplementary equation \@ref(eq:rankcor)). -->
+These ranks are averaged across all baits (see supplementary equation \@ref(eq:rankcor)).
 
@@
 
-To evaluate performance we use log2 geometric mean rank (GMR; $\bar{\gamma}$), defined as:
+To evaluate performance we use log2 geometric mean rank (log2(GMR); $\bar{\gamma}$), defined as:
 
@@
   theme_bw() +
+  labs(fill = "log2(GMR)") +
   theme(plot.title = element_text(size=8)) 
@@
 ```{r tc, echo = F, fig.cap= "Results of simulation 1 for all 4 methods.
-Log2 geometric mean rank (GMR) as a function of the number of differentially expressed URP genes (x-axis) and combined effect size (y axis); ($CE = \\|\\mathbf{T}_{\\alpha + \\alpha\\beta}\\mathbf{P}_{\\alpha + \\alpha\\beta}^T\\|_F$), lower GMR; dark blue, indicates better performance.
+Log2 geometric mean rank (log2(GMR)) as a function of the number of differentially expressed URP genes (x-axis) and combined effect size (y axis); ($CE = \\|\\mathbf{T}_{\\alpha + \\alpha\\beta}\\mathbf{P}_{\\alpha + \\alpha\\beta}^T\\|_F$), lower log2(GMR); dark blue, indicates better performance.
 ASCA model calculated per equation \\@ref(eq:ascamodel), PLS model per equation \\@ref(eq:plsmodel)."}
@@
 
-  One_Balance[[i]] <- ggplot(mRES[which(mRES$nCERs == orders[i]),], aes(x = Replicate_Number, y = log2(F1_Score), colour = Method)) +
+  One_Balance[[i]] <- ggplot(mRES[which(mRES$nCERs == orders[i]),], aes(x = Replicate_Number, y = log2(F1_Score), colour = Method, shape = Method)) +
   scale_color_viridis(discrete = T)+
@@
 
-# ps1 <- grid.arrange(ps, left = "Log2(GMR)", bottom = "Replicate Number                     ")
-
-a <- knitr::include_graphics("C:/Users/fwhite/Documents/replication_res.png")
+ps1 <- grid.arrange(ps, left = "log2(GMR)", bottom = "Replicate Number                     ")
+
+a <- knitr::include_graphics("replication_res.png")
 a
@@
 
-<!-- Ranked correlations are calculated from the correlation matrix $\mathbf{C} as: -->
-
-<!-- \begin{equation} -->
-<!-- \begin{aligned} -->
-<!-- RANKS  -->
-<!-- \end{aligned}(\#eq:rankcor) -->
-<!-- \end{equation} -->
-
+## Ranked correlations
+Ranked correlations are calculated as:
+
+\begin{equation}
+\begin{aligned}
+Rank_{\tilde{J}} = \frac{\sum_{b=1}^{B}rank_{b\tilde{j}}}{B}  
+\end{aligned}(\#eq:rankcor)
+\end{equation}
+
+Where *B* is the number of baits and $rank_{b\tilde{j}}$ is the ranked Pearson correlation coefficient between gene j and bait b in the context of $\tilde{J}$ non-bait genes.
+
+## Fisher transformed Pearson correlation coefficient
 To calculate Fisher transformed Pearson correlations (z) with Pearson's r:
@@
 
-<!--   r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}} -->
+<!-- $r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}}$ -->
 

### 2024-01-09 R/Paper_markdown.Rmd
24_01_09

@@
 output: 
-  # powerpoint_presentation
-  # ## keep_tex: true
-  # bookdown::pdf_document2:
     bookdown::word_document2:
@@
         twocolumn
- # # html_document:
-#  #   toc: true
-#  #   theme: united
-#  #   toc_depth: 3
-#  #   number_sections: true
 # bibliography: MASCARA_refs.json
 bibliography: "`r rbbt::bbt_write_bib('MASCARA_refs.json', overwrite = TRUE)`"
-# csl: oxford-scimed.csl
-#https://raw.githubusercontent.com/citation-style-language/styles/master/apa-6th-edition.csl
-# csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa-6th-edition.csl
 csl: vancouver.csl
@@
 library(knitr)
-# library(here)
 knitr::opts_chunk$set(echo = FALSE, warning = F, message = F, fig.pos = "H") #, root.dir=here()
-#pdf_document
-#powerpoint_presentation
-
-# author: Fred T.G. White^[1,2], Anna Heintz-Buschart^[1], Lemeng Dong^[2], Harro J. Bouwmeester^[2], Johan A. Westerhuis^[1], Age K. Smilde^[1]
-
-# author: "Fred T.G. White, Anna Heintz-Buschart, Lemeng Dong, Harro J. Bouwmeester, Johan A. Westerhuis, Age K. Smilde"
+
+if(!"renv" %in% installed.packages()[,1]){
+  install.packages("renv")
+}
+
+renv::restore()
 
@@
 Harsh experimental conditions can cause widespread transcriptional changes between groups.
-Discovering coexpression within a POI (here strigolactones) in this context is hampered by the dominant variance induced by the design.
+Discovering coexpression within a POI (here the strigolactone pathway) in this context is hampered by the dominant variance induced by the design.
 Minor changes in experimental conditions not controlled for may affect the plants, leading to small coordinated differences in genes within POIs and related pathways (RPs) between replicate plants in the same controlled experimental condition.
@@
 DE is used to determine which genes are expressed differently between two or more experimental conditions, highlighting known or novel pathways affected by the experimental conditions.
-CoE, on the other hand, aims at discovering genes that are part of a partially characterized POI.
+CoE, on the other hand, aims at discovering genes that are part of an already partially characterized POI.
 Here, the expression profiles of known genes of the POI (termed "baits") are used to detect novel pathway members with similar expression patterns (here termed "spikes").
@@
 Panel *A* represents a true hypothetical pathway including a pathway of interest (POI), related pathway (RP) and an unrelated pathway (URP).
-All three types of pathway are assumed to be affected (dis-)proportionally by the experimental design.
-URPs defined to be pathways not directly interacting with the POI.
+All three types of pathway are assumed to be affected by the experimental design.
+URPs are defined to be pathways not directly interacting with the POI.
 When dealing with experimentally designed data, we have to consider two key sources of variance: between-group variance and within-group variance.
@@
 Variance between replicates is caused by non-controlled variations in plant development and environmental conditions.
-One replicate within a group may be slightly bigger than the others and therefore respond slightly different to a treatment.
-This could results in slight differences in bait genes between the replicates, but also the genes that we want to identify with a co-expression analysis.
+One replicate plant within a group may be slightly bigger than the others and therefore respond slightly different to a treatment.
+This could result in slight differences in bait genes between the replicates, but also the genes that we want to identify with a co-expression analysis.
 There may be many other factors that are inducing this within-group variance, but it is quite likely that if a factor induces a change in the treatment response in one replicate, that this response difference occurs throughout the entire process or pathway in that one replicate.
@@
 
-Considering this within-group variance increases the strength of coexpression analysis, for example, in experiments where the intervention causes large between-group differences in many genes.
+Considering the within-group variance increases the strength of coexpression analysis, for example, in experiments where the intervention causes large between-group differences in many genes.
 Not making the distinction between types of variance results in *total CoE* associations driven by the differences between treatment groups as opposed to within group correlations.
@@
 
-<!-- , similar to MutRank (@poretskyMutRankShinyWebapplication2020), -->
-
 ```{r withinbetween, fig.cap= c("Correlation examples, Pearson’s correlation coefficient (r) indicated in facet titles.
@@
 
-For a successful coexpression analysis it is necessary to have a predefined set of baits that are already known to be involved in the mechanism under study.
+For a successful coexpression analysis it is necessary to have a predefined set of baits that are already known to be involved in the pathway under study.
 Methods like ranked correlations and partial least squares regression (PLS; @geladiPartialLeastsquaresRegression1986; @woldCollinearityProblemLinear1984) are able to investigate relationships between a set of baits and the other measured genes.
@@
 The most widely known and applied method that investigates variance sources is the analysis of variance (ANOVA; @searleLinearModels1971).
-ANOVA simultaneous component analysis (ASCA; @smildeANOVAsimultaneousComponentAnalysis2005) and its extension to imbalanced data via the generalised linear model (ASCA+; @thielASCAAPCAExtensions2017) have been developed to combine principal component analysis (PCA) with ANOVA and can therefore account for experimental design in dimensionality reduction.
-ASCA is of interest due to the variance partitioning, this is typically used to estimate the effect matrices that are calculated from the design and particularly the within-group variance structure which is maintained in the residuals of the model.
+ANOVA simultaneous component analysis (ASCA; @smildeANOVAsimultaneousComponentAnalysis2005) and its extension to imbalanced data via the generalised linear model (ASCA+; @thielASCAAPCAExtensions2017) have been developed to combine principal component analysis (PCA) with ANOVA and can therefore account for experimental design in dimension reduction.
+ASCA is of interest due to the variance partitioning.
+This is typically used to estimate the effect matrices that are calculated from the design and particularly the within-group variance structure which is maintained in the residuals of the model.
 Both PLS and ASCA have been used for DE (@augerGeneCoexpressionNetwork2022, @renRNABindingProteins2023, @nuedaDiscoveringGeneExpression2007, @jarmundALASCAPackageLongitudinal2022) and CoE (@guanSRGSSparsePartial2022, @nuedaFunctionalAssessmentTime2009).
@@
 The method capitalises on ASCA to account for an experimental design and the multivariate analysis capability of PLS2 to find uncharacterised POI genes using a set of baits.
-Instead of focusing on the variance induced by the experimental design, as is a typical use of ASCA, this variance is removed, and the remaining variance is further analysed to investigate the multivariate relationships between baits and the putative spikes in terms of within-group CoE.
+Instead of focusing on the variance induced by the experimental design, as is a typical use of ASCA, this variance is removed, and the remaining residual variance is further analysed to investigate the multivariate relationships between baits and the putative spikes in terms of within-group CoE.
 
@@
 At each combination of growth condition and time point, the root transcriptome of three new plants was harvested for RNA sequencing.
-In this dataset, the curated list of SL pathway genes (contained in Supplementary Table 1)  show a specific profile: low variance baseline expression across all time points in the P+ condition but activation and increasing expression over time with P starvation, data were generated and preprocessed to variance stabilised counts as per @haiderTranscriptomeAnalysisPhosphate2023.
+In this dataset, the curated list of SL pathway genes (contained in Supplementary Table 1)  shows a specific profile: low variance baseline expression across all time points in the P+ condition but activation and increasing expression over time with P starvation, data were generated and preprocessed to variance stabilised counts as per @haiderTranscriptomeAnalysisPhosphate2023.
 
@@
 
-The simulations are built upon the results of the ASCA analysis of the combined effect of growth condition and the interaction of growth condition and time (see Figure \@ref(fig:ASCA-1) for overview).
-The simulated data is created by simulating score and loading values (as in Figure \@ref(fig:simOverview)) for the combined effect ($\alpha + \alpha\beta$) ) explaining the difference in time profiles of the P+ and P- condition and the main time effect $\beta$, as well as, structured ambient variance ($S$).
+The simulations are built upon the results of the ASCA analysis of the combined effect of growth condition and the interaction of growth condition (g_c) and time.
+As can be seen in Figure \@ref(fig:ASCA-1) there is clear separation between samples due to the experimental factors growth condition (g_c) and time.
+The variation between the replicates is only small.
+The green loadings of URP genes are strongly related to the experimental variation due to g_c.
+The simulated data is created by simulating experimental variation consisting of the main time effect $beta$ and the combined effect ($\alpha + \alpha\beta$), and the Ambient/Residual variation which exists of the structured ambient variance ($S$) and residual variation ($E$) representing measurement error, modeled with normally distributed random values $e_{ij} = \mathcal{N}(0,\sigma^2)$.
+
+Each of the sources of variation except the residual variation is simulated as low dimensional variation created from 2 scores ($\mathbf{T}$) and loadings ($\mathbf{P}$).
+For the experimental variation, the scores matrices $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{T}_{\beta}$ each consist of two orthogonal score vectors and their relative levels are indicated by the colour.
+The different colours in Figure \@ref(fig:simOverview) represent that the scores are orthogonal to each other.
+Each small block consists of the same score value for each of the replicates within that condition.
+For the Ambient variation, the scores $\mathbf{T}_{S}$ are different for each replicate plant.
+The values in $\mathbf{T}_{S}$ are obtained from a normal distribution with mean 0 and variance 1, i.e.
+$\mathbf{T}_{S_{ik}} \sim \mathcal{N}(0,1)$.
+The $\mathbf{T}$ matrices represent the simulated scores for their respective factors.
+For blues, reds, greys and golds we have 4 different shades representing 4 unique values per factor level (total 8 unique values where reds and golds are sign flipped blues and greys respectively).
+Similarly oranges and pinks represent a growth condition independent time profile with four unique values.
+
+In the loading matrices $\mathbf{P}$, the different coloured blocks indicate different types of genes.
+The green block represents the bait genes of POI.
+The black genes represent “unknown” genes from POI or RP.
+Yellow and purple represent genes from URP that are affected by the experimental variation in different ways.
+The grey blocks represent very small background variation for those genes, while the white blocks represent that those genes are not affected by that source of background variation.
+
+In $\mathbf{P}_{\beta 1}^T$ the baits, spikes and other DE genes (all genes from POI, RP and URP) along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta 2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model.
+The remaining elements of $\mathbf{P}_{\alpha + \alpha \beta}^T$ and $\mathbf{P}_{\beta}^T$ are populated with zeroes.
+$\mathbf{P}_S$ only has values for the baits (green) and spike (black) genes.
+The loading values for the baits and spikes are obtained as $p_{S_jk}^{baits,spikes} \sim \mathcal{N}(0.5,0.2)$, while the loadings for the other genes are 0.
+
+To finalise, the response values all have an intercept $\mathbf{1}\mu_{j}^T$.
+Similarly to the real data, the gene expression values for each experimental condition are simulated with three independent replicates per time-condition-combination.
+Each simulated data Y set thus consists of 24 samples and 2000 genes.
+
+```{r simOverview, echo=FALSE, fig.cap="Coexpression simulation, simplified overview.
+Colours correspond to the same genes in the different loading vectors (*P*).
+Black; baits (known POI genes), green; spikes (unknown POI genes), yellow; DE URP genes, purple DE URP genes with a different profile.
+In the score vectors (T); Blues, reds, greys and golds represent experimental conditions, oranges and pinks; time.
+Each of the textures in the ambient/residual part represent a distinct normally distributed set of random values.", out.width = '100%'}
+knitr::include_graphics("Sim_SL_diagram_simplified.png")
+```
+
+<!-- The simulated data is created by simulating score and loading values (as in Figure \@ref(fig:simOverview)) for the combined effect ($\alpha + \alpha\beta$) ) explaining the difference in time profiles of the P+ and P- condition and the main time effect $\beta$, as well as, structured ambient variance ($S$).
 This is to mimic variance due to gene expression as a function of growth condition and time as well as variance due to coexpression of genes in the same  pathway on top of this experimentally induced DE variance.
+-->
 
@@
 
-The first simulation aims to determine how well the methods perform to find coexpressed genes in the presence of a dominant between-group variance structure and increasing numbers of differential URP genes (genes with high between-group correlation but no within-group correlation).
-The expression levels of differential transcripts is modeled with the scores and loadings of the growth condition and its interaction with time; termed here the combined effect ( $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{P}_{\alpha + \alpha\beta}^T$ ).
+The first simulation aims to determine how well the methods perform to find coexpressed genes in the presence of a dominant between-group variance structure and increasing numbers of differential URP genes (represented by the yellow block in Figure \@ref(fig:simOverview)) that only have high between-group variance but no within-group variance.
+
+In this simulation we have a set of 4 baits, 12 spike genes and a variable number URP genes.
+The issue to address is the presence of many differential URPs; those that have strong between group correlations but no within group correlation.
+We incrementally increase the number of genes in the yellow block and decrease number of white genes with 0 loadings.
+As well as the number of DE URPs we also independently control the size of the experimental variance through the parameter *d* in Equation in \@ref(eq:nCER).
+This approach controls the size of the experimental variance as well as the number of genes responding while keeping a fixed amount of ambient or within group variation.
+
+\begin{equation}
+\begin{aligned}
+\bm{Y} &= \bm{1\mu}^T + d(\bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T) + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\
+\end{aligned}(\#eq:nCER)
+\end{equation}
+
+<!-- \begin{equation} -->
+<!-- \begin{aligned} -->
+<!-- \bm{Y} &= \bm{1\mu}^T + \bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\ -->
+<!-- \text{where for each element}\\ -->
+<!-- &\bm{T}_{S_{ik}} \sim \mathcal{N}(0,1), \quad \bm{P}_{S_{jk}}^{\text{baits,spikes}} \sim \mathcal{N}(0.5,0.2) \\ -->
+<!-- \text{and:} &\\ -->
+<!-- \bm{P}_{S}^{\text{rest}} &= 0 -->
+<!-- \end{aligned}(\#eq:Sim2-bp) -->
+<!-- \end{equation} -->
+
+<!-- The expression levels of differential transcripts is modeled with the scores and loadings of the growth condition and its interaction with time; termed here the combined effect ( $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{P}_{\alpha + \alpha\beta}^T$ ).
 The time effect is modeled with scores and loadings ( $\mathbf{T}_{\beta}$  and $\mathbf{P}_{\beta}^T$ ).
 Also included is some random measurement noise, this is modeled as a matrix $\mathbf{E}$ with normally distributed random values $e_{ij} = \mathcal{N}(0,\sigma^2)$.
-
-On top of the between-group effect there is an additional correlation structure applied to the baits and spikes to create a model of coexpression that is independent to the variance induced by the experimental factors ($\mathbf{X}_S = \mathbf{T}_S\mathbf{P}_S^T$).
+-->
+
+<!-- On top of the between-group effect there is an additional correlation structure applied to the baits and spikes to create a model of coexpression that is independent to the variance induced by the experimental factors ($\mathbf{X}_S = \mathbf{T}_S\mathbf{P}_S^T$).
 This $\mathbf{X}_S$ matrix contains two component sources of variation which links the baits and spikes.
@@
 Each simulated data $\mathbf{Y}$ set thus consists of 24 samples and 2000 genes.
+-->
 
@@
 
-The scores matrices $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{T}_{\beta}$ each consist of two orthogonal score vectors and their relative levels are indicated by the colour.
+<!-- The scores matrices $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{T}_{\beta}$ each consist of two orthogonal score vectors and their relative levels are indicated by the colour.
 The different colours in Figure \@ref(fig:simOverview) represent that the scores are orthogonal to each other.
@@
 As displayed in Figure \@ref(fig:simOverview), the green bar represents 4 baits (or known pathway genes), the black bar represents the 12 spikes (“unknown” genes from POI or RP), yellow represents differential URP genes, and purple; a set of differential URPs with an alternative treatment response.
-
-In $\mathbf{P}_{\beta1}^T$ the baits, spikes and other DE genes (all genes from POI, RP and URP) along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model.
+-->
+
+<!-- In $\mathbf{P}_{\beta1}^T$ the baits, spikes and other DE genes (all genes from POI, RP and URP) along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model.
 The remaining elements of $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$ are populated with zeroes.
@@
 $\mathbf{E}$ represents random unstructured technical noise and the additional $\bm{1\mu}^T$ term in \@ref(eq:Sim2-bp) represents a mean value for each gene.
-
-```{r simOverview, echo=FALSE, fig.cap="Coexpression simulation, simplified overview.
-Colours correspond to the same genes in the different loading vectors (*P*).
-Black; baits (known POI genes), green; spikes (unknown POI genes), yellow; DE URP genes, purple DE URP genes with a different profile.
-In the score vectors (T); Blues, reds, greys and golds represent experimental conditions, oranges and pinks; time.
-Each of the textures in the ambient/residual part represent a distinct normally distributed set of random values.", out.width = '100%'}
-knitr::include_graphics("Sim_SL_diagram_simplified.png")
-```
-
-\begin{equation}
-\begin{aligned}
-\bm{Y} &= \bm{1\mu}^T + \bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\
-\text{where for each element}\\
-&\bm{T}_{S_{ik}} \sim \mathcal{N}(0,1), \quad \bm{P}_{S_{jk}}^{\text{baits,spikes}} \sim \mathcal{N}(0.5,0.2) \\
-\text{and:} &\\
-\bm{P}_{S}^{\text{rest}} &= 0
-\end{aligned}(\#eq:Sim2-bp)
-\end{equation}
-
-The issue to address is the presence of many differential URPs; those that have strong between group correlations but no within group correlation.
+-->
+
+<!-- The issue to address is the presence of many differential URPs; those that have strong between group correlations but no within group correlation.
 Here these differential genes relate  to the yellow bar in Figure \@ref(fig:simOverview).
-We incrementally increase the number of non-zero values and decrease the number of 0 values, which in figure \@ref(fig:simOverview) relates to increasing the size of the yellow bar, as well as varying the size (as quantified by the Frobenius norm) of the combined effect matrix through the scalar parameter *d* below;
-
-\begin{equation}
-\begin{aligned}
-\bm{Y} &= \bm{1\mu}^T + d(\bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T) + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\
-\end{aligned}(\#eq:nCER)
-\end{equation}
+We incrementally increase the number of non-zero values and decrease the number of 0 values, which in figure \@ref(fig:simOverview) relates to increasing the size of the yellow bar, as well as varying the size (as quantified by the Frobenius norm) of the combined effect matrix through the scalar parameter *d* below; -->
 
@@
 
-In order to evaluate the methods and indicate the levels of variance required for optimal performance we conduct a larger scale assay.
-Here we fix the number of DE URP genes to 30 which corresponds to the yellow bar size and the size of the within correlation effect matrix (\(\|\bm{T}_S\bm{P}_S^T\|_F = 0.025\)).
-We investigate a range of values for sizes of between variance (*d*) and random noise (*l*) across random structures (which correspond to repeated simulations of randomised values for all $\mathbf{T}$ matrices).
-Parameters were adjusted according to \@ref(eq:Test-2) on 20 randomised $\mathbf{T}$ matrices.
+The second simulation focuses on the effect of a different number of replicates.
+We investigate a range of values for sizes of between experimental variance (*d*) and random noise variation (*l*) across random structures (which correspond to 20 repeated simulations of randomised values for all $\mathbf{T}$ matrices).
 
@@
 
+With this setup we test the feasibility range of MASCARA with respect to the ratio of 3 key variance parameters namely:
+1. Within group correlation structure
+2. Between group correlation structure
+3. Random noise
+
+Here we set a fixed size of within group correlation structure (\(\|\bm{T}_S\bm{P}_{S}^T\|_F = 0.025\)) as well as number of non-pathway DE genes to 30 (yellow bar size Figure \@ref(fig:simOverview)) across all iterations of this test.
+We vary the sizes of the other two variance parameters (*d* and *l*).
+Furthermore we also show the effect of replicate number and demonstrate that MASCARA (along with the other methods) can be applied to data with generic profiles, not limited to the specific structures that were used to demonstrate the differences in coexpression analysis with simulation 1.
+Here we create 20 datasets with random combined effect and within group correlation structures.
+
+<!-- In order to evaluate the methods and indicate the levels of variance required for optimal performance we conduct a larger scale assay.
+Here we fix the number of DE URP genes to 30 which corresponds to the yellow bar size and the size of the within correlation effect matrix (\(\|\bm{T}_S\bm{P}_S^T\|_F = 0.025\)).
+We investigate a range of values for sizes of between variance (*d*) and random noise (*l*) across random structures (which correspond to repeated simulations of randomised values for all $\mathbf{T}$ matrices).
+Parameters were adjusted according to \@ref(eq:Test-2) on 20 randomised $\mathbf{T}$ matrices.
+-->
+
 ```{r, echo = FALSE}
@@
 
-Pearson's correlation coefficient *r* is calculated between all columns of \(\bm{Y}\).
-Ranks are assigned to all non-bait transcripts for each bait based on *r*, with the lowest rank for the highest correlation.
-These ranks are averaged across all baits (see supplementary equation \@ref(eq:rankcor)).
+Pearson's correlation coefficient $r_{b\tilde{j}}$ is calculated between all bait genes (b = 1..B) and all other genes ($\tilde{j} = 1...\tilde{J}$).
+Ranks $\tilde{r}_{b\tilde{j}}$ are assigned for each $r_{b\tilde{j}}$ such that the highest correlation for each bait gene b receives the lowest rank.
+These ranks are averaged across all baits B (see supplementary equation \@ref(eq:rankcor)).
 
@@
 
-where \(\mathbf{Y}\) is the response matrix of dimension \(I \times J\), \(\mathbf{X}\) is the model matrix of dimension \(I \times p\), containing the coded indicator levels for each of the levels for each factor and their interactions,
+where \(\mathbf{Y}\) is the response matrix of dimension \(I \times J\), \(\mathbf{X}\) is the model matrix of dimension \(I \times p\), containing the sumcoded indicator levels for each of the levels for each factor and their interactions,
  \(\mathbf{\Theta}\) is the parameter matrix of dimension \(p \times J\) and
@@
 \begin{aligned}
-\mathbf{M}_f=\left[0|\mathbf{X}_f |0\right]\mathbf{\hat{\Theta}} = \mathbf{X}_f \mathbf{\hat{\Theta}}_f
+\mathbf{M}_f=\left[0|\mathbf{X}_f |0\right]\left[0|\mathbf{\hat{\Theta}}_f |0\right]^T = \mathbf{X}_f \mathbf{\hat{\Theta}}_f
 \end{aligned}(\#eq:scim)
@@
 \mathbf{\hat{Y}} =& \mathbf{M}_{\alpha} + \mathbf{M}_{\alpha\beta}
-                =& \mathbf{T}_{\alpha}\mathbf{P}_{\alpha}^T + \mathbf{T}_{\alpha\beta}\mathbf{P}_{\alpha\beta}^T
                  =& \mathbf{T}_{\alpha + \alpha\beta}\mathbf{P}_{\alpha + \alpha\beta}^T
@@
 
-for $J_{spike}$ spikes. A lower $\gamma$ indicates better performance.
+for $J_{spike}$ spikes. A lower $\bar{\gamma}$ indicates better performance.
 
@@
 
-For this simulation we have some experimental variance for the combined effect and time ($\bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T$ and $\bm{T}_{\beta}\bm{P}_{\beta}^T$ respectively) a fixed amount of random noise \(\bm{E}\) and a fixed amount of structured residual variance $\bm{T}_S\bm{P}_{S}^T$, within which our 4 baits and 12 spikes share a correlation structure.
+For this simulation we have some experimental variance for the combined effect and time ($\bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T$ and $\bm{T}_{\beta}\bm{P}_{\beta}^T$ respectively) a fixed amount of random noise \(\bm{E}\) and a fixed amount of structured ambient/residual variance $\bm{T}_S\bm{P}_{S}^T$, within which our 4 baits and 12 spikes share a correlation structure.
 This reflects the situation expected from the real data example - drastic effect caused by nutrient deficiency, which also activates the SL pathway.
@@
 Correlations, ASCA and PLS, however, all showed the expected breakdown in performance due to the injection of more DE URP genes.
-This was the expected result as the between group variance is removed with MASCARA.
+This was the expected result as the experimental between group variance is removed with MASCARA.
 
@@
 
-The results from above were obtained from datasets with varying noise, with core aspects such as the underlying structure, the number of replicates, the strength of correlation between variables, and the size of the noise remained fixed.
-With the following, we test the feasibility range of MASCARA with respect to the ratio of 3 key variance parameters namely:
-
-1. Within group correlation structure
-2. Between group correlation structure
-3. Random noise
-
-Here we set a fixed size of within group correlation structure (\(\|\bm{T}_S\bm{P}_{S}^T\|_F = 0.025\)) as well as number of non-pathway DE genes to 30 (yellow bar size \@ref(fig:simOverview)) across all iterations of this test.
+<!-- The results from above were obtained from datasets with varying noise, with core aspects such as the underlying structure, the number of replicates, the strength of correlation between variables, and the size of the noise remained fixed.
+With the following, we test the feasibility range of MASCARA with respect to the ratio of 3 key variance parameters namely: -->
+
+<!-- 1. Within group correlation structure -->
+<!-- 2. Between group correlation structure -->
+<!-- 3. Random noise -->
+
+<!-- Here we set a fixed size of within group correlation structure (\(\|\bm{T}_S\bm{P}_{S}^T\|_F = 0.025\)) as well as number of non-pathway DE genes to 30 (yellow bar size \@ref(fig:simOverview)) across all iterations of this test.
 We vary the sizes of the other two variance parameters (*d* and *l*).
@@
 Here we create 20 datasets with random combined effect and within group correlation structures.
+-->
 
@@
 
-  One_Balance[[i]] <- ggplot(mRES[which(mRES$nCERs == orders[i]),], aes(x = Replicate_Number, y = log2(F1_Score), colour = Method, shape = Method)) +
+  One_Balance[[i]] <- ggplot(mRES[which(mRES$nCERs == orders[i]),], aes(x = Replicate_Number, y = log2(F1_Score), colour = Method, shape = Method, fill = Method)) +
   scale_color_viridis(discrete = T)+
+  # scale_fill_viridis(discrete = T)+
+
+        # scale_shape_manual(values = c(0:3)) +
     stat_summary() +
@@
   labs(subtitle = paste0("CE = ", main_lev[l], " E = ", nse[l]))+
-  theme_bw() +
+  theme_dark() +
       theme(axis.title = element_blank())
@@
 
-```{r repres, echo = F, fig.cap= "Simulation 2 results.
+```{r repres, echo = F, fig.cap= c("Simulation 2 results.
 Replications and variance type feasibility ranges.
@@
 Dots are median performance across 20 random structures.
-ASCA model calculated per equation \\@ref(eq:ascamodel), PLS model per equation \\@ref(eq:plsmodel)."}
+ASCA model calculated per equation \\@ref(eq:ascamodel), PLS model per equation \\@ref(eq:plsmodel).")}
 
@@
 
-ps1 <- grid.arrange(ps, left = "log2(GMR)", bottom = "Replicate Number                     ")
+# ps1 <- grid.arrange(ps, left = "log2(GMR)", bottom = "Replicate Number                     ")
 
@@
 It was expected that all geometric mean ranks converge towards a minimum with increasing numbers of replicates, see \@ref(fig:repres).
-Perfect detection of 12 spikes in top 12 candidates is achieved with a log2 geometric mean rank of 2.4, MASCARA was able to achieve this however the other methods were perturbed by the presence of the 30 differential URP genes.
+Perfect detection of 12 spikes in top 12 candidates is achieved with a log2 geometric mean rank of 2.4, MASCARA was able to achieve this, however, the other methods were perturbed by the presence of the 30 differential URP genes.
 With more replicates MASCARA always outperforms competitors.
-With higher combined effect size (CE) MASCARA outperforms the other methods but its performance is affected by random noise (E) size.
+With higher values of *d* (larger combined effect size) (CE) MASCARA outperforms the other methods but its performance is affected by amount of random noise (E) size indicated by *l*.
 
@@
 In the real data example more than 10% of of the total number of genes are differentially expressed.
-As shown in simulation 1 (Figure \@ref(fig:tc)) higher numbers of DE URP genes results in higher geometric mean rank of the spikes for correlations, ASCA and PLS.
+As shown in simulation 1 (Figure \@ref(fig:tc)) higher numbers of DE URP genes results in higher log2 geometric mean rank (log2(GMR)) of the spikes for correlations, ASCA and PLS.
 MASCARA is designed specifically to mitigate the effect of large numbers of DE URP genes.
@@
 
-It is apparent that MASCARA has a particular use case for coexpression analysis in data where there is a lot of experimentally induced between-group variance across many genes as well as some within-group correlation structure shared between pathway genes that is caused by independent non-controlled ambient factors.
+It is apparent that MASCARA has a particular use case for coexpression analysis in data where there is a lot of experimentally induced between-group variance across many genes as well as some within-group correlation structure shared between POI genes that is caused by independent non-controlled ambient factors.
 
 Here we illustrate an application of MASCARA to the dataset from @haiderTranscriptomeAnalysisPhosphate2023.
-For this we take 4 core strigolactone genes (Os11t0587000, Os04t0550600, Os01t0746400 and Os01t0700900 of the 9 used for illustrations in figures \@ref(fig:Dataset1-Overview) and \@ref(fig:ASCA-1)) as baits and investigate their relationship with top candidate genes.
+For this we take 4 core strigolactone genes (Os11t0587000, Os04t0550600, Os01t0746400 and Os01t0700900 of the 9 used for illustrations in Figures \@ref(fig:Dataset1-Overview) and \@ref(fig:ASCA-1)) as baits and investigate their relationship with top candidate genes.
 
@@
 In the real data from @haiderTranscriptomeAnalysisPhosphate2023 some plants were subjected to massive stress with the phosphate limitation.
-This is therefore reflected by large differences in the transcriptome compared to those with a normal supply of P.
+This is, therefore, reflected by large differences in the transcriptome compared to those with a normal supply of P.
 This limitation affected many more genes than those directly involved in SL biosynthesis.
-Furthermore, there appears to be a stronger experiment-independent relationship amongst the SL genes compared to their correlations with other DE genes, as displayed in Figure \@ref(fig:residualdists) where the blue distribution indicates residual correlations between the 9 POI transcripts from figures \@ref(fig:Dataset1-Overview) and \@ref(fig:ASCA-1) and the red indicates residual correlations between these 9 and top DE URP genes.
+Furthermore, there appears to be a stronger relationship amongst the SL genes in the ambient variation compared to their correlations with other DE URP genes, as displayed in Figure \@ref(fig:residualdists) where the blue distribution indicates correlations between residuals of the 9 POI transcripts from Figures \@ref(fig:Dataset1-Overview) and \@ref(fig:ASCA-1) and the red indicates correlations between residuals of these 9 and top DE URP genes.
 
@@
 
-Some example top candidate genes are shown in Figure \@ref(fig:realcandidates).
+Some example top coexpression candidate genes are shown in Figure \@ref(fig:realcandidates).
 The problem of dominant between group variation is again exemplified here.
@@
 MASCARA estimates effects through the GLM with a design matrix to remove the variance between conditions, then the residual matrix can be decomposed through PLS2, with feature ranking through target projections.
-This framework enables the investigation of multivariate relationships between a set of baits and the rest of the data.
+This framework enables the investigation of multivariate relationships between a set of bait genes and all other genes in the dataset.
 MASCARA thrives on the fact that within the typical experimental setup small systematic variance exists between replicates due to variations in ambient factors that affect the plant, and therefore the variance structures in the resulting data.
@@
 In a coexpression context, within certain assumed parameters of variance structure (dominant between-group, existing within-group), MASCARA outperformed competitors.
-Ranked correlations, ASCA (@smildeANOVAsimultaneousComponentAnalysis2005, @thielASCAAPCAExtensions2017) and PLS (@woldCollinearityProblemLinear1984) are all affected by dominant between-group variance, which is a common occurrence in experimental data, exemplified here by the dataset from @haiderTranscriptomeAnalysisPhosphate2023 but discussed elsewhere (@farahbodUntanglingEffectsCellular2020, @zhaoWeightedGeneCoexpression2010, @chowdhuryDifferentialCoExpressionAnalysis2020).
-
-The feature selection stage of MASCARA uses an average vector constructed from the loadings of the baits onto which to project the loadings of other genes, which is similar to target projection in @kvalheimInterpretationPartialLeast2010 although we construct the target vector from averaging the loadings in a PLS2 model as opposed to already having it defined with the univariate response in PLS1.
+Ranked correlations, ASCA (@smildeANOVAsimultaneousComponentAnalysis2005, @thielASCAAPCAExtensions2017) and PLS (@woldCollinearityProblemLinear1984) are all affected by dominant experimental between-group variance, which is a common occurrence in experimental data, exemplified here by the dataset from @haiderTranscriptomeAnalysisPhosphate2023 but discussed elsewhere (@farahbodUntanglingEffectsCellular2020, @zhaoWeightedGeneCoexpression2010, @chowdhuryDifferentialCoExpressionAnalysis2020).
+
+The feature selection stage of MASCARA uses an average vector constructed from the loadings of the baits onto which the loadings of other genes are projected.
+This approach is similar to target projection in @kvalheimInterpretationPartialLeast2010 although we construct the target vector from averaging the loadings in a PLS2 model as opposed to already having it defined with the univariate response in PLS1.
 There are many established methods for feature selection in PLS which include Variable Importance in Projection (VIP) and Selectivity Ratio (@rajalahtiBiomarkerDiscoveryMass2009), however these methods focus on the importance of variables in the construction of the model rather than a directional association with the response block as with the $\mathbf{\bar{q}}$ target projection stage.
-With this averaged target projection approach, as an extension of Kvalheim's target projection in PLS1 models, we can use the information in a set of baits to define a direction in the latent space which can then be used as an axis to determine the relatedness of other genes.
+With this averaged target projection approach, as an extension of Kvalheim's target projection in PLS1 models @kvalheimInterpretationPartialLeast2010, we can use the information in a set of baits to define a direction in the latent space which can then be used as an axis to determine the relatedness of other genes.
 This assumes moderate to high correlation in the residuals of the baits, which is a realistic assumption in coexpression analysis although should be verified in the selection of baits.
 
-Our analysis framework does not consider different correlation structures within the residual variance, this could be an issue especially in the cases similar to @caldanaHighdensityKineticAnalysis2011 where different combinations of experimental factors show differential within-group correlation structure.
-With MASCARA these will not be preferentially detected, as the focus is on detecting postively associating features.
+Our analysis framework does not consider different correlation structures within the residual variance, this could be an issue especially in the cases similar to @caldanaHighdensityKineticAnalysis2011 where different combinations of experimental factors show different within-group correlation structures.
+This type of relationship can be seen in Figure \@ref(fig:realcandidates) where the P- experiments show high positive correlation between the genes of the same pathway while the P+ experiments show a negative correlation.
+With MASCARA these will not be preferentially detected, as the focus is on detecting positively associating features.
 Ideally there would be enough samples to calculate robust correlations per group, to account for possible rewiring under different conditions, but in our case we focus on experiments where there are generally not enough replicates to do this.
@@
 In coexpression analysis, there is no need for class balance, it is in fact better to have more data points in the class where the pathway is induced (P- in the example dataset).
-When the mechanism of interest is only active in the case condition then the control samples are mostly redundant.
+When the POI is only active in the case condition then the control samples are mostly redundant.
 With more samples in the condition of interest the reliability of the coexpression analysis will improve.
@@
 The method is not limited to the calculation of positive associations, however performance in detecting negative correlations (indicative of inhibitory or negative feedback processes) or more complex relationships has not been tested within the current simulations.
-Our approach was based off the structure of the Haider RNAseq dataset, however the method is applicable across other organisms and omics types.
+Our approach was based on the structure of the study by Haider with RNAseq measurements of rice plants, however the method is applicable across other organisms and omics types.
 This holds under the assumption that the set of baits have a multivariate relationship between themselves and multiple other (unknown) genes.
@@
 \begin{aligned}
-Rank_{\tilde{J}} = \frac{\sum_{b=1}^{B}rank_{b\tilde{j}}}{B}  
+Rank_{\tilde{J}} = \frac{\sum_{b=1}^{B}\tilde{r}_{b\tilde{j}}}{B}  
 \end{aligned}(\#eq:rankcor)
@@
 <!-- MASCARA example run on $params: -->
-
-```{r}
-
-```
 

### 2024-01-09 R/Paper_markdown.Rmd
Update Paper_markdown.Rmd
23_01_09_02

@@
         twocolumn
-# bibliography: MASCARA_refs.json
-bibliography: "`r rbbt::bbt_write_bib('MASCARA_refs.json', overwrite = TRUE)`"
+bibliography: MASCARA_refs.json
 csl: vancouver.csl
@@
 
-renv::restore()
+renv::restore(prompt = FALSE)
+
+#bibliography: "`r rbbt::bbt_write_bib('MASCARA_refs.json', overwrite = TRUE, filter = ("accessed"))`"
 
@@
 
-Experiments in plant transcriptomics are usually designed to induce variation in a pathway of interest (POI).
+Experiments in plant transcriptomics are usually designed to induce variation in a pathway of interest.
 Harsh experimental conditions can cause widespread transcriptional changes between groups.
-Discovering coexpression within a POI (here the strigolactone pathway) in this context is hampered by the dominant variance induced by the design.
-Minor changes in experimental conditions not controlled for may affect the plants, leading to small coordinated differences in genes within POIs and related pathways (RPs) between replicate plants in the same controlled experimental condition.
-These systematic differences between replicates within the same experimental conditions can be used to detect genes from the same POI or RPs.
+Discovering coexpression within a pathway of interest (here the strigolactone pathway) in this context is hampered by the dominant variance induced by the design.
+Minor changes in experimental conditions not controlled for may affect the plants, leading to small coordinated differences in genes within pathways of interest and related pathways between replicate plants in the same controlled experimental condition.
+These systematic differences between replicates within the same experimental conditions can be used to detect genes from the same or related pathway.
 We introduce a novel framework “MASCARA” which combines ANOVA simultaneous component analysis and partial least squares to remove the experimentally induced variance and investigate multivariate relationships in the non-designed variance.
@@
 Figure \@ref(fig:nettype) outlines pathway nomenclature that will be used throughout this text.
-Panel *A* represents a true hypothetical pathway including a pathway of interest (POI), related pathway (RP) and an unrelated pathway (URP).
+Panel *A* represents hypothetical pathways including a pathway of interest (POI), related pathway (RP) and an unrelated pathway (URP).
 All three types of pathway are assumed to be affected by the experimental design.

### 2024-01-10 R/Paper_markdown.Rmd
renv_update

@@
 knitr::opts_chunk$set(echo = FALSE, warning = F, message = F, fig.pos = "H") #, root.dir=here()
-
-if(!"renv" %in% installed.packages()[,1]){
-  install.packages("renv")
-}
-
-renv::restore(prompt = FALSE)
+# install.pacakges()
+
+# if(!"renv" %in% installed.packages()[,1]){
+#   install.packages("renv")
+# }
+# 
+# renv::restore(prompt = FALSE)
+# renv::snapshot()
+
+# renv::dependencies()
+# renv::status()
 
@@
 library(biomformat)
-library(MetStaT)
 library(DiscriMiner)
@@
 library(ggbreak)
+library(rhmmer)
+```
+
+```{r}
+# renv::snapshot()  #change this
 
@@
 
-library(rhmmer)
 REF2 <- as.data.frame(read_tblout("Rice_Annotation/HMM_ITAG_1.txt"))

### 2024-01-11 R/23_02_02_real_data_rice.Rmd
24_11_01_1

@@
 knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, echo = F)
-#pdf_document
-#powerpoint_presentation
 ```
@@
 ```{r , echo=FALSE, include=FALSE}
-#SOURCE FUNCTIONS FROM EXTERNAL FILE
 source("DATA_SIM_FUNCS.R")
@@
 
-# png("MASCARA_heatmap.png")
-
 heat <- pheatmap::pheatmap(t(scale(t(h))),  #
@@
                            fontsize = 8,
-                             # annotation_row = anno,
                            main = "Scaled Gene Expression",silent = T)
 
-  # # heat
-  # library(ggplotify)
-  # heat2 <- as.ggplot(heat) + theme(plot.title = element_text(face = "plain"))
+png("MASCARA_heatmap.png", res = 150, height = 960, width = 960)
+grid::grid.newpage()
+grid::grid.draw(heat$gtable)
+dev.off()
 
-  # saveRDS(heat2, "heatmap_MASCARA_Fig_2.RDS")
-
-# png("MASCARA_heatmap.png")
-# heat
-# dev.off()
+grid::grid.newpage()
+grid::grid.draw(heat$gtable)
 
@@
 
-# Background
-
-The biological aim of the dataset is to find novel strigolactone (SL) genes.
-SLs are induced by phosphate starvation, but there are also a lot of more general phosphate deficiency related genes (PSIs) that are strongly affected by the experimental design.
-
-In this case where SLs are the pathway of interest that can make it difficult to detect novel pathway genes, since the experimental design is affecting the pathway of interest along with many other genes.
-
-Some SL genes are already characterised, we should use these to find novel pathway genes.
-
-# Methodological aim
-
-Find method that can preferentially detect members of a sub-network rather than more general experiment responders.
-
-The genes should respond to the experiment but also show some experiment independent association.
-
-# Data overview
-
 ```{r, echo = F, fig.cap="Top: distribution and mean sd relationship. Bottom: scores and loadings of the main effect; growth condition + interaction"}
@@
     ylab(paste0("PC2 ",ve[2])) +
-    # geom_point(data = df[which(df$Baits == "Other"),], alpha = 0.5) +
-    # geom_point(data = df[which(df$Baits == "POI"),], alpha = 1) +
-    # geom_point(data = df[which(df$Baits == "URP"),], alpha = 1) +
-geom_point(data = df) +
+
+geom_point(data = df, size = 5) +
     scale_x_continuous(guide = guide_axis(check.overlap = TRUE)) +
@@
 
-l <- l + scale_color_manual(values = c("Other" = "black", "POI" = "purple","URP" = "green"))# + scale_alpha(c(0.1,1,1))
+l <- l + scale_color_manual(values = c("Other" = "black", "POI" = "purple","URP" = "green")) + coord_fixed() + theme(aspect.ratio = 1,text=element_text(size=18))
 
@@
   ggplot(df, ...)+
-    geom_point() +
+    geom_point(size = 5) +
     xlab(paste0("PC1 ",ve[1])) +
@@
 s <- scoreplot(PCDE, meta2, ve1, aes(x= PC1, y = PC2, colour = g_c, shape = time)) + ggtitle("Scores")
+s <- s + coord_fixed() + theme(aspect.ratio = 1, text=element_text(size=18))
 
@@
 
-(dist_norm + plot_normal)/(s + l)
-
-```
-
-<!-- ```{r} -->
-<!-- png("rice_s_l.png") -->
-<!-- (s + l) -->
-<!-- dev.off() -->
-<!-- ``` -->
-
-```{r, include = F}
-source("DATA_SIM_FUNCS.R")
-
-pcd <- prcomp(res_ASCAplus_nb$residuals)
-residual_cands <- get_ASCA_cands(pcd)
-
-#HERE
-
-image(cor(res_ASCAplus_nb$residuals[,c(baits,baits2)]))
-```
-
-```{r, inlcude = FALSE}
-
-minT <-  res_ASCAplus_nb$decomposition$time
-tcd <- prcomp(minT)
-
-newdata <- res_ASCAplus_nb$decomposition$time + res_ASCAplus_nb$residuals
-PCDE <- predict(PCD, newdata)
-
-tcd <- prcomp(res_ASCAplus_nb$decomposition$time + res_ASCAplus_nb$residuals)
+# (dist_norm + plot_normal)/(s + l)
 
@@
 ```{r}
-#do permutation tests to see if the correlations in the "non experimentally interesting" data are significant between the strigolactone genes compared to random sets.
+(s + l)
+
+png("rice_s_l.png", width = 960, height = 480)
+(s + l)
+dev.off()
+```
+
+```{r}
+
+get_ASCA_cands2 <- function(PCD, meta, distance_calc= FALSE, baits = NULL, spikes = NULL, ret_candN = nrow(PCD$rotation)){
+  #############
+
+  if(distance_calc== TRUE){
+
+    cands <- ranked_dist(baits,PCD)
+
+  }else{
+    absload <- abs(data.matrix(PCD$rotation[,1:2])) %*% diag(summary(PCD)$importance[2,1:2])
+    combscore <- rowSums(absload[,1:2])
+
+    orderedload <- cbind(combscore, PCD$rotation[,1:2])
+    cands <- as.data.frame(orderedload[order(orderedload[,1], decreasing = T),])
+    colnames(cands) <- c("VIP", "PC1","PC2")
+
+    cands <- round(cands[1:ret_candN,], 4)
+
+  }
+
+  return(cands)
+
+}
 
@@
 
-p_n_b <- pathway[-which(pathway %in% baits)]
+ASCA_cands2 <- get_ASCA_cands2(a_c.PCD)
+ASCA_cands2 <- ASCA_cands2[which(ASCA_cands2[,2] < 0),]
+ASCA_cands2 <- ASCA_cands2[order(ASCA_cands2[,2]),]
+ASCA_cands2 <- ASCA_cands2[-which(rownames(ASCA_cands2) %in% pathway),]
 
-# sum(cor(res_ASCAplus_nb$residuals[,pathway]))
-# 
-# psi_ref <- PSI_ref[which(PSI_ref[,1] %in% colnames(res_ASCAplus_nb$residuals)),]
+ASCA_cands3 <- ASCA_cands2[-which(rownames(ASCA_cands2) %in% c("Os06t0570600-00","Os06t0570566-00", "Os09t0321200-00","Os06t0651900-01", "Os10t0444566-00")),]
 
-cor_val <- function(x,y){
+qqplot <- function(data, meta, ...){
 
-  y <- y[which(y %in% colnames(x))]
-  res <- (sum(cor(x[,y]))- length(y))/2   /length(y)
-  return(res)
+  df <- cbind.data.frame(meta,data)
+
+  df$time <- factor(df$time, levels = unique(meta$time))
+  df$growth_condition <- factor(df$growth_condition)
+
+  ggplot(df, ...) +
+    geom_point() +
+    theme_bw()
 
@@
 
-p_cor <- cor_val(res_ASCAplus_nb$residuals,sls)
-phos_cor <- cor_val(res_ASCAplus_nb$residuals, URPS)
+meta$time <- rep(c(1,3,7,8), each = 3, 2)
 
-phos_p_cor <- cor_val(res_ASCAplus_nb$residuals, c(sls,URPS))
+PSI_ref <-read.table("Rice_Annotation/PSI_genes.txt", sep = "\t", header = T)
+psi_ref <- PSI_ref[which(PSI_ref[,1] %in% colnames(res_ASCAplus_nb$residuals)),]
 
-rand_cor <- c()
-i <- NULL
-for(i in 1:1000){
+PSI_cands <- rownames(ASCA_cands[which(rownames(ASCA_cands) %in% psi_ref[,1]),])
+PSI_cands <- head(rownames(ASCA_cands[-which(rownames(ASCA_cands) %in% pathway),]),20)
 
-  rc <- cor_val(res_ASCAplus_nb$residuals, sample(colnames(res_ASCAplus_nb$residuals),13))
-  rand_cor <- c(rand_cor, rc)
+cand_b <- PSI_cands[10]
+cand_w <- baits[3]
 
-  }
-```
+cor_tw <- round(cor(X[,which(colnames(X) %in% cand_w)], X[,which(colnames(X) %in% baits[4])]), digits = 2)
 
-The pathway genes are responding to the experiment, but they are not the only responders, nor are they the strongest responders.
-Therefore it is interesting to see if there is any significant correlation in the residuals of the ASCA model.
+cor_tb <- round(cor(X[,which(colnames(X) %in% cand_b)], X[,which(colnames(X) %in% baits[4])]), digits = 2)
 
-# Results
+cor_rw <- round(cor(res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% cand_w)], res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% baits[4])]), digits = 2)
 
-## Permutation test: average correlations in residuals
+cor_rb <- round(cor(res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% cand_b)], res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% baits[4])]), digits = 2)
 
-```{r, fig.cap="red: pathway genes, blue: General phosphate genes"}
-hist(rand_cor, xlim = c(-0.5, 2), main = "Residual correlation permutation test", xlab = "Average correlation")
-abline(v = c(p_cor,phos_cor), col = c("red","blue"))
-# abline(v = phos_p_cor)
+within <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% cand_w)],
+                        "Bait" = X[,which(colnames(X) %in% baits[4])]),meta,
+       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + xlab(baits[4]) + ylab(cand_w) + labs(title = paste0("r = ",cor_tw)) + theme(aspect.ratio = 1)
+
+PSI_cands <- rownames(ASCA_cands[which(rownames(ASCA_cands) %in% psi_ref[,1]),])
+PSI_cands <- head(rownames(ASCA_cands[-which(rownames(ASCA_cands) %in% pathway),]),20)
+
+between <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% cand_b)],
+                        "Bait" = X[,which(colnames(X) %in% baits[4])]),meta,
+       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + xlab(baits[4]) + ylab(cand_b) + labs(title = paste0("r = ",cor_tb)) + theme(aspect.ratio = 1)
+
+# within + between + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A')) + plot_layout(heights = unit(c(5), c("cm")), widths = c(5))
+
+within_r <- qqplot(cbind.data.frame("Candidate" = res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% cand_w)],
+                        "Bait" = res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% baits[4])]),meta,
+       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + xlab(baits[4]) + ylab(cand_w)   + labs(title = paste0("r = ",cor_rw)) + theme(aspect.ratio = 1)#, alpha = (16- as.numeric(growth_condition))
+
+between_r <- qqplot(cbind.data.frame("Candidate" = res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% cand_b)],
+                        "Bait" = res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% baits[4])]),meta,
+       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + xlab(baits[4]) + ylab(cand_b)  + labs(title = paste0("r = ",cor_rb)) + theme(aspect.ratio = 1)
+
+(within + between)/(within_r + between_r) + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A','1')) #+ plot_layout(heights = unit(c(5), c("cm")), widths = c(5))
+
+png("within_between_example_2.png", res = 150, height = 960, width = 960)
+(within + between)/(within_r + between_r) + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A','1')) #+ plot_layout(heights = unit(c(5), c("cm")), widths = c(5))
+dev.off()
 
 ```
-
-The SL genes (in red) have a much higher average correlation than a random set in the residuals, however this is also true for the PSIs (in blue).
-
-```{r}
-
-all <- cbind.data.frame(meta[,c("growth_condition","time")], X)
-residuals <- cbind.data.frame(meta[,c("growth_condition","time")], res_ASCAplus_nb$residuals)
-```
-
-```{r}
-#get cands from residuals
-
-rPCD <- prcomp(data.matrix(residuals[,-c(1:2)]))
-
-# residual_cands_all <- ranked_dist(pathway, rPCD)
-# residual_cands_phos <- ranked_dist(PSI_ref[,1], rPCD)
-
-```
-
-```{r}
-#filter residuals for only n genes that show treatment effect i.e. set VIP cutoff value
-#how to do this with density?...
-
-#first quick and dirty attempt with top n candidates  -  then add density
-
-cutoff <- round(quantile(c(1:nrow(ASCA_cands)),0.1))
-res_filt <- residuals[,c(1:2,which(colnames(residuals) %in% rownames(ASCA_cands)[1:cutoff]))]
-
-rpcd <- prcomp(data.matrix(res_filt[,-c(1:2)]))
-
-ref$Baits <- as.character(ref$Baits)
-# ref$Baits[which(ref$Feature %in% PSI_ref[,1])] <- "PSI"
-ref$Baits <- factor(ref$Baits, levels = c("Other","POI","URP"))
-
-Fref <- ref[match(rownames(rpcd$rotation),ref$Feature),]
-
-```
-
-For the following analyses the data is prefiltered using the top 10% VIPs of the main effect model, since we expect the pathway to be induced by the experimental design.
-
-## PCA on ASCA residuals
-
-Since there is clearly an experiment independent relationship between the SL genes, it is interesting to see if there is any obvious structure in the residuals.
-The structure relating to the experimental design, in the samples, is not interesting since we should have removed that in the main effect. 
-Here we would like to see if the SL genes show any indication of clustering, and if that is in some way different to the PSIs.
-
-```{r}
-
-plots <- list()
-plots_p <- list()
-i <- NULL
-
-k <- c(1:16)
-l <- k[c(F,T)]
-k <- k[c(T,F)]
-
-importance <- paste0(round(summary(rpcd)$importance[2,] * 100, digits = 2), "%")
-
-for(i in 1:8){
-
-  plots[[i]] <- loadingplot(rpcd$rotation, meta, baits = Fref$Feature[which(Fref$Baits %in% c("POI","URP"))], c(importance[k[i]], importance[l[i]]), PCs =c(k[i],l[i]),  Fref, aes(x = !!sym(paste0("PC",k[i])),y = !!sym(paste0("PC",l[i])), colour = Baits, label = Description))
-
-}
-```
-
-```{r}
-wrap_plots(plots) + plot_layout(guides = "collect")
-```
-
-There isn't an obvious way to visually distinguish the two sets. But since the residual correlation analysis showed significant results we can investigate other methods on the residual matrix.
-
-## sPLS on ASCA residuals
-
-```{r}
-# resids <- residuals[,-c(1:2)]
-resids <- res_filt[,-c(1:2)]
-spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], resids[,which(colnames(resids) %in% baits)], all.outputs = T)
-
-sPLS_cands <- cbind.data.frame(abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X, spls_res$loadings$X)
-sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = T),]
-
-importance <- paste0(round(spls_res$prop_expl_var$X * 100, digits = 2), "%")
-
-Fref2 <- Fref[-which(Fref$Feature %in% baits),]
-Fref2 <- Fref2[match(rownames(sPLS_cands),Fref2$Feature),]
-
-p <- loadingplot(sPLS_cands, meta, baits = Fref2$Feature[which(Fref2$Baits %in% c("POI","URP"))], 
-            importance, PCs = c(1,2), ref = Fref2, aes(x = comp1, y = comp2, colour = Baits, label = Description))
-
-# ggplotly(p)
-
-```
-
-<!-- ```{r} -->
-<!-- #calculating direction of baits to determine direction of candidate selection spls loading space -->
-
-<!-- # loadingplot(sPLS_cands, meta, baits = Fref2$Feature[which(Fref2$Baits %in% c("Pathway","PSI"))],  -->
-<!-- #             importance, PCs = c(1,2), ref = Fref2, aes(x = comp1, y = comp2, colour = Baits, label = Description)) -->
-
-<!-- angles <- rbind.data.frame(spls_res$loadings$Y, colMeans(spls_res$loadings$Y)) -->
-
-<!-- x <- angles[4,] -->
-<!-- y <- angles[5,] -->
-
-<!-- #adapted from https://stackoverflow.com/questions/1897704/angle-between-two-vectors-in-r -->
-<!-- angle <- function(x,y){ -->
-
-<!--   x <- c(as.matrix(x)) -->
-<!--   y <- c(as.matrix(y)) -->
-
-<!--   dot.prod <- x%*%y  -->
-<!--   norm.x <- norm(x,type="2") -->
-<!--   norm.y <- norm(y,type="2") -->
-<!--   theta <- acos(dot.prod / (norm.x * norm.y)) -->
-<!--   return(as.numeric(theta)) -->
-<!-- } -->
-
-<!-- angles_loadings <- apply(sPLS_cands[,-1],1,angle, y = y) -->
-
-<!-- #define this better -->
-<!-- sPLS_cands <- cbind.data.frame(sPLS_cands, "angle" = angles_loadings, "score" = 3 * sPLS_cands$`abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X` * (1-angles_loadings)) -->
-
-<!-- sPLS_cands <- sPLS_cands[order(sPLS_cands$score, decreasing = T),] -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- avg_point <- angles[5,]/5 -->
-
-<!-- rot.mat <- matrix(c(0,-1,1,0),nrow = 2, byrow = T)  #90 degree clockwise rotation -->
-
-<!-- rotated_point <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat) -->
-
-<!-- ang <- angle(avg_point,c(1,0)) -->
-
-<!-- rot.mat2 <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation -->
-<!-- rotated_point2 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat2) -->
-
-<!-- # rot.mat3 <- matrix(c(cos(ang),sin(ang),-sin(ang),cos(ang)),nrow = 2, byrow = T) #anticlockwise rotation -->
-<!-- # rotated_point3 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat3) -->
-
-<!-- ggplot() +  -->
-<!--   geom_point(data = sPLS_cands, aes(x = comp1, y = comp2), alpha = 0.1) + -->
-<!--   geom_point(data = sPLS_cands[1:100,], aes(x = comp1, y = comp2)) + -->
-<!--   geom_segment(data = avg_point, aes(x = 0, y = 0, xend = comp1, yend = comp2)) + -->
-<!--   # geom_segment(data = rotated_point, aes(x = 0, y = 0, xend = V1, yend = V2)) + -->
-<!--   geom_segment(data = rotated_point2, aes(x = 0, y = 0, xend = V1, yend = V2)) + -->
-<!--   # geom_segment(data = rotated_point3, aes(x = 0, y = 0, xend = V1, yend = V2)) + -->
-
-<!--   # xlim(c(-0.1,0.1))+ -->
-<!--   # ylim(c(-0.1,0.1)) + -->
-
-<!--   theme_bw() -->
-<!--   # geom_smooth(method = "lm") -->
-
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- #rotated loadings -->
-<!-- ang <- angle(avg_point,c(1,0)) -->
-
-<!-- rot.mat2 <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation -->
-
-<!-- rotated_point2 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat2) -->
-
-<!-- rotate <- function(x,y){ -->
-<!--   x <- c(as.matrix(x)) -->
-<!--   x %*% y -->
-<!-- } -->
-
-<!-- loadings_r_baits <- t(apply(sPLS_cands[,c("comp1","comp2")],1,rotate, y = rot.mat2)) -->
-<!-- colnames(loadings_r_baits) <- c("comp1","comp2") -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- ggplot() +  -->
-<!--   geom_point(data = loadings_r_baits, aes(x = comp1, y = comp2), alpha = 0.1) + -->
-<!--   geom_point(data = loadings_r_baits[1:100,], aes(x = comp1, y = comp2)) + -->
-<!--   theme_bw() -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- #density  -->
-<!-- density <- kde2d(loadings_r[,1], loadings_r[,2], n = 200) -->
-<!-- image(as.matrix(density$z), col = hcl.colors(20)) -->
-
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- range.ang <- seq(from = -pi/6, to = pi/6, length.out = 60) -->
-
-<!-- ### -->
-
-<!-- i <- NULL -->
-<!-- L_R <- data.frame() -->
-<!-- LOADINGS_R <- list() -->
-
-<!-- for(i in 1:length(range.ang)){ -->
-
-<!-- ang <- range.ang[i] -->
-
-<!-- rot.mat <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation -->
-<!-- loadings_r <- t(apply(loadings_r_baits[,c("comp1","comp2")],1,rotate, y = rot.mat)) -->
-<!-- colnames(loadings_r) <- c("comp1","comp2") -->
-<!-- L_R <- rbind(L_R,loadings_r) -->
-<!--   LOADINGS_R[[i]] <- loadings_r -->
-
-<!-- } -->
-
-<!-- # colnames(LOADINGS_R) <- c("comp1","comp2") -->
-<!-- L_R$rad <- rep(range.ang, each = nrow(sPLS_cands)) -->
-<!-- L_R$frame <- rep(c(1:length(range.ang)), each = nrow(sPLS_cands)) -->
-
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- #get sd of original comp1 distribution -->
-<!-- sigma <- sd(sPLS_cands$comp1) -->
-
-<!-- big.T.sig <- function(x,y){ -->
-<!--   big <- length(which(x > 2*y)) -->
-<!--   return(big) -->
-<!-- } -->
-
-<!-- i <- NULL -->
-<!-- res <- c() -->
-<!-- for(i in 1:length(LOADINGS_R)){ -->
-
-<!--   r <- big.T.sig(LOADINGS_R[[i]][,1],sigma) -->
-<!--   res <- c(res,r) -->
-<!-- } -->
-
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- RES <- cbind.data.frame("Rad" = range.ang, "Density" = res) -->
-
-<!-- density <- ggplot(RES, aes(x = Rad, y = Density))+ -->
-<!--   geom_smooth() + theme_bw() + ggtitle("Number of points above 2*sigma of original rotation") -->
-
-<!-- density -->
-
-<!-- ``` -->
-
-<!-- ```{r} -->
-
-<!-- #taken from https://stackoverflow.com/questions/6836409/finding-local-maxima-and-minima -->
-<!-- inflect <- function(x, threshold = 1){ -->
-<!--   up   <- sapply(1:threshold, function(n) c(x[-(seq(n))], rep(NA, n))) -->
-<!--   down <-  sapply(-1:-threshold, function(n) c(rep(NA,abs(n)), x[-seq(length(x), length(x) - abs(n) + 1)])) -->
-<!--   a    <- cbind(x,up,down) -->
-<!--   list(minima = which(apply(a, 1, min) == a[,1]), maxima = which(apply(a, 1, max) == a[,1])) -->
-<!-- } -->
-
-<!-- ggres <- ggplot_build(density)$data[[1]] -->
-<!-- infs <- inflect(ggres$y) -->
-<!-- rads <- ggres$x[infs$maxima] -->
-
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- maxima.ang <- rads -->
-
-<!-- ### -->
-
-<!-- i <- NULL -->
-<!-- # L_R <- data.frame() -->
-<!-- sPLS_maxima_cands <- list() -->
-
-<!-- for(i in 1:length(maxima.ang)){ -->
-
-<!-- ang <- maxima.ang[i] -->
-
-<!-- rot.mat <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation -->
-<!-- loadings_r <- as.data.frame(t(apply(sPLS_cands[,c("comp1","comp2")],1,rotate, y = rot.mat))) -->
-<!-- colnames(loadings_r) <- c("comp1","comp2") -->
-<!-- # L_R <- rbind(L_R,loadings_r) -->
-<!--   sPLS_maxima_cands[[i]] <- loadings_r[order(loadings_r$comp1, decreasing = T),] -->
-
-<!-- } -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-
-<!-- ggplotly( -->
-<!--   ggplot() +  -->
-<!--   geom_point(data = sPLS_maxima_cands[[1]], aes(x = comp1, y = comp2), alpha = 0.1) + -->
-<!--   geom_point(data = sPLS_maxima_cands[[1]][1:20,], aes(x = comp1, y = comp2)) + -->
-<!--   theme_bw() -->
-<!-- ) -->
-
-<!-- ``` -->
-
-```{r}
-coexp_cands <- ranked_coexp(baits,X[,which(colnames(X) %in% colnames(resids))])
-
-residual_cands <- ranked_dist(baits[1:3], rpcd)
-
-residual_coexp <- ranked_coexp(baits, resids)
-
-```
-
-```{r}
-Methods <- c("sPLS_ASCA","PCA_DIST_ASCA","Correlations_ASCA", "Correlations")
-
-getrank <- function(RES,geneset,NOINFO = 26402){
-  # geneset <- 
-  # NOINFO = length(geneset) - length(ranks)
-  ranks <- which(rownames(RES) %in% geneset)
-  if(length(ranks) < length(geneset)){
-    ranks <- c(ranks,rep(nrow(RES) + 1, length(geneset) - length(ranks))) #NOINFO
-  }
-
-  Ranks <- cbind.data.frame("Geneset" = geneset, "Rank" = ranks)
-  return(Ranks)
-
-}
-
-psis <- PSI_ref[which(PSI_ref[,1] %in% colnames(resids)),1]
-GAs <- GAs[which(GAs %in% colnames(resids))]
-
-sPLS_ASCA <- rbind.data.frame(getrank(sPLS_cands,p_n_b),getrank(sPLS_cands,psis),getrank(sPLS_cands,GAs))
-sPLS_ASCA <- cbind.data.frame(sPLS_ASCA, "Method" = rep("sPLS_ASCA", nrow(sPLS_ASCA)))
-
-#####
-
-PCA_DIST_ASCA <- rbind.data.frame(getrank(residual_cands,p_n_b),getrank(residual_cands,psis),getrank(residual_cands,GAs))
-PCA_DIST_ASCA <- cbind.data.frame(PCA_DIST_ASCA, "Method" = rep("PCA_DIST_ASCA", nrow(PCA_DIST_ASCA)))
-
-Correlations <- rbind.data.frame(getrank(coexp_cands,p_n_b),getrank(coexp_cands,psis),getrank(coexp_cands,GAs))
-Correlations <- cbind.data.frame(Correlations, "Method" = rep("Correlations", nrow(Correlations)))
-
-Correlations_ASCA <- rbind.data.frame(getrank(residual_coexp,p_n_b),getrank(residual_coexp,psis),getrank(residual_coexp,GAs))
-Correlations_ASCA <- cbind.data.frame(Correlations_ASCA, "Method" = rep("Correlations_ASCA", nrow(Correlations_ASCA)))
-
-RES <- rbind.data.frame(sPLS_ASCA, PCA_DIST_ASCA, Correlations_ASCA, Correlations)
-
-RES$Geneset[which(RES$Geneset %in% p_n_b)] <- "SLs"
-RES$Geneset[which(RES$Geneset %in% psis)] <- "Phosphate"
-RES$Geneset[which(RES$Geneset %in% GAs)] <- "GAs"
-
-q <- ggplot(RES[-which(RES$Geneset == "GAs"),], aes(x = Method, y = Rank, fill = Geneset)) + geom_boxplot() + theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust=1))
-```
-
-## Methods compared
-
-```{r, fig.cap="Method Ranks on Real Data"}
-q
-
-```
-
-Ranked correlations outperforms the other methods at detecting SL genes, however it also detects PSIs as top candidates, due to the strong between correlations.
-Using the ASCA residuals to investigate the experiment independent associations helps to overcome this.
-Ranked correlations on ASCA residuals and sPLS on ASCA residuals both perform well at detecting the SLs and not detect the PSIs.
-The PSIs are ranked as least important with sPLS_ASCA, this would suggest the method is performing the required task best.
-
-We can explain the larger variation, and slightly worse ranking of sPLS_ASCA compared to Correlations_ASCA since with sPLS we rank the features based on VIP scores and therefore do not take the direction of the relationship into account, where at least 2 of the features used to evaluate the methods were detected through directional ranked correlations.
-
-Since looking at the distance between the SL genes in the PCA space did not yield the required results we drop this method at this point for simplicity.
-
-# QQplots of some top candidates
-
-Through assessing the relationship between one bait and the 9 top candidates of each method tested we highlight the problem with basic correlations, with 4 of the top 9 candidates showing little to no within group correlations.
-However it is difficult to determine which of the two residual based methods are performing best (within the top 20 candidates of the two there are 10 common features).
-Both are able to detect more complex relationships, after accounting for variance explained by the experimental design (plots below).
-
-It makes more sense to use sPLS as this models the response in a multivariate way as opposed to averaged pairwise rankings in the correlation approach.
-
-In this limited view we can see that sPLS seems to detect features that have a nonlinear relationship across the experiment, through linear relations in the ASCA residuals.
-However, sPLS as implemented through mixOmics shows an artefact (still uncharacterised) in previous simulations.
-
-# Questions
-
-- What metric to use in place of visualisation of pairwise relationships to measure method performance in this context?
-- If the relationship between SLs in the PCA of the ASCA residuals is unclear, how to simulate their relationship?
-- Could the filtering be automated or turned into an optimisation problem, instead of using arbitrary cutoff values?
-- Would automated selection of an "axis of interest" in the main effect model improve candidate selection - i.e.
-features need to behave in a more strictly similar manner (positive or negative) to baits?
-- Is it possible and would it make sense to project the baits into the latent space of the PLS model to define an axis/direction?
-
-```{r}
-
-qqplots_spls <- list()
-qqplots_coexp <- list()
-qqplots_pca <- list()
-qqplots_resid_cor <- list()
-
-qqplots_spls_SR <- list()
-
-i <- NULL
-for(i in 1:9){
-  qqplots_spls[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(sPLS_cands)[i])],
-                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
-       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time))
-
-  qqplots_coexp[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(coexp_cands)[i])],
-                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
-       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time))
-
-  qqplots_resid_cor[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(residual_coexp)[i])],
-                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
-       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time))
-
-}
-
-wrap_plots(qqplots_coexp) + plot_layout(guides = "collect") + plot_annotation(title = "Correlations")
-wrap_plots(qqplots_spls) + plot_layout(guides = "collect") + plot_annotation(title = "sPLS_ASCA")
-wrap_plots(qqplots_resid_cor) + plot_layout(guides = "collect") + plot_annotation(title = "Correlations_ASCA")
-
-```

### 2024-01-11 R/Paper_markdown.Rmd
24_11_01_1

@@
 
-#bibliography: "`r rbbt::bbt_write_bib('MASCARA_refs.json', overwrite = TRUE, filter = ("accessed"))`"
+#bibliography: "`r rbbt::bbt_write_bib('MASCARA_refs.json', overwrite = TRUE)`"
 
@@
 
-```{r}
-# renv::snapshot()  #change this
-
-```
-
 # Abstract
@@
 rownames(META) <- NULL
-# kable(META, format = "simple") #replace with CANDTAB like
 
@@
 levels(rdo[[2]]$data$time) <- c("1","3","7","8")
-
-# rdo <- list(META,rdo[[1]],rdo[[2]],rdo[[3]])
-# plot_grid(META)
-# # grid.newpage()
-# # grid.draw(META)
-```
-
-```{r, echo = F}
-# h <- readRDS("heat_comp.RDS")
-# levels(h[[2]]$time) <- c("1","3","7","8")
-# 
-# ann_colours <- list(g_c = c("P-" = "#F8766D","P+" = "lightblue"),
-#                     time = c("1" = "#fde725","3" = "#35b779","7" = "#31688e","8" = "#440154"))
-# 
-# heat <- pheatmap::pheatmap(t(scale(t(h[[1]]))),  #
-#                              cluster_rows = T,
-#                              cluster_cols = F, 
-#                              show_rownames = T, 
-#                              show_colnames = F,
-#                              annotation_col = h[[2]],
-#                            annotation_colors = ann_colours,
-#                            treeheight_row = 10,
-#                            fontsize = 8,
-#                              # annotation_row = anno,
-#                              main = "Strigolactone Gene Expression",silent = T)
-#   
-#   # heat
-#   library(ggplotify)
-#   heat2 <- as.ggplot(heat) + theme(plot.title = element_text(face = "plain"))
-
-# heat2 <- readRDS("heatmap_MASCARA_Fig_2.RDS")
 
@@
 knitr::include_graphics("rice_s_l.png")
-
 ```
@@
 
-<!-- The simulated data is created by simulating score and loading values (as in Figure \@ref(fig:simOverview)) for the combined effect ($\alpha + \alpha\beta$) ) explaining the difference in time profiles of the P+ and P- condition and the main time effect $\beta$, as well as, structured ambient variance ($S$).
-This is to mimic variance due to gene expression as a function of growth condition and time as well as variance due to coexpression of genes in the same  pathway on top of this experimentally induced DE variance.
--->
-
 ## Data simulation 1: Combined effect size and number of differential genes
@@
 
-<!-- For example the correlation between a SL gene $j_{sl}$ and a non-SL gene $j_q$ is calculated as: -->
-
-<!-- \begin{equation} -->
-<!-- \begin{aligned} -->
-<!--   r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}} -->
-<!-- \end{aligned}(\#eq:Pearson) -->
-<!-- \end{equation} -->
-
 ### ASCA+
@@
 
-<!-- The MASCARA pipeline in exemplified in Figure \@ref(fig:MASCARAOVERVIEW). -->
-
-<!-- ```{r MASCARAOVERVIEW, echo = F, fig.cap="MASCARA pipeline overview"} -->
-<!-- a <- knitr::include_graphics("MASCARA_Pipeline_overview.png") -->
-<!-- a -->
-<!-- ``` -->
-
 ## Performance evaluation; Log2 geometric mean rank
@@
 
-  #this predict method does matrix multiplication of the input (new data) and the rotation (loading) matrix
-  #projects the data into the PCA space
   PCDE <- predict(PCD, newdata)
@@
 a_c.PCD <- PCD
-
 ```
@@
 
-# f <- F1_plot(ASCA_cands, baits, 500)[[2]]
 ```
@@
 
-# #remove RP samples
-# X <- X[-which(meta$growth_condition == "PR"),]
-# meta <- meta[-which(meta$growth_condition == "PR"),]
-# 
 meta$time <- as.character(meta$time)
@@
 
-# REF <- readRDS("ref4ASCA_rice.RDS")
 ref <- cbind.data.frame(Feature = colnames(X), feature = colnames(X), Baits = colnames(X))
@@
 PSI_ref <-read.table("Rice_Annotation/PSI_genes.txt", sep = "\t", header = T)
-
-# baits <- SL_ref$feature
 
@@
 dens
-```
-
-```{r extra plot making}
-# ASCA_cands3 <- ASCA_cands2[-which(rownames(ASCA_cands2) %in% c("Os06t0570600-00","Os06t0570566-00", "Os09t0321200-00","Os06t0651900-01", "Os10t0444566-00")),]
-# h <- readRDS("heat_comp.RDS")
-# levels(h[[2]]$time) <- c("1","3","7","8")
-# 
-# # make row annotation
-# 
-# row_ann <- data.frame("Feature" = rep(c("SL","Other DE"),each = 9))
-# rownames(row_ann) <- c(pathway,rownames(ASCA_cands3)[1:9])
-# 
-# ann_colours <- list(g_c = c("P-" = "#F8766D","P+" = "lightblue"),
-#                     time = c("1" = "#fde725","3" = "#35b779","7" = "#31688e","8" = "#440154"),
-#                     Feature = c("SL" = "purple", "Other DE" = "lightgreen"))
-# 
-# h[[1]] <- t(X[,c(pathway,rownames(ASCA_cands3)[1:9])])  #psi_ref[,1]
-# 
-# heat <- pheatmap::pheatmap(t(scale(t(h[[1]]))),  #
-#                              cluster_rows = F,
-#                              cluster_cols = F,
-#                              show_rownames = T,
-#                              show_colnames = F,
-#                              annotation_col = h[[2]],
-#                            annotation_row = row_ann,
-#                            annotation_colors = ann_colours,
-#                            treeheight_row = 10,
-#                            fontsize = 8,
-#                              # annotation_row = anno,
-#                              main = "Scaled Gene Expression",silent = T)
-# 
-#   # heat
-#   library(ggplotify)
-#   heat2 <- as.ggplot(heat) + theme(plot.title = element_text(face = "plain"))
-#
-#   saveRDS(heat2, "heatmap_MASCARA_Fig_2.RDS")
-
-# 
-# qqplot <- function(data, meta, ...){
-# 
-#   df <- cbind.data.frame(meta,data)
-# 
-#   df$time <- factor(df$time, levels = unique(meta$time))
-#   df$growth_condition <- factor(df$growth_condition)
-# 
-# 
-#   ggplot(df, ...) +
-#     geom_point() +
-#     theme_bw()
-# 
-# 
-# }
-# 
-# meta$time <- rep(c(1,3,7,8), each = 3, 2)
-# 
-# cand_b <- PSI_cands[10]
-# cand_w <- baits[3]
-# 
-# cor_tw <- round(cor(X[,which(colnames(X) %in% cand_w)], X[,which(colnames(X) %in% baits[4])]), digits = 2)
-# 
-# cor_tb <- round(cor(X[,which(colnames(X) %in% cand_b)], X[,which(colnames(X) %in% baits[4])]), digits = 2)
-# 
-# cor_rw <- round(cor(res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% cand_w)], res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% baits[4])]), digits = 2)
-# 
-# cor_rb <- round(cor(res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% cand_b)], res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% baits[4])]), digits = 2)
-# 
-# 
-# within <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% cand_w)],
-#                         "Bait" = X[,which(colnames(X) %in% baits[4])]),meta,
-#        aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + xlab(baits[4]) + ylab(cand_w) + labs(title = paste0("r = ",cor_tw))
-# 
-# PSI_cands <- rownames(ASCA_cands[which(rownames(ASCA_cands) %in% psi_ref[,1]),])
-# PSI_cands <- head(rownames(ASCA_cands[-which(rownames(ASCA_cands) %in% pathway),]),20)
-# 
-# 
-# between <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% cand_b)],
-#                         "Bait" = X[,which(colnames(X) %in% baits[4])]),meta,
-#        aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + xlab(baits[4]) + ylab(cand_b) + labs(title = paste0("r = ",cor_tb))
-# 
-# # within + between + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A')) + plot_layout(heights = unit(c(5), c("cm")), widths = c(5))
-# 
-# 
-# 
-# 
-# within_r <- qqplot(cbind.data.frame("Candidate" = res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% cand_w)],
-#                         "Bait" = res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% baits[4])]),meta,
-#        aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + xlab(baits[4]) + ylab(cand_w)   + labs(title = paste0("r = ",cor_rw))#, alpha = (16- as.numeric(growth_condition))
-# 
-# 
-# between_r <- qqplot(cbind.data.frame("Candidate" = res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% cand_b)],
-#                         "Bait" = res_ASCAplus_nb$residuals[,which(colnames(res_ASCAplus_nb$residuals) %in% baits[4])]),meta,
-#        aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + xlab(baits[4]) + ylab(cand_b)  + labs(title = paste0("r = ",cor_rb))
-# 
-# (within + between)/(within_r + between_r) + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A','1')) #+ plot_layout(heights = unit(c(5), c("cm")), widths = c(5))
-
-#^^ C:/Users/fwhite/Documents/within_between_example_2.png
 ```

### 2024-01-11 R/run_first/2.plot_pathway_net_example.Rmd
24_11_01_1

@@
 library(animation)
+library(gASCA)
+library(matrixStats)
+library(patchwork)
+
+source("../DATA_SIM_FUNCS.R")
 ```
@@
 #windowsFonts() # to check which are loaded
+
+renv::status()
+
 ```
@@
 # plot residuals and full (with tweaks for problem demonstration)
+baits <- paste0("X_",c(1985:1988))   #501:510
+spikes <- paste0("X_",c(1989:2000))
 
@@
 spikes <- spikes[-c(1:5)]
+
+meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
+colnames(meta) <- c("growth_condition","time","interaction")
+meta <- as.data.frame(apply(meta,2,factor))  #9
+
+Experiment_responders <- 30
+
+sim_data <- Create_Core_DEV_2(nreps = 3, meta = meta, 
+                        EffectSize = c(X_a_ab = 5, time = 1, E = 1, mu = 1, Struc_E = 0.4),    # E = 10, Struc_E = 1
+                        struc_resid = TRUE, e_sigma = c(1.5,0.8,0.0), a_sigma = c(1,0.7,0.6,0.1),
+                        b_sigma = c(1,0.9), irr_spikes = TRUE, SCORE_SEED = 12, Experiment_responders = Experiment_responders,
+                        struc_seed = 127, score_plot = FALSE, plot = FALSE)  #12
+
+ar <- ASCA(sim_data[[1]], m = meta, Baits = baits, spikes = spikes, ncands = ncands, distance_calc = FALSE, Return_Model = FALSE)
 
@@
 cand_vals <- residuals[,1970:2000]    #rgs[which(rownames(rgs) %in% names(g_b)[1:50]),]
-
 adj <- cor(cand_vals)

### 2024-01-11 R/Paper_markdown.Rmd
24_01_11_2

@@
 source("DATA_SIM_FUNCS.R")
+source("MASCARA_FUNCS.R")
 ```
@@
 library(rhmmer)
+library(pracma)
 ```
@@
 
-<!-- \begin{equation} -->
-<!-- \begin{aligned} -->
-<!-- \bm{Y} &= \bm{1\mu}^T + \bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\ -->
-<!-- \text{where for each element}\\ -->
-<!-- &\bm{T}_{S_{ik}} \sim \mathcal{N}(0,1), \quad \bm{P}_{S_{jk}}^{\text{baits,spikes}} \sim \mathcal{N}(0.5,0.2) \\ -->
-<!-- \text{and:} &\\ -->
-<!-- \bm{P}_{S}^{\text{rest}} &= 0 -->
-<!-- \end{aligned}(\#eq:Sim2-bp) -->
-<!-- \end{equation} -->
-
-<!-- The expression levels of differential transcripts is modeled with the scores and loadings of the growth condition and its interaction with time; termed here the combined effect ( $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{P}_{\alpha + \alpha\beta}^T$ ).
-The time effect is modeled with scores and loadings ( $\mathbf{T}_{\beta}$  and $\mathbf{P}_{\beta}^T$ ).
-Also included is some random measurement noise, this is modeled as a matrix $\mathbf{E}$ with normally distributed random values $e_{ij} = \mathcal{N}(0,\sigma^2)$.
--->
-
-<!-- On top of the between-group effect there is an additional correlation structure applied to the baits and spikes to create a model of coexpression that is independent to the variance induced by the experimental factors ($\mathbf{X}_S = \mathbf{T}_S\mathbf{P}_S^T$).
-This $\mathbf{X}_S$ matrix contains two component sources of variation which links the baits and spikes.
-In this simulation we have a set of 4 baits that have positive loadings for the combined effect along with a variable number of other genes - "Combined Effect Responders", which are the genes from a simulated URP (indicated by the yellow bar in Figure \@ref(fig:simOverview)).
-For the structured ambient variance; 2 random vectors in \(\bm{T}_S\) (each column representing an effect caused by ambient conditions) which our baits and spikes contribute to (i.e.
-green and black bars in Figure \@ref(fig:simOverview) are non zero in \(\bm{P}_S\)).
-Similarly to the real data, the gene expression values for each experimental condition are simulated with three independent replicates per time-condition-combination.
-Each simulated data $\mathbf{Y}$ set thus consists of 24 samples and 2000 genes.
--->
-
 ```{r, echo = FALSE}
@@
 
-<!-- The scores matrices $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{T}_{\beta}$ each consist of two orthogonal score vectors and their relative levels are indicated by the colour.
-The different colours in Figure \@ref(fig:simOverview) represent that the scores are orthogonal to each other.
-Each small block consists of the same score value for each of the replicates within that condition.
-Loadings $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$  are simulated to create the baits and spikes.
-As displayed in Figure \@ref(fig:simOverview), the green bar represents 4 baits (or known pathway genes), the black bar represents the 12 spikes (“unknown” genes from POI or RP), yellow represents differential URP genes, and purple; a set of differential URPs with an alternative treatment response.
--->
-
-<!-- In $\mathbf{P}_{\beta1}^T$ the baits, spikes and other DE genes (all genes from POI, RP and URP) along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model.
-The remaining elements of $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$ are populated with zeroes.
-The $\mathbf{T}$ matrices represent the simulated scores for their respective factors.
-For blues, reds, greys and golds we have 4 different shades representing 4 unique values per factor level (total 8 unique values where reds and golds are sign flipped blues and greys respectively).
-Similarly oranges and pinks represent a growth condition independent time profile with four unique values.
-$\mathbf{E}$ represents random unstructured technical noise and the additional $\bm{1\mu}^T$ term in \@ref(eq:Sim2-bp) represents a mean value for each gene.
--->
-
-<!-- The issue to address is the presence of many differential URPs; those that have strong between group correlations but no within group correlation.
-Here these differential genes relate  to the yellow bar in Figure \@ref(fig:simOverview).
-We incrementally increase the number of non-zero values and decrease the number of 0 values, which in figure \@ref(fig:simOverview) relates to increasing the size of the yellow bar, as well as varying the size (as quantified by the Frobenius norm) of the combined effect matrix through the scalar parameter *d* below; -->
-
 ## Data simulation 2: Replication
@@
 Here we create 20 datasets with random combined effect and within group correlation structures.
-
-<!-- In order to evaluate the methods and indicate the levels of variance required for optimal performance we conduct a larger scale assay.
-Here we fix the number of DE URP genes to 30 which corresponds to the yellow bar size and the size of the within correlation effect matrix (\(\|\bm{T}_S\bm{P}_S^T\|_F = 0.025\)).
-We investigate a range of values for sizes of between variance (*d*) and random noise (*l*) across random structures (which correspond to repeated simulations of randomised values for all $\mathbf{T}$ matrices).
-Parameters were adjusted according to \@ref(eq:Test-2) on 20 randomised $\mathbf{T}$ matrices.
--->
 
@@
 Under the conditions of orthogonal scores; $\mathbf{t}_{k}^{T}\mathbf{t}_{k} = 0$ and normalised weights; $\mathbf{u}_{k}^T\mathbf{u}_{k{}} = 1$ and $\mathbf{v}_{k}^T\mathbf{v}_{k{}} = 1$.
-
-   <!-- c.
-Perform linear regression of \( \mathbf{\tilde{Y}}_k \) and \( \mathbf{B}_k \) onto their corresponding scores \( \bm{t}_k \) and \( \bm{w}_k \) to obtain the respective loadings \( \bm{p}_k \), and \( \bm{q}_k \).
--->
-   <!-- d.
-Deflate \( \mathbf{\tilde{Y}}_k \) and \( \mathbf{B}_k \) by removing the variance explained by the $k^{th}$ set of scores in the space of $\mathbf{\tilde{Y}}$\( \bm{t}_k \) and use the deflated matrices as inputs for the next iteration.
--->
-
-<!-- \begin{equation} -->
-<!-- \begin{aligned} -->
-<!-- \mathbf{\tilde{Y}_{k}} = \mathbf{\tilde{Y}}_{k-1} - \mathbf{t}_{k}\mathbf{p}^T_{k} \\ -->
-<!-- \text{and} \\ -->
-<!-- \mathbf{B_{k}} = \mathbf{B}_{k-1} - \mathbf{t}_{k}\mathbf{q}^T_{k} \\ -->
-<!-- \end{aligned} -->
-<!-- \end{equation} -->
 
@@
 source("DATA_SIM_FUNCS.R")
+source("MASCARA_FUNCS.R")
 
@@
 residuals <- cbind.data.frame(meta[,c("growth_condition","time")], res_ASCAplus_nb$residuals)
-```
-
-```{r}
-#get cands from residuals
-rPCD <- prcomp(data.matrix(residuals[,-c(1:2)]))
-```
-
-```{r}
+
 cutoff <- round(quantile(c(1:nrow(ASCA_cands)),0.1))
@@
 
-rpcd <- prcomp(data.matrix(res_filt[,-c(1:2)]))
-
-ref$Baits <- as.character(ref$Baits)
-ref$Baits[which(ref$Feature %in% PSI_ref[,1])] <- "PSI"
-ref$Baits <- factor(ref$Baits, levels = c("Other","Pathway","PSI"))
-
-Fref <- ref[match(rownames(rpcd$rotation),ref$Feature),]
-
-```
-
-```{r, include = F}
-importance <- paste0(round(summary(rpcd)$importance[2,] * 100, digits = 2), "%")
-```
-
-```{r}
-# resids <- residuals[,-c(1:2)]
 resids <- res_filt[,-c(1:2)]
-spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], resids[,which(colnames(resids) %in% baits)], all.outputs = T)
-
-sPLS_cands <- cbind.data.frame(abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X, spls_res$loadings$X * spls_res$prop_expl_var$X)
-sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = T),]
-
-importance <- paste0(round(spls_res$prop_expl_var$X * 100, digits = 2), "%")
-
-Fref2 <- Fref[-which(Fref$Feature %in% baits),]
-Fref2 <- Fref2[match(rownames(sPLS_cands),Fref2$Feature),]
-
-```
-
-```{r}
-#calculating direction of baits to determine direction of candidate selection spls loading space
-
-angles <- rbind.data.frame(spls_res$loadings$Y, colMeans(spls_res$loadings$Y))
-
-x <- angles[4,]
-y <- angles[5,]
-
-#adapted from https://stackoverflow.com/questions/1897704/angle-between-two-vectors-in-r
-angle <- function(x,y){
-
-  x <- c(as.matrix(x))
-  y <- c(as.matrix(y))
-
-  dot.prod <- x%*%y 
-  norm.x <- norm(x,type="2")
-  norm.y <- norm(y,type="2")
-  theta <- acos(dot.prod / (norm.x * norm.y))
-  return(as.numeric(theta))
-}
-
-angles_loadings <- apply(sPLS_cands[,-1],1,angle, y = y)
-
-#define this better
-sPLS_cands <- cbind.data.frame(sPLS_cands, "angle" = angles_loadings, "score" = 3 * sPLS_cands$`abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X` * (1-angles_loadings))
-
-sPLS_cands <- sPLS_cands[order(sPLS_cands$score, decreasing = T),]
-```
-
-```{r}
-avg_point <- angles[5,]/5
-
-rot.mat <- matrix(c(0,-1,1,0),nrow = 2, byrow = T)  #90 degree clockwise rotation
-
-rotated_point <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat)
-
-ang <- angle(avg_point,c(1,0))
-
-rot.mat2 <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation
-rotated_point2 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat2)
-
-```
-
-```{r}
-#rotated loadings
-ang <- angle(avg_point,c(1,0))
-
-rot.mat2 <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation
-
-rotated_point2 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat2)
-
-rotate <- function(x,y){
-  x <- c(as.matrix(x))
-  x %*% y
-}
-
-loadings_r_baits <- t(apply(sPLS_cands[,c("comp1","comp2")],1,rotate, y = rot.mat2))
-colnames(loadings_r_baits) <- c("comp1","comp2")
-```
-
-```{r}
-sPLS_bait_cands <- loadings_r_baits[order(loadings_r_baits[,1], decreasing = T),]
-
-```
-
-```{r}
-range.ang <- seq(from = -pi/6, to = pi/6, length.out = 60)
-
-###
-
-i <- NULL
-L_R <- data.frame()
-LOADINGS_R <- list()
-
-for(i in 1:length(range.ang)){
-
-ang <- range.ang[i]
-
-rot.mat <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation
-loadings_r <- t(apply(loadings_r_baits[,c("comp1","comp2")],1,rotate, y = rot.mat))
-colnames(loadings_r) <- c("comp1","comp2")
-L_R <- rbind(L_R,loadings_r)
-  LOADINGS_R[[i]] <- loadings_r
-
-}
-
-# colnames(LOADINGS_R) <- c("comp1","comp2")
-L_R$rad <- rep(range.ang, each = nrow(sPLS_cands))
-L_R$frame <- rep(c(1:length(range.ang)), each = nrow(sPLS_cands))
-
-```
-
-```{r}
-#get sd of original comp1 distribution
-sigma <- sd(sPLS_cands$comp1)
-
-big.T.sig <- function(x,y){
-  big <- length(which(x > 2.5*y))
-  return(big)
-}
-
-i <- NULL
-res <- c()
-for(i in 1:length(LOADINGS_R)){
-
-  r <- big.T.sig(LOADINGS_R[[i]][,1],sigma)
-  res <- c(res,r)
-}
-
-```
-
-```{r}
-RES <- cbind.data.frame("Rad" = range.ang, "Density" = res)
-
-density <- ggplot(RES, aes(x = Rad, y = Density))+
-  geom_smooth() + theme_bw() + ggtitle("Number of points above 2*sigma of original rotation")
-
-```
-
-```{r}
-
-#taken from https://stackoverflow.com/questions/6836409/finding-local-maxima-and-minima
-inflect <- function(x, threshold = 1){
-  up   <- sapply(1:threshold, function(n) c(x[-(seq(n))], rep(NA, n)))
-  down <-  sapply(-1:-threshold, function(n) c(rep(NA,abs(n)), x[-seq(length(x), length(x) - abs(n) + 1)]))
-  a    <- cbind(x,up,down)
-  list(minima = which(apply(a, 1, min) == a[,1]), maxima = which(apply(a, 1, max) == a[,1]))
-}
-
-ggres <- ggplot_build(density)$data[[1]]
-infs <- inflect(ggres$y)
-rads <- ggres$x[infs$maxima]
-
-```
-
-```{r}
-maxima.ang <- rads
-
-###
-
-i <- NULL
-sPLS_maxima_cands <- list()
-
-for(i in 1:length(maxima.ang)){
-
-ang <- maxima.ang[i]
-
-rot.mat <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation
-loadings_r <- as.data.frame(t(apply(loadings_r_baits[,c("comp1","comp2")],1,rotate, y = rot.mat)))
-colnames(loadings_r) <- c("comp1","comp2")
-sPLS_maxima_cands[[i]] <- loadings_r[order(loadings_r$comp1, decreasing = T),]
-
-}
-```
-
-```{r}
-Fref2 <- Fref[-which(Fref$Feature %in% baits),]
-Fref2 <- Fref2[match(rownames(sPLS_maxima_cands[[1]]),Fref2$Feature),]
-```
-
-```{r}
-Fref2 <- Fref[-which(Fref$Feature %in% baits),]
-Fref2 <- Fref2[match(rownames(sPLS_maxima_cands[[2]]),Fref2$Feature),]
 
@@
 
-residual_cands <- ranked_dist(baits[1:3], rpcd)
-
-residual_coexp <- ranked_coexp(baits, resids)
-
-```
-
-```{r}
-
-sPLS_cands <- sPLS_maxima_cands[[2]]
-
-Methods <- c("sPLS_ASCA","PCA_DIST_ASCA","Correlations_ASCA", "Correlations")
-
-getrank <- function(RES,geneset,NOINFO = 26402){
-
-  ranks <- which(rownames(RES) %in% geneset)
-  if(length(ranks) < length(geneset)){
-    ranks <- c(ranks,rep(nrow(RES) + 1, length(geneset) - length(ranks))) #NOINFO
-  }
-
-  Ranks <- cbind.data.frame("Geneset" = geneset, "Rank" = ranks)
-  return(Ranks)
-
-}
-
-psis <- PSI_ref[which(PSI_ref[,1] %in% colnames(resids)),1]
-
-```
-
-```{r}
-GA_ref <- read_xlsx("Rice_Annotation/GA pathway genes-IDs.xlsx")
-GA_ref$ID <- gsub("g","t",GA_ref$ID)
-GAs <- ref[which(gsub("-.*","",ref$Feature) %in% GA_ref$ID),1]
-
-GAs <- GAs[which(GAs %in% colnames(resids))]
-sPLS_ASCA <- rbind.data.frame(getrank(sPLS_cands,p_n_b),getrank(sPLS_cands,psis),getrank(sPLS_cands,GAs))
-sPLS_ASCA <- cbind.data.frame(sPLS_ASCA, "Method" = rep("sPLS_ASCA", nrow(sPLS_ASCA)))
-
-#####
-
-PCA_DIST_ASCA <- rbind.data.frame(getrank(residual_cands,p_n_b),getrank(residual_cands,psis),getrank(residual_cands,GAs))
-PCA_DIST_ASCA <- cbind.data.frame(PCA_DIST_ASCA, "Method" = rep("PCA_DIST_ASCA", nrow(PCA_DIST_ASCA)))
-
-Correlations <- rbind.data.frame(getrank(coexp_cands,p_n_b),getrank(coexp_cands,psis),getrank(coexp_cands,GAs))
-Correlations <- cbind.data.frame(Correlations, "Method" = rep("Correlations", nrow(Correlations)))
-
-Correlations_ASCA <- rbind.data.frame(getrank(residual_coexp,p_n_b),getrank(residual_coexp,psis),getrank(residual_coexp,GAs))
-Correlations_ASCA <- cbind.data.frame(Correlations_ASCA, "Method" = rep("Correlations_ASCA", nrow(Correlations_ASCA)))
-
-RES <- rbind.data.frame(sPLS_ASCA, PCA_DIST_ASCA, Correlations_ASCA, Correlations)
-
-RES$Geneset[which(RES$Geneset %in% p_n_b)] <- "SLs"
-RES$Geneset[which(RES$Geneset %in% psis)] <- "PSIs"
-RES$Geneset[which(RES$Geneset %in% GAs)] <- "GAs"
-
-RES$Method[which(RES$Method == "sPLS_ASCA")] <- "MASCARA"
-
-RES <- RES[which(RES$Method %in% c("Correlations","MASCARA")),]
-
-q <- ggplot(RES[-which(RES$Geneset == "GAs"),], aes(x = Method, y = Rank, fill = Geneset)) + geom_boxplot() + theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust=1))
-```
-
-```{r, fig.cap="Method Ranks on Real Data"}
-
-REF2 <- as.data.frame(read_tblout("Rice_Annotation/HMM_ITAG_1.txt"))
-
-f_p_m <- REF2[which(REF2$domain_name %in% rownames(sPLS_cands)[1:25]),]
-f_p_m <- f_p_m[order(f_p_m$sequence_evalue),]
-fpm <- f_p_m[!duplicated(f_p_m$domain_name),]
-fpm <- fpm[,c(1,3,19)]
-fpm[1,3] <- "Diterpenoid phytoalexin biosynthesis"
-fpm[12,3] <- "Vesicle-Associated Membrane Protein-Associated Protein"
-cands <- rownames(sPLS_cands)[1:25]
-cands <- cands[which(cands %in% fpm$domain_name)]
-rownames(fpm) <- fpm[,1]
-fpm <- fpm[cands,]
-
-```
-
-```{r}
-rownames(fpm) <- NULL
+sPLS_cands <-  MASCARA(data.matrix(resids), baits)[[1]]             #MASCARA
 
@@
 
-```{r realcandidates, fig.cap="Selected candidates from A-C ranked correlations and D-F MASCARA.
+```{r realcandidates, fig.cap="Example candidates from A-C ranked correlations and D-F MASCARA.
 X axes; transformed expression level of one of the bait genes (CCD8), y axes; expression levels of three example coexpression candidates for each method (see @haiderTranscriptomeAnalysisPhosphate2023 for preprocessing details)."}
@@
 qqplots_coexp <- list()
-qqplots_pca <- list()
-qqplots_resid_cor <- list()
-
-qqplots_spls_SR <- list()
 
 i <- NULL
-for(i in 1:9){
+for(i in 1:20){
   qqplots_spls[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(sPLS_cands)[i])],
@@
 
-  qqplots_resid_cor[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(residual_coexp)[i])],
-                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
-       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + ylab(rownames(residual_coexp)[i])
-
-}
-
-plot1 <- wrap_plots(qqplots_coexp[c(5,7,8)]) + plot_annotation(title = "Correlations")
-plot2 <- wrap_plots(qqplots_spls[c(1,5,8)]) + plot_annotation(title = "MASCARA")
-
-# plot1/plot2 + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A','1'))
+}
+
+plot1 <- wrap_plots(qqplots_coexp[c(6,8,10)]) + plot_annotation(title = "Correlations") #+ theme(aspect.ratio = 1)
+plot2 <- wrap_plots(qqplots_spls[c(2:4)]) + plot_annotation(title = "MASCARA") #+ theme(aspect.ratio = 1)
+
+# wrap_plots(qqplots_spls) + plot_layout(guides = "collect",tag_level = 'new')
+
+PLOT <- plot1/plot2 + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A','1'))
+# ggsave("within_between_example.png", PLOT)
 
@@
 
-We acknowledge funding by the Dutch Research Council (NWO/OCW) for the MiCRop Consortium program, Harnessing the second genome of plants (Grant number 024.004.014; to HB, LD, AKS, JAW), the Dutch Research Council (NWO-TTW grant 16873 Holland Innovative Potato; to HB and LD) and the Data Science Centre of the University of Amsterdam (to FW).
+We acknowledge funding by the Dutch Research Council (NWO/OCW) for the MiCRop Consortium program, Harnessing the second genome of plants (Grant number 024.004.014; to HB, LD, AKS, JAW), the Dutch Research Council (NWO-TTW) Holland Innovative Potato (grant 16873; to HB, LD, FW) and the Data Science Centre of the University of Amsterdam (to FW).
 FW would like to thank Frans van der Kloet, Fentaw Abegaz, Roel van der Ploeg for statistical discussions and Imran Haider for consultation on the strigolactone pathway.
@@
 \end{equation}
-
-<!-- $r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}}$ -->
-
-<!-- MASCARA example run on $params: -->
-
-<!-- Data was simulated: -->
-
-<!-- \begin{equation} -->
-<!-- \begin{aligned} -->
-<!-- \bm{Y} &= \bm{1\mu}^T + \bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T + \bm{Y}_{\beta} + \bm{E} \\ -->
-<!-- \text{where:} &\\ -->
-<!-- \bm{T}_{\alpha + \alpha\beta} &= \text{vec}(\bm{ab}^T) \quad \text{where:} \\ -->
-<!-- \bm{c}_{i,1:4} &\sim \mathcal{N}(0,1), \quad \bm{a}_{1:r} = [1,\ldots,1], \quad \bm{b}_{1:8} = [\bm{c}_{i},-\bm{c}_{i}]^T \\ -->
-<!-- \bm{T}_{\beta} &= \text{vec}(\bm{de}^T) \quad \text{where:} \\ -->
-<!-- \bm{f}_{1:4} &\sim \mathcal{N}(0,1.5), \quad \bm{d}_{1:r} = [1,\ldots,1], \quad \bm{e}_{1:8} = [\bm{f},\bm{f}]^T \\ -->
-<!-- \bm{P}_{\alpha + \alpha\beta}^{1:2} &= \begin{bmatrix}1.5_{1:4} & 1_{5:8} & 0.5_{9:12} & 0_{13:2000}\end{bmatrix} \\ -->
-<!-- \bm{P}_{\beta}^{1:2} &= \begin{bmatrix}0.1_{1:1000} & 0_{1001:2000}\end{bmatrix} \\ -->
-<!-- \bm{E} &\sim \mathcal{N}(0,0.75) -->
-<!-- \end{aligned}(\#eq:suppl-Sim-bp) -->
-<!-- \end{equation} -->
-
-<!-- genes 1:12 have non-zero loadings on PC1 of $svd({\bm{Y}_{\alpha + \alpha\beta}})$, these are considered genes of interest.
-These 12 genes correspond to the baits and spikes (spikes represent "unknown" genes of interest, used for method evaluation).
-Which of these genes are baits and which are spikes depends on the methods being tested and will be discussed in the Analysis and Feature Selection section.
-genes 501:510 have a value of 0.5 on PC2 of $svd({\bm{Y}_{\alpha + \alpha\beta}})$ and are considered to display a nonsense effect.
--->
-
-<!-- genes 1:1000 have a loadings of 0.1 on PC1 of $svd({\bm{Y}_\beta})$ and genes 1001:2000 have loadings of 0.1 on PC2 of $svd({\bm{Y}_\beta})$.
-To create $\bm{Y}$ the dot product of the scores and transposed loadings is calculated per factor, and added to a general mean value for each feature ($\bm{1\mu}^T$) as well as some normally distributed noise \(\bm{E}\) to create differences between the replicates, see equation \@ref(eq:Sim-bp) for an overview.
--->
-
-<!-- <!-- \End{multicols} -->
-
-<!-- sim2:  -->
-<!-- For this simulation we create 2 components in the combined effect (${\bm{Y}_{\alpha + \alpha\beta}}$), 2 for time ($\bm{Y}_\beta$).
-For the time effect, we create this similarly to simulation 1.
-For the combined effect: 2 random orthogonal scores.
-The loadings for these components contain several sets of genes, some with positive and some with negative loadings onto the 4 components.
--->
-
-<!-- Y = TP (ambient structure) This is regressed onto the concatenated scores of the combined effect and time to ensure orthogonality, creating relationships that are independent of the experimentally induced variance.
--->

### 2024-01-11 R/DATA_SIM_FUNCS.R

@@ -1862,121 +1862,17 @@ MASCARA4_test <- function(resids,ref, baits, spikes, ncands){
   ######
   
   sPLS_bait_cands <- loadings_r_baits[order(loadings_r_baits[,1], decreasing = TRUE),]
-  
-  
-  # #####
-  # 
-  # #loop over rotation matrices from -pi/4 -> pi/4
-  # #resolution of 100 (?)
-  # 
-  # range.ang <- seq(from = -pi/6, to = pi/6, length.out = 60)
-  # 
-  # ###
-  # 
-  # i <- NULL
-  # L_R <- data.frame()
-  # LOADINGS_R <- list()
-  # 
-  # for(i in 1:length(range.ang)){
-  #   
-  #   
-  #   ang <- range.ang[i]
-  #   
-  #   rot.mat <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = TRUE) #clockwise rotation
-  #   loadings_r <- t(apply(loadings_r_baits[,c("comp1","comp2")],1,rotate, y = rot.mat))
-  #   colnames(loadings_r) <- c("comp1","comp2")
-  #   L_R <- rbind(L_R,loadings_r)
-  #   LOADINGS_R[[i]] <- loadings_r
-  #   
-  #   
-  # }
-  # 
-  # # colnames(LOADINGS_R) <- c("comp1","comp2")
-  # L_R$rad <- rep(range.ang, each = nrow(sPLS_cands))
-  # L_R$frame <- rep(c(1:length(range.ang)), each = nrow(sPLS_cands))
-  # 
-  # 
-  # 
-  # #get sd of original comp1 distribution
-  # sigma <- sd(sPLS_cands$comp1)
-  # 
-  # big.T.sig <- function(x,y){
-  #   big <- length(which(x > 2.5*y))
-  #   return(big)
-  # }
-  # 
-  # i <- NULL
-  # res <- c()
-  # for(i in 1:length(LOADINGS_R)){
-  #   
-  #   r <- big.T.sig(LOADINGS_R[[i]][,1],sigma)
-  #   res <- c(res,r)
-  # }
-  # 
-  # 
-  # RES <- cbind.data.frame("Rad" = range.ang, "Density" = res)
-  # 
-  # density <- ggplot(RES, aes(x = Rad, y = Density))+
[…]

### 2024-01-11 R/MASCARA_FUNCS.R

@@ -1,3 +1,20 @@
+MASCARA <- function(resids, baits){
+  #target projection
+  spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
+                          resids[,which(colnames(resids) %in% baits)], ncomp = 2)
+  
+  #####
+  y <- colMeans(spls_res$Yloadings)
+  U <- spls_res$projection
+  UTP <- (dot(t(U),y)/dot(y,y))
+  
+  sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])
+  
+  return(list(sPLS_bait_cands,UTP,spls_res,y))
+}
+
+
+
 MASCARA4_test_1 <- function(resids, baits, spikes){
   #MIXOMICS
   spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], 

### 2024-01-22 R/23_02_02_real_data_rice.Rmd
24_01_22_1

@@
 ---
-title: "Haider Data Demo"
+title: "Haider Data Figures"
 author: "Fred White"

### 2024-01-22 R/MASCARA_Tutorial.Rmd
24_01_22_1

---
title: "MASCARA: Coexpression Analysis Tutorial"
author: "Fred White"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

```{r, echo=FALSE}
library(devtools)
library(ggplot2)
library(MetStaT)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)    
library(MUVR) 
library(pheatmap)
library(caret)
library(data.table)
library(plotly)
library(mixOmics)
library(DiscriMiner)
library(pracma)
library(matlib)
library(truncnorm)

source("DATA_SIM_FUNCS.R")
source("MASCARA_FUNCS.R")
```

Here we show a typical use case of MASCARA applied to a simulated dataset.

User must supply a dataframe of filtered + transformed data plus associated metadata.

```{r}

meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
colnames(meta) <- c("growth_condition","time","interaction")
meta <- as.data.frame(apply(meta,2,factor))  #9

Experiment_responders <- 30

set.seed(3)
# rnd_seed <- abs(rnorm(3, mean = 0.7, sd = 0.5))

baits <- paste0("X_",c(1985:1988))   #501:510
spikes <- paste0("X_",c(1989:2000))

sim_data <- Create_Core_DEV_2(nreps = 3, meta = meta, plot = TRUE, 
                        EffectSize = c(X_a_ab = 1.8, time = 1, E = 10, mu = 1, Struc_E = 1), 
                        struc_resid = TRUE, e_sigma = c(1.5,0.8,0.0), a_sigma = c(1,0.7,0.6,0.1),
                        b_sigma = c(1,0.9), irr_spikes = TRUE, SCORE_SEED = 12, 
                        Experiment_responders = Experiment_responders, struc_seed = 127)

```

```{r}
ar <- ASCA(sim_data[[1]], m = meta, Baits = baits, spikes = spikes, ncands = ncands, distance_calc = FALSE, Return_Model = FALSE)
```

```{r}
#classic
MASCARA_2 <- MASCARA4_test_2(ar[[3]], baits, spikes = spikes)
which(rownames(MASCARA_2[[2]]) %in% spikes)
```

### 2024-01-22 R/DATA_SIM_FUNCS.R

@@ -2191,7 +2191,7 @@ Create_Core_DEV_2 <- function(nreps, meta, irr_spikes = TRUE, struc_resid = FALS
                abs(rnorm((Experiment_responders + 4),mean = 1, sd = 0.25)), 
                targets_a_ab)
   
-  print(length(a_ab_p1))
+  #print(length(a_ab_p1))
   
   if(irr_spikes == FALSE){
     a_ab_p2 <- c(rep(0,length(a_ab_p1)))
@@ -2311,9 +2311,9 @@ Create_Core_DEV_2 <- function(nreps, meta, irr_spikes = TRUE, struc_resid = FALS
     norm_Struc_E <- norm(Struc_E, type = "F")
     
     
-    dim(Struc_E)
-    print(E[,1] %*% time[,1])
-    print(time[,1] %*% X_a_ab[,1])
+    #dim(Struc_E)
+    #print(E[,1] %*% time[,1])
+    #print(time[,1] %*% X_a_ab[,1])
     
     
     Non_Struc_E <- matrix(rnorm(dim(X_a_ab)[1] * dim(X_a_ab)[2], mean = 0, sd = noise_sd), nrow = nrow(meta))

### 2024-01-26 R/Paper_markdown.Rmd
24_01_22_1
24_01_26

@@
 knitr::opts_chunk$set(echo = FALSE, warning = F, message = F, fig.pos = "H") #, root.dir=here()
-# install.pacakges()
 
@@
 This can be done through, for example, varying the level of certain essential nutrients, knocking out/down a gene within a POI.
-To explore which pathways are affected by the intervention, gene expression can be measured through RNA sequencing (RNAseq).
+To explore the effects of an intervention, gene expression can be measured through RNA sequencing (RNAseq).
 Analysis of RNAseq data generally adheres to one of two approaches: differential expression analysis (DE; @andersDifferentialExpressionAnalysis2010; @robinsonEdgeRBioconductorPackage2010) or coexpression analysis (CoE; @langfelderWGCNAPackageWeighted2008; @tzfadiaCoExpNetVizComparativeCoExpression2016).
@@
 CoE, on the other hand, aims at discovering genes that are part of an already partially characterized POI.
-Here, the expression profiles of known genes of the POI (termed "baits") are used to detect novel pathway members with similar expression patterns (here termed "spikes").
+Here, the expression profiles of known genes of the POI (termed "baits") are used to detect novel pathway members with similar expression patterns (here termed "targets").
 
@@
 ```{r nettype, fig.cap= c("Pathway types exemplified, POI; pathway of interest, RP; related pathway, URP; unrelated pathway.
-A represents hypothetical biological pathways, B highlights the problem with detecting these pathways when dealing with data from designed experiments.")}
+Panel A represents hypothetical biological pathways  that are all affected by the experimental design and thus all contain DE genes, here the edges represent real connections between the nodes/genes.
+Panel B highlights the problem with detecting these pathways when dealing with data from designed experiments, edges represent putative connections defined by coexpression analysis; black is desired (within network detection), grey non-desired.")}
+
+#CHANGE caption
 
@@
 URPs are defined to be pathways not directly interacting with the POI.
-When dealing with experimentally designed data, we have to consider two key sources of variance: between-group variance and within-group variance.
+In the hypothetical example displayed in Figure \@ref(fig:nettype), taking the purple triangles as baits, the desired outcome of a coexpression analysis would be to detect all other purple and orange coloured genes and not the light green genes.
+However, these connections are not always preferentially detected, this is due to the presence of different sources of variance in the data.
+
+When dealing with experimentally designed data, we have to consider two key sources of variance: between-group and within-group.
 Between-group variance pertains to the differences between experimental groups (or conditions).
 This type of variance is of interest when conducting DE studies.
-Within-group variance, on the other hand, is the variance within experimental groups, this is not interesting for DE studies but may contain information useful to CoE.
+Within-group variance, on the other hand, is the variance within experimental groups (among replicates), this is not interesting for DE studies but may contain information useful to CoE.
 It is the excessive between-group variance that leads to the densely connected nature of panel *B* from \@ref(fig:nettype).
@@
 The within-group correlation is lower than the total in both cases, but there is a clear trend in *C*: P- condition in red.
-The within group correlation (r) is higher for the first transcript pair (*C*) than the second (*D*), which makes it possible to discern between genes within a POI or RP and genes that both respond similarly to a treatment (URP).
+The within group correlation (r) is higher for the first transcript pair (*C*) than the second (*D*), which makes it possible to discern between a POI gene or an URP gene.
 Hence, we need to distinguish *total CoE* which is calculated on the total variance and *within-group CoE* where the focus is on the shared variance within experimental groups.
@@
 This means that the within-group variance likely contains valuable information and by averaging out this variance, as is done in differential expression analysis, we are losing information.
-
-Considering the within-group variance increases the strength of coexpression analysis, for example, in experiments where the intervention causes large between-group differences in many genes.
 Not making the distinction between types of variance results in *total CoE* associations driven by the differences between treatment groups as opposed to within group correlations.
@@
 Due to such a strong effect of the experiment, many genes were affected on top of the desired SL genes.
-This masked the within-group correlation between the baits and the spikes (remaining unknown SL genes).
+This masked the within-group correlation between the baits and the targets (remaining unknown SL genes).
 This problematic issue was also highlighted recently by @saccentiWhatCanGo2023 where group sampling assumptions are investigated.
@@
 
-The baits are assumed to be part of the same POI and the remaining uncharacterised spikes (POI or RP genes) are assumed to be numerous.
-Therefore, both the baits and the remainder of the data could be reduced to linear combinations in order to investigate relationships between a set of baits and putative spikes.
+The baits are assumed to be part of the same POI and the remaining uncharacterised targets (POI or RP genes) are assumed to be numerous.
+Therefore, both the baits and the remainder of the data could be reduced to linear combinations in order to investigate relationships between a set of baits and putative targets.
 We introduce a new data analysis method called multivariate ASCA residual analysis (MASCARA).
@@
 The method capitalises on ASCA to account for an experimental design and the multivariate analysis capability of PLS2 to find uncharacterised POI genes using a set of baits.
-Instead of focusing on the variance induced by the experimental design, as is a typical use of ASCA, this variance is removed, and the remaining residual variance is further analysed to investigate the multivariate relationships between baits and the putative spikes in terms of within-group CoE.
+Instead of focusing on the variance induced by the experimental design, as is a typical use of ASCA, this variance is removed, and the remaining residual variance is further analysed to investigate the multivariate relationships between baits and the putative targets in terms of within-group CoE.
 
@@
 
-The problem in the real dataset is that the nutrient stress applied causes a major effect on many other genes besides our baits and spikes.
+The problem in the real dataset is that the nutrient stress applied causes a major effect on many other genes besides our baits and targets .
 This widespread transcriptional regulation can mask the coexpression patterns of interest due to the large number of genes that are upregulated under the stress condition, making it more difficult to detect genes that are responding to a lesser extent, for example the SL pathway.
@@
 
-In $\mathbf{P}_{\beta 1}^T$ the baits, spikes and other DE genes (all genes from POI, RP and URP) along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta 2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model.
+In $\mathbf{P}_{\beta 1}^T$ the baits, spikes (simulated targets; term used throughout simulations) and other DE genes (all genes from POI, RP and URP) along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta 2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model.
 The remaining elements of $\mathbf{P}_{\alpha + \alpha \beta}^T$ and $\mathbf{P}_{\beta}^T$ are populated with zeroes.
@@
 In this simulation we have a set of 4 baits, 12 spike genes and a variable number URP genes.
-The issue to address is the presence of many differential URPs; those that have strong between group correlations but no within group correlation.
+The issue to address is the presence of many differential URPs; those that have strong between group correlation but no within group correlation with the POI.
 We incrementally increase the number of genes in the yellow block and decrease number of white genes with 0 loadings.
@@
 
-1. Initialize \( \mathbf{\tilde{Y}}_0 \) to \( \mathbf{\tilde{Y}} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \).
-2. Iteratively perform the following steps for \( k = 1, 2, \ldots \):
+1. Initialise \( \mathbf{\tilde{Y}}_0 \) to \( \mathbf{\tilde{Y}} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \).
+2. Iteratively perform the following steps for \( k = 0, 1, 2, \ldots \):
    a.
@@
 The variance explained by all effect matrices of the experimental factors and interactions is removed.
-What remains in this \(\hat{\mathbf{E}}\) matrix is individual variance of each plant that is assumed to contain not just random technical variance but some form of structured variance induced by ambient (uncontrolled) effects.
+What remains in this \(\hat{\mathbf{E}}\) matrix is individual variance of each replicate that is assumed to contain not just random technical variance but some form of structured variance induced by ambient (uncontrolled) effects.
 The matrix \(\hat{\mathbf{E}}\) from equation \@ref(eq:ascaresid) is substituted as $\mathbf{Y}$ into PLS2  after being split into baits and remainder as per equation \@ref(eq:plspart).
@@
 Higher values in the target projection equate to stronger positive associations with the center of the baits.
-Large negative values can also be interesting candidates for strong negative associations, although non positive associations are not considered here.
+Large negative values can also be interesting candidates for strong negative associations, although only positive associations are considered here.
 This approach allows a general direction in the PLS space to be defined by multiple baits and assumes that these baits are highly correlated with one another.
@@
 
-To evaluate performance we use log2 geometric mean rank (log2(GMR); $\bar{\gamma}$), defined as:
-
-\begin{equation}
-\begin{aligned}
-\bar{\gamma} = log2((\prod_{j_{spike}=1}^{J_{spike}}\gamma_{j_{spike}})^{1/J_{spike}})
+To evaluate performance we use $log_{2}$ geometric mean rank ($log_{2}(GMR)$; $\bar{\gamma}$), defined as:
+
+\begin{equation}
+\begin{aligned}
+\bar{\gamma} = log_{2}((\prod_{j_{spike}=1}^{J_{spike}}\gamma_{j_{spike}})^{1/J_{spike}})
 \end{aligned}(\#eq:GMR)
@@
 
-Noise_Tests <- readRDS("Noise_Sim_Coexp_23_07_26_11.RDS")
+# Noise_Tests <- readRDS("Noise_Sim_Coexp_23_07_26_11.RDS")  
+Noise_Tests <- readRDS("Noise_Sim_Coexp_24_01_22.RDS")
 
@@
 This reflects the situation expected from the real data example - drastic effect caused by nutrient deficiency, which also activates the SL pathway.
-The POI/RP genes are expected to share some variance that is independent to the experiment - the 4 baits and 12 spikes load on to the structured part of the noise $\bm{T}_S\bm{P}_{S}^T$.
+The POI genes are expected to share some variance that is independent to the experiment, the 4 baits and 12 spikes load on to the structured part of the noise $\bm{T}_S\bm{P}_{S}^T$.
 
@@
 #parameters, 3-30 replicates per condition. 
-Rep_Tests <- readRDS("Noise_Sim_23_07_29_Replicate_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS
+# Rep_Tests <- readRDS("Noise_Sim_23_07_29_Replicate_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS
+
+Rep_Tests <- readRDS("Noise_Sim_24_01_22_Replicate_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS
+
 names(Rep_Tests) <- c("ASCA", "PLS","Correlations","MASCARA")
@@
 These are thought to be carried out by cytochrome p450s.
-The top candidates from MASCARA contain at least three (putative) cytochrome p450 genes that have not yet been biologically investigated in SL biosynthesis, as well as, a methyltransferase and several putative carotenoid associated genes.
+The top candidates from MASCARA contain at least three (putative) cytochrome p450 genes that have not yet been biologically investigated in SL biosynthesis, as well as, a methyltransferase and several putative carotenoid synthesis genes.
 MASCARA also detected genes linked to Gibberellins (GA) preferentially to general phosphate starvation genes.

### 2024-01-26 R/run_first/2.plot_pathway_net_example.Rmd
24_01_26

@@
 ---
-title: "network_plot_work"
+title: "network_plot"
 author: "Fred White"
@@
 
-renv::status()
+# renv::status()
 
@@
 
-VIP <- rep(5,length(V(g)))
+VIP <- rep(10,length(V(g)))
 
@@
 type[-which(type %in% c("POI","RP"))] <- "URP"
+
+temp <- paste0("X_",c(1991:1992,1989,1985))   #501:510
+
+Baits <- names(V(g))
+Baits[-which(Baits %in% temp)] <- "Other"
+Baits[which(Baits %in% temp)] <- "baits"
 
@@
 
+tidynet <- tidynet %>%
+  activate(nodes) %>%
+  mutate(Baits = Baits)
+
+tidynet <- tidynet %>%
+  activate(nodes) %>%
+  mutate(ID = names(V(g)))
+
 colours = c("POI" = "purple", "RP" = "orange", "URP" = "lightgreen")
+
+# shape = c("Baits" = 18, "other" = 1)
 
@@
   geom_node_point(
-    aes(size = VIP, color = type), show.legend=FALSE
+    aes(size = VIP, color = type, shape = Baits), show.legend=FALSE         #CHANGE THIS
   ) +
   geom_node_point(
-    aes(color = type)
+    aes(color = type)    #, shape = Baits
   ) +
   scale_color_manual(values = colours) + 
+    scale_shape_manual(values = c(17,16)) +   #18,20
 
@@
 
+realplot
 ```
@@
 
+#set specific weights in adj matrix   CHANGE THIS
+
 cand_vals <- data[,which(colnames(data) %in% names(V(g)))]
@@
 
+adj[type == "URP", type == "URP"]  <- 1
+adj[type %in% c("POI","RP"), type %in% c("POI","RP")] <- 1
+
+adj[type %in% c("POI","RP"), type == "URP"] <- -1
+adj[type == "URP", type %in% c("POI","RP")] <- -1
+```
+
+```{r}
 g <- graph.adjacency(
@@
 g <- simplify(g, remove.multiple=TRUE, remove.loops=FALSE)
+```
+
+```{r}
 
@@
 colours = c("POI" = "purple", "RP" = "orange", "URP" = "lightgreen")  
-
-edge_colours <- c("+ve" = "grey", "-ve" = "blue")
+# shape = c("Baits" = 18, "other" = 1)
+edge_colours <- c("+ve" = "black", "-ve" = "grey")
 
@@
 badplot <- ggraph(tidynet, layout = coords) + 
-  geom_edge_link(aes(colour = association, alpha = weight), show.legend=FALSE) + 
+  geom_edge_link(aes(colour = association, alpha = weight), show.legend=FALSE) +   #, linetype = "dashed"
   scale_edge_color_manual(values = edge_colours, guide = "none") +
   geom_node_point(
-    aes(size = VIP, color = type), show.legend=FALSE
+    aes(size = VIP, color = type, shape = Baits), show.legend=FALSE
   ) +
@@
   scale_color_manual(values = colours) + 
+      scale_shape_manual(values = c(17,16)) +   #18,20
+  # scale_edge_linetype_manual(values = "dashed") +
+
   theme_void()
@@
 
-png("networktypes.png", height = 360)
+png("../networktypes.png", height = 3600, width = 4800, res = 700)
 realplot + badplot + plot_annotation("Pathway Types", tag_levels = 'A') + plot_layout(guides = "collect")

### 2024-01-26 R/run_first/3.sim1_CE_size.R

@@ -25,70 +25,56 @@ library(mixOmics)
 library(pracma)
 library(pls)
 
-source("../DATA_SIM_FUNCS.R")
+# source("../DATA_SIM_FUNCS.R")
+
+source("DATA_SIM_FUNCS_DEV.R")
+
 F1_plot <- F1_plot_no_plot
 nreps <- 3
 ncands <- 500
 
 Create_Core_DEV <- Create_Core_DEV_2
 
-sPLSr <- function(X, meta, baits, spikes, ncands, TITLE = NULL){
+
+splsR <- function(resids, baits, spikes){
+  #vip
+  spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
+                         resids[,which(colnames(resids) %in% baits)], ncomp = 2)
   
   
-  splsR <- spls(X[,-which(colnames(X) %in% baits)],X[,which(colnames(X) %in% baits)], ncomp = 2)
+  #####
+  y <- colMeans(spls_res$Yloadings)
+  U <- spls_res$projection
   
-  # splsR <- simpls.fit(X[,-which(colnames(X) %in% baits)],X[,which(colnames(X) %in% baits)], ncomp = 2)
+  x <- Yloadings(spls_res)
+  ve_y <- colSums(x^2)/nrow(x)
   
-  sPLS_cands <- cbind.data.frame(abs(splsR$loadings$X) %*% splsR$prop_expl_var$X, splsR$loadings$X)
-  sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = T),]
+  sPLS_cands <- cbind.data.frame(abs(spls_res$projection) %*% ve_y, spls_res$projection)
+  sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = TRUE),]
   
-  F1_scores <- F1_plot(sPLS_cands, spikes, ncands, TITLE = paste0(": ",TITLE))
+  RP <- prod(which(rownames(sPLS_cands) %in% spikes))^(1/length(spikes))
   
-  return(list(F1_scores,sPLS_cands))
-  
+  return(list(RP, sPLS_cands))
 }
 
 
 
-MASCARA4_test <- function(resids,ref, baits, spikes, ncands){
-  
-  Fref <- ref
-  
-  spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], 
[…]

### 2024-01-26 R/run_first/4.sim2_rep.R

@@ -24,7 +24,10 @@ library(caret)
 library(mixOmics)
 library(pracma)
 
-source("../DATA_SIM_FUNCS.R")
+# source("../DATA_SIM_FUNCS.R")
+
+source("DATA_SIM_FUNCS_DEV.R")
+
 
 Create_Core_DEV <- Create_Core_DEV_2
 
@@ -33,63 +36,50 @@ F1_plot <- F1_plot_no_plot
 alphaN <- 2
 betaN <- 4
 ncands <- 500
-# baits <- paste0("X_",c(1989:1996))   #501:510,
-# baits <- paste0("X_",c(1985:1996))   #501:510
 
 
-
-MASCARA4_test <- function(resids,ref, baits, spikes, ncands){
+splsR <- function(resids, baits, spikes){
+  #vip
+  spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
+                         resids[,which(colnames(resids) %in% baits)], ncomp = 2)
   
-  Fref <- ref
-  
-  spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], 
-                   resids[,which(colnames(resids) %in% baits)], all.outputs = T)
-  
-  sPLS_cands <- cbind.data.frame(abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X, spls_res$loadings$X) #  * spls_res$prop_expl_var$X
-  sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = T),]
-  
-  importance <- paste0(round(spls_res$prop_expl_var$X * 100, digits = 2), "%")
-  
-  Fref2 <- Fref[-which(Fref$Feature %in% baits),]
-  Fref2 <- Fref2[match(rownames(sPLS_cands),Fref2$Feature),]
   
   #####
+  y <- colMeans(spls_res$Yloadings)
+  U <- spls_res$projection
   
-  y <- colMeans(spls_res$loadings$Y)
+  x <- Yloadings(spls_res)
+  ve_y <- colSums(x^2)/nrow(x)
   
-  U <- spls_res$loadings$X
+  sPLS_cands <- cbind.data.frame(abs(spls_res$projection) %*% ve_y, spls_res$projection)
+  sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = TRUE),]
   
[…]

### 2024-02-08 R/MASCARA_Tutorial.Rmd
24_02_08

@@
 
-User must supply a dataframe of filtered + transformed data plus associated metadata.
+User must supply a dataframe of filtered + transformed data plus associated metadata, with samples in the rows and genes in the columns.
 
 ```{r}
-
 meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
 colnames(meta) <- c("growth_condition","time","interaction")
-meta <- as.data.frame(apply(meta,2,factor))  #9
+meta <- as.data.frame(apply(meta,2,factor))  
+knitr::kable(head(meta))
+```
 
+Data is assumed to be approximately normally distributed:
+
+```{r setting simulation parameters}
 Experiment_responders <- 30
@@
 set.seed(3)
-# rnd_seed <- abs(rnorm(3, mean = 0.7, sd = 0.5))
 
-baits <- paste0("X_",c(1985:1988))   #501:510
+baits <- paste0("X_",c(1985:1988))  
 spikes <- paste0("X_",c(1989:2000))
@@
                         b_sigma = c(1,0.9), irr_spikes = TRUE, SCORE_SEED = 12, 
-                        Experiment_responders = Experiment_responders, struc_seed = 127)
+                        Experiment_responders = Experiment_responders, struc_seed = 127)[[1]] 
+hist(sim_data)
+```
+
+We calculate an ASCA model on the data to partition the experimental and residual variance:
+
+```{r}
+ar <- ASCA(sim_data, m = meta, Baits = baits, spikes = spikes, ncands = ncands, distance_calc = FALSE, Return_Model = FALSE)
+resids <- ar[[3]]
+```
+
+Before continuing it is a good idea to check that the set of baits are indeed correlated in the residual variance:
+
+```{r}
+cor(resids[,which(colnames(resids) %in% baits)])
+```
+In this case they are, in the case of real data we expect these values to be somewhat lower (values ~ 0.6 are reason enough to continue).
+
+Another sanity check at this stage is to look at the baits residual variance against eachother:
+
+```{r}
+
+plot(resids[,baits[1]], resids[,baits[2]])
 
@@
 
+We can see above that there is a relationship between the examples.
+
+From here we take the residuals and calculate a PLS2 model with a predefined set of baits using the SIMPLS algorithm:
+
 ```{r}
-ar <- ASCA(sim_data[[1]], m = meta, Baits = baits, spikes = spikes, ncands = ncands, distance_calc = FALSE, Return_Model = FALSE)
+
+spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
+                          resids[,which(colnames(resids) %in% baits)], ncomp = 2)
+
 ```
 
+Selecting features with $\mathbf{\bar{q}}$ target projection:
+
 ```{r}
-#classic
-MASCARA_2 <- MASCARA4_test_2(ar[[3]], baits, spikes = spikes)
-which(rownames(MASCARA_2[[2]]) %in% spikes)
+q_bar <- colMeans(spls_res$Yloadings)
+R <- spls_res$projection
+R_TP <- (dot(t(R),q_bar)/dot(q_bar,q_bar))
+
+Candidates <- as.data.frame(R_TP[order(R_TP, decreasing = TRUE)])
 ```
+
+Now we have our ordered candidate list, top of the list represents features that have the strongest positive association with the baits.

### 2024-02-08 R/Paper_markdown.Rmd
24_02_08

@@
 #     address: Department 2, Street, City, State, Zip
+
+## CHANGE THIS   
+#"one comment already, since I am looking at it right now.
+In the readme “~/MASCARA/R” , the repo (clone) most likely doesn’t live there.
+Maybe start from ‘clone this repo’ and then let people set the wd to where they put the repo.
+Or put the snippet from the readme into an r script that you let them open in rstudio to make sure they’re in the right place" AHB
 ```
@@
 Panel A represents hypothetical biological pathways  that are all affected by the experimental design and thus all contain DE genes, here the edges represent real connections between the nodes/genes.
-Panel B highlights the problem with detecting these pathways when dealing with data from designed experiments, edges represent putative connections defined by coexpression analysis; black is desired (within network detection), grey non-desired.")}
+Panel B highlights the problem with detecting these pathways when dealing with data from designed experiments, here edges represent putative connections defined by coexpression analysis; black is desired (within network detection), grey non-desired.")}
 
@@
 
-1. Initialise \( \mathbf{\tilde{Y}}_0 \) to \( \mathbf{\tilde{Y}} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \).
-2. Iteratively perform the following steps for \( k = 0, 1, 2, \ldots \):
-   a.
+<!-- 1. Initialise \( \mathbf{\tilde{Y}}_0 \) to \( \mathbf{\tilde{Y}} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \). -->
+<!-- 2. Iteratively perform the following steps for \( k = 0, 1, 2, \ldots \): -->
+<!--    a.
 Compute the singular value decomposition (SVD) of \( \mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k} \) to obtain the left and right singular vectors; weights \( \mathbf{u}_k \) and \( \mathbf{v}_k \) respectively.
-   b.
+-->
+<!--    b.
 Project \( \mathbf{\tilde{Y}}_{k} \) and \( \mathbf{B}_{k} \) onto the first left and right singular vectors (\( \mathbf{u}_k \), \( \mathbf{v}_k \)) to obtain the scores \( \bm{t}_k = \mathbf{\tilde{Y}}_{k} \mathbf{u}_k \) and \( \bm{w}_k = \mathbf{B}_{k} \mathbf{v}_k \).
-
-Under the conditions of orthogonal scores; $\mathbf{t}_{k}^{T}\mathbf{t}_{k} = 0$ and normalised weights; $\mathbf{u}_{k}^T\mathbf{u}_{k{}} = 1$ and $\mathbf{v}_{k}^T\mathbf{v}_{k{}} = 1$.
-
-PLS2 aims to maximize the covariance between the scores \( \bm{t}_k \) and \( \bm{w}_k \) at each iteration.
+-->
+
+1. For the $1^{st}$ component:
+   a. Compute the left singular vector $\mathbf{u}_1$ of $\mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k}$
+   b. $\mathbf{t}_1 = \mathbf{\tilde{Y}}\mathbf{u}; \mathbf{T} = \left [\mathbf{t}_1\right]$
+   c. $\mathbf{p}_1 = \frac{\mathbf{\tilde{Y}}^T\mathbf{t}^t}{\mathbf{t}^t\mathbf{t}}$
+2. For the following components:
+   a. Compute the left singular vector $\mathbf{u}_k$ of $(\mathbf{I} - \mathbf{P}(\mathbf{P}^T\mathbf{P})^{-1}\mathbf{P})\mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k}$
+   b. $\mathbf{t}_k = \mathbf{\tilde{Y}}\mathbf{u}_k; \mathbf{T} = \left [\mathbf{T}|\mathbf{t}_k\right]$
+   c. $\mathbf{p}_k = \frac{\mathbf{\tilde{Y}}^T\mathbf{t}_{k}^t}{\mathbf{t}_{k}^t\mathbf{t}_{k}}; \mathbf{P} = \left [\mathbf{P}|\mathbf{p}_k\right]$
+
+Under the conditions of orthogonal scores; $\mathbf{t}_{k}^{T}\mathbf{t}_{k} = 0$ and normalised weights; $\mathbf{u}_{k}^T\mathbf{u}_{k{}} = 1$ .
+Loadings for B can be obtained with a least squares step $\mathbf{Q} = (\mathbf{T}^T\mathbf{T})^{-1}\mathbf{T}^T\mathbf{B}$.
+
 In the following simulations and application the number of components calculated by PLS2 is set to 2 to create an interpretable model.
+
+<!-- and $\mathbf{v}_{k}^T\mathbf{v}_{k{}} = 1$ -->
+<!-- PLS2 aims to maximize the covariance between the scores \( \bm{t}_k \) and \( \bm{w}_k \) at each iteration. -->
 
@@
 In ASCA+ we calculate the VIP scores in the latent space of \(\bm{Y}_{\alpha + \alpha\beta}\) as described in @thielASCAAPCAExtensions2017.
-In PLS we use the VE from the latent space of \(\bm{B}\) using the weights ($\mathbf{U}$) from the latent space of \(\bm{Y}\).
+In PLS we use the VE from the latent space of \(\bm{B}\) using the weights ($\mathbf{U}$) from the latent space of the predictor matrix \(\bm{Y}\).
 

### 2024-02-08 R/run_first/2.plot_pathway_net_example.Rmd
24_02_08

@@
 ```{r}
+#sparsify -1 connections
+
+set.seed(22)
+
+# adj[which(adj == -1)] <- adj[which(adj == -1)] * sample(c(1,0,0,0))
+
+# adj[which(adj == -1, arr.ind = T)] <- adj[which(adj == -1,arr.ind = T)] * sample(c(1,0,0,0))
+
+# W <- matrix(sample(c(1,1,0,0,0),length(baits) * length(spikes), replace = T), nrow = length(spikes))
+# 
+# 
+# adj[which(rownames(adj) %in% spikes),which(colnames(adj) %in% baits)] <- adj[which(rownames(adj) %in% spikes),which(colnames(adj) %in% baits)] * W
+
+```
+
+```{r}
 g <- graph.adjacency(
@@
 
-png("../networktypes.png", height = 3600, width = 4800, res = 700)
+# png("../networktypes.png", height = 3600, width = 4800, res = 700)
 realplot + badplot + plot_annotation("Pathway Types", tag_levels = 'A') + plot_layout(guides = "collect")
-dev.off()
+# dev.off()
 

### 2024-02-08 R/Paper_markdown.Rmd
24_02_08_c

@@
 
-# Fred T.G.
-White^[Biosystems Data Analysis Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands, Plant Hormone Biology Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands *], Anna Heintz-Buschart^[Biosystems Data Analysis Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands], Lemeng Dong^[Plant Hormone Biology Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands], Harro J.
-Bouwmeester^[Plant Hormone Biology Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands], Johan A.
-Westerhuis^[Biosystems Data Analysis Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands], Age K.
-Smilde^[Biosystems Data Analysis Group, Swammerdam Institute for Life Sciences, University of Amsterdam, Amsterdam, The Netherlands]
-
-# - name: Fred T.G. White
-#     email: f.t.g.white@uva.nl
-#     affiliation: 
-#       - University of Amsterdam
-#       - Another example
-#     corresponding: alice@example.com
-#   - name: Bob Security
-#     email: bob@example.com
-#     affiliation: 
-#       - Another University
-#       - Some Institute of Technology
-# address:
-#   - code: Univeristy of Amsterdam
-#     address: Department 1, Street, City, State, Zip
-#   - code: Another example
-#     address: Department 2, Street, City, State, Zip
-
-## CHANGE THIS   
-#"one comment already, since I am looking at it right now.
-In the readme “~/MASCARA/R” , the repo (clone) most likely doesn’t live there.
-Maybe start from ‘clone this repo’ and then let people set the wd to where they put the repo.
-Or put the snippet from the readme into an r script that you let them open in rstudio to make sure they’re in the right place" AHB
 ```
@@
 
-<!-- For the rest of the manuscript we take the strigolactones as our example biological pathway. -->
-
 ```{r nettype, fig.cap= c("Pathway types exemplified, POI; pathway of interest, RP; related pathway, URP; unrelated pathway.
@@
 Panel B highlights the problem with detecting these pathways when dealing with data from designed experiments, here edges represent putative connections defined by coexpression analysis; black is desired (within network detection), grey non-desired.")}
-
-#CHANGE caption
 
@@
 
-<!-- \onecolumn ### -->
-
 \End{multicols}
@@
 \twocolumn
-
-<!-- \Begin{multicols}{2} -->
 
@@
 
-<!-- 1. Initialise \( \mathbf{\tilde{Y}}_0 \) to \( \mathbf{\tilde{Y}} \) and \( \mathbf{B}_0 \) to \( \mathbf{B} \). -->
-<!-- 2. Iteratively perform the following steps for \( k = 0, 1, 2, \ldots \): -->
-<!--    a.
-Compute the singular value decomposition (SVD) of \( \mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k} \) to obtain the left and right singular vectors; weights \( \mathbf{u}_k \) and \( \mathbf{v}_k \) respectively.
--->
-<!--    b.
-Project \( \mathbf{\tilde{Y}}_{k} \) and \( \mathbf{B}_{k} \) onto the first left and right singular vectors (\( \mathbf{u}_k \), \( \mathbf{v}_k \)) to obtain the scores \( \bm{t}_k = \mathbf{\tilde{Y}}_{k} \mathbf{u}_k \) and \( \bm{w}_k = \mathbf{B}_{k} \mathbf{v}_k \).
--->
-
 1. For the $1^{st}$ component:
@@
 
-<!-- and $\mathbf{v}_{k}^T\mathbf{v}_{k{}} = 1$ -->
-<!-- PLS2 aims to maximize the covariance between the scores \( \bm{t}_k \) and \( \bm{w}_k \) at each iteration. -->
-
 ### Variable importance in projection (VIP)
@@
 
-# Noise_Tests <- readRDS("Noise_Sim_Coexp_23_07_26_11.RDS")  
 Noise_Tests <- readRDS("Noise_Sim_Coexp_24_01_22.RDS")
@@
 
-<!-- The results from above were obtained from datasets with varying noise, with core aspects such as the underlying structure, the number of replicates, the strength of correlation between variables, and the size of the noise remained fixed.
-With the following, we test the feasibility range of MASCARA with respect to the ratio of 3 key variance parameters namely: -->
-
-<!-- 1. Within group correlation structure -->
-<!-- 2. Between group correlation structure -->
-<!-- 3. Random noise -->
-
-<!-- Here we set a fixed size of within group correlation structure (\(\|\bm{T}_S\bm{P}_{S}^T\|_F = 0.025\)) as well as number of non-pathway DE genes to 30 (yellow bar size \@ref(fig:simOverview)) across all iterations of this test.
-We vary the sizes of the other two variance parameters (*d* and *l*).
-Furthermore we also show the effect of replicate number and demonstrate that MASCARA (along with the other methods) can be applied to data with generic profiles, not limited to the specific structures that were used to demonstrate the differences in coexpression analysis above.
-Here we create 20 datasets with random combined effect and within group correlation structures.
--->
-
 ```{r, echo = F}
@@
 #parameters, 3-30 replicates per condition.
-# Rep_Tests <- readRDS("Noise_Sim_23_07_29_Replicate_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS
-
-Rep_Tests <- readRDS("Noise_Sim_24_01_22_Replicate_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS
+
+Rep_Tests <- readRDS("Noise_Sim_24_01_22_Replicate_Tests.RDS")
 
@@
   scale_color_viridis(discrete = T)+
-  # scale_fill_viridis(discrete = T)+
-
-        # scale_shape_manual(values = c(0:3)) +
     stat_summary() +
@@
 MASCARA accounts for the different types of (experimental and ambient) variance and enables finer scale detection of within group variance structure.
+Code and data for this project are available at https://github.com/BiosystemsDataAnalysis/MASCARA.
 
@@
 
-```{r}
+```{r fig.cap="Feature selection simulation, selectivity ratio; SR, target projection; TP and variable importance in projection (VIP) compareed at 3 levels of noise, indicated in facet titles."}
 Rep_Tests <- readRDS("Noise_Sim_23_07_30_FS_Tests.RDS")  #_3  Noise_Sim_23_06_23_Replicate_Tests.RDS

### 2024-02-09 R/run_first/2.plot_pathway_net_example.Rmd
renv

@@
 
-type <- V(g)$name
+Net <- V(g)$name
 ```
@@
 ```{r}
-id <- c(1:length(type))
+id <- c(1:length(Net))
 
@@
 
-type[which(type %in% baits)] <- "POI"
-type[which(type %in% spikes)] <- "RP"
+Net[which(Net %in% baits)] <- "POI"
+Net[which(Net %in% spikes)] <- "RP"
 
-type[-which(type %in% c("POI","RP"))] <- "URP"
+Net[-which(Net %in% c("POI","RP"))] <- "URP"
 
@@
 
-Baits <- names(V(g))
-Baits[-which(Baits %in% temp)] <- "Other"
-Baits[which(Baits %in% temp)] <- "baits"
+Gene <- names(V(g))
+Gene[-which(Gene %in% temp)] <- "Other"
+Gene[which(Gene %in% temp)] <- "Baits"
 
@@
   activate(nodes) %>%
-  mutate(type = type)
+  mutate(Net = Net)
 
@@
   activate(nodes) %>%
-  mutate(Baits = Baits)
+  mutate(Gene = Gene)
 
@@
   geom_node_point(
-    aes(size = VIP, color = type, shape = Baits), show.legend=FALSE         #CHANGE THIS
+    aes(size = VIP, color = Net, shape = Gene), show.legend=FALSE        #CHANGE THIS
   ) +
   geom_node_point(
-    aes(color = type)    #, shape = Baits
+    aes(color = Net, shape = Gene)    #, shape = Baits
   ) +
@@
 
-adj[type == "URP", type == "URP"]  <- 1
-adj[type %in% c("POI","RP"), type %in% c("POI","RP")] <- 1
+adj[Net == "URP", Net == "URP"]  <- 1
+adj[Net %in% c("POI","RP"), Net %in% c("POI","RP")] <- 1
 
-adj[type %in% c("POI","RP"), type == "URP"] <- -1
-adj[type == "URP", type %in% c("POI","RP")] <- -1
+adj[Net %in% c("POI","RP"), Net == "URP"] <- -1
+adj[Net == "URP", Net %in% c("POI","RP")] <- -1
 ```
@@
 
-#define correctly ordered type type
-type <- V(g)$name
+#define correctly ordered type Net
+Net <- V(g)$name
 
@@
   geom_node_point(
-    aes(size = VIP, color = type, shape = Baits), show.legend=FALSE
+    aes(size = VIP, color = Net, shape = Gene), show.legend=FALSE
   ) +
   geom_node_point(
-    aes(color = type)
+    aes(color = Net, shape = Gene)
   ) +

### 2024-02-09 R/23_02_02_real_data_rice.Rmd
24_02_09
TIFFs and formatting

@@
 
-row_ann <- data.frame("Gene" = rep(c("SL","URP"),each = 9))
+row_ann <- data.frame("Pathway" = rep(c("SL","URP"),each = 9))
 rownames(row_ann) <-  c(sls,URPS)         #c(pathway,rownames(ASCA_cands3)[1:9])
@@
                     time = c("1" = "#fde725","3" = "#35b779","7" = "#31688e","8" = "#440154"),
-                    Gene = c("SL" = "purple", "URP" = "lightgreen"))
+                    Pathway = c("SL" = "purple", "URP" = "lightgreen"))
 
@@
 
-png("MASCARA_heatmap.png", res = 150, height = 960, width = 960)
+tiff("Fig3.tif", res = 150, height = 960, width = 960)   #MASCARA_heatmap.png
 grid::grid.newpage()
@@
 
-png("rice_s_l.png", width = 960, height = 480)
+tiff("Fig4.tif", width = 960, height = 480)   #rice_s_l.png
 (s + l)
@@
 
-png("within_between_example_2.png", res = 150, height = 960, width = 960)
+tiff("Fig2.tif", res = 150, height = 960, width = 960)
 (within + between)/(within_r + between_r) + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A','1')) #+ plot_layout(heights = unit(c(5), c("cm")), widths = c(5))

### 2024-02-09 R/Paper_markdown.Rmd
24_02_09
TIFFs and formatting

@@
 Panel A represents hypothetical biological pathways  that are all affected by the experimental design and thus all contain DE genes, here the edges represent real connections between the nodes/genes.
-Panel B highlights the problem with detecting these pathways when dealing with data from designed experiments, here edges represent putative connections defined by coexpression analysis; black is desired (within network detection), grey non-desired.")}
-
-a <- knitr::include_graphics("networktypes.png")
+Panel B highlights the problem with detecting these pathways when dealing with data from designed experiments, here edges represent putative connections defined by coexpression analysis; black is desired (within network detection), grey is non-desired.")}
+
+a <- knitr::include_graphics("Fig1.tif")
 a
@@
 
-a <- knitr::include_graphics("within_between_example_2.png")
+a <- knitr::include_graphics("Fig2.tif")   
 a
@@
 
-heat2 <- knitr::include_graphics("MASCARA_heatmap.png")
+heat2 <- knitr::include_graphics("Fig3.tif")
 heat2
@@
 
-```{r ASCA-1, echo = F, out.width = '100%', fig.cap=c("Scores and loadings of the ASCA model of the combined effect (condition + condition:time interaction), model indicates samples and genes relative positions in the designed variance.
-*g_c* indicates growth condition, *time* indicates days after starvation induction.")}
-knitr::include_graphics("rice_s_l.png")
+```{r ASCA-1, echo = F, out.width = '100%', fig.cap=c("Scores and loadings of the ASCA model of the combined effect (condition + condition:time interaction).
+Scores: *g_c* indicates growth condition, *time* indicates days after starvation induction.
+Loadings: POI; SL genes and URP; other highly differential genes (indicated in purple and green respectively) show upregulation in the P- condition compared  to P+.")}
+knitr::include_graphics("Fig4.tif")
 ```
@@
 
-The problem in the real dataset is that the nutrient stress applied causes a major effect on many other genes besides our baits and targets .
+The problem in the real dataset is that the nutrient stress applied causes a major effect on many other genes besides our baits and targets.
 This widespread transcriptional regulation can mask the coexpression patterns of interest due to the large number of genes that are upregulated under the stress condition, making it more difficult to detect genes that are responding to a lesser extent, for example the SL pathway.
 
-The simulations are built upon the results of the ASCA analysis of the combined effect of growth condition and the interaction of growth condition (g_c) and time.
-As can be seen in Figure \@ref(fig:ASCA-1) there is clear separation between samples due to the experimental factors growth condition (g_c) and time.
+The simulations are built upon the results of the ASCA analysis of the combined effect of growth condition (g_c) and the interaction of growth condition and time.
+As can be seen in Figure \@ref(fig:ASCA-1) there is clear separation between samples due to the experimental factors growth condition and time.
 The variation between the replicates is only small.
 The green loadings of URP genes are strongly related to the experimental variation due to g_c.
-The simulated data is created by simulating experimental variation consisting of the main time effect $beta$ and the combined effect ($\alpha + \alpha\beta$), and the Ambient/Residual variation which exists of the structured ambient variance ($S$) and residual variation ($E$) representing measurement error, modeled with normally distributed random values $e_{ij} = \mathcal{N}(0,\sigma^2)$.
+The simulated data is created by simulating experimental variation consisting of the main time effect $\beta$ and the combined effect ($\alpha + \alpha\beta$), and the Ambient/Residual variation which exists of the structured ambient variance ($\mathbf{S}$) and residual variation ($\mathbf{E}$) representing measurement error, modeled with normally distributed random values $e_{ij} = \mathcal{N}(0,\sigma^2)$.
 
@@
 The values in $\mathbf{T}_{S}$ are obtained from a normal distribution with mean 0 and variance 1, i.e.
-$\mathbf{T}_{S_{ik}} \sim \mathcal{N}(0,1)$.
+$\mathbf{t}_{S_{ik}} \sim \mathcal{N}(0,1)$.
 The $\mathbf{T}$ matrices represent the simulated scores for their respective factors.
@@
 The green block represents the bait genes of POI.
-The black genes represent “unknown” genes from POI or RP.
+The black genes represent “unknown” target genes from POI or RP.
 Yellow and purple represent genes from URP that are affected by the experimental variation in different ways.
@@
 
-In $\mathbf{P}_{\beta 1}^T$ the baits, spikes (simulated targets; term used throughout simulations) and other DE genes (all genes from POI, RP and URP) along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta 2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model.
+In $\mathbf{P}_{\beta 1}^T$ the baits, targets and other DE genes (all genes from POI, RP and URP) along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta 2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model.
 The remaining elements of $\mathbf{P}_{\alpha + \alpha \beta}^T$ and $\mathbf{P}_{\beta}^T$ are populated with zeroes.
-$\mathbf{P}_S$ only has values for the baits (green) and spike (black) genes.
-The loading values for the baits and spikes are obtained as $p_{S_jk}^{baits,spikes} \sim \mathcal{N}(0.5,0.2)$, while the loadings for the other genes are 0.
+$\mathbf{P}_S$ only has values for the baits (green) and target (black) genes.
+The loading values for the baits and targets are obtained as $p_{S_jk}^{baits,targets} \sim \mathcal{N}(0.5,0.2)$, while the loadings for the other genes are 0.
 
@@
 ```{r simOverview, echo=FALSE, fig.cap="Coexpression simulation, simplified overview.
-Colours correspond to the same genes in the different loading vectors (*P*).
-Black; baits (known POI genes), green; spikes (unknown POI genes), yellow; DE URP genes, purple DE URP genes with a different profile.
-In the score vectors (T); Blues, reds, greys and golds represent experimental conditions, oranges and pinks; time.
+Colours correspond to the same genes in the different loading vectors ($\\mathbf{P}$).
+Black; baits (known POI genes), green; targets (unknown POI genes), yellow; DE URP genes, purple DE URP genes with a different profile.
+In the score vectors ($\\mathbf{T}$); Blues, reds, greys and golds represent experimental conditions, oranges and pinks; time.
 Each of the textures in the ambient/residual part represent a distinct normally distributed set of random values.", out.width = '100%'}
-knitr::include_graphics("Sim_SL_diagram_simplified.png")
+knitr::include_graphics("Fig5.tif")
 ```
@@
 
-In this simulation we have a set of 4 baits, 12 spike genes and a variable number URP genes.
+In this simulation we have a set of 4 baits, 12 target genes and a variable number URP genes.
 The issue to address is the presence of many differential URPs; those that have strong between group correlation but no within group correlation with the POI.
@@
 ```{r, echo = F}
-#create "ref" for feature spike information
+#create "ref" for feature target information
 
@@
 
-Pearson's correlation coefficient $r_{b\tilde{j}}$ is calculated between all bait genes (b = 1..B) and all other genes ($\tilde{j} = 1...\tilde{J}$).
-Ranks $\tilde{r}_{b\tilde{j}}$ are assigned for each $r_{b\tilde{j}}$ such that the highest correlation for each bait gene b receives the lowest rank.
-These ranks are averaged across all baits B (see supplementary equation \@ref(eq:rankcor)).
+Pearson's correlation coefficient $r_{b\tilde{j}}$ is calculated between all bait genes $b = 1..B$ and all other genes ($\tilde{j} = 1...\tilde{J}$).
+Ranks $\tilde{r}_{b\tilde{j}}$ are assigned for each $r_{b\tilde{j}}$ such that the highest correlation for each bait gene $b$ receives the lowest rank.
+These ranks are averaged across all baits $B$ (see supplementary equation \@ref(eq:rankcor)).
 
@@
 1. For the $1^{st}$ component:
-   a. Compute the left singular vector $\mathbf{u}_1$ of $\mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k}$
+   a. Compute the left singular vector $\mathbf{u}_1$ of $\mathbf{\tilde{Y}}^{T} \mathbf{B}$
    b. $\mathbf{t}_1 = \mathbf{\tilde{Y}}\mathbf{u}; \mathbf{T} = \left [\mathbf{t}_1\right]$
-   c. $\mathbf{p}_1 = \frac{\mathbf{\tilde{Y}}^T\mathbf{t}^t}{\mathbf{t}^t\mathbf{t}}$
-2. For the following components:
-   a. Compute the left singular vector $\mathbf{u}_k$ of $(\mathbf{I} - \mathbf{P}(\mathbf{P}^T\mathbf{P})^{-1}\mathbf{P})\mathbf{\tilde{Y}}_{k}^{T} \mathbf{B}_{k}$
+   c. $\mathbf{p}_1 = \frac{\mathbf{\tilde{Y}}^T\mathbf{t}^t}{\mathbf{t}^t\mathbf{t}}; \mathbf{P} = \left [\mathbf{p}_1\right]$
+2. For the following components $k = 2...K$:
+   a. Compute the left singular vector $\mathbf{u}_k$ of $(\mathbf{I} - \mathbf{P}(\mathbf{P}^T\mathbf{P})^{-1}\mathbf{P})\mathbf{\tilde{Y}}^{T} \mathbf{B}$
    b. $\mathbf{t}_k = \mathbf{\tilde{Y}}\mathbf{u}_k; \mathbf{T} = \left [\mathbf{T}|\mathbf{t}_k\right]$
@@
 
-Under the conditions of orthogonal scores; $\mathbf{t}_{k}^{T}\mathbf{t}_{k} = 0$ and normalised weights; $\mathbf{u}_{k}^T\mathbf{u}_{k{}} = 1$ .
-Loadings for B can be obtained with a least squares step $\mathbf{Q} = (\mathbf{T}^T\mathbf{T})^{-1}\mathbf{T}^T\mathbf{B}$.
+Under the conditions of orthogonal scores; $\mathbf{t}_{k}^{T}\mathbf{t}_{k+1} = 0$ and normalised weights; $\mathbf{u}_{k}^T\mathbf{u}_{k{}} = 1$ .
+Loadings for $\mathbf{B}$ can be obtained with a least squares step $\mathbf{Q} = (\mathbf{T}^T\mathbf{T})^{-1}\mathbf{T}^T\mathbf{B}$.
 
@@
 
-Where k=1...K represents the number of components in the respective models, j indexes a gene,  $H$ is substituted for the loadings $\mathbf{P}_{\alpha + \alpha\beta}^T$ or weights $\mathbf{U}$ (for ASCA+ and PLS respectively) and VE is the variance explained, where $VE_{k}$ is the amount of variance explained by component k and $VE_{total}$ is the variance explained by the whole model.
+Where $k=1...K$ represents the number of components in the respective models, $j$ indexes a gene,  $H$ is substituted for the loadings $\mathbf{P}_{\alpha + \alpha\beta}^T$ or weights $\mathbf{U}$ (for ASCA+ and PLS respectively) and VE is the variance explained, where $VE_{k}$ is the amount of variance explained by component k and $VE_{total}$ is the variance explained by the whole model.
 In ASCA+ we calculate the VIP scores in the latent space of \(\bm{Y}_{\alpha + \alpha\beta}\) as described in @thielASCAAPCAExtensions2017.
@@
 \begin{aligned}
-\bar{\gamma} = log_{2}((\prod_{j_{spike}=1}^{J_{spike}}\gamma_{j_{spike}})^{1/J_{spike}})
+\bar{\gamma} = log_{2}((\prod_{j_{target}=1}^{J_{target}}\gamma_{j_{target}})^{1/J_{target}})
 \end{aligned}(\#eq:GMR)
@@
 
-for $J_{spike}$ spikes. A lower $\bar{\gamma}$ indicates better performance.
+for $J_{target}$ targets. A lower $\bar{\gamma}$ indicates better performance.
 
@@
 
-For this simulation we have some experimental variance for the combined effect and time ($\bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T$ and $\bm{T}_{\beta}\bm{P}_{\beta}^T$ respectively) a fixed amount of random noise \(\bm{E}\) and a fixed amount of structured ambient/residual variance $\bm{T}_S\bm{P}_{S}^T$, within which our 4 baits and 12 spikes share a correlation structure.
+For this simulation we have some experimental variance for the combined effect and time ($\bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T$ and $\bm{T}_{\beta}\bm{P}_{\beta}^T$ respectively) a fixed amount of random noise \(\bm{E}\) and a fixed amount of structured ambient/residual variance $\bm{T}_S\bm{P}_{S}^T$, within which our 4 baits and 12 targets share a correlation structure.
 This reflects the situation expected from the real data example - drastic effect caused by nutrient deficiency, which also activates the SL pathway.
-The POI genes are expected to share some variance that is independent to the experiment, the 4 baits and 12 spikes load on to the structured part of the noise $\bm{T}_S\bm{P}_{S}^T$.
+The POI genes are expected to share some variance that is independent to the experiment, the 4 baits and 12 targets load on to the structured part of the noise $\bm{T}_S\bm{P}_{S}^T$.
 
@@
 
+# tiff("Fig6.tif", res = 150, height = 960, width = 960)
+# wrap_plots(plots) + plot_layout(guides = "collect")
+# dev.off()
 ```
@@
 
+# tiff("Fig7.tif")
 # ps1 <- grid.arrange(ps, left = "log2(GMR)", bottom = "Replicate Number                     ")
-
-a <- knitr::include_graphics("replication_res.png")
+# dev.off()
+
+a <- knitr::include_graphics("Fig7.tif")
 a
@@
 It was expected that all geometric mean ranks converge towards a minimum with increasing numbers of replicates, see \@ref(fig:repres).
-Perfect detection of 12 spikes in top 12 candidates is achieved with a log2 geometric mean rank of 2.4, MASCARA was able to achieve this, however, the other methods were perturbed by the presence of the 30 differential URP genes.
+Perfect detection of 12 targets in top 12 candidates is achieved with a log2 geometric mean rank of 2.4, MASCARA was able to achieve this, however, the other methods were perturbed by the presence of the 30 differential URP genes.
 With more replicates MASCARA always outperforms competitors.
@@
 In the real data example more than 10% of of the total number of genes are differentially expressed.
-As shown in simulation 1 (Figure \@ref(fig:tc)) higher numbers of DE URP genes results in higher log2 geometric mean rank (log2(GMR)) of the spikes for correlations, ASCA and PLS.
+As shown in simulation 1 (Figure \@ref(fig:tc)) higher numbers of DE URP genes results in higher log2 geometric mean rank (log2(GMR)) of the targets for correlations, ASCA and PLS.
 MASCARA is designed specifically to mitigate the effect of large numbers of DE URP genes.
@@
 dens
+
+# tiff("Fig8.tif")
+# dens
+# dev.off()
+
 ```
@@
 
-a <- knitr::include_graphics("within_between_example.png")
+# tiff("Fig9.tif")
+# PLOT
+# dev.off()
+
+a <- knitr::include_graphics("Fig9.tiff")
 a
@@
 Here we have similar but reduced simulation set up to the replicate test section; 3 levels of unstructured noise; matrix F-norm sizes (as indicated in the panels).
-In each of the 3 tests the $\mathbf{T}_S$ is randomised 30 times resulting in different correlation structures between baits and spikes.
+In each of the 3 tests the $\mathbf{T}_S$ is randomised 30 times resulting in different correlation structures between baits and targets.
 Target projection using $\mathbf{\bar{q}}$ significantly outperforms the conventional approaches.

### 2024-02-09 R/run_first/2.plot_pathway_net_example.Rmd
24_02_09
TIFFs and formatting

@@
 
-# png("../networktypes.png", height = 3600, width = 4800, res = 700)
+tiff("../Fig1.tif", height = 3600, width = 4800, res = 700)
 realplot + badplot + plot_annotation("Pathway Types", tag_levels = 'A') + plot_layout(guides = "collect")
-# dev.off()
+dev.off()
 

### 2025-01-20 R/23_02_02_real_data_rice.Rmd
25_01_20
MILP arc consistency KEGG network

@@
 ```{r, echo = FALSE, include=FALSE}
+# getwd()
 library(devtools)
 library(ggplot2)
-library(MetStaT)
 library(gASCA)
@@
 library(doParallel)    
-library(MUVR) 
 library(pheatmap)
-library(caret)
 library(data.table)
 library(plotly)
-library(mixOmics)
-library(DiscriMiner)
+library(pls)
+library(pracma)
+
 ```
@@
 ASCA_cands2 <- get_ASCA_cands2(a_c.PCD)
+
+ac_top <- ASCA_cands2[1:(nrow(ASCA_cands2)/2),]
+
 ASCA_cands2 <- ASCA_cands2[which(ASCA_cands2[,2] < 0),]
@@
 ```
+
+```{r}
+
+spikes <- sls[-which(sls %in% baits)]
+resids <- res_ASCAplus_nb$residuals[,rownames(ac_top)]
+
+spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
+                          resids[,which(colnames(resids) %in% baits)], ncomp = 2)
+
+  #####
+  y <- colMeans(spls_res$Yloadings)
+  U <- spls_res$projection
+  UTP <- (dot(t(U),y)/dot(y,y))
+
+  sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])
+
+```
+
+```{r}
+library(riceidconverter)
+
+X <- X[,which(colnames(X) %in% colnames(resids))]
+
+symbol <- RiceIDConvert(unique(colnames(X)),'TRANSCRIPTID',toType = 'SYMBOL')
+symbol$SYMBOL <- gsub("LOC","osa:",symbol$SYMBOL)
+symbol <- symbol[grep("^osa",symbol$SYMBOL),]
+# KOIDs <- read.table("https://rest.kegg.jp/link/osa/ko")
+path <- read.table("https://rest.kegg.jp/link/osa/pathway")
+# module <- read.table("https://rest.kegg.jp/link/osa/module")
+
+colnames(path) <- c("PATH","SYMBOL")
+merged <- merge(symbol, path, by = "SYMBOL")
+
+pnum <- as.data.frame(table(merged$PATH))
+pnum <- pnum[-which(pnum$Freq < 4),]
+
+merged <- merged[which(merged$PATH %in% pnum$Var1),]
+
+if (!requireNamespace("KEGGREST", quietly = TRUE)) {
+  install.packages("BiocManager")
+  BiocManager::install("KEGGREST")
+}
+
+library(KEGGREST)
+
+kegg_ids <- unique(gsub("path:","",merged$PATH))
+
+# - STUPID FUNCTION BELOW -  ??
+
+# Function to get pathway description for each ID
+get_kegg_description <- function(kegg_id) {
+  pathway_info <- keggGet(kegg_id)  # Query KEGG
+  if (!is.null(pathway_info[[1]])) {
+    return(pathway_info[[1]]$NAME)  # Extract the name/description
+  } else {
+    return(NA)  # Return NA if the ID is invalid
+  }
+}
+
+# Apply the function to all KEGG IDs
+kegg_descriptions <- sapply(kegg_ids, get_kegg_description)
+
+# Combine IDs and descriptions into a data frame
+kegg_mapping <- data.frame(
+  KEGG_ID = kegg_ids,
+  Description = kegg_descriptions,
+  stringsAsFactors = FALSE
+)
+
+# Print the mapping
+print(kegg_mapping)
+
+kegg_DESC <- t(do.call(cbind,kegg_descriptions))
+
+X <- X[,which(colnames(X) %in% merged$TRANSCRIPTID)]       #from ~25k to 4k IDs
+
+cands <- sPLS_bait_cands[which(rownames(sPLS_bait_cands) %in% colnames(X)),, drop = FALSE]
+cands <- cbind(rownames(cands), cands)
+colnames(cands) <- c("TRANSCRIPTID","UTP")
+
+cands <- merge(cands,merged, by = "TRANSCRIPTID")
+
+cands <- cands[order(cands$UTP, decreasing = TRUE),]
+
+kegg_DESC[gsub("path:","",as.data.frame(table(cands$PATH[c(nrow(cands):(nrow(cands) - 50))]))[,1]),]
+
+kegg_DESC[gsub("path:","",as.data.frame(table(cands$PATH[1:50]))[,1]),]
+
+resids <- resids[,which(colnames(resids) %in% colnames(X))]
+
+res_cor <- cor(resids)
+```
+```{r}
+
+diag(res_cor) <- NA
+
+threshold <- quantile(res_cor, 0.999, na.rm = TRUE)
+
+inds <- which(res_cor > threshold, arr.ind = TRUE)
+
+cor_vals <- res_cor[inds]
+
+results <- data.frame(
+  Row = inds[, 1],
+  Column = inds[, 2],
+  Correlation = cor_vals
+)
+
+if (!is.null(rownames(res_cor))) {
+  results$Variable1 <- rownames(res_cor)[results$Row]
+  results$Variable2 <- colnames(res_cor)[results$Column]
+}
+
+results <- results[results$Row < results$Column, ]
+results <- results[order(-results$Correlation), ]
+
+length(unique(c(results$Variable1, results$Variable2)))
+
+res1 <- merge(results,merged, by.x = "Variable1", by.y = "TRANSCRIPTID")
+res1 <- merge(res1,merged, by.x = "Variable2", by.y = "TRANSCRIPTID")
+res1 <- res1[which(res1$`PATH.x` == res1$`PATH.y`),]
+
+cands <- cands[order(cands$UTP, decreasing = TRUE),]
+
+kegg_DESC[gsub("path:","",as.data.frame(table(res1$`PATH.x`))[,1]),]
+
+same_path <- as.data.frame(table(res1$`PATH.x`))
+same_path <- same_path[which(same_path[,2] > 5 & same_path[,2] < 50),]
+
+kegg_DESC[gsub("path:","",same_path[,1]),]
+
+saveRDS(same_path, "path_example.RDS")
+
+```

### 2025-01-20 R/Caldana_application.Rmd
25_01_20
MILP arc consistency KEGG network

---
title: "Caldana_metabolomics_application"
author: "Fred White"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
# library(readxl)
library(openxlsx)
library(ggplot2)
library(patchwork)
library(gASCA)
library(pls)
library(pracma)
library(pheatmap)

source("DATA_SIM_FUNCS.R")
source("MASCARA_FUNCS.R")
```

```{r}
data <- as.data.frame(read.xlsx("Data_Caldana/Caldana-et-al_Normalized-metabolic-data.xlsx"))

```

```{r}
# meta <- cbind.data.frame(colnames(data), t(data[1:2,]))
# meta <- meta[-1,]

# meta$Temperature <- gsub("-.*","",meta$Temperature)
# 
# meta$Light <- gsub(".*-","",meta$Light)

meta <- read.table("Caldana_meta.csv", sep = ",", header = T)
meta$ID <- paste(meta$Temperature, meta$Light, meta$Time, meta$Replicate, sep = "_")
```

```{r, warning=FALSE}
data <- data[-c(1,2),]
Metabolites <- data[,1]
data <- (apply(data[,-c(1)],2, as.numeric))
rownames(data) <- Metabolites

colnames(data) <- meta$ID

data[which(is.na(data), arr.ind = T)] <- 0
```

```{r}
#filtering?

```

```{r}
data <- log2(data + 1)

# data <- data[-grep("Similar.to.Adenine..2TMS",rownames(data)),]
# Metabolites <- Metabolites[-grep("Similar.to.Adenine..2TMS",Metabolites)]
# data <- data[-grep("Similar.to.Adenine..2TMS",rownames(data)),]
# Metabolites <- Metabolites[-grep("Similar.to.Adenine..2TMS",Metabolites)]
```

```{r}
PCA <- prcomp(t(data))#, scale = F, center = F)

A <- PCA$x
B <- PCA$rotation

df <- cbind.data.frame(meta[,c("Temperature","Light","Time")], A)

df$Light <- factor(df$Light, levels = c("D","LL","L","HL"))
# colnames(df)[-which(colnames(df) %in% colnames(meta))] <- paste0("PC",colnames(df)[-which(colnames(df) %in% colnames(meta))])

# pdf(paste0(Sys.Date(),"_feature_loadings_project_full2.pdf"))
scores <- ggplot(df, aes(x = PC1,y = PC2, colour = Temperature, shape = (Light), size = Time)) +
  geom_point(alpha = 0.5) +
  ggtitle("Scores") +
  theme_bw()

#CHANGE

df <- cbind.data.frame(Metabolites,B)

# do.call(c,lapply(Data,colnames))

# pdf(paste0(Sys.Date(),"_feature_loadings_project_full2.pdf"))
loadings <- ggplot(df, aes(x = PC1,y = PC2, alpha = 0.5)) +
  geom_point(alpha = 0.5) + 
  scale_color_manual(values = colours) +
  ggtitle("Loadings") +
  theme_bw()

scores + loadings  + plot_annotation(title = "PCA")

```

```{r}
# #PERMANOVA
# options(knitr.kable.NA = '')
# library(vegan)
# #calculate euclidean distance between samples
# #this matrix should be of the dimensions: number of samples * number of samples
# 
# dist_matrix <- dist(t(data))
# 
# 
# 
# permanova_re <- adonis2(dist_matrix ~ Temperature * Light * Time, 
#                             meta[,c("Temperature","Light","Time")], permutations = 999, method = "euclidean", 
#                             sqrt.dist = FALSE, add = FALSE, by = "terms", 
#                             parallel = 4)
# 
# 
# 
# kable(permanova_re)
#

```

We calculate an ASCA model on the data to partition the experimental and residual variance:

```{r}
ar <- ASCA_decompose(d = apply(meta[,-c(4,5)],2,as.factor), x = t(data), f = "Temperature * Light * Time")

#unexpectedly different from PERMANOVA R2
norm(ar$residuals, type = "F")

norm(ar$decomposition$Temperature, type = "F")

resids <- ar$residuals
```

```{r}
#select some baits (based on PESCA model ?..)

# cors <-

p1 <- pheatmap(cov(resids), show_colnames = F, fontsize_row = 8)
```

```{r}

p2 <- pheatmap(cov(t(data[p1$tree_row$order,])), show_colnames = F, fontsize_row = 8, cluster_rows = F, cluster_cols = F)

# p2 <- pheatmap(cor(t(data[])), show_colnames = F, fontsize_row = 8)

```

Observe the clustering of certain amino acids with high residual correlations, for example Lysine, Tyrosine, Leucine and Isoleucine.

```{r}
baits <- c("Lysine","Tyrosine","Leucine")

```

```{r}
Y_hat <-  t(data) - resids

# time plays a larger role in caldana data -
# Y_hat_nt <-  t(data) - resids - ar$decomposition$Time

Y_hat_r <- cov(t(data))

E_r <- cov(resids)

# plot(Y_hat_r[upper.tri(Y_hat_r)], E_r[upper.tri(E_r)], xlab = "Between Correlations", ylab = "Within Correlations")
# 
# plot(Y_hat_r[], E_r[], xlab = "Between Correlations", ylab = "Within Correlations")

#alpha by norm(resid)/norm(data-resid)

#shape by quantiles below

ratio <- resids^2/Y_hat^2

#make this signless ?
ratio2 <- abs(E_r)/abs(Y_hat_r)

```

```{r}
library(reshape2)

df <- cbind.data.frame(melt(Y_hat_r), melt(E_r)[,3])

colnames(df)[c(3:4)] <- c("Total","Within")

df$Ratio <- abs(df$Within)/abs(df$Total)

df$Ratio2 <- abs(df$Total)/abs(df$Within)

df <- df[-which(df$Var1 == df$Var2),]

#subset higher within examples
df$Covariance <- df$Within

#subset higher Total examples
df$Covariance[which(abs(df$Total) > quantile(abs(df$Total),0.99))] <- "Total"

df$Covariance[which(abs(df$Within) > quantile(abs(df$Within),0.99))] <- "Within"

df$Covariance[-which(df$Covariance %in% c("Within","Total"))] <- "none"

```

```{r}
ggplot(df, aes(x = Total, y = Within)) +
  geom_point(aes(alpha = 0.1)) +
  geom_point(data = df[which(df$Covariance %in% c("Within","Total")),],
aes(alpha = 1, size = 1.2, colour = Covariance)) +
  scale_color_manual(values = c("lightgreen","purple")) + 
     # geom_abline(intercept = 0,slope = 1)
  guides(alpha = "none", size = "none") +
  theme_bw()

```

```{r}
head(df[order(abs(df$Within), decreasing = T),],20)
head(df[order(abs(df$Total), decreasing = T),],20)
df[which(df$Covariance == "Total"),]

```

```{r}
which(ratio2 > quantile(ratio2,0.999),arr.ind = T)
which(ratio2 < quantile(ratio2,0.001),arr.ind = T)

#check sign or signless ratio

#make above plot but make the high ratio standout
```

```{r}
#MASCARA PLOTS

```

Before continuing it is a good idea to check that the set of baits are indeed correlated in the residual variance:

```{r}
cor(resids[,which(colnames(resids) %in% baits)])
```
In this case they are, in the case of real data we expect these values to be somewhat lower (values ~ 0.5 are reason enough to continue).

Another sanity check at this stage is to look at the baits residual variance against each other:

```{r}

plot(resids[,baits[1]], resids[,baits[2]])

```

We can see above that there is a relationship between the examples.

From here we take the residuals and calculate a PLS2 model with a predefined set of baits using the SIMPLS algorithm:

```{r}

spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
                          resids[,which(colnames(resids) %in% baits)], ncomp = 2)

```

Selecting features with $\mathbf{\bar{q}}$ target projection:

```{r}
q_bar <- colMeans(spls_res$Yloadings)
R <- spls_res$projection
R_TP <- (dot(t(R),q_bar)/dot(q_bar,q_bar))

Candidates <- as.data.frame(R_TP[order(R_TP, decreasing = TRUE)])
```

Now we have our ordered candidate list, top of the list represents features that have the strongest positive association with the baits.

### 2025-01-20 R/MASCARA_Tutorial.Rmd
24_02_08_b
24_02_23
Package?
25_01_20
MILP arc consistency KEGG network

@@
 library(doParallel)    
-library(MUVR) 
+# library(MUVR) 
 library(pheatmap)
@@
 library(plotly)
-library(mixOmics)
-library(DiscriMiner)
+# library(mixOmics)
+# library(DiscriMiner)
 library(pracma)
@@
 ```{r}
-ar <- ASCA(sim_data, m = meta, Baits = baits, spikes = spikes, ncands = ncands, distance_calc = FALSE, Return_Model = FALSE)
-resids <- ar[[3]]
+ar <- ASCA_decompose(d = meta[,1:2], x = sim_data, f = "growth_condition + time + growth_condition:time")
+resids <- ar$residuals
 ```

### 2025-01-20 R/MASCARA_b_s_sim.Rmd
25_01_20
MILP arc consistency KEGG network

---
title: "sim_b_s"
author: "Fred White"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r , echo=FALSE, include=FALSE}
source("DATA_SIM_FUNCS.R")
source("MASCARA_FUNCS.R")
```

```{r, echo = FALSE, include=FALSE}
library(devtools)
library(ggplot2)
# library(MetStaT)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)
# library(MUVR)
library(pheatmap)
# library(caret)
# library(mixOmics)
library(cowplot)
library(reticulate)
library(biomformat)
# library(DiscriMiner)
library(viridis)
library(directlabels)
library(ggpmisc)
library(DescTools)
library(data.table)
library(plotly)
# library(ggbreak)
# library(rhmmer)
library(pracma)
library(pls)
```

```{r,echo = F}

# Noise_Tests <- readRDS("Noise_Sim_Coexp_24_01_22.RDS")
# getwd()
Noise_Tests <- readRDS("Noise_Sim_b_s_24_12_04.RDS")    #Noise_Sim_Coexp_24_05_24.RDS
# Noise_Tests <- Noise_Tests[1:4]
names(Noise_Tests) <- c("ASCA", "PLS", "Correlations", "MASCARA")

Noise_Tests <- Noise_Tests[c("ASCA", "PLS", "Correlations", "MASCARA")]
Noise_Tests <- lapply(Noise_Tests, narm, keep_col1 = T)

bx <- c(2:15)
sx <-c(2:50)

POI <- expand_grid(bx,sx)

s_B_res <- data.matrix(do.call(rbind,lapply(Noise_Tests, function(x) x[1,])))

s_B_res <- cbind.data.frame(POI, t(s_B_res))
s_B_res <- melt(s_B_res, id.vars = c("bx","sx"))
colnames(s_B_res)[3:4] <- c("Method","GMR")

  i <- NULL
for(i in 1:nrow(s_B_res)){

  s_B_res$GMR[i] <- s_B_res$GMR[i]/ (prod(c(1:s_B_res$sx[i]))^(1/s_B_res$sx[i]))

}

s_B_res$GMR <- log2(s_B_res$GMR)
max_value <- max(s_B_res$GMR)
min_value <- min(s_B_res$GMR)

library(viridis)

ggplot(s_B_res, aes(x = bx, y = sx)) +
  geom_tile(aes(fill = GMR)) +
  # scale_fill_gradient(low = "blue", high = "yellow", limits = c(min_value,max_value))+
  scale_fill_viridis_c(limits = c(min_value, max_value)) +
  facet_wrap(~ Method) +
  labs(title = "Num baits & spikes",
       x = "Baits",
       y = "Spikes",
       color = "GMR")

```

### 2025-01-20 R/Paper_markdown.Rmd
25_01_20
MILP arc consistency KEGG network

@@
 header-includes:
-  - \usepackage{multicol}
   - \usepackage{amsmath}
@@
 library(ggplot2)
-library(MetStaT)
+# library(MetStaT)
 library(gASCA)
@@
 library(doParallel)
-library(MUVR)
+# library(MUVR)
 library(pheatmap)
-library(caret)
-library(mixOmics)
+# library(caret)
+# library(mixOmics)
 library(cowplot)
@@
 library(biomformat)
-library(DiscriMiner)
+# library(DiscriMiner)
 library(viridis)
@@
 library(plotly)
-library(ggbreak)
-library(rhmmer)
+# library(ggbreak)
+# library(rhmmer)
 library(pracma)
+library(pls)
 ```
@@
 
-From a plant biology point of view, the between-group and the within-group variance can be quite different.
+From a biological point of view, the between-group and the within-group variance can be quite different.
 However, both may contain valuable biological information.
@@
 This could result in slight differences in bait genes between the replicates, but also the genes that we want to identify with a co-expression analysis.
+loblolly pine cones @liExtensiveVariationDroughtInduced2021
 There may be many other factors that are inducing this within-group variance, but it is quite likely that if a factor induces a change in the treatment response in one replicate, that this response difference occurs throughout the entire process or pathway in that one replicate.
@@
 For a successful coexpression analysis it is necessary to have a predefined set of baits that are already known to be involved in the pathway under study.
-Methods like ranked correlations and partial least squares regression (PLS; @geladiPartialLeastsquaresRegression1986; @woldCollinearityProblemLinear1984) are able to investigate relationships between a set of baits and the other measured genes.
+Methods like ranked correlations WGCNA  and partial least squares regression (PLS; @geladiPartialLeastsquaresRegression1986; @woldCollinearityProblemLinear1984) are able to investigate relationships between a set of baits and the other measured genes.
 In ranked correlations the relationships are quantified at a bivariate level.
@@
 
-To explore the utility of MASCARA, the method will be compared to ASCA, PLS2 and correlation analysis illustrated using an example RNAseq dataset (from @haiderTranscriptomeAnalysisPhosphate2023).
+To explore the utility of MASCARA, the method will be compared to ASCA, PLS2 and correlation analysis illustrated using an example RNAseq dataset (from @haiderTranscriptomeAnalysisPhosphate2023) as well as an LCMS dataset from @caldanaHighdensityKineticAnalysis2011.
 Furthermore, this dataset is used as a guide to several simulation studies to validate the development of a novel approach to coexpression analysis.
@@
 baits <- paste0("X_",c(1989:2000))   #501:510
-
 meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
@@
 
-Noise_Tests <- readRDS("Noise_Sim_Coexp_24_01_22.RDS")
-
-names(Noise_Tests) <- c("ASCA", "PLS", "Correlations", "MASCARA")
-
-Noise_Tests <- Noise_Tests[c("ASCA", "PLS", "Correlations","MASCARA")]
+# Noise_Tests <- readRDS("Noise_Sim_Coexp_24_01_22.RDS")
+
+Noise_Tests <- readRDS("Noise_Sim_Coexp_24_06_13.RDS")    #Noise_Sim_Coexp_24_05_24.RDS
+
+names(Noise_Tests) <- c("ASCA", "PLS", "Correlations", "MASCARA", "WGCNA")
+
+Noise_Tests <- Noise_Tests[c("ASCA", "PLS", "Correlations", "WGCNA", "MASCARA")]
 Noise_Tests <- lapply(Noise_Tests, narm, keep_col1 = T)
@@
 
-for(i in 2:4){
+for(i in 2:5){
   res <- Noise_Tests[[i]][12,]
@@
 
-mRES$Noise <- rep(ERs, 4)
-mRES$Main <- rep(main_lev, 4)
+mRES$Noise <- rep(ERs, 5)
+mRES$Main <- rep(main_lev, 5)
 ```
@@
 
-mRES$Size_num <- rep(y, 4)
+mRES$Size_num <- rep(y, 5)
 ```
@@
 
-Rep_Tests <- readRDS("Noise_Sim_24_01_22_Replicate_Tests.RDS")
-
-names(Rep_Tests) <- c("ASCA", "PLS","Correlations","MASCARA")
+# Rep_Tests <- readRDS("Noise_Sim_24_01_22_Replicate_Tests.RDS")
+
+Rep_Tests <- readRDS("Noise_Sim_24_12_08_Replicate_Tests_3.RDS") #Noise_Sim_24_12_04_Replicate_Tests.RDS  #Noise_Sim_24_06_13_Replicate_Tests.RDS Noise_Sim_24_05_24_Replicate_Tests.RDS
+
+names(Rep_Tests) <- c("ASCA", "PLS","Correlations","MASCARA", "WGCNA")
 
@@
 
-y <- rep(c(0.0004,0.002, 0.01),3)      
-x <- rep(25,9)
-nse <- rep(c(0.1,0.2,0.3), each = 3)
-DS <- c(1:9)
+# y <- rep(c(0.0004,0.002, 0.01),3)      
+# x <- rep(25,9)
+# nse <- rep(c(0.1,0.2,0.3), each = 3)
+
+# y <- rep(c(0.008333333, 0.016666667, 0.025000000),3)      
+# x <- rep(30,9)  #100
+# nse <- rep(c(0.2,0.4,0.6), each = 3)
+
+m_l <- c(1:5)/10
+
+x <- rep(30,25)  #100  number of Experiment responders (CE) genes  URP
+
+y <- m_l/x[1]     #rep(c(0.008333333, 0.016666667, 0.025000000))
+y <- rep(y,length(y))
+
+nse <- rep(m_l, each = length(m_l))  #rep(c(0.1,0.2, 0.3, 0.4, 0.5,0.6), each = length(y))
+
+DS <- c(1:25)
 
@@
 
-for(i in 2:4){
+for(i in 2:5){
   res <- Rep_Tests[[i]][12,]
@@
 
-orders <- c(7,8,9,4,5,6,1,2,3)
-
-max <- max(log2(mRES$F1_Score) + 1)
+# orders <- c(7,8,9,4,5,6,1,2,3)      # - WARNING -
+orders <- c(21,22,23,24,25,
+            16,17,18,19,20,
+            11,12,13,14,15,
+            6,7,8,9,10,
+            1,2,3,4,5)
+
+max <- max(log2(mRES$F1_Score) )
 i <- NULL
@@
   labs(subtitle = paste0("CE = ", main_lev[l], " E = ", nse[l]))+
-  theme_dark() +
+  theme_bw() +
       theme(axis.title = element_blank())
@@
 
-# tiff("Fig7.tif")
+# tiff("Fig7_new1.tif",width = 960*4, height = 960*4,pointsize = 12, res = 300)
 # ps1 <- grid.arrange(ps, left = "log2(GMR)", bottom = "Replicate Number                     ")
@@
 
-a <- knitr::include_graphics("Fig7.tif")
+a <- knitr::include_graphics("Fig7_new1.tif")
 a
+
+getwd()
 
@@
 
+```{r}
+      # F_norms <- c(do.call(c,lapply(res_ASCAplus_nb$decomposition, function(x) norm(x[, c(pathway, non_pathway)], type = "F"))), 
+      #            residual = norm(res_ASCAplus_nb$residuals[, c(pathway, non_pathway)], type = "F"))
+      #
+
+      F_norms <- c(do.call(c,lapply(res_ASCAplus_nb$decomposition, norm, type = "F")), 
+                 residual = norm(res_ASCAplus_nb$residuals, type = "F"))
+
+MASCARA <- function(resids, baits){
+  #target projection
+  spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
+                          resids[,which(colnames(resids) %in% baits)], ncomp = 2)
+
+  #####
+  y <- colMeans(spls_res$Yloadings)
+  U <- spls_res$projection
+  UTP <- (dot(t(U),y)/dot(y,y))
+
+  sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])
+
+  return(list(sPLS_bait_cands,UTP,spls_res,y))
+}
+
+m_res_full <- MASCARA(res_ASCAplus_nb$residuals,baits)
+m_rest <- m_res_full[[3]]$scores %*% t(m_res_full[[3]]$loadings)
+m_baits <- m_res_full[[3]]$Yscores %*% t(m_res_full[[3]]$Yloadings)
+
+m_comb <- cbind.data.frame(m_rest, m_baits)
+
+deflated_perc <- sum(m_res_full[[3]]$Xvar)/m_res_full[[3]]$Xtotvar
+
+      F_norms <- c(do.call(c,lapply(res_ASCAplus_nb$decomposition, norm, type = "F")), 
+                 residual = norm(res_ASCAplus_nb$residuals, type = "F"))
+
+F_norms <- c(F_norms, 
+             "CE" = norm(res_ASCAplus_nb$decomposition$growth_condition +
+                           res_ASCAplus_nb$decomposition$`growth_condition:time`, type = "F"),
+             "modelled" = deflated_perc * F_norms["residual"],
+             "noise" = (1-deflated_perc) * F_norms["residual"] )
+
+F_norms[5:7] / sum(F_norms[5:7])
+
+```
+
 In the real data from @haiderTranscriptomeAnalysisPhosphate2023 some plants were subjected to massive stress with the phosphate limitation.
@@
 
+@tajiImportantRolesDrought2002
+
+@zintaDynamicsMetabolicResponses2018
+
 ```{r residualdists, fig.cap=c("Distributions of Fisher transformed Pearson correlations (supplementary equation \\@ref(eq:Fisher)) in residuals from the ASCA model on the data from @haiderTranscriptomeAnalysisPhosphate2023.
 Blue indicates correlations between SL pathway genes, red indicates correlations between SL pathway genes and top 10% DE URP genes (upregulated in P- over P+).")}
+library(FNN)
 
@@
 
+```
+
+```{r}
+GA_ref <- read_xlsx("Rice_Annotation/GA pathway genes-IDs.xlsx")
+GA_ref$ID <- gsub("g","t",GA_ref$ID)
+GAs <- ref[which(gsub("-.*","",ref$Feature) %in% GA_ref$ID),1]
+GAs <- GAs[which(GAs %in% colnames(res_ASCAplus_nb$residuals))]
+
+# pathway <- GAs
+
+PSI_ref <-read.table("Rice_Annotation/PSI_genes.txt", sep = "\t", header = T)
+
+# Load necessary libraries
+library(reshape2)  # For melt()
+library(ggplot2)   # For ggplot()
+
+# Define the function to calculate Fisher-transformed correlation values
+cor_val <- function(x){
+  cor_res <- cor(x[])
+  cor_res <- 0.5 * log((1 + cor_res) / (1 - cor_res))
+  return(cor_res)
+}
+
+# Calculate correlations for res_ASCAplus_nb$residuals
+# test_residuals <- cor_val(res_ASCAplus_nb$residuals[, pathway])
+
+# pathway <- baits
+
+ASCA_cands2_filtered <- ASCA_cands2[!rownames(ASCA_cands2) %in% pathway, ]
+
+# non_pathway <- PSI_ref[,1]
+# non_pathway <- non_pathway[which(non_pathway %in% rownames(ASCA_cands2_filtered))]
+
+non_pathway <- rownames(ASCA_cands2_filtered[which(ASCA_cands2_filtered$VIP > quantile(ASCA_cands2_filtered$VIP, 0.9)),])#[1:1253]
+
+test2_residuals <- cor_val(res_ASCAplus_nb$residuals[, c(pathway, non_pathway)])
+test2_residuals[upper.tri(test2_residuals, diag = TRUE)] <- NA
+
+mt_residuals <- melt(test2_residuals, na.rm = TRUE)
+
+mt_residuals$group <- "Other"
+mt_residuals$group[mt_residuals$Var1 %in% pathway & mt_residuals$Var2 %in% pathway] <- "SL-SL"
+mt_residuals$group[mt_residuals$Var1 %in% non_pathway & mt_residuals$Var2 %in% pathway] <- "SL-Other"
+
+mt_residuals <- mt_residuals[mt_residuals$group %in% c("SL-SL", "SL-Other"), ]
+colnames(mt_residuals)[3] <- "Correlation"
+
+mt_residuals$Dataset <- "Within"
+
+# test_X <- cor_val(X[, pathway])
+
+test2_X <- cor_val(X[, c(pathway, non_pathway)])
+test2_X[upper.tri(test2_X, diag = TRUE)] <- NA
+
+mt_X <- melt(test2_X, na.rm = TRUE)
+
+mt_X$group <- "Other"
+mt_X$group[mt_X$Var1 %in% pathway & mt_X$Var2 %in% pathway] <- "SL-SL"
+mt_X$group[mt_X$Var1 %in% non_pathway & mt_X$Var2 %in% pathway] <- "SL-Other"
+
+mt_X <- mt_X[mt_X$group %in% c("SL-SL", "SL-Other"), ]
+colnames(mt_X)[3] <- "Correlation"
+
+mt_X$Dataset <- "Total"
+
+#deflated residuals:
+# MASCARA <- function(resids, baits){
+#   #target projection
+#   spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
+#                           resids[,which(colnames(resids) %in% baits)], ncomp = 3)
+#   
+#   #####
+#   y <- colMeans(spls_res$Yloadings)
+#   U <- spls_res$projection
+#   UTP <- (dot(t(U),y)/dot(y,y))
+#   
+#   sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])
+#   
+#   return(list(sPLS_bait_cands,UTP,spls_res,y))
+# }
+
+# m_res <- MASCARA(res_ASCAplus_nb$residuals[, c(pathway, non_pathway)],baits)[[3]]
+# 
+# m_rest <- m_res$scores %*% t(m_res$loadings)
+# m_baits <- m_res$Yscores %*% t(m_res$Yloadings)
+# 
+# m_comb <- cbind.data.frame(m_rest, m_baits)
+# 
+# test_m <- cor_val(m_comb[, c(pathway, non_pathway)])
+# test_m[upper.tri(test_m, diag = TRUE)] <- NA
+# 
+# mt_m <- melt(test_m, na.rm = TRUE)
+# 
+# mt_m$group <- "Other"
+# mt_m$group[mt_m$Var1 %in% pathway & mt_m$Var2 %in% pathway] <- "SL-SL"
+# mt_m$group[mt_m$Var1 %in% non_pathway & mt_m$Var2 %in% pathway] <- "SL-Other"
+# 
+# mt_m <- mt_m[mt_m$group %in% c("SL-SL", "SL-Other"), ]
+# colnames(mt_m)[3] <- "Correlation"
+# 
+# mt_m$Dataset <- "Deflated"
+#
+
+library(ggpubr)
+
+mt_combined <- rbind(mt_residuals, mt_X)   #   , mt_m
+
+mt_combined$Group_Dataset <- paste(mt_combined$group, mt_combined$Dataset, sep = "_")
+
+mt_combined$Dataset <- factor(mt_combined$Dataset, levels = c("Total", "Within", "Deflated"))
+
+dens1 <- ggplot(mt_combined, aes(x = group, y = Correlation, fill = group)) +
+  geom_violin() +
+  theme_bw() +
+  labs(fill = "Group") +
+  ggtitle("Boxplot of Fisher-Transformed Correlations by Variance Partition") +
+  xlab("Group") +
+  ylab("Correlation") +
+  facet_wrap(~ Dataset) +
+  stat_compare_means() +
+  theme(
+    plot.title = element_text(hjust = 0.5),
+    legend.position = "none"  # Remove legend if 'fill' is the same as 'x'
+  )
+
+print(dens1)
+
+dens <- ggplot(mt_combined, aes(Correlation, fill = group)) + geom_histogram(bins = 140) + theme_bw()  + facet_wrap(~ Dataset)
+
+dens
+
+# 
+# xkl <- KL.divergence(mt_combined$Correlation[which(mt_combined$Group_Dataset == "SL-Other_Total")], mt_combined$Correlation[which(mt_combined$Group_Dataset == "SL-SL_Total")])
+# 
+# rkl <- KL.divergence(mt_combined$Correlation[which(mt_combined$Group_Dataset == "SL-Other_Within")], mt_combined$Correlation[which(mt_combined$Group_Dataset == "SL-SL_Within")])
+# 
+# plot(xkl, rkl)
+# abline(a=0,b=1)
 ```
@@
 
-a <- knitr::include_graphics("Fig9.tiff")
+a <- knitr::include_graphics("Fig9.tif")
 a

### 2025-01-20 R/WGCNA_testing.Rmd
25_01_20
MILP arc consistency KEGG network

---
title: "WGCNA_test"
author: "Fred White"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(WGCNA)
library(matrixStats)
library(gASCA)
library(pracma)
source("DATA_SIM_FUNCS.R")
#getwd()
```

```{r}
Create_Core_DEV <- Create_Core_DEV_2

baits <- paste0("X_",c(1985:1988))   #501:510
spikes <- paste0("X_",c(1989:2000))

meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
colnames(meta) <- c("growth_condition","time","interaction")

```

```{r}

y <- rep(c(0.0004,0.002, 0.01),3)

x <- rep(30,9)  #100

nse <- rep(c(0.1,0.3,0.5), each = 3)

# y <- c(0.1, 2, 4)
# x <- c(25,125,250)

main_lev <- x*y
ERs <- x

nreps <- 3

a <- 6
i <- 19
    sim_data <- Create_Core_DEV(nreps = nreps, meta = meta, plot = FALSE, 
                    EffectSize = c(X_a_ab = main_lev[a], time = 0.2, E = nse[a], mu = 0.2, Struc_E = 0.05), 
                    struc_resid = T, e_sigma = c(1.5,0.8,0.0), a_sigma = c(1.5,0.7,0.6,0.1),
                    b_sigma = c(1,0), irr_spikes = FALSE, SCORE_SEED = (250021 + i), Experiment_responders = ERs[a], ts = 3 + i)

```

```{r}
data <- sim_data[[1]]

```

#variance partitioning

```{r}
ar <- ASCA_decompose(d = apply(meta[,c(1,2)],2,as.factor), x = data, f = "growth_condition * time")

#unexpectedly different from PERMANOVA R2
norm(ar$residuals, type = "F")

# norm(ar$decomposition$Temperature, type = "F")

resids <- ar$residuals

# data <- resids
```

```{r}
spt <- pickSoftThreshold(data, moreNetworkConcepts = TRUE) 
# spt$powerEstimate
```

```{r}
softPower <- 6
adjacency <- adjacency(data, power = softPower)

TOM <- TOMsimilarity(adjacency)
```

```{r}
par(mfrow = c(1,2))
plot(spt$fitIndices[,1], -sign(spt$fitIndices[,3])*spt$fitIndices[,2],
      xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
      main = paste("Scale independence"));
  text(spt$fitIndices[,1], -sign(spt$fitIndices[,3])*spt$fitIndices[,2],
       labels=spt$fitIndices[,1],cex=0.9,col="red");
  # this line corresponds to using an R^2 cut-off of h
  abline(h=0.90,col="red")

# Mean connectivity as a function of the soft-thresholding power
plot(spt$fitIndices[,1], spt$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
text(spt$fitIndices[,1], spt$fitIndices[,5], labels=spt$fitIndices[,1], cex=0.9,col="red")

```

```{r}
TOM.dissimilarity <- 1-TOM
```

```{r}
geneTree <- hclust(as.dist(TOM.dissimilarity), method = "average")

sizeGrWindow(12,9)
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
labels = FALSE, hang = 0.04)
```

```{r}
Modules <- cutreeDynamic(dendro = geneTree, distM = TOM.dissimilarity, deepSplit = 2, pamRespectsDendro = FALSE, minClusterSize = 12)
table(Modules)

ModuleColors <- labels2colors(Modules) #assigns each module number a color
table(ModuleColors)

plotDendroAndColors(geneTree, ModuleColors,"Module",
dendroLabels = FALSE, hang = 0.03,
addGuide = TRUE, guideHang = 0.05,
main = "Gene dendrogram and module colors")
```

```{r}
MElist <- moduleEigengenes(data, colors = ModuleColors)
MEs <- MElist$eigengenes
head(MEs)
ME.dissimilarity = 1-cor(MElist$eigengenes, use="complete")
```

```{r}
METree = hclust(as.dist(ME.dissimilarity), method = "average") #Clustering eigengenes
par(mar = c(0,4,2,0)) #setting margin sizes
par(cex = 0.6);#scaling the graphic
plot(METree)
abline(h=.25, col = "red") #a height of .25 corresponds to correlation of .75

```

```{r}
merge <- mergeCloseModules(data, ModuleColors, cutHeight = .25)
# The merged module colors, assigning one color to each module
mergedColors = merge$colors
# Eigengenes of the new merged modules
mergedMEs = merge$newMEs

#check modules for POIs, URPs, 
#WGCNA also fails to discern between POIs and URPs under our simulated conditions as can be seen by module membership 
#although the miminum module number choice is arbitrarily chosen
tail(merge$colors,50)
```

```{r}

sKME <- signedKME(data, mergedMEs)

```

#selecting features

```{r}
# #calculate input data relationship to precalculated eigengenes
# sKME <- signedKME(data, mergedMEs)

#select coexpression candidates based on the cluster membership of a set of baits
#qbar target projection from MASCARA
q_bar <- colMeans(sKME[baits,])
R <- sKME[!rownames(sKME) %in% baits,]
R_TP <- (dot(t(R),q_bar)/dot(q_bar,q_bar))

#ordered list of candidates 
Candidates <- as.data.frame(R_TP[order(R_TP, decreasing = TRUE)])
```

<!-- ```{r} -->
<!-- plotDendroAndColors(geneTree, cbind(ModuleColors, mergedColors), -->
<!-- c("Original Module", "Merged Module"), -->
<!-- dendroLabels = FALSE, hang = 0.03, -->
<!-- addGuide = TRUE, guideHang = 0.05, -->
<!-- main = "Gene dendrogram and module colors for original and merged modules") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- nGenes = ncol(data) -->
<!-- nSamples = nrow(data) -->

<!-- #  -->
<!-- # module.trait.correlation = cor(mergedMEs, datTraits, use = "p") #p for pearson correlation coefficient -->
<!-- # module.trait.Pvalue = corPvalueStudent(module.trait.correlation, nSamples) #calculate the p-value associated with the correlation -->

<!-- ``` -->

### 2025-01-20 R/run_first/1.pattern_examples_MASCARA_pipeline.Rmd
25_01_20
MILP arc consistency KEGG network

@@
 sim_data <- Create_Core_DEV_2(nreps = 3, meta = meta, 
-                        EffectSize = c(X_a_ab = 5, time = 1, E = 10, mu = 1, Struc_E = 1),    # E = 1, Struc_E = 0.4
+                        EffectSize = c(X_a_ab = 5, time = 1, E = 5, mu = 1, Struc_E = 0.9),    # E = 1, Struc_E = 0.4
                         struc_resid = TRUE, e_sigma = c(1.5,0.8,0.0), a_sigma = c(1,0.7,0.6,0.1),
                         b_sigma = c(1,0.9), irr_spikes = TRUE, SCORE_SEED = 12, Experiment_responders = Experiment_responders,
-                        struc_seed = 127, score_plot = FALSE, plot = FALSE)
+                        struc_seed = 1230, score_plot = FALSE, plot = FALSE)
 
@@
   geom_point(aes(colour = growth_condition, shape = time, size = 20)) +
-  ggtitle("Scores") + 
+    scale_shape_manual(values=c(18,19,20,21))+  #3, 16, 17
+
+  guides(size = FALSE) +
+  ggtitle("Residual Scores") + 
   theme_bw()
+```
 
+```{r}
 colnames(yloadings) <- colnames(loadings_r_baits)
@@
 ```{r}
-loadings_r_baits$ID[-which(loadings_r_baits$ID %in% c(baits,spikes))] <- "Other"
+loadings_r_baits$ID[-which(loadings_r_baits$ID %in% c(baits,spikes))] <- "Unrelated"
+set.seed(213)
+loadings_r_baits$ID[which(loadings_r_baits$ID %in% spikes)] <- sample(c("Pathway","Related"),length(which(loadings_r_baits$ID %in% spikes)),replace = T)
+loadings_r_baits$ID[which(loadings_r_baits$ID %in% baits)] <- "Baits"
 
-loadings_r_baits$ID[which(loadings_r_baits$ID %in% spikes)] <- "Spikes"
-loadings_r_baits$ID[which(loadings_r_baits$ID %in% baits)] <- "Baits"
+# sample(loadings_r_baits$ID[which(loadings_r_baits$ID == "Pathway")],6) <- "Related"
 ```
@@
 
-  geom_segment(x = 0, y = 0, xend = line[1], yend = line[2], lineend = "round", linewidth = 2, colour = "purple", arrow = arrow()) +    
-  scale_shape_manual(values=c(3, 16, 17))+
+  geom_segment(x = 0, y = 0, xend = line[1], yend = line[2], lineend = "round", linewidth = 2, colour = "steelblue", arrow = arrow()) +    
+  scale_shape_manual(values=c(17,16,16,15))+  #3, 16, 17
 
-    scale_color_manual(values=c('#E69F00',"black", '#56B4E9'))+
-      geom_point(data = loadings_r_baits[which(loadings_r_baits$ID %in% c("Baits","Spikes")),], aes(size = 15)) +
+    scale_color_manual(values=c('purple',  'purple',"orange","lightgreen"))+
+      geom_point(data = loadings_r_baits[which(loadings_r_baits$ID %in% c("Baits","Pathway","Related")),], aes(size = 15)) +
+    guides(size = FALSE) +
+
   theme_bw()
 
-b
+a + b
 ```

### 2025-01-20 R/run_first/2.plot_pathway_net_example.Rmd
25_01_20
MILP arc consistency KEGG network

@@
 
-Net <- V(g)$name
+Pathway  <- V(g)$name
 ```
@@
 ```{r}
-id <- c(1:length(Net))
+id <- c(1:length(Pathway ))
 
@@
 
-Net[which(Net %in% baits)] <- "POI"
-Net[which(Net %in% spikes)] <- "RP"
+Pathway [which(Pathway  %in% baits)] <- "POI"
+Pathway [which(Pathway  %in% spikes)] <- "Related"
 
-Net[-which(Net %in% c("POI","RP"))] <- "URP"
+Pathway [-which(Pathway  %in% c("POI","Related"))] <- "Unrelated"
 
@@
 
-Gene <- names(V(g))
-Gene[-which(Gene %in% temp)] <- "Other"
-Gene[which(Gene %in% temp)] <- "Baits"
+Feature <- names(V(g))
+Feature[-which(Feature %in% temp)] <- "Other"
+
+Feature[which(Pathway  %in% c("POI","Related"))] <- "Targets"
+
+Feature[which(names(V(g)) %in% temp)] <- "Baits"
+
+Feature <- factor(Feature, levels = c("Baits","Targets","Other"))
+
+# Feature[which(names(V(g)) %in% spikes)] <- "Targets"
 
@@
   activate(nodes) %>%
-  mutate(Net = Net)
+  mutate(Pathway  = Pathway )
 
@@
   activate(nodes) %>%
-  mutate(Gene = Gene)
+  mutate(Feature = Feature)
 
@@
 
-colours = c("POI" = "purple", "RP" = "orange", "URP" = "lightgreen")
+colours = c("POI" = "purple", "Related" = "orange", "Unrelated" = "lightgreen")
 
@@
 
-edge_colours <- c("+ve" = "grey", "-ve" = "blue")
+edge_colours <- c("+ve" = "#091EB7", "-ve" = "blue")
 
@@
 realplot <- ggraph(tidynet, layout = coords) + 
-  geom_edge_link(aes(colour = association, alpha = weight), show.legend=FALSE) +  
+  geom_edge_link(aes(colour = association), show.legend=FALSE) +  
   scale_edge_color_manual(values = edge_colours, guide = "none") +
   geom_node_point(
-    aes(size = VIP, color = Net, shape = Gene), show.legend=FALSE        #CHANGE THIS
+    aes(size = VIP, color = Pathway , shape = Feature), show.legend=FALSE        #CHANGE THIS
   ) +
   geom_node_point(
-    aes(color = Net, shape = Gene)    #, shape = Baits
+    aes(color = Pathway , shape = Feature)    #, shape = Baits
   ) +
   scale_color_manual(values = colours) + 
-    scale_shape_manual(values = c(17,16)) +   #18,20
+    scale_shape_manual(values = c(17,16,15)) +   #18,20
 
@@
 
-adj[Net == "URP", Net == "URP"]  <- 1
-adj[Net %in% c("POI","RP"), Net %in% c("POI","RP")] <- 1
+adj[Pathway  == "Unrelated", Pathway  == "Unrelated"]  <- 1
+adj[Pathway  %in% c("POI","Related"), Pathway  %in% c("POI","Related")] <- 1
 
-adj[Net %in% c("POI","RP"), Net == "URP"] <- -1
-adj[Net == "URP", Net %in% c("POI","RP")] <- -1
+adj[Pathway  %in% c("POI","Related"), Pathway  == "Unrelated"] <- -1
+adj[Pathway  == "Unrelated", Pathway  %in% c("POI","Related")] <- -1
 ```
@@
 
-#define correctly ordered type Net
-Net <- V(g)$name
+#define correctly ordered type Pathway 
+Pathway  <- V(g)$name
 
@@
   activate(nodes) %>%
-  mutate(Net = Net)
+  mutate(Pathway  = Pathway )
 
-colours = c("POI" = "purple", "RP" = "orange", "URP" = "lightgreen")  
+colours = c("POI" = "purple", "Related" = "orange", "Unrelated" = "lightgreen")  
 # shape = c("Baits" = 18, "other" = 1)
@@
   geom_node_point(
-    aes(size = VIP, color = Net, shape = Gene), show.legend=FALSE
+    aes(size = VIP, color = Pathway , shape = Feature), show.legend=FALSE
   ) +
   geom_node_point(
-    aes(color = Net, shape = Gene)
+    aes(color = Pathway , shape = Feature)
   ) +
   scale_color_manual(values = colours) + 
-      scale_shape_manual(values = c(17,16)) +   #18,20
+      scale_shape_manual(values = c(17,16,15)) +   #18,20
   # scale_edge_linetype_manual(values = "dashed") +

### 2025-01-20 R/DATA_SIM_FUNCS.R

@@ -2160,6 +2160,57 @@ MASCARA_DEV <- function(resids,ref, baits){
 
 
 
+MASCARA4_test <- function(resids, baits, spikes){
+  #target projection
+  spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
+                         resids[,which(colnames(resids) %in% baits)], ncomp = 2)
+  
+  #####
+  y <- colMeans(spls_res$Yloadings)
+  U <- spls_res$projection
+  UTP <- (dot(t(U),y)/dot(y,y))
+  
+  sPLS_bait_cands <- as.data.frame(UTP[order(UTP, decreasing = TRUE)])
+  RP <- prod(which(rownames(sPLS_bait_cands) %in% spikes))^(1/length(spikes))
+  
+  
+  return(list(RP, sPLS_bait_cands,UTP))
+}
+
+WGCNA4_test <- function(data, baits, spikes){
+  
+  softPower <- 6
+  adjacency <- adjacency(data, power = softPower)
+  
+  TOM <- TOMsimilarity(adjacency)
+  TOM.dissimilarity <- 1-TOM
+  geneTree <- hclust(as.dist(TOM.dissimilarity), method = "average")
+  Modules <- cutreeDynamic(dendro = geneTree, distM = TOM.dissimilarity, deepSplit = 2, pamRespectsDendro = FALSE, minClusterSize = 12)
+  
+  MElist <- moduleEigengenes(data, colors = ModuleColors)
+  MEs <- MElist$eigengenes
+  ME.dissimilarity = 1-cor(MElist$eigengenes, use="complete")
+  
+  merge <- mergeCloseModules(data, ModuleColors, cutHeight = .25)
+  
+  mergedMEs = merge$newMEs
+  
+  
+  sKME <- signedKME(data, mergedMEs)
+  
+  #feature selection
+  q_bar <- colMeans(sKME[baits,])
+  R <- sKME[!rownames(sKME) %in% baits,]
+  R_TP <- (dot(t(R),q_bar)/dot(q_bar,q_bar))
+  
+  Candidates <- as.data.frame(R_TP[order(R_TP, decreasing = TRUE)])
+  
+  RP <- prod(which(rownames(Candidates) %in% spikes))^(1/length(spikes))
+  
[…]

### 2025-01-20 R/Rice_Annotation/ID_conv.R

@@ -1,0 +1,642 @@
+library(riceidconverter)
+library(devtools)
+library(ggplot2)
+library(MetStaT)
+library(gASCA)
+library(patchwork)
+library(readxl)
+library(ggfortify)
+library(ggrepel)
+library(grid)
+library(gridExtra)
+library(tidyr)
+library(tidyverse)
+library(reshape2)
+library(scales)
+library(matrixStats)
+library(DESeq2)
+library(MASS)
+library(glmnet)
+library(foreach)
+library(doParallel)    
+library(MUVR) 
+library(pheatmap)
+library(caret)
+library(mixOmics)
+library(pracma)
+library(pls)
+library(WGCNA)
+library(DESeq2)
+
+
+
+setwd("C:/Users/fwhite/Documents/Github/MASCARA/R")  # - WARNING -
+
+
+source("DATA_SIM_FUNCS.R")
+
+
+# ref <- read.table("R/Rice_Annotation/Oryza_sativa.IRGSP-1.0.50.gtf", sep = "\t")
+# ref <- ref[which(ref$V3 == "transcript"),]
+# ref$ID <- gsub(".*gene_id ","",ref$V9)
+# ref$ID <- gsub(";.*","",ref$ID)
+# 
+# ref$transcript <- gsub(".*transcript_id ","",ref$V9)
+# ref$transcript <- gsub(";.*","",ref$transcript)
+# 
+# ref$biotype <- gsub(".*transcript_biotype ","",ref$V9)
+# ref$biotype <- gsub(";.*","",ref$biotype)
+#
+
+
+X <- readRDS("Data_Haider/Rice_Counts4ASCA.RDS")
+meta <- readRDS("Data_Haider/Rice_Meta4ASCA.RDS")
+colnames(meta)[3] <- c("time")
+
+meta$time <- as.character(meta$time)
+
+meta$time[which(meta$time == "8")] <- "1"
+meta$time[which(meta$time == "10")] <- "3"
+meta$time[which(meta$time == "14")] <- "7"
+meta$time[which(meta$time == "15")] <- "8"
+
+
+X <- t(assay(X))
+
+meta <- meta[order(meta$time),]
+meta <- meta[order(meta$growth_condition),]
+
+X <- X[match(meta$ID, rownames(X)),]
+
+
+
+#X <- X[,which(colnames(X) %in% rownames(sPLS_bait_cands))]
+
[…]

### 2025-01-20 R/run_first/3.sim1_CE_size.R

@@ -24,6 +24,7 @@ library(caret)
 library(mixOmics)
 library(pracma)
 library(pls)
+library(WGCNA)
 
 # source("../DATA_SIM_FUNCS.R")
 
@@ -57,6 +58,41 @@ splsR <- function(resids, baits, spikes){
   return(list(RP, sPLS_cands))
 }
 
+WGCNA4_test <- function(data, baits, spikes){
+  
+  softPower <- 6
+  adjacency <- adjacency(data, power = softPower)
+  
+  TOM <- TOMsimilarity(adjacency)
+  TOM.dissimilarity <- 1-TOM
+  geneTree <- hclust(as.dist(TOM.dissimilarity), method = "average")
+  Modules <- cutreeDynamic(dendro = geneTree, distM = TOM.dissimilarity, deepSplit = 2, pamRespectsDendro = FALSE, minClusterSize = 12)
+  ModuleColors <- labels2colors(Modules)
+  MElist <- moduleEigengenes(data, colors = ModuleColors)
+  MEs <- MElist$eigengenes
+  ME.dissimilarity = 1-cor(MElist$eigengenes, use="complete")
+  
+  merge <- mergeCloseModules(data, ModuleColors, cutHeight = .25)
+  # mergedColors = merge$colors
+  
+  mergedMEs = merge$newMEs
+  
+  
+  sKME <- signedKME(data, mergedMEs)
+  
+  #feature selection
+  q_bar <- colMeans(sKME[baits,])
+  R <- sKME[!rownames(sKME) %in% baits,]
+  R_TP <- (dot(t(R),q_bar)/dot(q_bar,q_bar))
+  
+  Candidates <- as.data.frame(R_TP[order(R_TP, decreasing = TRUE)])
+  
+  RP <- prod(which(rownames(Candidates) %in% spikes))^(1/length(spikes))
+  
+  return(list(RP, Candidates, mergedMEs, sKME))
+  
+  
+}
 
 
 
@@ -162,6 +198,9 @@ PLS_RES <- data.frame(matrix(nrow = ncands, ncol = length(X_funced)))
 
 MASCARA_RES  <- data.frame(matrix(nrow = ncands, ncol = length(X_funced)))
 
[…]

### 2025-01-20 R/run_first/sim_num_baits_spikes.R

@@ -27,8 +27,11 @@ library(WGCNA)
 
 
 # source("../DATA_SIM_FUNCS.R")
+# setwd("C:/Users/fwhite/Documents/GitHub/MASCARA/R")
 
-source("DATA_SIM_FUNCS_DEV.R")
+setwd("/zfs/omics/personal/fwhite/MASCARA/R")
+
+source("DATA_SIM_FUNCS.R")
 
 
 Create_Core_DEV <- Create_Core_DEV_2
@@ -89,7 +92,10 @@ WGCNA4_test <- function(data, baits, spikes){
   TOM <- TOMsimilarity(adjacency)
   TOM.dissimilarity <- 1-TOM
   geneTree <- hclust(as.dist(TOM.dissimilarity), method = "average")
-  Modules <- cutreeDynamic(dendro = geneTree, distM = TOM.dissimilarity, deepSplit = 2, pamRespectsDendro = FALSE, minClusterSize = 12)
+  browser()
+  Modules <- cutreeDynamic(dendro = geneTree, distM = TOM.dissimilarity, deepSplit = 2, pamRespectsDendro = FALSE, minClusterSize = 4)
+  
+  # if(length(Modules > 15))
   ModuleColors <- labels2colors(Modules)
   MElist <- moduleEigengenes(data, colors = ModuleColors)
   MEs <- MElist$eigengenes
@@ -118,126 +124,78 @@ WGCNA4_test <- function(data, baits, spikes){
 }
 
 
-baits <- paste0("X_",c(1985:1988))   #501:510
-spikes <- paste0("X_",c(1989:2000))
 
+# # spikes <- paste0("X_",c(1977:1988))
+# baits <- paste0("X_",c(1985:1988))   #501:510
+# spikes <- paste0("X_",c(1989:2000))
+# 
 
-Baits <- c(baits,spikes)
-
-
-nreps <- 3
-meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = nreps))
-colnames(meta) <- c("growth_condition","time")
-meta$ID <- paste(meta$growth_condition, meta$time, sep = "_")
-
+meta <- cbind.data.frame(rep(c(1,-1), each = 12), rep(c(1:4), each = 3), rep(c(-1:-4,1:4), each = 3))
+colnames(meta) <- c("growth_condition","time","interaction")
 
 
[…]

### 2025-01-20 R/sim2_dev_24_1_04.R

@@ -1,0 +1,719 @@
+library(devtools)
+library(ggplot2)
+library(MetStaT)
+library(gASCA)
+library(patchwork)
+library(readxl)
+library(ggfortify)
+library(ggrepel)
+library(grid)
+library(gridExtra)
+library(tidyr)
+library(tidyverse)
+library(reshape2)
+library(scales)
+library(matrixStats)
+library(DESeq2)
+library(MASS)
+library(glmnet)
+library(foreach)
+library(doParallel)    
+library(MUVR) 
+library(pheatmap)
+library(caret)
+library(mixOmics)
+library(pracma)
+library(WGCNA)
+library(parallel)
+
+
+# source("../DATA_SIM_FUNCS.R")
+
+source("/zfs/omics/personal/fwhite/MASCARA/DATA_SIM_FUNCS.R")
+
+
+# Create_Core_DEV <- Create_Core_DEV_2
+
+F1_plot <- F1_plot_no_plot
+# nreps <- 3
+alphaN <- 2
+betaN <- 4
+ncands <- 500
+
+
+splsR <- function(resids, baits, spikes){
+  #vip
+  spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
+                         resids[,which(colnames(resids) %in% baits)], ncomp = 2)
+  
+  
+  #####
+  y <- colMeans(spls_res$Yloadings)
+  U <- spls_res$projection
+  
+  x <- Yloadings(spls_res)
+  ve_y <- colSums(x^2)/nrow(x)
+  
+  sPLS_cands <- cbind.data.frame(abs(spls_res$projection) %*% ve_y, spls_res$projection)
+  sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = TRUE),]
+  
+  RP <- prod(which(rownames(sPLS_cands) %in% spikes))^(1/length(spikes))
+  
+  return(list(RP, sPLS_cands))
+}
+
+
+
+
+MASCARA4_test <- function(resids, baits, spikes){
+  #target projection
+  spls_res <- simpls.fit(resids[,-which(colnames(resids) %in% baits)],
+                         resids[,which(colnames(resids) %in% baits)], ncomp = 2)
+  
+  #####
+  y <- colMeans(spls_res$Yloadings)
+  U <- spls_res$projection
+  UTP <- (dot(t(U),y)/dot(y,y))
+  
[…]
